% intro.tex
\clearpage
{
  \fontsize{11}{13.5}\selectfont
  \setlength{\parindent}{1.5em}
\section*{Foreword}

The short book of Esther is quite different and substantially longer in the Greek Septuagint (LXX) compared with the Hebrew Masoretic Text (MT). Traditionally Roman Catholic and Eastern Orthodox Bibles have included the longer text of the Septuagint while Bibles used by Protestants generally have used the shorter Masoretic Text. The differences are significant. The Hebrew text is remarkable in that God is nowhere explicitly mentioned, and although his action is everywhere evident it is left to the reader to understand this. The additions in the Greek text, on the other hand, are very explicit about the hand of God.

\subsection*{Other text traditions}
There is no single authoritative ``Septuagint'' since different ancient manuscripts have their own variants, mostly trivial changes of spelling or similar minor, inconsequential variations. There is, however, another significant variant of the text of Esther called the Alpha Text (or AT) that is found four medieval manuscripts. This has been published (in the original Greek) in Hanhart's {\textit Septuaginta:
Vetus Testamentum Graecum Auctoritate Academiae Scientiarum Gottingensis editum VIII.3: Esther} [Göttingen:
Vandenhoeck \& Ruprecht, 1966].

Until now the only readily available source for English speakers to access this text has been the excellent New English Translation of the Septuagint (NETS) Esther. This can be found at \url{https://ccat.sas.upenn.edu/nets/edition/17-esther-nets.pdf}. Although it is readily accessible and liberally licensed, the NETS Esther is protected by copyright meaning it cannot be freely redistributed or used to make derivative works without jumping through the hoops necessary to obtain copyright permission. For many independent scholars that is prohibitive. For this document, I have produced (with the assistance of generative AI) a new translation of the Alpha text from Hanhart's Greek text (as obtained from a Logos Bible software module) under a Creative Commons Attribution licence. You are essentially free to do whatever you like with the text. (But please let me know if you intend to create a commercial product from it.)

Discussions of the historical background of these three text traditions are beyond the scope of this work. Interested readers may wish to consult M.V. Fox. 1990. “The Alpha Text of the Greek Esther.” Textus, 15, Pp. 27-54.

This document presents these three textual traditions side by side using the World English Bible (WEB) translation for the first two columns and my own for the third. The left column translates the Hebrew MT and the right hand column the Greek LXX. The WEB translation was selected because it is in the public domain, meaning this document can also be made freely available and free to copy.

\subsection*{Chapter and verse divisions}

Verse numbering of the Greek version of Esther in English Bibles usually follows the ordering Jerome used in the Vulgate. This provides the Hebrew version of Esther as found in the MT first, with all the additional material from the Greek versions as additional chapters at the end. This breaks up the narrative of the LXX version so much as to make it a bit of a puzzle to read. This edition presents the text in the original order of the Greek with the parallel presentation of the MT and LXX making it easy to identify which parts are present in which sources. The WEB file obtained from eBible.org for the Greek text achieves this ordering in some places by assigning only one verse number to an extended passage. This edition uses a slightly different approach where the additions are identified using the letters A to F (corresponding to Jerome's additional chapters) and verses assigned numbers within those additional blocks. Additions A and C have been assigned their own chapter (with A before chapter 1 and C between chapters 4 and 5). Other additions form part of an existing chapter. For example, addition B is inserted into chapter 3 between 3:13 and 3:14, with the additional verses labelled as B:1 to B:6.

The versification of the Alpha Text is as it appears in the Göttingen edition, but some verses have been subdivided to make corresponding text better match the parallel versions, but without changing the order even when it is different from the other versions.

\subsection*{Spelling}
The World English Bible is available in a British English edition as well as its standard American English edition. Both versions are available for the MT and LXX columns of this document. The third column translated from the Alpha Text also comes in US and British versions, with the only difference being scripted spelling adjustments. This is the \ifbritish{British}\else{American}\fi\ English version.

\bigskip
\noindent Glen Prideaux\\
March 2026
}