\input{preamble.tex}
\begin{document}
\input{intro}
\section*{Jeremiah (Parallel) — LXX order}
\ChapterHeading{1}
\begin{paracol}{2}
\VersePair{1:1}{The word of God which came to Jeremias the son of Chelcias, of the priests, who dwelt in Anathoth in the land of Benjamin:}{1:1}{The words of Jeremiah the son of Hilkiah, one of the priests who were in Anathoth in the land of Benjamin.}
\VersePair{1:2}{accordingly as the word of God came to him in the days of Josias son of Amos king of Juda, in the thirteenth year of his reign.}{1:2}{Yahweh’s word came to him in the days of Josiah the son of Amon, king of Judah, in the thirteenth year of his reign. \footnote{1:2: “Yahweh” is God’s proper Name, sometimes rendered “LORD” (all caps) in other translations.}}
\VersePair{1:3}{And it was in the days of Joakim son of Josias king of Juda, until the eleventh year of Sedekias king of Juda, even until the captivity of Jerusalem in the fifth month.}{1:3}{It came also in the days of Jehoiakim the son of Josiah, king of Judah, to the end of the eleventh year of Zedekiah, the son of Josiah, king of Judah, to the carrying away of Jerusalem captive in the fifth month.}
\VersePair{1:4}{And the word of the Lord came to him, saying,}{1:4}{Now Yahweh’s word came to me, saying,}
\VersePair{1:5}{Before I formed thee in the belly, I knew thee; and before thou camest forth from the womb, I sanctified thee; I appointed thee a prophet to the nations.}{1:5}{{\raggedright “Before I formed you in the womb, I knew you. \poemline{2}{Before you were born, I sanctified you.}\poemline{2}{I have appointed you a prophet to the nations.”}}}
\VersePair{1:6}{And I said, O Lord, thou that art supreme Lord, behold, I know not how to speak, for I am a child.}{1:6}{Then I said, “Ah, Lord Yahweh! Behold, I don’t know how to speak; for I am a child.” \footnote{1:6: The word translated “Lord” is “Adonai.”} \footnote{1:6: “Behold”, from “\texthebrew{הִנֵּה}”, means look at, take notice, observe, see, or gaze at. It is often used as an interjection.}}
\VersePair{1:7}{And the Lord said to me, Say not, I am a child: for thou shalt go to all to whomsoever I shall send thee, and according to all the words that I shall command thee, thou shalt speak.}{1:7}{But Yahweh said to me, “Don’t say, ‘I am a child;’ for you must go to whomever I send you, and you must say whatever I command you.}
\VersePair{1:8}{Be not afraid before them: for I am with thee to deliver thee, saith the Lord.}{1:8}{Don’t be afraid because of them, for I am with you to rescue you,” says Yahweh.}
\VersePair{1:9}{And the Lord stretched forth his hand to me, and touched my mouth: and the Lord said to me, Behold, I have put my words into thy mouth.}{1:9}{Then Yahweh stretched out his hand and touched my mouth. Then Yahweh said to me, “Behold, I have put my words in your mouth.}
\VersePair{1:10}{Behold, I have appointed thee this day over nations and over kingdoms, to root out, and to pull down, and to destroy, and to rebuild, and to plant.}{1:10}{Behold, I have today set you over the nations and over the kingdoms, to uproot and to tear down, to destroy and to overthrow, to build and to plant.”}
\VersePair{1:11}{And the word of the Lord came to me, saying, What seest thou? And I said, A rod of an almond tree.}{1:11}{Moreover Yahweh’s word came to me, saying, “Jeremiah, what do you see?” I said, “I see a branch of an almond tree.”}
\VersePair{1:12}{And the Lord said to me, Thou hast well seen: for I have watched over my words to perform them.}{1:12}{Then Yahweh said to me, “You have seen well; for I watch over my word to perform it.”}
\VersePair{1:13}{And the word of the Lord came to me a second time, saying, What seest thou? And I said, A caldron on the fire; and the face of it is toward the north.}{1:13}{Yahweh’s word came to me the second time, saying, “What do you see?” I said, “I see a boiling cauldron; and it is tipping away from the north.”}
\VersePair{1:14}{And the Lord said to me, From the north shall flame forth evils upon all the inhabitants of the land.}{1:14}{Then Yahweh said to me, “Out of the north, evil will break out on all the inhabitants of the land.}
\VersePair{1:15}{For, behold, I call together all the kingdoms of the earth from the north, saith the Lord; and they shall come, and shall set each one his throne at the entrance of the gates of Jerusalem, and against all the walls round about her, and against all the cities of Juda.}{1:15}{{\raggedright For behold, I will call all the families of the kingdoms of the north,” says Yahweh. \poemline{1}{“They will come, and they will each set his throne at the entrance of the gates of Jerusalem,}\poemline{2}{and against all its walls all around, and against all the cities of Judah.}}}
\VersePair{1:16}{And I will speak to them in judgment, concerning all their iniquity, forasmuch as they have forsaken me, and sacrificed to strange gods, and worshipped the works of their own hands.}{1:16}{{\raggedright I will utter my judgments against them concerning all their wickedness, \poemline{2}{in that they have forsaken me,}\poemline{2}{and have burned incense to other gods,}\poemline{2}{and worshiped the works of their own hands.}}}
\VersePair{1:17}{And do thou gird up thy loins, and stand up, and speak all the words that I shall command thee: be not afraid of their face, neither be thou alarmed before them; for I am with thee to deliver thee, saith the Lord.}{1:17}{“You therefore put your belt on your waist, arise, and say to them all that I command you. Don’t be dismayed at them, lest I dismay you before them.}
\VersePair{1:18}{Behold, I have made thee this day as a strong city, and as a brazen wall, strong against all the kings of Juda, and the princes thereof, and the people of the land.}{1:18}{For behold, I have made you today a fortified city, an iron pillar, and bronze walls against the whole land— against the kings of Judah, against its princes, against its priests, and against the people of the land.}
\VersePair{1:19}{And they shall fight against thee; but they shall by no means prevail against thee; because I am with thee, to deliver thee, saith the Lord.}{1:19}{They will fight against you, but they will not prevail against you; for I am with you ”, says Yahweh, “to rescue you.”}
\end{paracol}
\ChapterHeading{2}
\begin{paracol}{2}
\VersePair{2:2}{And he said, Thus saith the Lord, I remember the kindness of thy youth, and the love of thine espousals, in following the Holy One of Israel, saith the Lord.}{2:1-2:2}{{\raggedright Yahweh’s word came to me, saying, “Go and proclaim in the ears of Jerusalem, saying, ‘Yahweh says, \poemline{1}{“I remember for you the kindness of your youth,}\poemline{2}{your love as a bride,}\poemline{1}{how you went after me in the wilderness,}\poemline{2}{in a land that was not sown.}}}
\VersePair{2:3}{Israel was the holy people to the Lord, and the first-fruits of his increase: all that devoured him shall offend; evils shall come upon them, saith the Lord.}{2:3}{{\raggedright Israel was holiness to Yahweh, \poemline{2}{the first fruits of his increase.}\poemline{1}{All who devour him will be held guilty.}\poemline{2}{Evil will come on them,”’ says Yahweh.”}}}
\VersePair{2:4}{Hear the word of the Lord, O house of Jacob, and every family of the house of Israel.}{2:4}{Hear Yahweh’s word, O house of Jacob, and all the families of the house of Israel!}
\VersePair{2:5}{Thus saith the Lord, What trespass have your fathers found in me, that they have revolted far from me, and gone after vanities, and become vain?}{2:5}{{\raggedright Yahweh says, \poemline{1}{“What unrighteousness have your fathers found in me,}\poemline{2}{that they have gone far from me,}\poemline{1}{and have walked after worthless vanity,}\poemline{2}{and have become worthless?}}}
\VersePair{2:6}{And they said not, Where is the Lord, who brought us up out of the land of Egypt, who guided us in the wilderness, in an untried and trackless land, in a land which no man at all went through, and no man dwelt there?}{2:6}{{\raggedright They didn’t say, ‘Where is Yahweh who brought us up out of the land of Egypt, \poemline{2}{who led us through the wilderness,}\poemline{2}{through a land of deserts and of pits,}\poemline{2}{through a land of drought and of the shadow of death,}\poemline{2}{through a land that no one passed through,}\poemline{2}{and where no man lived?’}}}
\VersePair{2:7}{And I brought you to Carmel, that ye should eat the fruits thereof, and the good thereof; and ye went in, and defiled my land, and made mine heritage an abomination.}{2:7}{{\raggedright I brought you into a plentiful land \poemline{2}{to eat its fruit and its goodness;}\poemline{1}{but when you entered, you defiled my land,}\poemline{2}{and made my heritage an abomination.}}}
\VersePair{2:8}{The priests said not, Where is the Lord? and they that held by the law knew me not: the shepherds also sinned against me, and the prophets prophesied by Baal, and went after that which profited not.}{2:8}{{\raggedright The priests didn’t say, ‘Where is Yahweh?’ \poemline{2}{and those who handle the law didn’t know me.}\poemline{1}{The rulers also transgressed against me,}\poemline{2}{and the prophets prophesied by Baal}\poemline{2}{and followed things that do not profit.}}}
\VersePair{2:9}{Therefore I will yet plead with you, and will plead with your children's children.}{2:9}{{\raggedright “Therefore I will yet contend with you,” says Yahweh, \poemline{2}{“and I will contend with your children’s children.}}}
\VersePair{2:10}{For go to the isles of the Chettians, and see; and send to Kedar, and observe accurately, and see if such things have been done;}{2:10}{{\raggedright For pass over to the islands of Kittim, and see. \poemline{2}{Send to Kedar, and consider diligently,}\poemline{2}{and see if there has been such a thing.}}}
\VersePair{2:11}{if the nations will change their gods, though they are not gods: but my people have changed their glory, for that from which they shall not be profited.}{2:11}{{\raggedright Has a nation changed its gods, \poemline{2}{which really are no gods?}\poemline{2}{But my people have changed their glory for that which doesn’t profit.}}}
\VersePair{2:12}{The heaven is amazed at this, and is very exceedingly horror-struck, saith the Lord.}{2:12}{{\raggedright “Be astonished, you heavens, at this \poemline{2}{and be horribly afraid.}\poemline{2}{Be very desolate,” says Yahweh.}}}
\VersePair{2:13}{For my people has committed two faults, and evil ones: they have forsaken me, the fountain of water of life, and hewn out for themselves broken cisterns, which will not be able to hold water.}{2:13}{{\raggedright “For my people have committed two evils: \poemline{2}{they have forsaken me, the spring of living waters,}\poemline{2}{and cut out cisterns for themselves: broken cisterns that can’t hold water.}}}
\VersePair{2:14}{Is Israel a servant, or a home-born slave? why has he become a spoil?}{2:14}{{\raggedright Is Israel a slave? \poemline{2}{Is he born into slavery?}\poemline{2}{Why has he become a captive?}}}
\VersePair{2:15}{The lions roared upon him, and uttered their voice, which have made his land a wilderness: and his cities are broken down, that they should not be inhabited.}{2:15}{{\raggedright The young lions have roared at him and raised their voices. \poemline{2}{They have made his land waste.}\poemline{2}{His cities are burned up, without inhabitant.}}}
\VersePair{2:16}{Also the children of Memphis and Taphnas have known thee, and mocked thee.}{2:16}{The children also of Memphis and Tahpanhes have broken the crown of your head.}
\VersePair{2:17}{Has not thy forsaking me brought these things upon thee? saith the Lord thy God.}{2:17}{{\raggedright “Haven’t you brought this on yourself, \poemline{2}{in that you have forsaken Yahweh your God, when he led you by the way? \footnote{2:17: The Hebrew word rendered “God” is “\texthebrew{אֱלֹהִ֑ים}” (Elohim).}}}}
\VersePair{2:18}{And now what hast thou to do with the way of Egypt, to drink the water of Geon? and what hast thou to do with the way of the Assyrians, to drink the water of rivers?}{2:18}{{\raggedright Now what do you gain by going to Egypt, to drink the waters of the Shihor? \poemline{2}{Or why do you go on the way to Assyria, to drink the waters of the River? \footnote{2:18: i.e., the Euphrates River}}}}
\VersePair{2:19}{Thine apostasy shall correct thee, and thy wickedness shall reprove thee: know then, and see, that thy forsaking me has been bitter to thee, saith the Lord thy God; and I have taken no pleasure in thee, saith the Lord thy God.}{2:19}{{\raggedright “Your own wickedness will correct you, \poemline{2}{and your backsliding will rebuke you.}\poemline{1}{Know therefore and see that it is an evil and bitter thing,}\poemline{2}{that you have forsaken Yahweh your God,}\poemline{2}{and that my fear is not in you,” says the Lord, Yahweh of Armies.}}}
\VersePair{2:20}{For of old thou hast broken thy yoke, and plucked asunder thy bands; and thou hast said, I will not serve thee, but will go upon every high hill, and under every shady tree, there will I indulge in my fornication.}{2:20}{{\raggedright “For long ago I broke off your yoke, \poemline{2}{and burst your bonds.}\poemline{1}{You said, ‘I will not serve;’}\poemline{2}{for on every high hill and under every green tree you bowed yourself,}\poemline{2}{playing the prostitute.}}}
\VersePair{2:21}{Yet I planted thee a fruitful vine, entirely of the right sort: how art thou a strange vine turned to bitterness!}{2:21}{{\raggedright Yet I had planted you a noble vine, \poemline{2}{a pure and faithful seed.}\poemline{2}{How then have you turned into the degenerate branches of a foreign vine to me?}}}
\VersePair{2:22}{Though thou shouldest wash thyself with nitre, and multiply to thyself soap, still thou art stained by thine iniquities before me, saith the Lord.}{2:22}{{\raggedright For though you wash yourself with lye, \poemline{2}{and use much soap,}\poemline{2}{yet your iniquity is marked before me,” says the Lord Yahweh.}}}
\VersePair{2:23}{How wilt thou say, I am not polluted, and have not gone after Baal? behold thy ways in the burial-ground, and know what thou hast done: her voice has howled in the evening:}{2:23}{{\raggedright “How can you say, ‘I am not defiled. \poemline{2}{I have not gone after the Baals ’?}\poemline{1}{See your way in the valley.}\poemline{2}{Know what you have done.}\poemline{1}{You are a swift dromedary traversing her ways,}}}
\VersePair{2:24}{she has extended her ways over the waters of the desert; she was hurried along by the lusts of her soul; she is given up to them, who will turn her back? none that seek her shall be weary; at the time of her humiliation they shall find her.}{2:24}{{\raggedright a wild donkey used to the wilderness, that sniffs the wind in her craving. \poemline{2}{When she is in heat, who can turn her away?}\poemline{2}{All those who seek her will not weary themselves. In her month, they will find her.}}}
\VersePair{2:25}{Withdraw thy foot from a rough way, and thy throat from thirst: but she said, I will strengthen myself: for she loved strangers, and went after them.}{2:25}{{\raggedright “Keep your feet from being bare, \poemline{2}{and your throat from thirst.}\poemline{1}{But you said, ‘It is in vain.}\poemline{2}{No, for I have loved strangers,}\poemline{2}{and I will go after them.’}}}
\VersePair{2:26}{As is the shame of a thief when he is caught, so shall the children of Israel be ashamed; they, and their kings, and their princes, and their priests, and their prophets.}{2:26}{{\raggedright As the thief is ashamed when he is found, \poemline{2}{so the house of Israel is ashamed —}\poemline{2}{they, their kings, their princes, their priests, and their prophets,}}}
\VersePair{2:27}{They said to a stock, Thou art my father; and to a stone, Thou hast begotten me: and they have turned their backs to me, and not their faces: yet in the time of their afflictions they will say, Arise, and save us.}{2:27}{{\raggedright who tell wood, ‘You are my father,’ \poemline{2}{and a stone, ‘You have given birth to me,’}\poemline{1}{for they have turned their back to me,}\poemline{2}{and not their face,}\poemline{2}{but in the time of their trouble they will say, ‘Arise, and save us!’}}}
\VersePair{2:28}{And where are thy gods, which thou madest for thyself? will they arise and save in the time of thine affliction? for according to the number of thy cities were thy gods, O Juda; and according to the number of the streets of Jerusalem they sacrificed to Baal.}{2:28}{{\raggedright “But where are your gods that you have made for yourselves? \poemline{2}{Let them arise, if they can save you in the time of your trouble,}\poemline{2}{for you have as many gods as you have towns, O Judah.}}}
\VersePair{2:29}{Wherefore do ye speak unto me? ye all have been ungodly, and ye all have transgressed against me, saith the Lord.}{2:29}{{\raggedright “Why will you contend with me? \poemline{2}{You all have transgressed against me,” says Yahweh.}}}
\VersePair{2:30}{In vain have I smitten your children; ye have not received correction: a sword has devoured your prophets as a destroying lion; yet ye feared not.}{2:30}{{\raggedright “I have struck your children in vain. \poemline{2}{They received no correction.}\poemline{1}{Your own sword has devoured your prophets,}\poemline{2}{like a destroying lion.}}}
\VersePair{2:31}{Hear ye the word of the Lord: thus saith the Lord, Have I been a wilderness or a dry land to Israel? wherefore has my people said, We will not be ruled over, and will not come to thee any more?}{2:31}{{\raggedright Generation, consider Yahweh’s word. \poemline{2}{Have I been a wilderness to Israel?}\poemline{2}{Or a land of thick darkness?}\poemline{1}{Why do my people say, ‘We have broken loose.}\poemline{2}{We will come to you no more ’?}}}
\VersePair{2:32}{Will a bride forget her ornaments, or a virgin her girdle? but my people has forgotten me days without number.}{2:32}{{\raggedright “Can a virgin forget her ornaments, \poemline{2}{or a bride her attire?}\poemline{2}{Yet my people have forgotten me for days without number.}}}
\VersePair{2:33}{What fair device wilt thou yet employ in thy ways, so as to seek love? it shall not be so; moreover thou hast done wickedly in corrupting thy ways;}{2:33}{{\raggedright How well you prepare your way to seek love! \poemline{2}{Therefore you have even taught the wicked women your ways.}}}
\VersePair{2:34}{and in thine hands has been found the blood of innocent souls; I have not found them in holes, but on every oak.}{2:34}{{\raggedright Also the blood of the souls of the innocent poor is found in your skirts. \poemline{2}{You didn’t find them breaking in,}\poemline{2}{but it is because of all these things.}}}
\VersePair{2:35}{Yet thou saidst, I am innocent: only let his wrath be turned away from me. Behold, I will plead with thee, whereas thou sayest, I have not sinned.}{2:35}{{\raggedright “Yet you said, ‘I am innocent. \poemline{2}{Surely his anger has turned away from me.’}\poemline{1}{“Behold, I will judge you,}\poemline{2}{because you say, ‘I have not sinned.’}}}
\VersePair{2:36}{For thou hast been so exceedingly contemptuous as to repeat thy ways; but thou shalt be ashamed of Egypt, as thou wast ashamed of Assur.}{2:36}{{\raggedright Why do you go about so much to change your ways? \poemline{2}{You will be ashamed of Egypt also,}\poemline{2}{as you were ashamed of Assyria.}}}
\VersePair{2:37}{For thou shalt go forth thence also with thine hands upon thine head; for the Lord has rejected thine hope, and thou shalt not prosper in it.}{2:37}{{\raggedright You will also leave that place with your hands on your head; \poemline{2}{for Yahweh has rejected those in whom you trust,}\poemline{2}{and you won’t prosper with them.}}}
\end{paracol}
\ChapterHeading{3}
\begin{paracol}{2}
\VersePair{3:1}{If a man put away his wife, and she depart from him, and become another man's, shall she return to him any more at all? shall not that woman be utterly defiled? yet thou hast gone a-whoring with many shepherds, and hast returned to me, saith the Lord.}{3:1}{“They say, ‘If a man puts away his wife, and she goes from him, and becomes another man’s, should he return to her again?’ Wouldn’t that land be greatly polluted? But you have played the prostitute with many lovers; yet return again to me,” says Yahweh.}
\VersePair{3:2}{Lift up thine eyes to look straight forward, and see where thou hast not been utterly defiled. Thou hast sat for them by the wayside as a deserted crow, and hast defiled the land with thy fornications and thy wickedness.}{3:2}{“Lift up your eyes to the bare heights, and see! Where have you not been lain with? You have sat waiting for them by the road, as an Arabian in the wilderness. You have polluted the land with your prostitution and with your wickedness.}
\VersePair{3:3}{And thou didst retain many shepherds for a stumbling-block to thyself: thou hadst a whore's face, thou didst become shameless toward all.}{3:3}{Therefore the showers have been withheld and there has been no latter rain; yet you have had a prostitute’s forehead and you refused to be ashamed.}
\VersePair{3:4}{Hast thou not called me as it were a home, and the father and guide of thy virgin-time?}{3:4}{Will you not from this time cry to me, ‘My Father, you are the guide of my youth!’?}
\VersePair{3:5}{Will God's anger continue for ever, or be preserved to the end? Behold, thou hast spoken and done these bad things, and hadst power to do them.}{3:5}{“‘Will he retain his anger forever? Will he keep it to the end?’ Behold, you have spoken and have done evil things, and have had your way.”}
\VersePair{3:6}{And the Lord said to me in the days of Josias the king, Hast thou seen what things the house of Israel has done to me? they have gone on every high mountain, and under every shady tree, and have committed fornication there.}{3:6}{Moreover, Yahweh said to me in the days of Josiah the king, “Have you seen that which backsliding Israel has done? She has gone up on every high mountain and under every green tree, and has played the prostitute there.}
\VersePair{3:7}{And I said after she had committed all these acts of fornication, Turn again to me. Yet she returned not. And faithless Juda saw her faithlessness.}{3:7}{I said after she had done all these things, ‘She will return to me;’ but she didn’t return, and her treacherous sister Judah saw it.}
\VersePair{3:8}{And I saw that (for all the sins of which she was convicted, wherein the house of Israel committed adultery, and I put her away, and gave into her hands a bill of divorcement,) yet faithless Juda feared not, but went and herself also committed fornication.}{3:8}{I saw when, for this very cause, that backsliding Israel had committed adultery, I had put her away and given her a certificate of divorce, yet treacherous Judah, her sister, had no fear, but she also went and played the prostitute.}
\VersePair{3:9}{And her fornication was nothing accounted of; and she committed adultery with wood and stone.}{3:9}{Because she took her prostitution lightly, the land was polluted, and she committed adultery with stones and with wood.}
\VersePair{3:10}{And for all these things faithless Juda turned not to me with all her heart, but falsely.}{3:10}{Yet for all this her treacherous sister, Judah, has not returned to me with her whole heart, but only in pretense,” says Yahweh.}
\VersePair{3:11}{And the Lord said to me, Israel has justified himself more than faithless Juda.}{3:11}{Yahweh said to me, “Backsliding Israel has shown herself more righteous than treacherous Judah.}
\VersePair{3:12}{Go and read these words toward the north, and thou shalt say, Return to me, O house of Israel, saith the Lord; and I will not set my face against you: for I am merciful, saith the Lord, and I will not be angry with you for ever.}{3:12}{Go, and proclaim these words toward the north, and say, ‘Return, you backsliding Israel,’ says Yahweh; ‘I will not look in anger on you, for I am merciful,’ says Yahweh. ‘I will not keep anger forever.}
\VersePair{3:13}{Nevertheless, know thine iniquity, that thou hast sinned against the Lord thy God, and hast scattered thy ways to strangers under every shady tree, but thou didst not hearken to my voice, saith the Lord.}{3:13}{Only acknowledge your iniquity, that you have transgressed against Yahweh your God, and have scattered your ways to the strangers under every green tree, and you have not obeyed my voice,’” says Yahweh.}
\VersePair{3:14}{Turn, ye children that have revolted, saith the Lord; for I will rule over you: and I will take you one of a city, and two of a family, and I will bring you in to Sion:}{3:14}{“Return, backsliding children,” says Yahweh, “for I am a husband to you. I will take one of you from a city, and two from a family, and I will bring you to Zion.}
\VersePair{3:15}{and I will give you shepherds after my heart, and they shall certainly tend you with knowledge.}{3:15}{I will give you shepherds according to my heart, who will feed you with knowledge and understanding.}
\VersePair{3:16}{And it shall come to pass that when ye are multiplied and increased upon the land, saith the Lord, in those days they shall say no more, The ark of the covenant of the Holy One of Israel: it shall not come to mind; it shall not be named; neither shall it be visited; nor shall this be done any more.}{3:16}{It will come to pass, when you are multiplied and increased in the land in those days,” says Yahweh, “they will no longer say, ‘the ark of Yahweh’s covenant!’ It will not come to mind. They won’t remember it. They won’t miss it, nor will another be made.}
\VersePair{3:17}{In those days and at that time they shall call Jerusalem the throne of the Lord; and all the nations shall be gathered to it: and they shall not walk any more after the imaginations of their evil heart.}{3:17}{At that time they will call Jerusalem ‘Yahweh’s Throne;’ and all the nations will be gathered to it, to Yahweh’s name, to Jerusalem. They will no longer walk after the stubbornness of their evil heart.}
\VersePair{3:18}{In those days the house of Juda shall come together to the house of Israel, and they shall come, together, from the land of the north, and from all the countries, to the land, which I caused their fathers to inherit.}{3:18}{In those days the house of Judah will walk with the house of Israel, and they will come together out of the land of the north to the land that I gave for an inheritance to your fathers.}
\VersePair{3:19}{And I said, So be it, Lord, for thou saidst, I will set thee among children, and will give thee a choice land, the inheritance of the Almighty God of the Gentiles: and I said, Ye shall call me Father; and ye shall not turn away from me.}{3:19}{“But I said, ‘How I desire to put you among the children, and give you a pleasant land, a goodly heritage of the armies of the nations!’ and I said, ‘You shall call me “My Father ”, and shall not turn away from following me.’}
\VersePair{3:20}{But as a wife acts treacherously against her husband, so has the house of Israel dealt treacherously against me, saith the Lord.}{3:20}{“Surely as a wife treacherously departs from her husband, so you have dealt treacherously with me, house of Israel,” says Yahweh.}
\VersePair{3:21}{A voice from the lips was heard, even of weeping and supplication of the children of Israel: for they have dealt unrighteously in their ways, they have forgotten God their Holy One.}{3:21}{A voice is heard on the bare heights, the weeping and the petitions of the children of Israel; because they have perverted their way, they have forgotten Yahweh their God.}
\VersePair{3:22}{Turn, ye children that are given to turning, and I will heal your bruises. Behold, we will be thy servants; for thou art the Lord our God.}{3:22}{Return, you backsliding children, and I will heal your backsliding. “Behold, we have come to you; for you are Yahweh our God.}
\VersePair{3:23}{Truly the hills and the strength of the mountains were a lying refuge: but by the Lord our God is the salvation of Israel.}{3:23}{Truly help from the hills, the tumult on the mountains, is in vain. Truly the salvation of Israel is in Yahweh our God.}
\VersePair{3:24}{But shame has consumed the labours of our fathers from our youth; their sheep and their calves, and their sons and their daughters.}{3:24}{But the shameful thing has devoured the labor of our fathers from our youth, their flocks and their herds, their sons and their daughters.}
\VersePair{3:25}{We have lain down in our shame, and our disgrace has covered us: because we and our fathers have sinned before our God, from our youth until this day; and we have not hearkened to the voice of the Lord our God.}{3:25}{Let us lie down in our shame, and let our confusion cover us; for we have sinned against Yahweh our God, we and our fathers, from our youth even to this day. We have not obeyed Yahweh our God’s voice.”}
\end{paracol}
\ChapterHeading{4}
\begin{paracol}{2}
\VersePair{4:1}{If Israel will return to me, saith the Lord, he shall return: and if he will remove his abominations out of his mouth, and fear before me, and swear,}{4:1}{“If you will return, Israel,” says Yahweh, “if you will return to me, and if you will put away your abominations out of my sight; then you will not be removed;}
\VersePair{4:2}{The Lord lives, with truth, in judgment and righteousness, then shall nations bless by him, and by him they shall praise God in Jerusalem.}{4:2}{and you will swear, ‘As Yahweh lives,’ in truth, in justice, and in righteousness. The nations will bless themselves in him, and they will glory in him.”}
\VersePair{4:3}{For thus saith the Lord to the men of Juda, and to the inhabitants of Jerusalem, Break up fresh ground for yourselves, and sow not among thorns.}{4:3}{For Yahweh says to the men of Judah and to Jerusalem, “Break up your fallow ground, and don’t sow among thorns.}
\VersePair{4:4}{Circumcise yourselves to your God, and circumcise your hardness of heart, ye men of Juda, and inhabitants of Jerusalem: lest my wrath go forth as fire, and burn, and there be none to quench it, because of the evil of your devices.}{4:4}{Circumcise yourselves to Yahweh, and take away the foreskins of your heart, you men of Judah and inhabitants of Jerusalem; lest my wrath go out like fire, and burn so that no one can quench it, because of the evil of your doings.}
\VersePair{4:5}{Declare ye in Juda, and let it be heard in Jerusalem: say ye, Sound the trumpet in the land; cry ye aloud: say ye, Gather yourselves together, and let us enter into the fortified cities.}{4:5}{Declare in Judah, and publish in Jerusalem; and say, ‘Blow the trumpet in the land!’ Cry aloud and say, ‘Assemble yourselves! Let’s go into the fortified cities!’}
\VersePair{4:6}{Gather up your wares and flee to Sion: hasten, stay not: for I will bring evils from the north, and great destruction.}{4:6}{Set up a standard toward Zion. Flee for safety! Don’t wait; for I will bring evil from the north, and a great destruction.”}
\VersePair{4:7}{The lion is gone up from his lair, he has roused himself to the destruction of the nations, and has gone forth out of his place, to make the land desolate; and the cities shall be destroyed, so as to be without inhabitant.}{4:7}{A lion has gone up from his thicket, and a destroyer of nations. He is on his way. He has gone out from his place, to make your land desolate, that your cities be laid waste, without inhabitant.}
\VersePair{4:8}{For these things gird yourselves with sackclothes, and lament, and howl: for the anger of the Lord is not turned away from you.}{4:8}{For this, clothe yourself with sackcloth, lament and wail; for the fierce anger of Yahweh hasn’t turned back from us.}
\VersePair{4:9}{And it shall come to pass in that day, saith the Lord, that the heart of the king shall perish, and the heart of the princes; and the priests shall be amazed, and the prophets shall wonder.}{4:9}{“It will happen at that day,” says Yahweh, “that the heart of the king will perish, along with the heart of the princes. The priests will be astonished, and the prophets will wonder.”}
\VersePair{4:10}{And I said, O sovereign Lord, verily thou hast greatly deceived this people and Jerusalem, saying, There shall be peace; whereas behold, the sword has reached even to their soul.}{4:10}{Then I said, “Ah, Lord Yahweh! Surely you have greatly deceived this people and Jerusalem, saying, ‘You will have peace;’ whereas the sword reaches to the heart.”}
\VersePair{4:11}{At that time they shall say to this people and to Jerusalem, There is a spirit of error in the wilderness: the way of the daughter of my people is not to purity, nor to holiness.}{4:11}{At that time it will be said to this people and to Jerusalem, “A hot wind blows from the bare heights in the wilderness toward the daughter of my people, not to winnow, nor to cleanse.}
\VersePair{4:12}{But a spirit of full vengeance shall come upon me; and now I declare my judgments against them.}{4:12}{A full wind from these will come for me. Now I will also utter judgments against them.”}
\VersePair{4:13}{Behold, he shall come up as a cloud, and his chariots as a tempest: his horses are swifter than eagles. Woe unto us! for we are in misery.}{4:13}{Behold, he will come up as clouds, and his chariots will be as the whirlwind. His horses are swifter than eagles. Woe to us! For we are ruined.}
\VersePair{4:14}{Cleanse thine heart from wickedness, O Jerusalem, that thou mayest be saved: how long will thy grievous thoughts be within thee?}{4:14}{Jerusalem, wash your heart from wickedness, that you may be saved. How long will your evil thoughts lodge within you?}
\VersePair{4:15}{For a voice of one publishing from Dan shall come, and trouble out of mount Ephraim shall be heard of.}{4:15}{For a voice declares from Dan, and publishes evil from the hills of Ephraim:}
\VersePair{4:16}{Remind ye the nations; behold, they are come: proclaim it in Jerusalem, that bands are approaching from a land afar off, and have uttered their voice against the cities of Juda.}{4:16}{“Tell the nations, behold, publish against Jerusalem, ‘Watchers come from a far country, and raise their voice against the cities of Judah.}
\VersePair{4:17}{As keepers of a field, they have surrounded her; because thou, saith the Lord, hast neglected me.}{4:17}{As keepers of a field, they are against her all around, because she has been rebellious against me,’” says Yahweh.}
\VersePair{4:18}{Thy ways and thy devices have brought these things upon thee; this is thy wickedness, for it is bitter, for it has reached to thy heart.}{4:18}{“Your way and your doings have brought these things to you. This is your wickedness, for it is bitter, for it reaches to your heart.”}
\VersePair{4:19}{I am pained in my bowels, my bowels, and the sensitive powers of my heart; my soul is in great commotion, my heart is torn: I will not be silent, for my soul has heard the sound of a trumpet, the cry of war, and of distress: it calls on destruction;}{4:19}{My anguish, my anguish! I am pained at my very heart! My heart trembles within me. I can’t hold my peace, because you have heard, O my soul, the sound of the trumpet, the alarm of war.}
\VersePair{4:20}{for all the land is distressed: suddenly my tabernacle is distressed, my curtains have been rent asunder.}{4:20}{Destruction on destruction is decreed, for the whole land is laid waste. Suddenly my tents are destroyed, and my curtains gone in a moment.}
\VersePair{4:21}{How long shall I see fugitives, and hear the sound of the trumpet?}{4:21}{How long will I see the standard and hear the sound of the trumpet?}
\VersePair{4:22}{For the princes of my people have not known me, they are foolish and unwise children: they are wise to do evil, but how to do good they have not known.}{4:22}{“For my people are foolish. They don’t know me. They are foolish children, and they have no understanding. They are skillful in doing evil, but they don’t know how to do good.”}
\VersePair{4:23}{I looked upon the earth, and, behold, it was not; and to the sky, and there was no light in it.}{4:23}{I saw the earth and, behold, it was waste and void, and the heavens, and they had no light.}
\VersePair{4:24}{I beheld the mountains, and they trembled, and I saw all the hills in commotion.}{4:24}{I saw the mountains, and behold, they trembled, and all the hills moved back and forth.}
\VersePair{4:25}{I looked, and, behold, there was no man, and all the birds of the sky were scared.}{4:25}{I saw, and behold, there was no man, and all the birds of the sky had fled.}
\VersePair{4:26}{I saw, and, behold, Carmel was desert, and all the cities were burnt with fire at the presence of the Lord, and at the presence of his fierce anger they were utterly destroyed.}{4:26}{I saw, and behold, the fruitful field was a wilderness, and all its cities were broken down at the presence of Yahweh, before his fierce anger.}
\VersePair{4:27}{Thus saith the Lord, The whole land shall be desolate; but I will not make a full end.}{4:27}{For Yahweh says, “The whole land will be a desolation; yet I will not make a full end.}
\VersePair{4:28}{For these things let the earth mourn, and let the sky be dark above: for I have spoken, and I will not repent; I have purposed, and I will not turn back from it.}{4:28}{For this the earth will mourn, and the heavens above be black, because I have spoken it. I have planned it, and I have not repented, neither will I turn back from it.”}
\VersePair{4:29}{The whole land has recoiled from the noise of the horseman and the bent bow; they have gone into the caves, and have hidden themselves in the groves, and have gone up upon the rocks: every city was abandoned, no man dwelt in them.}{4:29}{Every city flees for the noise of the horsemen and archers. They go into the thickets and climb up on the rocks. Every city is forsaken, and not a man dwells therein.}
\VersePair{4:30}{And what wilt thou do? Though thou clothe thyself with scarlet, and adorn thyself with golden ornaments; though thou adorn thine eyes with stibium, thy beauty will be in vain: thy lovers have rejected thee, they seek thy life.}{4:30}{You, when you are made desolate, what will you do? Though you clothe yourself with scarlet, though you deck yourself with ornaments of gold, though you enlarge your eyes with makeup, you make yourself beautiful in vain. Your lovers despise you. They seek your life.}
\VersePair{4:31}{For I have heard thy groaning as the voice of a woman in travail, as of her that brings forth her first child; the voice of the daughter of Zion shall fail through weakness, and she shall lose the strength of her hands, saying, Woe is me! for my soul faints because of the slain.}{4:31}{For I have heard a voice as of a woman in travail, the anguish as of her who gives birth to her first child, the voice of the daughter of Zion, who gasps for breath, who spreads her hands, saying, “Woe is me now! For my soul faints before the murderers.”}
\end{paracol}
\ChapterHeading{5}
\begin{paracol}{2}
\VersePair{5:1}{Run ye about in the streets of Jerusalem, and see, and know, and seek in her broad places, if ye can find one, if there is any one that does judgment, and seeks faithfulness; and I will pardon them, saith the Lord.}{5:1}{“Run back and forth through the streets of Jerusalem, and see now, and know, and seek in its wide places, if you can find a man, if there is anyone who does justly, who seeks truth, then I will pardon her.}
\VersePair{5:2}{The Lord lives, they say; do they not therefore swear falsely?}{5:2}{Though they say, ‘As Yahweh lives,’ surely they swear falsely.”}
\VersePair{5:3}{O Lord, thine eyes are upon faithfulness: thou hast scourged them, but they have not grieved; thou hast consumed them; but they would not receive correction: they have made their faces harder than a rock; and they would not return.}{5:3}{O Yahweh, don’t your eyes look on truth? You have stricken them, but they were not grieved. You have consumed them, but they have refused to receive correction. They have made their faces harder than a rock. They have refused to return.}
\VersePair{5:4}{Then I said, It may be they are poor; for they are weak, for they know not the way of the Lord, or the judgment of God.}{5:4}{Then I said, “Surely these are poor. They are foolish; for they don’t know Yahweh’s way, nor the law of their God.}
\VersePair{5:5}{I will go to the rich men, and will speak to them; for they have known the way of the Lord, and the judgment of God: but, behold, with one consent they have broken the yoke, they have burst the bonds.}{5:5}{I will go to the great men and will speak to them, for they know the way of Yahweh, and the law of their God.” But these with one accord have broken the yoke, and burst the bonds.}
\VersePair{5:6}{Therefore has a lion out of the forest smitten them, and a wolf has destroyed them even to their houses, and a leopard has watched against their cities: all that go forth from them shall be hunted: for they have multiplied their ungodliness, they have strengthened themselves in their revoltings.}{5:6}{Therefore a lion out of the forest will kill them. A wolf of the evenings will destroy them. A leopard will watch against their cities. Everyone who goes out there will be torn in pieces, because their transgressions are many and their backsliding has increased.}
\VersePair{5:7}{In what way shall I forgive thee for these things? Thy sons have forsaken me, and sworn by them that are no gods: and I fed them to the full, and they committed adultery, and lodged in harlots'houses.}{5:7}{“How can I pardon you? Your children have forsaken me, and sworn by what are no gods. When I had fed them to the full, they committed adultery, and assembled themselves in troops at the prostitutes’houses.}
\VersePair{5:8}{They became as wanton horses: they neighed each one after his neighbour's wife.}{5:8}{They were as fed horses roaming at large. Everyone neighed after his neighbor’s wife.}
\VersePair{5:9}{Shall I not visit for these things? saith the Lord: and shall not my soul be avenged on such a nation as this.}{5:9}{Shouldn’t I punish them for these things?” says Yahweh. “Shouldn’t my soul be avenged on such a nation as this?}
\VersePair{5:10}{Go up upon her battlements, and break them down; but make not a full end: leave her buttresses; for they are the Lord's.}{5:10}{“Go up on her walls, and destroy, but don’t make a full end. Take away her branches, for they are not Yahweh’s.}
\VersePair{5:11}{For the house of Israel have indeed dealt treacherously against me, saith the Lord: the house of Juda also}{5:11}{For the house of Israel and the house of Judah have dealt very treacherously against me,” says Yahweh.}
\VersePair{5:12}{have lied to their Lord, and they have said, These things are not so; no evils shall come upon us; and we shall not see sword or famine.}{5:12}{They have denied Yahweh, and said, “It is not he. Evil won’t come on us. We won’t see sword or famine.}
\VersePair{5:13}{Our prophets became wind, and the word of the Lord was not in them.}{5:13}{The prophets will become wind, and the word is not in them. Thus it will be done to them.”}
\VersePair{5:14}{Therefore thus saith the Lord Almighty, Because ye have spoken this word, behold, I have made my words in thy mouth fire, and this people wood, and it shall devour them.}{5:14}{Therefore Yahweh, the God of Armies says, “Because you speak this word, behold, I will make my words in your mouth fire, and this people wood, and it will devour them.}
\VersePair{5:15}{Behold, I will bring upon you a nation from far, O house of Israel, saith the Lord; a nation the sound of whose language one shall not understand.}{5:15}{Behold, I will bring a nation on you from far away, house of Israel,” says Yahweh. “It is a mighty nation. It is an ancient nation, a nation whose language you don’t know and don’t understand what they say.}
\VersePair{5:16}{They are all mighty men:}{5:16}{Their quiver is an open tomb. They are all mighty men.}
\VersePair{5:17}{and they shall devour your harvest, and your bread; and shall devour your sons, and your daughters; and they shall devour your sheep, and your calves, and devour your vineyards, and your fig-plantations, and your olive yards: and they shall utterly destroy your strong cities, wherein ye trusted, with the sword.}{5:17}{They will eat up your harvest and your bread, which your sons and your daughters should eat. They will eat up your flocks and your herds. They will eat up your vines and your fig trees. They will beat down your fortified cities in which you trust with the sword.}
\VersePair{5:18}{And it shall come to pass in those days, saith the Lord thy God, that I will not utterly destroy you.}{5:18}{“But even in those days,” says Yahweh, “I will not make a full end of you.}
\VersePair{5:19}{And it shall come to pass, when ye shall say, Wherefore has the Lord our God done all these things to us? that thou shalt say to them, Because ye served strange gods in your land, so shall ye serve strangers in a land that is not yours.}{5:19}{It will happen when you say, ‘Why has Yahweh our God done all these things to us?’ Then you shall say to them, ‘Just as you have forsaken me and served foreign gods in your land, so you will serve strangers in a land that is not yours.’}
\VersePair{5:20}{Proclaim these things to the house of Jacob, and let them be heard in the house of Juda.}{5:20}{“Declare this in the house of Jacob, and publish it in Judah, saying,}
\VersePair{5:21}{Hear ye now these things, O foolish and senseless people; who have eyes, and see not; and have ears, and hear not:}{5:21}{‘Hear this now, foolish people without understanding, who have eyes, and don’t see, who have ears, and don’t hear:}
\VersePair{5:22}{will ye not be afraid of me? saith the Lord; and will ye not fear before me, who have set the sand for a bound to the sea, as a perpetual ordinance, and it shall not pass it: yea, it shall rage, but not prevail; and its waves shall roar, but not pass over it.}{5:22}{Don’t you fear me?’ says Yahweh; ‘Won’t you tremble at my presence, who have placed the sand for the bound of the sea by a perpetual decree, that it can’t pass it? Though its waves toss themselves, yet they can’t prevail. Though they roar, they still can’t pass over it.’}
\VersePair{5:23}{But this people has a disobedient and rebellious heart; and they have turned aside and gone back:}{5:23}{“But this people has a revolting and a rebellious heart. They have revolted and gone.}
\VersePair{5:24}{and they have not said in their heart, Let us fear now the Lord our God, who gives us the early and latter rain, according to the season of the fulfilment of the ordinance of harvest, and has preserved it for us.}{5:24}{They don’t say in their heart, ‘Let’s now fear Yahweh our God, who gives rain, both the former and the latter, in its season, who preserves to us the appointed weeks of the harvest.’}
\VersePair{5:25}{Your transgressions have turned away these things, and your sins have removed good things from you.}{5:25}{“Your iniquities have turned away these things, and your sins have withheld good from you.}
\VersePair{5:26}{For among my people were found ungodly men; and they have set snares to destroy men, and have caught them.}{5:26}{For wicked men are found among my people. They watch, as fowlers lie in wait. They set a trap. They catch men.}
\VersePair{5:27}{As a snare which has been set is full of birds, so are their houses full of deceit: therefore have they grown great, and become rich:}{5:27}{As a cage is full of birds, so are their houses full of deceit. Therefore they have become great, and grew rich.}
\VersePair{5:28}{and they have transgressed the rule of judgment; they have not judged the cause of the orphan, nor have they judged the cause of the widow.}{5:28}{They have grown fat. They shine; yes, they excel in deeds of wickedness. They don’t plead the cause, the cause of the fatherless, that they may prosper; and they don’t defend the rights of the needy.}
\VersePair{5:29}{Shall I not visit for these things? saith the Lord: and shall not my soul be avenged on such a nation as this?}{5:29}{“Shouldn’t I punish for these things?” says Yahweh. “Shouldn’t my soul be avenged on such a nation as this?}
\VersePair{5:30}{Shocking and horrible deeds have been done on the land;}{5:30}{“An astonishing and horrible thing has happened in the land.}
\VersePair{5:31}{the prophets utter unrighteous prophecies, and the priests have clapped their hands: and my people has loved to have it thus: and what will ye do for the future?}{5:31}{The prophets prophesy falsely, and the priests rule by their own authority; and my people love to have it so. What will you do in the end of it?}
\end{paracol}
\ChapterHeading{6}
\begin{paracol}{2}
\VersePair{6:1}{Strengthen yourselves, ye children of Benjamin, to flee out of the midst of Jerusalem, and sound an alarm with the trumpet in Thecue, and set up a signal over Bæthacharma: for evil threatens from the north, and a great destruction is coming.}{6:1}{“Flee for safety, you children of Benjamin, out of the middle of Jerusalem! Blow the trumpet in Tekoa and raise up a signal on Beth Haccherem, for evil looks out from the north with a great destruction.}
\VersePair{6:2}{And thy pride, O daughter of Sion, shall be taken away.}{6:2}{I will cut off the beautiful and delicate one, the daughter of Zion.}
\VersePair{6:3}{The shepherds and their flocks shall come to her; and they shall pitch their tents against her round about, and shall feed their flocks each with his hand.}{6:3}{Shepherds with their flocks will come to her. They will pitch their tents against her all around. They will feed everyone in his place.”}
\VersePair{6:4}{Prepare yourselves for war against her; rise up, and let us go up against her at noon. Woe to us! for the day has gone down, for the shadows of the day fail.}{6:4}{“Prepare war against her! Arise! Let’s go up at noon. Woe to us! For the day declines, for the shadows of the evening are stretched out.}
\VersePair{6:5}{Rise, and let us go up against her by night, and destroy her foundations.}{6:5}{Arise! Let’s go up by night, and let’s destroy her palaces.”}
\VersePair{6:6}{For thus saith the Lord, Hew down her trees, array a numerous force against Jerusalem. O false city; there is all oppression in her.}{6:6}{For Yahweh of Armies said, “Cut down trees, and cast up a mound against Jerusalem. This is the city to be visited. She is filled with oppression within herself.}
\VersePair{6:7}{As a cistern cools water, so her wickedness cools her, ungodliness and misery shall be heard in her, as continually before her.}{6:7}{As a well produces its waters, so she produces her wickedness. Violence and destruction is heard in her. Sickness and wounds are continually before me.}
\VersePair{6:8}{Thou shalt be chastened, O Jerusalem, with pain and the scourge, lest my soul depart from thee; lest I make thee a desert land, which shall not be inhabited.}{6:8}{Be instructed, Jerusalem, lest my soul be alienated from you, lest I make you a desolation, an uninhabited land.”}
\VersePair{6:9}{For thus saith the Lord, Glean, glean thoroughly as a vine the remnant of Israel: turn back your hands as a grape-gatherer to his basket.}{6:9}{Yahweh of Armies says, “They will thoroughly glean the remnant of Israel like a vine. Turn again your hand as a grape gatherer into the baskets.”}
\VersePair{6:10}{To whom shall I speak, and testify, that he may hearken? behold, thine ears are uncircumcised, and they shall not be able to hear: behold, the word of the Lord is become to them a reproach, they will not at all desire it.}{6:10}{To whom should I speak and testify, that they may hear? Behold, their ear is uncircumcised, and they can’t listen. Behold, Yahweh’s word has become a reproach to them. They have no delight in it.}
\VersePair{6:11}{And I allowed my wrath to come to the full, yet I kept it in, and did not utterly destroy them: I will pour it out on the children without, and on the assembly of young men together: for man and woman shall be taken together, the old man with him that is full of days.}{6:11}{{\raggedright Therefore I am full of Yahweh’s wrath. I am weary with holding it in. \poemline{1}{“Pour it out on the children in the street,}\poemline{2}{and on the assembly of young men together;}\poemline{1}{for even the husband with the wife will be taken,}\poemline{2}{the aged with him who is full of days.}}}
\VersePair{6:12}{And their houses shall be turned to others, with their fields and their wives together: for I will stretch out my hand upon the inhabitants of this land, saith the Lord.}{6:12}{{\raggedright Their houses will be turned to others, \poemline{2}{their fields and their wives together;}\poemline{1}{for I will stretch out my hand on the inhabitants of the land, says Yahweh.”}}}
\VersePair{6:13}{For from the least of them even to the greatest they have all committed iniquity; from the priest even to the false prophet they have all wrought falsely.}{6:13}{{\raggedright “For from their least even to their greatest, everyone is given to covetousness. \poemline{2}{From the prophet even to the priest, everyone deals falsely.}}}
\VersePair{6:14}{And they healed the breach of my people imperfectly, making light of it, and saying, Peace, peace: and where is peace?}{6:14}{{\raggedright They have healed also the hurt of my people superficially, \poemline{2}{saying, ‘Peace, peace!’ when there is no peace.}}}
\VersePair{6:15}{They were ashamed because they failed; yet they were not ashamed as those who are truly ashamed, and they knew not their own disgrace: therefore shall they utterly fall when they do fall, and in the time of visitation shall they perish, said the Lord.}{6:15}{{\raggedright Were they ashamed when they had committed abomination? \poemline{2}{No, they were not at all ashamed, neither could they blush.}\poemline{1}{Therefore they will fall among those who fall.}\poemline{2}{When I visit them, they will be cast down,” says Yahweh.}}}
\VersePair{6:16}{Thus saith the Lord, Stand ye in the ways, and see, and ask for the old paths of the Lord; and see what is the good way, and walk in it, and ye shall find purification for your souls. But they said, We will not walk in them.}{6:16}{Yahweh says, “Stand in the ways and see, and ask for the old paths, ‘Where is the good way?’ and walk in it, and you will find rest for your souls. But they said, ‘We will not walk in it.’}
\VersePair{6:17}{I have set watchmen over you, saying, Hear ye the sound of the trumpet. But they said, We will not hear it.}{6:17}{I set watchmen over you, saying, ‘Listen to the sound of the trumpet!’ But they said, ‘We will not listen!’}
\VersePair{6:18}{Therefore have the nations heard, and they that feed their flocks.}{6:18}{Therefore hear, you nations, and know, congregation, what is among them.}
\VersePair{6:19}{Hear, O earth: behold, I will bring evils upon this people, even the fruit of their rebellions; for they have not heeded my words, and they have rejected my law.}{6:19}{Hear, earth! Behold, I will bring evil on this people, even the fruit of their thoughts, because they have not listened to my words; and as for my law, they have rejected it.}
\VersePair{6:20}{Wherefore do ye bring me frankincense from Saba, and cinnamon from a land afar off? your whole-burnt-offerings are not acceptable, and your sacrifices have not been pleasant to me.}{6:20}{To what purpose does frankincense from Sheba come to me, and the sweet cane from a far country? Your burnt offerings are not acceptable, and your sacrifices are not pleasing to me.”}
\VersePair{6:21}{Therefore thus saith the Lord, Behold, I will bring weakness upon this people, and the fathers and sons shall be weak together; the neighbour and his friend shall perish.}{6:21}{Therefore Yahweh says, “Behold, I will lay stumbling blocks before this people. The fathers and the sons together will stumble against them. The neighbor and his friend will perish.”}
\VersePair{6:22}{Thus saith the Lord, Behold, a people comes from the north, and nations shall be stirred up from the end of the earth.}{6:22}{Yahweh says, “Behold, a people comes from the north country. A great nation will be stirred up from the uttermost parts of the earth.}
\VersePair{6:23}{They shall lay hold on bow and spear; the people is fierce, and will have no mercy; their voice is as the roaring sea; they shall array themselves for war against thee as fire on horses and chariots, O daughter of Sion.}{6:23}{They take hold of bow and spear. They are cruel, and have no mercy. Their voice roars like the sea, and they ride on horses, everyone set in array, as a man to the battle, against you, daughter of Zion.”}
\VersePair{6:24}{We have heard the report of them: our hands are weakened: anguish has seized us, the pangs as of a woman in travail.}{6:24}{We have heard its report. Our hands become feeble. Anguish has taken hold of us, and pains as of a woman in labor.}
\VersePair{6:25}{Go not forth into the field, and walk not in the ways; for the sword of the enemy lingers round about.}{6:25}{Don’t go out into the field or walk by the way; for the sword of the enemy and terror are on every side.}
\VersePair{6:26}{O daughter of my people, gird thyself with sackcloth: sprinkle thyself with ashes; make for thyself pitiable lamentation, as the mourning for a beloved son: for misery will come suddenly upon you.}{6:26}{Daughter of my people, clothe yourself with sackcloth, and wallow in ashes! Mourn, as for an only son, most bitter lamentation, for the destroyer will suddenly come on us.}
\VersePair{6:27}{I have caused thee to be tried among tried nations, and thou shalt know me when I have tried their way.}{6:27}{“I have made you a tester of metals and a fortress among my people, that you may know and try their way.}
\VersePair{6:28}{They are all disobedient, walking perversely: they are brass and iron; they are all corrupted.}{6:28}{They are all grievous rebels, going around to slander. They are bronze and iron. All of them deal corruptly.}
\VersePair{6:29}{The bellows have failed from the fire, the lead has failed: the silversmith works at his trade in vain; their wickedness is not consumed.}{6:29}{The bellows blow fiercely. The lead is consumed in the fire. In vain they go on refining, for the wicked are not plucked away.}
\VersePair{6:30}{Call ye them reprobate silver, because the Lord has rejected them.}{6:30}{Men will call them rejected silver, because Yahweh has rejected them.”}
\end{paracol}
\ChapterHeading{7}
\begin{paracol}{2}
\VersePair{7:2}{Hear ye the word of the Lord, all Judea.}{7:1-7:2}{The word that came to Jeremiah from Yahweh, saying, “Stand in the gate of Yahweh’s house, and proclaim this word there, and say, ‘Hear Yahweh’s word, all you of Judah, who enter in at these gates to worship Yahweh.’”}
\VersePair{7:3}{Thus saith the Lord God of Israel, Correct your ways and your devices, and I will cause you to dwell in this place.}{7:3}{Yahweh of Armies, the God of Israel says, “Amend your ways and your doings, and I will cause you to dwell in this place.}
\VersePair{7:4}{Trust not in yourselves with lying words, for they shall not profit you at all, saying, It is the temple of the Lord, the temple of the Lord.}{7:4}{Don’t trust in lying words, saying, ‘Yahweh’s temple, Yahweh’s temple, Yahweh’s temple, are these.’}
\VersePair{7:5}{For if ye thoroughly correct your ways and your practices, and do indeed execute judgment between a man and his neighbour;}{7:5}{For if you thoroughly amend your ways and your doings, if you thoroughly execute justice between a man and his neighbor;}
\VersePair{7:6}{and oppress not the stranger, and the orphan, and the widow, and shed not innocent blood in this place, and go not after strange gods to your hurt:}{7:6}{if you don’t oppress the foreigner, the fatherless, and the widow, and don’t shed innocent blood in this place, and don’t walk after other gods to your own hurt,}
\VersePair{7:7}{then will I cause you to dwell in this place, in the land which I gave to your fathers of old and for ever.}{7:7}{then I will cause you to dwell in this place, in the land that I gave to your fathers, from of old even forever more.}
\VersePair{7:8}{But whereas ye have trusted in lying words, whereby ye shall not be profited;}{7:8}{Behold, you trust in lying words that can’t profit.}
\VersePair{7:9}{and ye murder, and commit adultery, and steal, and swear falsely, and burn incense to Baal, and are gone after strange gods whom ye know not,}{7:9}{Will you steal, murder, commit adultery, swear falsely, burn incense to Baal, and walk after other gods that you have not known,}
\VersePair{7:10}{so that it is evil with you; yet have ye come, and stood before me in the house, whereon my name is called, and ye have said, We have refrained from doing all these abominations.}{7:10}{then come and stand before me in this house, which is called by my name, and say, ‘We are delivered,’ that you may do all these abominations?}
\VersePair{7:11}{Is my house, whereon my name is called, a den of robbers in your eyes? And, behold, I have seen it, saith the Lord.}{7:11}{Has this house, which is called by my name, become a den of robbers in your eyes? Behold, I myself have seen it,” says Yahweh.}
\VersePair{7:12}{For go ye to my place which is in Selo, where I caused my name to dwell before, and see what I did to it because of the wickedness of my people Israel.}{7:12}{“But go now to my place which was in Shiloh, where I caused my name to dwell at the first, and see what I did to it for the wickedness of my people Israel.}
\VersePair{7:13}{And now, because ye have done all these deeds, and I spoke to you, but ye hearkened not to me; and I called you, but ye answered not;}{7:13}{Now, because you have done all these works,” says Yahweh, “and I spoke to you, rising up early and speaking, but you didn’t hear; and I called you, but you didn’t answer;}
\VersePair{7:14}{therefore I also will do to the house whereon my name is called, wherein ye trust, and to the place which I gave to you and to your fathers, as I did to Selo.}{7:14}{therefore I will do to the house which is called by my name, in which you trust, and to the place which I gave to you and to your fathers, as I did to Shiloh.}
\VersePair{7:15}{And I will cast you out of my sight, as I cast away your brethren, all the seed of Ephraim.}{7:15}{I will cast you out of my sight, as I have cast out all your brothers, even the whole offspring of Ephraim. \footnote{7:15: or, seed}}
\VersePair{7:16}{Therefore pray not thou for this people, and intercede not for them to be pitied, yea, pray not, and approach me not for them: for I will not hearken unto thee.}{7:16}{“Therefore don’t pray for this people. Don’t lift up a cry or prayer for them or make intercession to me; for I will not hear you.}
\VersePair{7:17}{Seest thou not what they do in the cities of Juda, and in the streets of Jerusalem?}{7:17}{Don’t you see what they do in the cities of Judah and in the streets of Jerusalem?}
\VersePair{7:18}{Their children gather wood, and their fathers kindle a fire, and their women knead dough, to make cakes to the host of heaven; and they have poured out drink-offerings to strange gods, that they might provoke me to anger.}{7:18}{The children gather wood, and the fathers kindle the fire, and the women knead the dough, to make cakes to the queen of the sky, and to pour out drink offerings to other gods, that they may provoke me to anger.}
\VersePair{7:19}{Do they provoke me to anger? saith the Lord: do they not provoke themselves, that their faces may be ashamed?}{7:19}{Do they provoke me to anger?” says Yahweh. “Don’t they provoke themselves, to the confusion of their own faces?”}
\VersePair{7:20}{Therefore thus saith the Lord; Behold, my anger and wrath shall be poured out upon this place, and upon the men, and upon the cattle, and upon every tree of their field, and upon the fruits of the land; and it shall burn, and not be quenched.}{7:20}{Therefore the Lord Yahweh says: “Behold, my anger and my wrath will be poured out on this place, on man, on animal, on the trees of the field, and on the fruit of the ground; and it will burn and will not be quenched.”}
\VersePair{7:21}{Thus saith the Lord, Gather your whole-burnt-offerings with your meat-offerings, and eat flesh.}{7:21}{Yahweh of Armies, the God of Israel says: “Add your burnt offerings to your sacrifices and eat meat.}
\VersePair{7:22}{For I spoke not to your fathers, and commanded them not in the day wherein I brought them up out of the land of Egypt, concerning whole-burnt-offerings and sacrifice:}{7:22}{For I didn’t speak to your fathers or command them in the day that I brought them out of the land of Egypt concerning burnt offerings or sacrifices;}
\VersePair{7:23}{but I commanded them this thing, saying, Hear ye my voice, and I will be to you a God, and ye shall be to me a people: and walk ye in all my ways, which I shall command you, that it may be well with you.}{7:23}{but this thing I commanded them, saying, ‘Listen to my voice, and I will be your God, and you shall be my people. Walk in all the way that I command you, that it may be well with you.’}
\VersePair{7:24}{But they hearkened not to me, and their ear gave no heed, but they walked in the imaginations of their evil heart, and went backward, and not forward;}{7:24}{But they didn’t listen or turn their ear, but walked in their own counsels and in the stubbornness of their evil heart, and went backward, and not forward.}
\VersePair{7:25}{from the day that their fathers went forth out of the land of Egypt, even until this day. And I sent to you all my servants, the prophets, by day and early in the morning: yea, I sent them,}{7:25}{Since the day that your fathers came out of the land of Egypt to this day, I have sent to you all my servants the prophets, daily rising up early and sending them.}
\VersePair{7:26}{but they hearkened not to me, and their ear gave no heed; and they made their neck harder than their fathers.}{7:26}{Yet they didn’t listen to me or incline their ear, but made their neck stiff. They did worse than their fathers.}
\VersePair{7:27}{Therefore thou shalt speak this word to them;}{7:27}{“You shall speak all these words to them, but they will not listen to you. You shall also call to them, but they will not answer you.}
\VersePair{7:28}{This is the nation which has not hearkened to the voice of the Lord, nor received correction: truth has failed from their mouth.}{7:28}{You shall tell them, ‘This is the nation that has not listened to Yahweh their God’s voice, nor received instruction. Truth has perished, and is cut off from their mouth.’}
\VersePair{7:29}{Cut off thine hair, and cast it away, and take up a lamentation on thy lips; for the Lord has reprobated and rejected the generation that does these things.}{7:29}{Cut off your hair, and throw it away, and take up a lamentation on the bare heights; for Yahweh has rejected and forsaken the generation of his wrath.}
\VersePair{7:30}{For the children of Juda have wrought evil before me, saith the Lord; they have set their abominations in the house on which my name is called, to defile it.}{7:30}{“For the children of Judah have done that which is evil in my sight,” says Yahweh. “They have set their abominations in the house which is called by my name, to defile it.}
\VersePair{7:31}{And they have built the altar of Tapheth, which is in the valley of the son of Ennom, to burn their sons and their daughters with fire; which I did not command them to do, neither did I design it in my heart.}{7:31}{They have built the high places of Topheth, which is in the valley of the son of Hinnom, to burn their sons and their daughters in the fire, which I didn’t command, nor did it come into my mind.}
\VersePair{7:32}{Therefore, behold, the days come, saith the Lord, when they shall no more say, The altar of Tapheth, and the valley of the son of Ennom, but, The valley of the slain; and they shall bury in Tapheth, for want of room.}{7:32}{Therefore behold, the days come ”, says Yahweh, “that it will no more be called ‘Topheth’or ‘The valley of the son of Hinnom ’, but ‘The valley of Slaughter ’; for they will bury in Topheth until there is no place to bury.}
\VersePair{7:33}{And the dead bodies of this people shall be for food to the birds of the sky, and to the wild beasts of the earth; and there shall be none to drive them away.}{7:33}{The dead bodies of this people will be food for the birds of the sky, and for the animals of the earth. No one will frighten them away.}
\VersePair{7:34}{And I will destroy out of the cities of Juda, and the streets of Jerusalem, the voice of them that make merry, and the voice of them that rejoice, the voice of the bridegroom, and the voice of the bride; for the whole land shall become a desolation.}{7:34}{Then I will cause to cease from the cities of Judah and from the streets of Jerusalem the voice of mirth and the voice of gladness, the voice of the bridegroom and the voice of the bride; for the land will become a waste.”}
\end{paracol}
\ChapterHeading{8}
\begin{paracol}{2}
\VersePair{8:1}{At that time, saith the Lord, they shall bring out the bones of the kings of Juda, and the bones of his princes, and the bones of the priests, and the bones of the prophets, and the bones of the inhabitants of Jerusalem, out of their graves;}{8:1}{“At that time,” says Yahweh, “they will bring the bones of the kings of Judah, the bones of his princes, the bones of the priests, the bones of the prophets, and the bones of the inhabitants of Jerusalem, out of their graves.}
\VersePair{8:2}{and they shall spread them out to the sun, and the moon, and to all the stars, and to all the host of heaven, which they have loved, and which they have served, and after which they have walked, and to which they have held, and which they have worshipped; they shall not be mourned for, neither shall they be buried; but they shall be for an example on the face of the earth,}{8:2}{They will spread them before the sun, the moon, and all the army of the sky, which they have loved, which they have served, after which they have walked, which they have sought, and which they have worshiped. They will not be gathered or be buried. They will be like dung on the surface of the earth.}
\VersePair{8:3}{because they chose death rather than life, even to all the remnant that are left of that family, in every place whither I shall drive them out.}{8:3}{Death will be chosen rather than life by all the residue that remain of this evil family, that remain in all the places where I have driven them,” says Yahweh of Armies.}
\VersePair{8:4}{For thus saith the Lord, Shall not he that falls arise? or he that turns away, shall he not turn back again?}{8:4}{{\raggedright “Moreover you shall tell them, ‘Yahweh says: \poemline{1}{“‘Do men fall, and not rise up again?}\poemline{2}{Does one turn away, and not return?}}}
\VersePair{8:5}{Wherefore has this my people turned away with a shameless revolting, and strengthened themselves in their wilfulness, and refused to return?}{8:5}{{\raggedright Why then have the people of Jerusalem fallen back by a perpetual backsliding? \poemline{2}{They cling to deceit.}\poemline{2}{They refuse to return.}}}
\VersePair{8:6}{Hearken, I pray you, and hear: will they not speak thus, There is no man that repents of his wickedness, saying, What have I done? the runner has failed from his course, as a tired horse in his neighing.}{8:6}{{\raggedright I listened and heard, but they didn’t say what is right. \poemline{2}{No one repents of his wickedness, saying, “What have I done?”}\poemline{1}{Everyone turns to his course,}\poemline{2}{as a horse that rushes headlong in the battle.}}}
\VersePair{8:7}{Yea, the stork in the heaven knows her time, also the turtle-dove and wild swallow; the sparrows observe the times of their coming in; but this my people knows not the judgments of the Lord.}{8:7}{{\raggedright Yes, the stork in the sky knows her appointed times. \poemline{2}{The turtledove, the swallow, and the crane observe the time of their coming;}\poemline{2}{but my people don’t know Yahweh’s law.}}}
\VersePair{8:8}{How will ye say, We are wise, and the law of the Lord is with us? In vain have the scribes used a false pen.}{8:8}{{\raggedright “‘How do you say, “We are wise, and Yahweh’s law is with us ”? \poemline{2}{But, behold, the false pen of the scribes has made that a lie.}}}
\VersePair{8:9}{The wise men are ashamed, and alarmed, and taken; because they have rejected the word of the Lord; what wisdom is there in them?}{8:9}{{\raggedright The wise men are disappointed. \poemline{2}{They are dismayed and trapped.}\poemline{1}{Behold, they have rejected Yahweh’s word.}\poemline{2}{What kind of wisdom is in them?}}}
\VersePair{8:10}{Therefore will I give their wives to others, and their fields to new inheritors; and they shall gather their fruits, saith the Lord.}{8:10}{{\raggedright Therefore I will give their wives to others \poemline{2}{and their fields to those who will possess them.}\poemline{1}{For everyone from the least even to the greatest is given to covetousness;}\poemline{2}{from the prophet even to the priest everyone deals falsely.}}}
\VersePair{8:11}{}{8:11}{{\raggedright They have healed the hurt of the daughter of my people slightly, saying, \poemline{2}{“Peace, peace,” when there is no peace.}}}
\VersePair{8:12}{}{8:12}{{\raggedright Were they ashamed when they had committed abomination? \poemline{2}{No, they were not at all ashamed.}\poemline{2}{They couldn’t blush.}\poemline{1}{Therefore they will fall among those who fall.}\poemline{2}{In the time of their visitation they will be cast down, says Yahweh.}}}
\VersePair{8:13}{There are no grapes on the vines, and there are no figs on the fig-trees, and the leaves have fallen off.}{8:13}{{\raggedright “‘I will utterly consume them, says Yahweh. \poemline{2}{No grapes will be on the vine,}\poemline{2}{no figs on the fig tree,}\poemline{2}{and the leaf will fade.}\poemline{1}{The things that I have given them}\poemline{2}{will pass away from them.’”}}}
\VersePair{8:14}{Why do we sit still? assemble yourselves, and let us enter into the strong cities, and let us be cast out there: for God has cast us out, and made us drink water of gall, because we have sinned before him.}{8:14}{{\raggedright “Why do we sit still? \poemline{2}{Assemble yourselves!}\poemline{2}{Let’s enter into the fortified cities,}\poemline{2}{and let’s be silent there;}\poemline{1}{for Yahweh our God has put us to silence,}\poemline{2}{and given us poisoned water to drink,}\poemline{2}{because we have sinned against Yahweh.}}}
\VersePair{8:15}{We assembled for peace, but there was no prosperity; for a time of healing, but behold anxiety.}{8:15}{{\raggedright We looked for peace, but no good came; \poemline{2}{and for a time of healing, and behold, dismay!}}}
\VersePair{8:16}{We shall hear the neighing of his swift horses out of Dan: the whole land quaked at the sound of the neighing of his horses; and he shall come, and devour the land and the fulness of it; the city, and them that dwell in it.}{8:16}{{\raggedright The snorting of his horses is heard from Dan. \poemline{2}{The whole land trembles at the sound of the neighing of his strong ones;}\poemline{1}{for they have come, and have devoured the land and all that is in it,}\poemline{2}{the city and those who dwell therein.”}}}
\VersePair{8:17}{For, behold, I send forth against you deadly serpents, which cannot be charmed, and they shall bite you}{8:17}{{\raggedright “For, behold, I will send serpents, \poemline{2}{adders among you,}\poemline{2}{which will not be charmed;}\poemline{2}{and they will bite you,” says Yahweh.}}}
\VersePair{8:18}{mortally with the pain of your distressed heart.}{8:18}{{\raggedright Oh that I could comfort myself against sorrow! \poemline{2}{My heart is faint within me.}}}
\VersePair{8:19}{Behold, there is a sound of the cry of the daughter of my people from a land afar off: Is not the Lord in Sion? is there not a king there? because they have provoked me with their graven images, and with strange vanities.}{8:19}{{\raggedright Behold, the voice of the cry of the daughter of my people from a land that is very far off: \poemline{2}{“Isn’t Yahweh in Zion?}\poemline{2}{Isn’t her King in her?”}\poemline{1}{“Why have they provoked me to anger with their engraved images,}\poemline{2}{and with foreign idols?”}}}
\VersePair{8:20}{The summer is gone, the harvest is past, and we are not saved.}{8:20}{{\raggedright “The harvest is past. \poemline{2}{The summer has ended,}\poemline{2}{and we are not saved.”}}}
\VersePair{8:21}{For the breach of the daughter of my people I have been saddened: in my perplexity pangs have seized upon me as of a woman in travail.}{8:21}{{\raggedright For the hurt of the daughter of my people, I am hurt. \poemline{2}{I mourn.}\poemline{2}{Dismay has taken hold of me.}}}
\VersePair{8:22}{And is there no balm in Galaad, or is there no physician there? why has not the healing of the daughter of my people taken place?}{8:22}{{\raggedright Is there no balm in Gilead? \poemline{2}{Is there no physician there?}\poemline{1}{Why then isn’t the health of the daughter of my people recovered?}}}
\VersePair{8:23}{Who will give water to my head, and a fountain of tears to my eyes? then would I weep for this my people day and night, even for the slain of the daughter of my people.}{9:1}{{\raggedright Oh that my head were waters, \poemline{2}{and my eyes a spring of tears,}\poemline{1}{that I might weep day and night}\poemline{2}{for the slain of the daughter of my people!}}}
\end{paracol}
\ChapterHeading{9}
\begin{paracol}{2}
\VersePair{9:1}{Who would give me a most distant lodge in the wilderness, that I might leave my people, and depart from them? for they all commit adultery, an assembly of treacherous men.}{9:2}{{\raggedright Oh that I had in the wilderness \poemline{2}{a lodging place of wayfaring men,}\poemline{1}{that I might leave my people}\poemline{2}{and go from them!}\poemline{1}{For they are all adulterers,}\poemline{2}{an assembly of treacherous men.}}}
\VersePair{9:2}{And they have bent their tongue like a bow: falsehood and not faithfulness has prevailed upon the earth; for they have gone on from evil to evil, and have not known me, saith the Lord.}{9:3}{{\raggedright “They bend their tongue, \poemline{2}{as their bow, for falsehood.}\poemline{1}{They have grown strong in the land,}\poemline{2}{but not for truth;}\poemline{1}{for they proceed from evil to evil,}\poemline{2}{and they don’t know me,” says Yahweh.}}}
\VersePair{9:3}{Beware ye each of his neighbour, and trust ye not in your brethren: for every one will surely supplant, and every friend will walk craftily.}{9:4}{{\raggedright “Everyone beware of his neighbor, \poemline{2}{and don’t trust in any brother;}\poemline{1}{for every brother will utterly supplant,}\poemline{2}{and every neighbor will go around like a slanderer.}}}
\VersePair{9:4}{Every one will mock his friend; they will not speak truth: their tongue has learned to speak falsehoods; they have committed iniquity, they ceased not, so as to return.}{9:5}{{\raggedright Friends deceive each other, \poemline{2}{and will not speak the truth.}\poemline{1}{They have taught their tongue to speak lies.}\poemline{2}{They weary themselves committing iniquity.}}}
\VersePair{9:5}{There is usury upon usury, and deceit upon deceit: they would not know me, saith the Lord.}{9:6}{{\raggedright Your habitation is in the middle of deceit. \poemline{2}{Through deceit, they refuse to know me,” says Yahweh.}}}
\VersePair{9:6}{Therefore thus saith the Lord, Behold, I will try them with fire, and prove them; for I will do thus because of the wickedness of the daughter of my people.}{9:7}{{\raggedright Therefore Yahweh of Armies says, \poemline{1}{“Behold, I will melt them and test them;}\poemline{2}{for how should I deal with the daughter of my people?}}}
\VersePair{9:7}{Their tongue is a wounding arrow; the words of their mouth are deceitful: one speaks peaceably to his neighbour, but in himself retains enmity.}{9:8}{{\raggedright Their tongue is a deadly arrow. \poemline{2}{It speaks deceit.}\poemline{1}{One speaks peaceably to his neighbor with his mouth,}\poemline{2}{but in his heart, he waits to ambush him.}}}
\VersePair{9:8}{Shall I not visit for these things? saith the Lord: and shall not my soul be avenged on such a people as this?}{9:9}{{\raggedright Shouldn’t I punish them for these things?” says Yahweh. \poemline{2}{“Shouldn’t my soul be avenged on a nation such as this?}}}
\VersePair{9:9}{Take up a lamentation for the mountains, and a mournful dirge for the paths of the wilderness, for they are desolate for want of men; they heard not the sound of life from the birds of the sky, nor the cattle: they were amazed, they are gone.}{9:10}{{\raggedright I will weep and wail for the mountains, \poemline{2}{and lament for the pastures of the wilderness,}\poemline{1}{because they are burned up, so that no one passes through;}\poemline{2}{Men can’t hear the voice of the livestock.}\poemline{1}{Both the birds of the sky and the animals have fled.}\poemline{2}{They are gone.}}}
\VersePair{9:10}{And I will remove the inhabitants of Jerusalem, and make it a dwelling-place of dragons; and I will utterly waste the cities of Juda, so that they shall not be inhabited.}{9:11}{{\raggedright “I will make Jerusalem heaps, \poemline{2}{a dwelling place of jackals.}\poemline{1}{I will make the cities of Judah a desolation,}\poemline{2}{without inhabitant.”}}}
\VersePair{9:11}{Who is the wise man, that he may understand this? and he that has the word of the mouth of the Lord addressed to him, let him tell you wherefore the land has been destroyed, has been ravaged by fire like a desert, so that no one passes through it.}{9:12}{Who is wise enough to understand this? Who is he to whom the mouth of Yahweh has spoken, that he may declare it? Why has the land perished and burned up like a wilderness, so that no one passes through?}
\VersePair{9:12}{And the Lord said to me, Because they have forsaken my law, which I set before them, and have not hearkened to my voice;}{9:13}{Yahweh says, “Because they have forsaken my law which I set before them, and have not obeyed my voice or walked in my ways,}
\VersePair{9:13}{but went after the lusts of their evil heart, and after the idols which their fathers taught them to worship:}{9:14}{but have walked after the stubbornness of their own heart and after the Baals, which their fathers taught them.”}
\VersePair{9:14}{therefore thus saith the Lord God of Israel, Behold, I will feed them with trouble and will cause them to drink water of gall:}{9:15}{Therefore Yahweh of Armies, the God of Israel, says, “Behold, I will feed them, even this people, with wormwood and give them poisoned water to drink.}
\VersePair{9:15}{and I will scatter them among the nations, to them whom neither they nor their fathers knew; and I will send a sword upon them, until I have consumed them with it.}{9:16}{I will scatter them also among the nations, whom neither they nor their fathers have known. I will send the sword after them, until I have consumed them.”}
\VersePair{9:16}{Thus saith the Lord, Call ye the mourning women, and let them come; and send to the wise women, and let them utter their voice;}{9:17}{{\raggedright Yahweh of Armies says, \poemline{1}{“Consider, and call for the mourning women, that they may come.}\poemline{2}{Send for the skillful women, that they may come.}}}
\VersePair{9:17}{and let them take up a lamentation for you, and let your eyes pour down tears, and your eyelids drop water.}{9:18}{{\raggedright Let them make haste \poemline{2}{and take up a wailing for us,}\poemline{1}{that our eyes may run down with tears}\poemline{2}{and our eyelids gush out with waters.}}}
\VersePair{9:18}{For a voice of lamentation has been heard in Sion, How are we become wretched! we are greatly ashamed, for we have forsaken the land, and have abandoned our tabernacles!}{9:19}{{\raggedright For a voice of wailing is heard out of Zion, \poemline{2}{‘How we are ruined!}\poemline{1}{We are greatly confounded}\poemline{2}{because we have forsaken the land,}\poemline{2}{because they have cast down our dwellings.’”}}}
\VersePair{9:19}{Hear now, ye women, the word of God, and let your ears receive the words of his mouth, and teach your daughters lamentation, and every woman her neighbour a dirge.}{9:20}{{\raggedright Yet hear Yahweh’s word, you women. \poemline{2}{Let your ear receive the word of his mouth.}\poemline{1}{Teach your daughters wailing.}\poemline{2}{Everyone teach her neighbor a lamentation.}}}
\VersePair{9:20}{For death has come up through your windows, it has entered into our land, to destroy the infants without, and the young men from the streets.}{9:21}{{\raggedright For death has come up into our windows. \poemline{2}{It has entered into our palaces}\poemline{1}{to cut off the children from outside,}\poemline{2}{and the young men from the streets.}}}
\VersePair{9:21}{And the carcases of the men shall be for an example on the face of the field of your land, like grass after the mower, and there shall be none to gather them.}{9:22}{{\raggedright Speak, “Yahweh says, \poemline{1}{“‘The dead bodies of men will fall as dung on the open field,}\poemline{2}{and as the handful after the harvester.}\poemline{2}{No one will gather them.’”}}}
\VersePair{9:22}{Thus saith the Lord, Let not the wise man boast in his wisdom, and let not the strong man boast in his strength, and let not the rich man boast in his wealth;}{9:23}{{\raggedright Yahweh says, \poemline{1}{“Don’t let the wise man glory in his wisdom.}\poemline{2}{Don’t let the mighty man glory in his might.}\poemline{2}{Don’t let the rich man glory in his riches.}}}
\VersePair{9:23}{but let him that boasts boast in this, the understanding and knowing that I am the Lord that exercise mercy, and judgment, and righteousness, upon the earth; for in these things is my pleasure, saith the Lord.}{9:24}{{\raggedright But let him who glories glory in this, \poemline{2}{that he has understanding, and knows me,}\poemline{1}{that I am Yahweh who exercises loving kindness, justice, and righteousness in the earth,}\poemline{2}{for I delight in these things,” says Yahweh.}}}
\VersePair{9:24}{Behold, the days come, saith the Lord, when I will visit upon all the circumcised their uncircumcision;}{9:25}{“Behold, the days come,” says Yahweh, “that I will punish all those who are circumcised only in their flesh:}
\VersePair{9:25}{on Egypt, and on Idumea, and on Edom, and on the children of Ammon, and on the children of Moab, and on every one that shaves his face round about, even them that dwell in the wilderness; for all the Gentiles are uncircumcised in flesh, and all the house of Israel are uncircumcised in their hearts.}{9:26}{Egypt, Judah, Edom, the children of Ammon, Moab, and all who have the corners of their hair cut off, who dwell in the wilderness, for all the nations are uncircumcised, and all the house of Israel are uncircumcised in heart.”}
\end{paracol}
\ChapterHeading{10}
\begin{paracol}{2}
\VersePair{10:1}{Hear ye the word of the Lord, which he has spoken to you, O house of Israel.}{10:1}{Hear the word which Yahweh speaks to you, house of Israel!}
\VersePair{10:2}{Thus saith the Lord, Learn ye not the ways of the heathen, and be not alarmed at the signs of the sky; for they are alarmed at them, falling on their faces.}{10:2}{{\raggedright Yahweh says, \poemline{1}{“Don’t learn the way of the nations,}\poemline{2}{and don’t be dismayed at the signs of the sky;}\poemline{2}{for the nations are dismayed at them.}}}
\VersePair{10:3}{For the customs of the nations are vain; it is a tree cut out of the forest, the work of the carpenter, or a molten image.}{10:3}{{\raggedright For the customs of the peoples are vanity; \poemline{2}{for one cuts a tree out of the forest,}\poemline{2}{the work of the hands of the workman with the ax.}}}
\VersePair{10:4}{They are beautified with silver and gold, they fix them with hammers and nails;}{10:4}{{\raggedright They deck it with silver and with gold. \poemline{2}{They fasten it with nails and with hammers,}\poemline{2}{so that it can’t move.}}}
\VersePair{10:5}{they will set them up that they may not move; it is wrought silver, they will not walk,}{10:5}{{\raggedright They are like a palm tree, of turned work, \poemline{2}{and don’t speak.}\poemline{1}{They must be carried,}\poemline{2}{because they can’t move.}\poemline{1}{Don’t be afraid of them;}\poemline{2}{for they can’t do evil,}\poemline{2}{neither is it in them to do good.”}}}
\VersePair{10:6}{}{10:6}{{\raggedright There is no one like you, Yahweh. \poemline{2}{You are great,}\poemline{2}{and your name is great in might.}}}
\VersePair{10:7}{}{10:7}{{\raggedright Who shouldn’t fear you, \poemline{2}{King of the nations?}\poemline{2}{For it belongs to you.}\poemline{1}{Because among all the wise men of the nations,}\poemline{2}{and in all their royal estate,}\poemline{2}{there is no one like you.}}}
\VersePair{10:8}{}{10:8}{{\raggedright But they are together brutish and foolish, \poemline{2}{instructed by idols!}\poemline{2}{It is just wood.}}}
\VersePair{10:9}{it is forged silver brought from Tharsis, gold will come from Mophaz, and the work of goldsmiths: they are all the works of craftsmen, they will clothe themselves with blue and scarlet.}{10:9}{{\raggedright There is silver beaten into plates, which is brought from Tarshish, \poemline{2}{and gold from Uphaz,}\poemline{2}{the work of the engraver and of the hands of the goldsmith.}\poemline{1}{Their clothing is blue and purple.}\poemline{2}{They are all the work of skillful men.}}}
\VersePair{10:10}{}{10:10}{{\raggedright But Yahweh is the true God. \poemline{2}{He is the living God,}\poemline{2}{and an everlasting King.}\poemline{1}{At his wrath, the earth trembles.}\poemline{2}{The nations aren’t able to withstand his indignation.}}}
\VersePair{10:11}{Thus shall ye say to them, Let the gods which have not made heaven and earth perish from off the earth, and from under this sky.}{10:11}{“You shall say this to them: ‘The gods that have not made the heavens and the earth will perish from the earth, and from under the heavens.’”}
\VersePair{10:12}{It is the Lord that made the earth by his strength, who set up the world by his wisdom, and by his understanding stretched out the sky,}{10:12}{{\raggedright God has made the earth by his power. \poemline{2}{He has established the world by his wisdom,}\poemline{2}{and by his understanding has he stretched out the heavens.}}}
\VersePair{10:13}{and set abundance of waters in the sky, and brought up clouds from the ends of the earth; he made lightnings for the rain, and brought forth light out of his treasures.}{10:13}{{\raggedright When he utters his voice, \poemline{2}{the waters in the heavens roar,}\poemline{2}{and he causes the vapors to ascend from the ends of the earth.}\poemline{1}{He makes lightnings for the rain,}\poemline{2}{and brings the wind out of his treasuries.}}}
\VersePair{10:14}{Every man is deprived of knowledge, every goldsmith is confounded because of his graven images; for he has cast false gods, there is no breath in them.}{10:14}{{\raggedright Every man has become brutish and without knowledge. \poemline{2}{Every goldsmith is disappointed by his engraved image;}\poemline{1}{for his molten image is falsehood,}\poemline{2}{and there is no breath in them.}}}
\VersePair{10:15}{They are vain works, wrought in mockery; in the time of their visitation they shall perish.}{10:15}{{\raggedright They are vanity, a work of delusion. \poemline{2}{In the time of their visitation they will perish.}}}
\VersePair{10:16}{Such is not the portion of Jacob; for he that formed all things, he is his inheritance; the Lord is his name.}{10:16}{{\raggedright The portion of Jacob is not like these; \poemline{2}{for he is the maker of all things;}\poemline{1}{and Israel is the tribe of his inheritance.}\poemline{2}{Yahweh of Armies is his name.}}}
\VersePair{10:17}{He has gathered thy substance from without that lodged in choice vessels.}{10:17}{{\raggedright Gather up your wares out of the land, \poemline{2}{you who live under siege.}}}
\VersePair{10:18}{For thus saith the Lord, Behold, I will overthrow the inhabitants of this land with affliction, that thy plague may be discovered.}{10:18}{{\raggedright For Yahweh says, \poemline{2}{“Behold, I will sling out the inhabitants of the land at this time,}\poemline{2}{and will distress them, that they may feel it.”}}}
\VersePair{10:19}{Alas for thy ruin! thy plague is grievous: and I said, Surely this is thy wound, and it has overtaken thee.}{10:19}{{\raggedright Woe is me because of my injury! \poemline{2}{My wound is serious;}\poemline{1}{but I said,}\poemline{2}{“Truly this is my grief, and I must bear it.”}}}
\VersePair{10:20}{Thy tabernacle is in a ruinous state, it has perished; and all thy curtains have been torn asunder: my children and my cattle are no more: there is no more any place for my tabernacle, nor place for my curtains.}{10:20}{{\raggedright My tent has been destroyed, \poemline{2}{and all my cords are broken.}\poemline{1}{My children have gone away from me, and they are no more.}\poemline{2}{There is no one to spread my tent any more,}\poemline{2}{to set up my curtains.}}}
\VersePair{10:21}{For the shepherds have become foolish, and have not sought the Lord; therefore the whole pasture has failed, and the sheep have been scattered.}{10:21}{{\raggedright For the shepherds have become brutish, \poemline{2}{and have not inquired of Yahweh.}\poemline{1}{Therefore they have not prospered,}\poemline{2}{and all their flocks have scattered.}}}
\VersePair{10:22}{Behold, there comes a sound of a noise, and a great earthquake from the land of the north, to make the cities of Juda a desolation, and a resting-place for ostriches.}{10:22}{{\raggedright The voice of news, behold, it comes, \poemline{2}{and a great commotion out of the north country,}\poemline{1}{to make the cities of Judah a desolation,}\poemline{2}{a dwelling place of jackals.}}}
\VersePair{10:23}{I know, O Lord, that man's way is not his own; neither shall a man go, and direct his going.}{10:23}{{\raggedright Yahweh, I know that the way of man is not in himself. \poemline{2}{It is not in man who walks to direct his steps.}}}
\VersePair{10:24}{Chasten us, O Lord, but with judgment; and not in wrath, lest thou make us few.}{10:24}{{\raggedright Yahweh, correct me, but gently; \poemline{2}{not in your anger,}\poemline{2}{lest you reduce me to nothing.}}}
\VersePair{10:25}{Pour out thy wrath upon the nations that have not known thee, and upon the families that have not called upon thy name: for they have devoured Jacob, and consumed him, and have made his pasture desolate.}{10:25}{{\raggedright Pour out your wrath on the nations that don’t know you, \poemline{2}{and on the families that don’t call on your name;}\poemline{1}{for they have devoured Jacob.}\poemline{2}{Yes, they have devoured him, consumed him,}\poemline{2}{and have laid waste his habitation.}}}
\end{paracol}
\ChapterHeading{11}
\begin{paracol}{2}
\VersePair{11:1}{The word that came to Jeremias from the Lord, saying,}{11:1}{The word that came to Jeremiah from Yahweh, saying,}
\VersePair{11:2}{Hear ye the words of this covenant, and thou shalt speak to the men of Juda, and to the dwellers in Jerusalem;}{11:2}{“Hear the words of this covenant, and speak to the men of Judah, and to the inhabitants of Jerusalem;}
\VersePair{11:3}{and thou shalt say to them, Thus saith the Lord God of Israel, Cursed is the man, who shall not hearken to the words of this covenant,}{11:3}{and say to them, Yahweh, the God of Israel says: ‘Cursed is the man who doesn’t hear the words of this covenant,}
\VersePair{11:4}{which I commanded your fathers, in the day wherein I brought them up out of the land of Egypt, out of the iron furnace, saying, Hearken to my voice, and do all things that I shall command you; so shall ye be to me a people, and I will be to you a God;}{11:4}{which I commanded your fathers in the day that I brought them out of the land of Egypt, out of the iron furnace,’ saying, ‘Obey my voice and do them, according to all which I command you; so you shall be my people, and I will be your God;}
\VersePair{11:5}{that I may confirm mine oath, which I sware to your fathers, to give them a land flowing with milk and honey, as it is this day. Then I answered and said, So be it, O Lord.}{11:5}{that I may establish the oath which I swore to your fathers, to give them a land flowing with milk and honey,’ as it is today.” Then I answered, and said, “Amen, Yahweh.”}
\VersePair{11:6}{And the Lord said to me, Read these words in the cities of Juda, and in the streets of Jerusalem, saying, Hear ye the words of this covenant, and do them.}{11:6}{Yahweh said to me, “Proclaim all these words in the cities of Judah, and in the streets of Jerusalem, saying, ‘Hear the words of this covenant, and do them.}
\VersePair{11:7}{}{11:7}{For I earnestly protested to your fathers in the day that I brought them up out of the land of Egypt, even to this day, rising early and protesting, saying, “Obey my voice.”}
\VersePair{11:8}{But they did them not.}{11:8}{Yet they didn’t obey, nor turn their ear, but everyone walked in the stubbornness of their evil heart. Therefore I brought on them all the words of this covenant, which I commanded them to do, but they didn’t do them.’”}
\VersePair{11:9}{And the Lord said to me, A conspiracy is found among the men of Juda, and among the dwellers in Jerusalem.}{11:9}{Yahweh said to me, “A conspiracy is found among the men of Judah, and among the inhabitants of Jerusalem.}
\VersePair{11:10}{They are turned aside to the iniquities of their fathers that were of old, who would not hearken to my words: and, behold, they go after strange gods, to serve them: and the house of Israel and the house of Juda have broken my covenant, which I made with their fathers.}{11:10}{They have turned back to the iniquities of their forefathers, who refused to hear my words. They have gone after other gods to serve them. The house of Israel and the house of Judah have broken my covenant which I made with their fathers.}
\VersePair{11:11}{Therefore thus saith the Lord, Behold, I bring evils upon this people, out of which they shall not be able to come forth; and they shall presently cry to me, but I will not hearken to them.}{11:11}{Therefore Yahweh says, ‘Behold, I will bring evil on them which they will not be able to escape; and they will cry to me, but I will not listen to them.}
\VersePair{11:12}{And the cities of Juda and the dwellers in Jerusalem shall go, and cry to the gods to whom they burn incense; which shall not deliver them in the time of their troubles.}{11:12}{Then the cities of Judah and the inhabitants of Jerusalem will go and cry to the gods to which they offer incense, but they will not save them at all in the time of their trouble.}
\VersePair{11:13}{For according to the number of thy cities were thy gods, O Juda; and according to the number of the streets of Jerusalem have ye set up altars to burn incense to Baal.}{11:13}{For according to the number of your cities are your gods, Judah; and according to the number of the streets of Jerusalem you have set up altars to the shameful thing, even altars to burn incense to Baal.’}
\VersePair{11:14}{And thou, pray not for this people, and intercede not for them in supplication and prayer: for I will not hear in the day in which they call upon me, in the day of their affliction.}{11:14}{“Therefore don’t pray for this people. Don’t lift up cry or prayer for them; for I will not hear them in the time that they cry to me because of their trouble.}
\VersePair{11:15}{Why has my beloved wrought abomination in my house? will prayers and holy offerings take away thy wickedness from thee, or shalt thou escape by these things?}{11:15}{{\raggedright What has my beloved to do in my house, \poemline{2}{since she has behaved lewdly with many,}\poemline{2}{and the holy flesh has passed from you?}\poemline{1}{When you do evil,}\poemline{2}{then you rejoice.”}}}
\VersePair{11:16}{The Lord called thy name a fair olive tree, of a goodly shade in appearance, at the noise of its being lopped, fire was kindled against it; great is the affliction coming upon thee: her branches are become good for nothing.}{11:16}{{\raggedright Yahweh called your name, “A green olive tree, \poemline{2}{beautiful with goodly fruit.”}\poemline{1}{With the noise of a great roar he has kindled fire on it,}\poemline{2}{and its branches are broken.}}}
\VersePair{11:17}{And the Lord that planted thee has pronounced evils against thee, because of the iniquity of the house of Israel and the house of Juda, whatsoever they have done against themselves to provoke me to anger by burning incense to Baal.}{11:17}{For Yahweh of Armies, who planted you, has pronounced evil against you, because of the evil of the house of Israel and of the house of Judah, which they have done to themselves in provoking me to anger by offering incense to Baal.}
\VersePair{11:18}{O Lord, teach me, and I shall know: then I saw their practices.}{11:18}{Yahweh gave me knowledge of it, and I knew it. Then you showed me their doings.}
\VersePair{11:19}{But I as an innocent lamb led to the slaughter, knew not: against me they devised an evil device, saying, Come and let us put wood into his bread, and let us utterly destroy him from off the land of the living, and let his name not be remembered any more.}{11:19}{{\raggedright But I was like a gentle lamb that is led to the slaughter. I didn’t know that they had devised plans against me, saying, \poemline{1}{“Let’s destroy the tree with its fruit,}\poemline{2}{and let’s cut him off from the land of the living,}\poemline{2}{that his name may be no more remembered.”}}}
\VersePair{11:20}{O Lord, that judgest righteously, trying the reins and hearts, let me see thy vengeance taken upon them, for to thee I have declared my cause.}{11:20}{{\raggedright But, Yahweh of Armies, who judges righteously, \poemline{2}{who tests the heart and the mind,}\poemline{1}{I will see your vengeance on them;}\poemline{2}{for to you I have revealed my cause.}}}
\VersePair{11:21}{Therefore thus saith the Lord concerning the men of Anathoth, that seek my life, that say, Thou shalt not prophesy at all in the name of the Lord, but if thou dost, thou shalt die by our hands:}{11:21}{“Therefore Yahweh says concerning the men of Anathoth, who seek your life, saying, ‘You shall not prophesy in Yahweh’s name, that you not die by our hand ’—}
\VersePair{11:22}{behold, I will visit them: their young men shall die by the sword; and their sons and their daughters shall die of famine:}{11:22}{therefore Yahweh of Armies says, ‘Behold, I will punish them. The young men will die by the sword. Their sons and their daughters will die by famine.}
\VersePair{11:23}{and there shall be no remnant left of them; for I will bring evil upon the dwellers in Anathoth, in the year of their visitation.}{11:23}{There will be no remnant to them, for I will bring evil on the men of Anathoth, even the year of their visitation.’”}
\end{paracol}
\ChapterHeading{12}
\begin{paracol}{2}
\VersePair{12:1}{Righteous art thou, O Lord, that I may make my defence to thee, yea, I will speak to thee of judgments. Why is it that the way of ungodly men prospers? that all that deal very treacherously are flourishing?}{12:1}{{\raggedright You are righteous, Yahweh, \poemline{2}{when I contend with you;}\poemline{1}{yet I would like to plead a case with you.}\poemline{2}{Why does the way of the wicked prosper?}\poemline{2}{Why are they all at ease who deal very treacherously?}}}
\VersePair{12:2}{Thou hast planted them, and they have taken root; they have begotten children, and become fruitful; thou art near to their mouth, and far from their reins.}{12:2}{{\raggedright You have planted them. Yes, they have taken root. \poemline{2}{They grow. Yes, they produce fruit.}\poemline{1}{You are near in their mouth,}\poemline{2}{and far from their heart.}}}
\VersePair{12:3}{But thou, Lord, knowest me; thou hast proved my heart before thee; purify them for the day of their slaughter.}{12:3}{{\raggedright But you, Yahweh, know me. \poemline{2}{You see me, and test my heart toward you.}\poemline{1}{Pull them out like sheep for the slaughter,}\poemline{2}{and prepare them for the day of slaughter.}}}
\VersePair{12:4}{How long shall the land mourn, and the grass of the field wither, for the wickedness of them that dwell in it? the beasts and birds are utterly destroyed; because the people said, God shall not see our ways.}{12:4}{{\raggedright How long will the land mourn, \poemline{2}{and the herbs of the whole country wither?}\poemline{1}{Because of the wickedness of those who dwell therein,}\poemline{2}{the animals and birds are consumed;}\poemline{1}{because they said,}\poemline{2}{“He won’t see our latter end.”}}}
\VersePair{12:5}{Thy feet run, and they cause thee to faint; how wilt thou prepare to ride upon horses? and thou hast been confident in the land of thy peace? how wilt thou do in the roaring of Jordan?}{12:5}{{\raggedright “If you have run with the footmen, \poemline{2}{and they have wearied you,}\poemline{2}{then how can you contend with horses?}\poemline{1}{Though in a land of peace you are secure,}\poemline{2}{yet how will you do in the pride of the Jordan?}}}
\VersePair{12:6}{For even thy brethren and the house of thy father, even these have dealt treacherously with thee; and they have cried out, they are gathered together in pursuit of thee; trust not thou in them, though they shall speak fair words to thee.}{12:6}{{\raggedright For even your brothers, and the house of your father, \poemline{2}{even they have dealt treacherously with you!}\poemline{2}{Even they have cried aloud after you!}\poemline{1}{Don’t believe them,}\poemline{2}{though they speak beautiful words to you.}}}
\VersePair{12:7}{I have forsaken mine house, I have left mine heritage; I have given my beloved one into the hands of her enemies.}{12:7}{{\raggedright “I have forsaken my house. \poemline{2}{I have cast off my heritage.}\poemline{2}{I have given the dearly beloved of my soul into the hand of her enemies.}}}
\VersePair{12:8}{My inheritance has become to me as a lion in a forest; she has uttered her voice against me; therefore have I hated her.}{12:8}{{\raggedright My heritage has become to me as a lion in the forest. \poemline{2}{She has uttered her voice against me.}\poemline{2}{Therefore I have hated her.}}}
\VersePair{12:9}{Is not my inheritance to me a hyæna's cave, or a cave round about her? Go ye, gather together all the wild beasts of the field, and let them come to devour her.}{12:9}{{\raggedright Is my heritage to me as a speckled bird of prey? \poemline{2}{Are the birds of prey against her all around?}\poemline{1}{Go, assemble all the animals of the field.}\poemline{2}{Bring them to devour.}}}
\VersePair{12:10}{Many shepherds have destroyed my vineyard, they have defiled my portion, they have made my desirable portion a trackless wilderness;}{12:10}{{\raggedright Many shepherds have destroyed my vineyard. \poemline{2}{They have trodden my portion under foot.}\poemline{2}{They have made my pleasant portion a desolate wilderness.}}}
\VersePair{12:11}{it is made a complete ruin: for my sake the whole land has been utterly ruined, because there is none that lays the matter to heart.}{12:11}{{\raggedright They have made it a desolation. \poemline{2}{It mourns to me, being desolate.}\poemline{1}{The whole land is made desolate,}\poemline{2}{because no one cares.}}}
\VersePair{12:12}{The ravagers are come to every passage in the wilderness: for the sword of the Lord will devour from one end of the land to the other: no flesh has any peace.}{12:12}{{\raggedright Destroyers have come on all the bare heights in the wilderness; \poemline{2}{for the sword of Yahweh devours from the one end of the land even to the other end of the land.}\poemline{2}{No flesh has peace.}}}
\VersePair{12:13}{Sow wheat, and reap thorns; their portions shall not profit them: be ashamed of your boasting, because of reproach before the Lord.}{12:13}{{\raggedright They have sown wheat, \poemline{2}{and have reaped thorns.}\poemline{1}{They have exhausted themselves,}\poemline{2}{and profit nothing.}\poemline{1}{You will be ashamed of your fruits,}\poemline{2}{because of Yahweh’s fierce anger.”}}}
\VersePair{12:14}{For thus saith the Lord concerning all the evil neighbours that touch mine inheritance, which I have divided to my people Israel; Behold, I will draw them away from their land, and I will cast out Juda from the midst of them.}{12:14}{Yahweh says, “Concerning all my evil neighbors, who touch the inheritance which I have caused my people Israel to inherit: Behold, I will pluck them up from off their land, and will pluck up the house of Judah from among them.}
\VersePair{12:15}{And it shall come to pass, after I have cast them out, that I will return, and have mercy upon them, and will cause them to dwell every one in his inheritance, and every one in his land.}{12:15}{It will happen that after I have plucked them up, I will return and have compassion on them. I will bring them again, every man to his heritage, and every man to his land.}
\VersePair{12:16}{And it shall be, if they will indeed learn the way of my people, to swear by my name, saying, The Lord lives; as they taught my people to swear by Baal; then shall that nation be built in the midst of my people.}{12:16}{It will happen, if they will diligently learn the ways of my people, to swear by my name, ‘As Yahweh lives;’ even as they taught my people to swear by Baal, then they will be built up in the middle of my people.}
\VersePair{12:17}{But if they will not return, then will I cut off that nation with utter ruin and destruction.}{12:17}{But if they will not hear, then I will pluck up that nation, plucking up and destroying it,” says Yahweh.}
\end{paracol}
\ChapterHeading{13}
\begin{paracol}{2}
\VersePair{13:1}{Thus saith the Lord, Go and procure for thyself a linen girdle, and put it about thy loins, and let it not be put in water.}{13:1}{Yahweh said to me, “Go, and buy yourself a linen belt, and put it on your waist, and don’t put it in water.”}
\VersePair{13:2}{So I procured the girdle according to the word of the Lord, and put it about my loins.}{13:2}{So I bought a belt according to Yahweh’s word, and put it on my waist.}
\VersePair{13:3}{And the word of the Lord came to me, saying,}{13:3}{Yahweh’s word came to me the second time, saying,}
\VersePair{13:4}{Take the girdle that is upon thy loins, and arise, and go to the Euphrates, and hide it there in a hole of the rock.}{13:4}{“Take the belt that you have bought, which is on your waist, and arise, go to the Euphrates, and hide it there in a cleft of the rock.”}
\VersePair{13:5}{So I went, and hid it by the Euphrates, as the Lord commanded me.}{13:5}{So I went and hid it by the Euphrates, as Yahweh commanded me.}
\VersePair{13:6}{And it came to pass after many days, that the Lord said to me, Arise, go to the Euphrates, and take thence the girdle, which I commanded thee to hide there.}{13:6}{After many days, Yahweh said to me, “Arise, go to the Euphrates, and take the belt from there, which I commanded you to hide there.”}
\VersePair{13:7}{So I went to the river Euphrates, and dug, and took the girdle out of the place where I had buried it: and, behold, it was rotten, utterly good for nothing.}{13:7}{Then I went to the Euphrates, and dug, and took the belt from the place where I had hidden it; and behold, the belt was ruined. It was profitable for nothing.}
\VersePair{13:8}{And the word of the Lord came to me, saying, Thus saith the Lord,}{13:8}{Then Yahweh’s word came to me, saying,}
\VersePair{13:9}{Thus will I mar the pride of Juda, and the pride of Jerusalem;}{13:9}{“Yahweh says, ‘In this way I will ruin the pride of Judah, and the great pride of Jerusalem.}
\VersePair{13:10}{even this great pride of the men that will not hearken to my words, and have gone after strange gods, to serve them, and to worship them: and they shall be as this girdle, which can be used for nothing.}{13:10}{This evil people, who refuse to hear my words, who walk in the stubbornness of their heart, and have gone after other gods to serve them and to worship them, will even be as this belt, which is profitable for nothing.}
\VersePair{13:11}{For as a girdle cleaves about the loins of a man, so have I caused to cleave to myself the house of Israel, and the whole house of Juda; that they might be to me a famous people, and a praise, and a glory: but they did not hearken to me.}{13:11}{For as the belt clings to the waist of a man, so I have caused the whole house of Israel and the whole house of Judah to cling to me,’ says Yahweh; ‘that they may be to me for a people, for a name, for praise, and for glory; but they would not hear.’}
\VersePair{13:12}{And thou shalt say to this people, Every bottle shall be filled with wine: and it shall come to pass, if they shall say to thee, Shall we not certainly know that every bottle shall be filled with wine? that thou shalt say to them,}{13:12}{“Therefore you shall speak to them this word: ‘Yahweh, the God of Israel says, “Every container should be filled with wine.”’ They will tell you, ‘Do we not certainly know that every container should be filled with wine?’}
\VersePair{13:13}{Thus saith the Lord, Behold, I will fill the inhabitants of this land, and their kings the sons of David that sit upon their throne, and the priests, and the prophets, and Juda and all the dwellers in Jerusalem, with strong drink.}{13:13}{Then tell them, ‘Yahweh says, “Behold, I will fill all the inhabitants of this land, even the kings who sit on David’s throne, the priests, the prophets, and all the inhabitants of Jerusalem, with drunkenness.}
\VersePair{13:14}{And I will scatter them a man and his brother, and their fathers and their sons together: I will not have compassion, saith the Lord, and I will not spare, neither will I pity to save them from destruction.}{13:14}{I will dash them one against another, even the fathers and the sons together,” says Yahweh: “I will not pity, spare, or have compassion, that I should not destroy them.”’”}
\VersePair{13:15}{Hear ye, and give ear, and be not proud: for the Lord has spoken.}{13:15}{{\raggedright Hear, and give ear. \poemline{2}{Don’t be proud,}\poemline{2}{for Yahweh has spoken.}}}
\VersePair{13:16}{Give glory to the Lord your God, before he cause darkness, and before your feet stumble on the dark mountains, and ye shall wait for light, and behold the shadow of death, and they shall be brought into darkness.}{13:16}{{\raggedright Give glory to Yahweh your God, \poemline{2}{before he causes darkness,}\poemline{2}{and before your feet stumble on the dark mountains,}\poemline{1}{and while you look for light,}\poemline{2}{he turns it into the shadow of death,}\poemline{2}{and makes it deep darkness.}}}
\VersePair{13:17}{But if ye will not hearken, your soul shall weep in secret because of pride, and your eyes shall pour down tears, because the Lord's flock is sorely bruised.}{13:17}{{\raggedright But if you will not hear it, \poemline{2}{my soul will weep in secret for your pride.}\poemline{1}{My eye will weep bitterly,}\poemline{2}{and run down with tears,}\poemline{2}{because Yahweh’s flock has been taken captive.}}}
\VersePair{13:18}{Say ye to the king and the princes, Humble yourselves, and sit down; for your crown of glory is removed from your head.}{13:18}{{\raggedright Say to the king and to the queen mother, \poemline{2}{“Humble yourselves.}\poemline{1}{Sit down, for your crowns have come down,}\poemline{2}{even the crown of your glory.}}}
\VersePair{13:19}{The cities toward the south were shut, and there was none to open them: Juda is removed into captivity, they have suffered a complete removal.}{13:19}{{\raggedright The cities of the South are shut up, \poemline{2}{and there is no one to open them.}\poemline{1}{Judah is carried away captive: all of them.}\poemline{2}{They are wholly carried away captive.}}}
\VersePair{13:20}{Lift up thine eyes, O Jerusalem, and behold them that come from the north; where is the flock that was given thee, the sheep of thy glory?}{13:20}{{\raggedright Lift up your eyes, \poemline{2}{and see those who come from the north.}\poemline{1}{Where is the flock that was given to you,}\poemline{2}{your beautiful flock?}}}
\VersePair{13:21}{What wilt thou say when they shall visit thee, for thou didst teach them lessons for rule against thyself; shall not pangs seize thee as a woman in travail?}{13:21}{{\raggedright What will you say when he sets over you as head those whom you have yourself taught to be friends to you? \poemline{2}{Won’t sorrows take hold of you, as of a woman in travail?}}}
\VersePair{13:22}{And if thou shouldest say in thine heart, Wherefore have these things happened to me? Because of the abundance of thine iniquity have thy skirts been discovered, that thine heels might be exposed.}{13:22}{{\raggedright If you say in your heart, \poemline{2}{“Why have these things come on me?”}\poemline{1}{Your skirts are uncovered because of the greatness of your iniquity,}\poemline{2}{and your heels suffer violence.}}}
\VersePair{13:23}{If the Ethiopian shall change his skin, or the leopardess her spots, then shall ye be able to do good, having learnt evil.}{13:23}{{\raggedright Can the Ethiopian change his skin, \poemline{2}{or the leopard his spots?}\poemline{1}{Then may you also do good,}\poemline{2}{who are accustomed to do evil.}}}
\VersePair{13:24}{So I scattered them as sticks carried by the wind into the wilderness.}{13:24}{{\raggedright “Therefore I will scatter them \poemline{2}{as the stubble that passes away}\poemline{2}{by the wind of the wilderness.}}}
\VersePair{13:25}{Thus is thy lot, and the reward of your disobedience to me, saith the Lord; as thou didst forget me, and trust in lies,}{13:25}{{\raggedright This is your lot, \poemline{2}{the portion measured to you from me,” says Yahweh,}\poemline{1}{“because you have forgotten me,}\poemline{2}{and trusted in falsehood.”}}}
\VersePair{13:26}{I also will expose thy skirts upon thy face, and thy shame shall be seen;}{13:26}{{\raggedright Therefore I will also uncover your skirts on your face, \poemline{2}{and your shame will appear.}}}
\VersePair{13:27}{thine adultery also, and thy neighing, and the looseness of thy fornication: on the hills and in the fields I have seen thine abominations. Woe to thee, O Jerusalem, for thou hast not been purified so as to follow me; how long yet shall it be?}{13:27}{{\raggedright I have seen your abominations, even your adulteries \poemline{2}{and your neighing, the lewdness of your prostitution,}\poemline{2}{on the hills in the field.}\poemline{1}{Woe to you, Jerusalem!}\poemline{2}{You will not be made clean.}\poemline{2}{How long will it yet be?”}}}
\end{paracol}
\ChapterHeading{14}
\begin{paracol}{2}
\VersePair{14:1}{And the Word of the Lord came to Jeremias concerning the Drought.}{14:1}{This is Yahweh’s word that came to Jeremiah concerning the drought:}
\VersePair{14:2}{Judea has mourned, and her gates are emptied, and are darkened upon the earth; and the cry of Jerusalem is gone up.}{14:2}{{\raggedright “Judah mourns, \poemline{2}{and its gates languish.}\poemline{1}{They sit in black on the ground.}\poemline{2}{The cry of Jerusalem goes up.}}}
\VersePair{14:3}{And her nobles have sent their little ones to the water: they came to the wells, and found no water: and brought back their vessels empty.}{14:3}{{\raggedright Their nobles send their little ones to the waters. \poemline{2}{They come to the cisterns,}\poemline{2}{and find no water.}\poemline{1}{They return with their vessels empty.}\poemline{2}{They are disappointed and confounded,}\poemline{2}{and cover their heads.}}}
\VersePair{14:4}{And the labours of the land failed, because there was no rain: the husbandmen were ashamed, they covered their heads.}{14:4}{{\raggedright Because of the ground which is cracked, \poemline{2}{because no rain has been in the land,}\poemline{1}{the plowmen are disappointed.}\poemline{2}{They cover their heads.}}}
\VersePair{14:5}{And hinds calved in the field, and forsook it, because there was no grass.}{14:5}{{\raggedright Yes, the doe in the field also calves and forsakes her young, \poemline{2}{because there is no grass.}}}
\VersePair{14:6}{The wild asses stood by the forests, and snuffed up the wind; their eyes failed, because there was no grass.}{14:6}{{\raggedright The wild donkeys stand on the bare heights. \poemline{2}{They pant for air like jackals.}\poemline{1}{Their eyes fail,}\poemline{2}{because there is no vegetation.}}}
\VersePair{14:7}{Our sins have risen up against us: O Lord, do thou for us for thine own sake; for our sins are many before thee; for we have sinned against thee.}{14:7}{{\raggedright Though our iniquities testify against us, \poemline{2}{work for your name’s sake, Yahweh;}\poemline{1}{for our rebellions are many.}\poemline{2}{We have sinned against you.}}}
\VersePair{14:8}{O Lord, thou art the hope of Israel, and deliverest us in time of troubles; why art thou become as a sojourner upon the land, or as one born in the land, yet turning aside for a resting-place?}{14:8}{{\raggedright You hope of Israel, \poemline{2}{its Savior in the time of trouble,}\poemline{1}{why should you be as a foreigner in the land,}\poemline{2}{and as a wayfaring man who turns aside to stay for a night?}}}
\VersePair{14:9}{Wilt thou be as a man asleep, or as a strong man that cannot save? yet thou art among us, O Lord, and thy name is called upon us; forget us not.}{14:9}{{\raggedright Why should you be like a scared man, \poemline{2}{as a mighty man who can’t save?}\poemline{1}{Yet you, Yahweh, are in the middle of us,}\poemline{2}{and we are called by your name.}\poemline{2}{Don’t leave us.}}}
\VersePair{14:10}{Thus saith the Lord to this people, They have loved to wander, and they have not spared, therefore God has not prospered them; now will he remember their iniquity.}{14:10}{{\raggedright Yahweh says to this people: \poemline{1}{“Even so they have loved to wander.}\poemline{2}{They have not restrained their feet.}\poemline{1}{Therefore Yahweh does not accept them.}\poemline{2}{Now he will remember their iniquity,}\poemline{2}{and punish them for their sins.”}}}
\VersePair{14:11}{And the Lord said to me, Pray not for this people for their good:}{14:11}{Yahweh said to me, “Don’t pray for this people for their good.}
\VersePair{14:12}{for though they fast, I will not hear their supplication; and though they offer whole-burnt-offerings and sacrifices, I will take no pleasure in them: for I will consume them with sword, and with famine, and with pestilence.}{14:12}{When they fast, I will not hear their cry; and when they offer burnt offering and meal offering, I will not accept them; but I will consume them by the sword, by famine, and by pestilence.”}
\VersePair{14:13}{And I said, O ever living Lord! behold, their prophets prophesy, and say, Ye shall not see a sword, nor shall famine be among you; for I will give truth and peace on the land, and in this place.}{14:13}{Then I said, “Ah, Lord Yahweh! Behold, the prophets tell them, ‘You will not see the sword, neither will you have famine; but I will give you assured peace in this place.’”}
\VersePair{14:14}{Then the Lord said to me, The prophets prophesy lies in my name: I sent them not, and I commanded them not, and I spoke not to them: for they prophesy to you false visions, and divinations, and auguries, and devices of their own heart.}{14:14}{Then Yahweh said to me, “The prophets prophesy lies in my name. I didn’t send them. I didn’t command them. I didn’t speak to them. They prophesy to you a lying vision, divination, and a thing of nothing, and the deceit of their own heart.}
\VersePair{14:15}{Therefore thus saith the Lord concerning the prophets that prophesy lies in my name, and I sent them not, who say, Sword and famine shall not be upon this land; they shall die by a grievous death, and the prophets shall be consumed by famine.}{14:15}{Therefore Yahweh says concerning the prophets who prophesy in my name, but I didn’t send them, yet they say, ‘Sword and famine will not be in this land.’ Those prophets will be consumed by sword and famine.}
\VersePair{14:16}{And the people to whom they prophesy, they also shall be cast out in the streets of Jerusalem, because of the sword and famine; and there shall be none to bury them: their wives also, and their sons, and their daughters shall die thus; and I will pour out their wickedness upon them.}{14:16}{The people to whom they prophesy will be cast out in the streets of Jerusalem because of the famine and the sword. They will have no one to bury them — them, their wives, their sons, or their daughters, for I will pour their wickedness on them.}
\VersePair{14:17}{And thou shalt speak this word to them; Let your eyes shed tears day and night, and let them not cease: for the daughter of my people has been sorely bruised, and her plague is very grievous.}{14:17}{{\raggedright “You shall say this word to them: \poemline{1}{“‘Let my eyes run down with tears night and day,}\poemline{2}{and let them not cease;}\poemline{1}{for the virgin daughter of my people is broken with a great breach,}\poemline{2}{with a very grievous wound.}}}
\VersePair{14:18}{If I go forth into the plain, then behold the slain by the sword! and if I enter into the city, then behold the distress of famine! for priest and prophet have gone to a land which they knew not.}{14:18}{{\raggedright If I go out into the field, \poemline{2}{then behold, the slain with the sword!}\poemline{1}{If I enter into the city,}\poemline{2}{then behold, those who are sick with famine!}\poemline{1}{For both the prophet and the priest go about in the land,}\poemline{2}{and have no knowledge.’”}}}
\VersePair{14:19}{Hast thou utterly rejected Juda? and has thy soul departed from Sion? wherefore hast thou smitten us, and there is no healing for us? we waited for peace, but there was no prosperity; for a time of healing, and behold trouble!}{14:19}{{\raggedright Have you utterly rejected Judah? \poemline{2}{Has your soul loathed Zion?}\poemline{1}{Why have you struck us, and there is no healing for us?}\poemline{2}{We looked for peace, but no good came;}\poemline{2}{and for a time of healing, and behold, dismay!}}}
\VersePair{14:20}{We know, O Lord, our sins, and the iniquities of our fathers: for we have sinned before thee.}{14:20}{{\raggedright We acknowledge, Yahweh, our wickedness, \poemline{2}{and the iniquity of our fathers;}\poemline{2}{for we have sinned against you.}}}
\VersePair{14:21}{Refrain for thy name's sake, destroy not the throne of thy glory: remember, break not thy covenant with us.}{14:21}{{\raggedright Do not abhor us, for your name’s sake. \poemline{2}{Do not disgrace the throne of your glory.}\poemline{2}{Remember, and don’t break your covenant with us.}}}
\VersePair{14:22}{Is there any one among the idols of the Gentiles that can give rain? and will the sky yield his fulness at their bidding? Art not thou he? we will even wait on thee, O Lord: for thou hast made all these things.}{14:22}{{\raggedright Are there any among the vanities of the nations that can cause rain? \poemline{2}{Or can the sky give showers?}\poemline{2}{Aren’t you he, Yahweh our God?}\poemline{1}{Therefore we will wait for you;}\poemline{2}{for you have made all these things.}}}
\end{paracol}
\ChapterHeading{15}
\begin{paracol}{2}
\VersePair{15:1}{And the Lord said to me, Though Moses and Samuel stood before my face, my soul could not be toward them: dismiss this people, and let them go forth.}{15:1}{Then Yahweh said to me, “Though Moses and Samuel stood before me, yet my mind would not turn toward this people. Cast them out of my sight, and let them go out!}
\VersePair{15:2}{And it shall be, if they say to thee, Whither shall we go forth? then thou shalt say to them, Thus saith the Lord; As many as are for death, to death; and as many as are for famine, to famine; and as many as are for the sword, to the sword; and as many as are for captivity, to captivity.}{15:2}{{\raggedright It will happen when they ask you, ‘Where shall we go out?’ then you shall tell them, ‘Yahweh says: \poemline{1}{“Such as are for death, to death;}\poemline{1}{such as are for the sword, to the sword;}\poemline{1}{such as are for the famine, to the famine;}\poemline{1}{and such as are for captivity, to captivity.”’}}}
\VersePair{15:3}{And I will punish them with four kinds of death, saith the Lord, the sword to slay, and the dogs to tear, and the wild beasts of the earth, and the birds of the sky to devour and destroy.}{15:3}{“I will appoint over them four kinds,” says Yahweh: “the sword to kill, the dogs to tear, the birds of the sky, and the animals of the earth, to devour and to destroy.}
\VersePair{15:4}{And I will deliver them up for distress to all the kingdoms of the earth, because of Manasses son of Ezekias king of Juda, for all that he did in Jerusalem.}{15:4}{I will cause them to be tossed back and forth among all the kingdoms of the earth, because of Manasseh, the son of Hezekiah, king of Judah, for that which he did in Jerusalem.}
\VersePair{15:5}{Who will spare thee, O Jerusalem? and who will fear for thee? or who will turn back to ask for thy welfare?}{15:5}{{\raggedright For who will have pity on you, Jerusalem? \poemline{2}{Who will mourn you?}\poemline{2}{Who will come to ask of your welfare?}}}
\VersePair{15:6}{Thou hast turned away from me, saith the Lord, thou wilt go back: therefore I will stretch out my hand, and will destroy thee, and will no more spare them.}{15:6}{{\raggedright You have rejected me,” says Yahweh. \poemline{2}{“You have gone backward.}\poemline{1}{Therefore I have stretched out my hand against you}\poemline{2}{and destroyed you.}\poemline{2}{I am weary of showing compassion.}}}
\VersePair{15:7}{And I will completely scatter them; in the gates of my people they are bereaved of children: they have destroyed my people because of their iniquities.}{15:7}{{\raggedright I have winnowed them with a fan in the gates of the land. \poemline{2}{I have bereaved them of children.}\poemline{1}{I have destroyed my people.}\poemline{2}{They didn’t return from their ways.}}}
\VersePair{15:8}{Their widows have been multiplied more than the sand of the sea: I have brought young men against the mother, even distress at noon-day: I have suddenly cast upon her trembling and anxiety.}{15:8}{{\raggedright Their widows are increased more than the sand of the seas. \poemline{2}{I have brought on them against the mother of the young men a destroyer at noonday.}\poemline{2}{I have caused anguish and terrors to fall on her suddenly.}}}
\VersePair{15:9}{She that bore seven is spent; her soul has fainted under trouble; her sun is gone down while it is yet noon; she is ashamed and disgraced: I will give the remnant of them to the sword before their enemies.}{15:9}{{\raggedright She who has borne seven languishes. \poemline{2}{She has given up the spirit.}\poemline{1}{Her sun has gone down while it was yet day.}\poemline{2}{She has been disappointed and confounded.}\poemline{2}{I will deliver their residue to the sword before their enemies,” says Yahweh.}}}
\VersePair{15:10}{Woe is me, my mother! thou hast born me as some man of strife, and at variance with the whole earth; I have not helped others, nor has any one helped me; my strength has failed among them that curse me.}{15:10}{{\raggedright Woe is me, my mother, that you have borne me, a man of strife, \poemline{2}{and a man of contention to the whole earth!}\poemline{1}{I have not lent, neither have men lent to me;}\poemline{2}{yet every one of them curses me.}}}
\VersePair{15:11}{Be it so, Lord, in their prosperity; surely I stood before thee in the time of their calamities, and in the time of their affliction, for their good against the enemy.}{15:11}{{\raggedright Yahweh said, \poemline{1}{“Most certainly I will strengthen you for good.}\poemline{2}{Most certainly I will cause the enemy to make supplication to you in the time of evil}\poemline{2}{and in the time of affliction.}}}
\VersePair{15:12}{Will iron be known? whereas thy strength is a brazen covering.}{15:12}{{\raggedright Can one break iron, \poemline{2}{even iron from the north, and bronze?}}}
\VersePair{15:13}{Yea, I will give thy treasures for a spoil as a recompence, because of all thy sins, and that in all thy borders.}{15:13}{{\raggedright I will give your substance and your treasures for a plunder without price, \poemline{2}{and that for all your sins,}\poemline{2}{even in all your borders.}}}
\VersePair{15:14}{And I will enslave thee to thine enemies round about, in a land which thou hast not known; for a fire has been kindled out of my wrath; it shall burn upon you.}{15:14}{{\raggedright I will make them to pass with your enemies into a land which you don’t know; \poemline{2}{for a fire is kindled in my anger,}\poemline{2}{which will burn on you.”}}}
\VersePair{15:15}{O Lord, remember me, and visit me, and vindicate me before them that persecute me; do not bear long with them; know how I have met with reproach for thy sake, from those who set at nought thy words;}{15:15}{{\raggedright Yahweh, you know. \poemline{2}{Remember me, visit me,}\poemline{2}{and avenge me of my persecutors.}\poemline{1}{You are patient, so don’t take me away.}\poemline{2}{Know that for your sake I have suffered reproach.}}}
\VersePair{15:16}{consume them; and thy word shall be to me for the joy and gladness of my heart: for thy name has been called upon me, O Lord Almighty.}{15:16}{{\raggedright Your words were found, \poemline{2}{and I ate them.}\poemline{1}{Your words were to me a joy and the rejoicing of my heart,}\poemline{2}{for I am called by your name, Yahweh, God of Armies.}}}
\VersePair{15:17}{I have not sat in the assembly of them as they mocked, but I feared because of thy power: I sat alone, for I was filled with bitterness.}{15:17}{{\raggedright I didn’t sit in the assembly of those who make merry and rejoice. \poemline{2}{I sat alone because of your hand,}\poemline{2}{for you have filled me with indignation.}}}
\VersePair{15:18}{Why do they that grieve me prevail against me? my wound is severe; whence shall I be healed? it is indeed become to me as deceitful water, that has no faithfulness.}{15:18}{{\raggedright Why is my pain perpetual, \poemline{2}{and my wound incurable,}\poemline{2}{which refuses to be healed?}\poemline{1}{Will you indeed be to me as a deceitful brook,}\poemline{2}{like waters that fail?}}}
\VersePair{15:19}{Therefore thus saith the Lord, If thou wilt return, then will I restore thee, and thou shalt stand before my face: and if thou wilt bring forth the precious from the worthless, thou shalt be as my mouth: and they shall return to thee; but thou shalt not return to them.}{15:19}{{\raggedright Therefore Yahweh says, \poemline{1}{“If you return, then I will bring you again,}\poemline{2}{that you may stand before me;}\poemline{1}{and if you take out the precious from the vile,}\poemline{2}{you will be as my mouth.}\poemline{1}{They will return to you,}\poemline{2}{but you will not return to them.}}}
\VersePair{15:20}{And I will make thee to this people as a strong brazen wall; and they shall fight against thee, but they shall by no means prevail against thee;}{15:20}{{\raggedright I will make you to this people a fortified bronze wall. \poemline{2}{They will fight against you,}\poemline{2}{but they will not prevail against you;}\poemline{1}{for I am with you to save you}\poemline{2}{and to deliver you,” says Yahweh.}}}
\VersePair{15:21}{for I am with thee to save thee, and to deliver thee out of the hand of wicked men; and I will ransom thee out of the hand of pestilent men.}{15:21}{{\raggedright “I will deliver you out of the hand of the wicked, \poemline{2}{and I will redeem you out of the hand of the terrible.”}}}
\end{paracol}
\ChapterHeading{16}
\begin{paracol}{2}
\VersePair{16:1}{And thou shalt not take a wife, saith the Lord God of Israel:}{16:1}{Then Yahweh’s word came to me, saying,}
\VersePair{16:2}{and there shall be no son born to thee, nor daughter in this place.}{16:2}{“You shall not take a wife, neither shall you have sons or daughters, in this place.”}
\VersePair{16:3}{For thus saith the Lord concerning the sons and concerning the daughters that are born in this place, and concerning their mothers that have born them, and concerning their fathers that have begotten them in this land;}{16:3}{For Yahweh says concerning the sons and concerning the daughters who are born in this place, and concerning their mothers who bore them, and concerning their fathers who became their father in this land:}
\VersePair{16:4}{They shall die of grievous death; they shall not be lamented, nor buried; they shall be for an example on the face of the earth; and they shall be for the wild beasts of the land, and for the birds of the sky: they shall fall by the sword, and shall be consumed with famine.}{16:4}{“They will die grievous deaths. They will not be lamented, neither will they be buried. They will be as dung on the surface of the ground. They will be consumed by the sword and by famine. Their dead bodies will be food for the birds of the sky and for the animals of the earth.”}
\VersePair{16:5}{Thus saith the Lord, Enter not into their mourning feast, and go not to lament, and mourn not for them: for I have removed my peace from this people.}{16:5}{For Yahweh says, “Don’t enter into the house of mourning. Don’t go to lament. Don’t bemoan them, for I have taken away my peace from this people,” says Yahweh, “even loving kindness and tender mercies.}
\VersePair{16:6}{They shall not bewail them, nor make cuttings for them, and they shall not shave themselves for them:}{16:6}{Both great and small will die in this land. They will not be buried. Men won’t lament for them, cut themselves, or make themselves bald for them.}
\VersePair{16:7}{and there shall be no bread broken in mourning for them for consolation over the dead: they shall not give one to drink a cup for consolation over his father or his mother.}{16:7}{Men won’t break bread for them in mourning, to comfort them for the dead. Men won’t give them the cup of consolation to drink for their father or for their mother.}
\VersePair{16:8}{Thou shalt not enter into the banquet-house, to sit with them to eat and to drink.}{16:8}{“You shall not go into the house of feasting to sit with them, to eat and to drink.”}
\VersePair{16:9}{For thus saith the Lord God of Israel; Behold, I will make to cease out of this place before your eyes, and in your days, the voice of joy, and the voice of gladness, the voice of the bridegroom, and the voice of the bride.}{16:9}{For Yahweh of Armies, the God of Israel says: “Behold, I will cause to cease out of this place, before your eyes and in your days, the voice of mirth and the voice of gladness, the voice of the bridegroom and the voice of the bride.}
\VersePair{16:10}{And it shall come to pass, when thou shalt report to this people all these words, and they shall say to thee, Wherefore has the Lord pronounced against us all these evils? what is our unrighteousness? and what is our sin which we have sinned before the Lord our God?}{16:10}{It will happen, when you tell this people all these words, and they ask you, ‘Why has Yahweh pronounced all this great evil against us?’ or ‘What is our iniquity?’ or ‘What is our sin that we have committed against Yahweh our God?’}
\VersePair{16:11}{Then thou shalt say to them, Because your fathers forsook me, saith the Lord, and went after strange gods and served them, and worshipped them, and forsook me, and kept not my law;}{16:11}{then you shall tell them, ‘Because your fathers have forsaken me,’ says Yahweh, ‘and have walked after other gods, have served them, have worshiped them, have forsaken me, and have not kept my law.}
\VersePair{16:12}{(and ye sinned worse than your fathers; for, behold, ye walk every one after the lusts of your own evil heart, so as not to hearken to me);}{16:12}{You have done evil more than your fathers, for behold, you each walk after the stubbornness of his evil heart, so that you don’t listen to me.}
\VersePair{16:13}{therefore I will cast you off from this good land into a land which neither ye nor your fathers have known; and ye shall serve there other gods, who shall have no mercy upon you.}{16:13}{Therefore I will cast you out of this land into the land that you have not known, neither you nor your fathers. There you will serve other gods day and night, for I will show you no favor.’}
\VersePair{16:14}{Therefore, behold, the days come, saith the Lord, when they shall no more say, The Lord lives, that brought up the children of Israel out of the land of Egypt;}{16:14}{“Therefore behold, the days come,” says Yahweh, “that it will no more be said, ‘As Yahweh lives, who brought up the children of Israel out of the land of Egypt;’}
\VersePair{16:15}{but, The Lord lives, who brought up the house of Israel from the land of the north, and from all the countries whither they were thrust out: and I will restore them to their own land, which I gave to their fathers.}{16:15}{but, ‘As Yahweh lives, who brought up the children of Israel from the land of the north, and from all the countries where he had driven them.’ I will bring them again into their land that I gave to their fathers.}
\VersePair{16:16}{Behold, I will send many fishers, saith the Lord, and they shall fish them; and afterward I will send many hunters, and they shall hunt them upon every mountain, and upon every hill, and out of the holes of the rocks.}{16:16}{“Behold, I will send for many fishermen,” says Yahweh, “and they will fish them up. Afterward I will send for many hunters, and they will hunt them from every mountain, from every hill, and out of the clefts of the rocks.}
\VersePair{16:17}{For mine eyes are upon all their ways; and their iniquities have not been hidden from mine eyes.}{16:17}{For my eyes are on all their ways. They are not hidden from my face. Their iniquity isn’t concealed from my eyes.}
\VersePair{16:18}{And I will recompense their mischiefs doubly, and their sins, whereby they have profaned my land with the carcases of their abominations, and with their iniquities, whereby they have trespassed against mine inheritance.}{16:18}{First I will recompense their iniquity and their sin double, because they have polluted my land with the carcasses of their detestable things, and have filled my inheritance with their abominations.”}
\VersePair{16:19}{O Lord, thou art my strength, and mine help, and my refuge in days of evil: to thee the Gentiles shall come from the end of the earth, and shall say, How vain were the idols which our fathers procured to themselves, and there is no help in them.}{16:19}{{\raggedright Yahweh, my strength, my stronghold, \poemline{2}{and my refuge in the day of affliction,}\poemline{1}{the nations will come to you from the ends of the earth,}\poemline{2}{and will say,}\poemline{1}{“Our fathers have inherited nothing but lies,}\poemline{2}{vanity and things in which there is no profit.}}}
\VersePair{16:20}{Will a man make gods for himself, whereas these are no gods?}{16:20}{{\raggedright Should a man make to himself gods \poemline{2}{which yet are no gods?”}}}
\VersePair{16:21}{Therefore, behold, I will at this time manifest my hand to them, and will make known to them my power; and they shall know that my name is the Lord.}{16:21}{{\raggedright “Therefore behold, I will cause them to know, \poemline{2}{this once I will cause them to know my hand and my might.}\poemline{2}{Then they will know that my name is Yahweh.”}}}
\end{paracol}
\ChapterHeading{17}
\begin{paracol}{2}
\VersePair{17:1}{}{17:1}{{\raggedright “The sin of Judah is written with a pen of iron, \poemline{2}{and with the point of a diamond.}\poemline{1}{It is engraved on the tablet of their heart,}\poemline{2}{and on the horns of your altars.}}}
\VersePair{17:2}{}{17:2}{{\raggedright Even their children remember their altars \poemline{2}{and their Asherah poles by the green trees on the high hills.}}}
\VersePair{17:3}{}{17:3}{{\raggedright My mountain in the field, \poemline{2}{I will give your substance and all your treasures for a plunder,}\poemline{2}{and your high places, because of sin, throughout all your borders.}}}
\VersePair{17:4}{}{17:4}{{\raggedright You, even of yourself, will discontinue from your heritage that I gave you. \poemline{2}{I will cause you to serve your enemies in the land which you don’t know,}\poemline{2}{for you have kindled a fire in my anger which will burn forever.”}}}
\VersePair{17:5}{Cursed is the man who trusts in man, and will lean his arm of flesh upon him, while his heart departs from the Lord.}{17:5}{{\raggedright Yahweh says: \poemline{1}{“Cursed is the man who trusts in man,}\poemline{2}{relies on strength of flesh,}\poemline{2}{and whose heart departs from Yahweh.}}}
\VersePair{17:6}{And he shall be as the wild tamarisk in the desert: he shall not see when good comes; but he shall dwell in barren places, and in the wilderness, in a salt land which is not inhabited.}{17:6}{{\raggedright For he will be like a bush in the desert, \poemline{2}{and will not see when good comes,}\poemline{1}{but will inhabit the parched places in the wilderness,}\poemline{2}{an uninhabited salt land.}}}
\VersePair{17:7}{But blessed is the man who trusts in the Lord, and whose hope the Lord shall be.}{17:7}{{\raggedright “Blessed is the man who trusts in Yahweh, \poemline{2}{and whose confidence is in Yahweh.}}}
\VersePair{17:8}{And he shall be as a thriving tree by the waters, and he shall cast forth his root toward a moist place: he shall not fear when heat comes, and there shall be upon him shady branches: he shall not fear in a year of drought, and he shall not fail to bear fruit.}{17:8}{{\raggedright For he will be as a tree planted by the waters, \poemline{2}{who spreads out its roots by the river,}\poemline{1}{and will not fear when heat comes,}\poemline{2}{but its leaf will be green,}\poemline{1}{and will not be concerned in the year of drought.}\poemline{2}{It won’t cease from yielding fruit.}}}
\VersePair{17:9}{The heart is deep beyond all things, and it is the man, and who can know him?}{17:9}{{\raggedright The heart is deceitful above all things \poemline{2}{and it is exceedingly corrupt.}\poemline{2}{Who can know it?}}}
\VersePair{17:10}{I the Lord try the hearts, and prove the reins, to give to every one according to his ways, and according to the fruits of his devices.}{17:10}{{\raggedright “I, Yahweh, search the mind. \poemline{2}{I try the heart,}\poemline{1}{even to give every man according to his ways,}\poemline{2}{according to the fruit of his doings.”}}}
\VersePair{17:11}{The partridge utters her voice, she gathers eggs which she did not lay; so is a man gaining his wealth unjustly; in the midst of his days his riches shall leave him, and at his latter end he will be a fool.}{17:11}{{\raggedright As the partridge that sits on eggs which she has not laid, \poemline{2}{so is he who gets riches, and not by right.}\poemline{1}{In the middle of his days, they will leave him.}\poemline{2}{At his end, he will be a fool.}}}
\VersePair{17:12}{An exalted throne of glory is our sanctuary.}{17:12}{{\raggedright A glorious throne, set on high from the beginning, \poemline{2}{is the place of our sanctuary.}}}
\VersePair{17:13}{O Lord, the hope of Israel, let all that have left thee be ashamed, let them that have revolted be written on the earth, because they have forsaken the fountain of life, the Lord.}{17:13}{{\raggedright Yahweh, the hope of Israel, \poemline{2}{all who forsake you will be disappointed.}\poemline{1}{Those who depart from me will be written in the earth,}\poemline{2}{because they have forsaken Yahweh,}\poemline{2}{the spring of living waters.}}}
\VersePair{17:14}{Heal me, O Lord, and I shall be healed; save me, and I shall be saved; for thou art my boast.}{17:14}{{\raggedright Heal me, O Yahweh, and I will be healed. \poemline{2}{Save me, and I will be saved;}\poemline{2}{for you are my praise.}}}
\VersePair{17:15}{Behold, they say to me, Where is the word of the Lord? let it come.}{17:15}{{\raggedright Behold, they ask me, \poemline{2}{“Where is Yahweh’s word?}\poemline{2}{Let it be fulfilled now.”}}}
\VersePair{17:16}{But I have not been weary of following thee, nor have I desired the day of man; thou knowest; the words that proceed out of my lips are before thy face.}{17:16}{{\raggedright As for me, I have not hurried from being a shepherd after you. \poemline{2}{I haven’t desired the woeful day. You know.}\poemline{2}{That which came out of my lips was before your face.}}}
\VersePair{17:17}{Be not to me a stranger, but spare me in the evil day.}{17:17}{{\raggedright Don’t be a terror to me. \poemline{2}{You are my refuge in the day of evil.}}}
\VersePair{17:18}{Let them that persecute me be ashamed, but let me not be ashamed: let them be alarmed, but let me not be alarmed: bring upon them the evil day, crush them with double destruction.}{17:18}{{\raggedright Let them be disappointed who persecute me, \poemline{2}{but don’t let me be disappointed.}\poemline{1}{Let them be dismayed,}\poemline{2}{but don’t let me be dismayed.}\poemline{1}{Bring on them the day of evil,}\poemline{2}{and destroy them with double destruction.}}}
\VersePair{17:19}{Thus saith the Lord; Go and stand in the gates of the children of thy people, by which the kings of Juda enter, and by which they go out, and in all the gates of Jerusalem:}{17:19}{Yahweh said this to me: “Go and stand in the gate of the children of the people, through which the kings of Judah come in and by which they go out, and in all the gates of Jerusalem.}
\VersePair{17:20}{and thou shalt say to them, Hear the word of the Lord, ye kings of Juda, and all Judea, and all Jerusalem, all who go in at these gates:}{17:20}{Tell them, ‘Hear Yahweh’s word, you kings of Judah, all Judah, and all the inhabitants of Jerusalem, that enter in by these gates:}
\VersePair{17:21}{thus saith the Lord; Take heed to your souls, and take up no burdens on the sabbath-day, and go not forth through the gates of Jerusalem;}{17:21}{Yahweh says, “Be careful, and bear no burden on the Sabbath day, nor bring it in by the gates of Jerusalem.}
\VersePair{17:22}{and carry forth no burdens out of your houses on the sabbath-day, and ye shall do no work: sanctify the sabbath-day, as I commanded your fathers.}{17:22}{Don’t carry a burden out of your houses on the Sabbath day. Don’t do any work, but make the Sabbath day holy, as I commanded your fathers.}
\VersePair{17:23}{But they hearkened not, and inclined not their ear, but stiffened their neck more than their fathers did, so as not to hear me, and not to receive correction.}{17:23}{But they didn’t listen. They didn’t turn their ear, but made their neck stiff, that they might not hear, and might not receive instruction.}
\VersePair{17:24}{And it shall come to pass, if ye will hearken to me, saith the Lord, to carry in no burdens through the gates of this city on the sabbath-day, and to sanctify the sabbath-day, so as to do no work upon it,}{17:24}{It will happen, if you diligently listen to me,” says Yahweh, “to bring in no burden through the gates of this city on the Sabbath day, but to make the Sabbath day holy, to do no work therein;}
\VersePair{17:25}{that there shall enter through the gates of this city kings and princes sitting on the throne of David, and riding on their chariots and horses, they, and their princes, the men of Juda, and the dwellers in Jerusalem: and this city shall be inhabited for ever.}{17:25}{then there will enter in by the gates of this city kings and princes sitting on David’s throne, riding in chariots and on horses, they and their princes, the men of Judah and the inhabitants of Jerusalem; and this city will remain forever.}
\VersePair{17:26}{And men shall come out of the cities of Juda, and from round about Jerusalem, and out of the land of Benjamin, and out of the plain country, and from the hill country, and from the south country, bringing whole-burnt-offerings, and sacrifices, and incense, and manna, and frankincense, bringing praise to the house of the Lord.}{17:26}{They will come from the cities of Judah, and from the places around Jerusalem, from the land of Benjamin, from the lowland, from the hill country, and from the South, bringing burnt offerings, sacrifices, meal offerings, and frankincense, and bringing sacrifices of thanksgiving to Yahweh’s house.}
\VersePair{17:27}{But it shall come to pass, if ye will not hearken to me to sanctify the sabbath-day, to bear no burdens, nor go in with them by the gates of Jerusalem on the sabbath-day; then will I kindle a fire in the gates thereof, and it shall devour the streets of Jerusalem, and shall not be quenched.}{17:27}{But if you will not listen to me to make the Sabbath day holy, and not to bear a burden and enter in at the gates of Jerusalem on the Sabbath day, then I will kindle a fire in its gates, and it will devour the palaces of Jerusalem. It will not be quenched.”’”}
\end{paracol}
\ChapterHeading{18}
\begin{paracol}{2}
\VersePair{18:1}{The word that came from the Lord to}{18:1}{The word which came to Jeremiah from Yahweh, saying,}
\VersePair{18:2}{Jeremias, saying, Arise, and go down to the potter's house, and there thou shalt hear my words.}{18:2}{“Arise, and go down to the potter’s house, and there I will cause you to hear my words.”}
\VersePair{18:3}{So I went down to the potter's house, and behold, he was making a vessel on the stones.}{18:3}{Then I went down to the potter’s house, and behold, he was making something on the wheels.}
\VersePair{18:4}{And the vessel which he was making with his hands fell: so he made it again another vessel, as it seemed good to him to make it.}{18:4}{When the vessel that he made of the clay was marred in the hand of the potter, he made it again another vessel, as seemed good to the potter to make it.}
\VersePair{18:5}{And the word of the Lord came to me, saying,}{18:5}{Then Yahweh’s word came to me, saying,}
\VersePair{18:6}{Shall I not be able, O house of Israel, to do to you as this potter? behold, as the clay of the potter are ye in my hands.}{18:6}{“House of Israel, can’t I do with you as this potter?” says Yahweh. “Behold, as the clay in the potter’s hand, so are you in my hand, house of Israel.}
\VersePair{18:7}{If I shall pronounce a decree upon a nation, or upon a kingdom, to cut them off, and to destroy them;}{18:7}{At the instant I speak concerning a nation, and concerning a kingdom, to pluck up and to break down and to destroy it,}
\VersePair{18:8}{and that nation turn from all their sins, then will I repent of the evils which I purposed to do to them.}{18:8}{if that nation, concerning which I have spoken, turns from their evil, I will repent of the evil that I thought to do to them.}
\VersePair{18:9}{And if I shall pronounce a decree upon a nation and kingdom, to rebuild and to plant it;}{18:9}{At the instant I speak concerning a nation, and concerning a kingdom, to build and to plant it,}
\VersePair{18:10}{and they do evil before me, so as not to hearken to my voice, then will I repent of the good which I spoke of, to do it to them.}{18:10}{if they do that which is evil in my sight, that they don’t obey my voice, then I will repent of the good with which I said I would benefit them.}
\VersePair{18:11}{And now say to the men of Juda, and to the inhabitants of Jerusalem, Behold, I prepare evils against you, and devise a device against you: let every one turn now from his evil way, and amend your practices.}{18:11}{“Now therefore, speak to the men of Judah, and to the inhabitants of Jerusalem, saying, ‘Yahweh says: “Behold, I frame evil against you, and devise a plan against you. Everyone return from his evil way now, and amend your ways and your doings.”’}
\VersePair{18:12}{And they said, We will quit ourselves like men, for we will pursue our perverse ways, and we will perform each the lusts of his evil heart.}{18:12}{But they say, ‘It is in vain; for we will walk after our own plans, and we will each follow the stubbornness of his evil heart.’”}
\VersePair{18:13}{Therefore thus saith the Lord; Enquire now among the nations, who has heard such very horrible things as the virgin of Israel has done?}{18:13}{{\raggedright Therefore Yahweh says: \poemline{1}{“Ask now among the nations,}\poemline{2}{‘Who has heard such things?’}\poemline{2}{The virgin of Israel has done a very horrible thing.}}}
\VersePair{18:14}{Will fertilising streams fail to flow from a rock, or snow fail from Libanus? will water violently impelled by the wind turn aside?}{18:14}{{\raggedright Will the snow of Lebanon fail from the rock of the field? \poemline{2}{Will the cold waters that flow down from afar be dried up?}}}
\VersePair{18:15}{For my people have forgotten me, they have offered incense in vain, and they fail in their ways, leaving the ancient tracks, to enter upon impassable paths;}{18:15}{{\raggedright For my people have forgotten me. \poemline{2}{They have burned incense to false gods.}\poemline{1}{They have been made to stumble in their ways}\poemline{2}{in the ancient paths,}\poemline{2}{to walk in byways, in a way not built up,}}}
\VersePair{18:16}{to make their land a desolation, and a perpetual hissing; all that go through it shall be amazed, and shall shake their heads.}{18:16}{{\raggedright to make their land an astonishment, \poemline{2}{and a perpetual hissing.}\poemline{1}{Everyone who passes by it will be astonished,}\poemline{2}{and shake his head.}}}
\VersePair{18:17}{I will scatter them before their enemies like an east wind; I will shew them the day of their destruction.}{18:17}{{\raggedright I will scatter them as with an east wind before the enemy. \poemline{2}{I will show them the back, and not the face,}\poemline{2}{in the day of their calamity.}}}
\VersePair{18:18}{Then they said, Come, and let us devise a device against Jeremias; for the law shall not perish from the priest, nor counsel from the wise, nor the word from the prophet. Come, and let us smite him with the tongue, and we will hear all his words.}{18:18}{Then they said, “Come! Let’s devise plans against Jeremiah; for the law won’t perish from the priest, nor counsel from the wise, nor the word from the prophet. Come, and let’s strike him with the tongue, and let’s not give heed to any of his words.”}
\VersePair{18:19}{Hear me, O Lord, and hear the voice of my pleading.}{18:19}{{\raggedright Give heed to me, Yahweh, \poemline{2}{and listen to the voice of those who contend with me.}}}
\VersePair{18:20}{Forasmuch as evil is rewarded for good; for they have spoken words against my soul, and they have hidden the punishment they meant for me; remember that I stood before thy face, to speak good for them, to turn away thy wrath from them.}{18:20}{{\raggedright Should evil be recompensed for good? \poemline{2}{For they have dug a pit for my soul.}\poemline{1}{Remember how I stood before you to speak good for them,}\poemline{2}{to turn away your wrath from them.}}}
\VersePair{18:21}{Therefore do thou deliver their sons to famine, and gather them to the power of the sword: let their women be childless and widows; and let their men be cut off by death, and their young men fall by the sword in war.}{18:21}{{\raggedright Therefore deliver up their children to the famine, \poemline{2}{and give them over to the power of the sword.}\poemline{1}{Let their wives become childless and widows.}\poemline{2}{Let their men be killed}\poemline{2}{and their young men struck by the sword in battle.}}}
\VersePair{18:22}{Let there be a cry in their houses: thou shalt bring upon them robbers suddenly: for they have formed a plan to take me, and have hidden snares for me.}{18:22}{{\raggedright Let a cry be heard from their houses \poemline{2}{when you bring a troop suddenly on them;}\poemline{1}{for they have dug a pit to take me}\poemline{2}{and hidden snares for my feet.}}}
\VersePair{18:23}{And thou, Lord, knowest all their deadly counsel against me: account not their iniquities guiltless, and blot not out their sins from before thee: let their weakness come before thee; deal with them in the time of thy wrath.}{18:23}{{\raggedright Yet, Yahweh, you know all their counsel against me to kill me. \poemline{2}{Don’t forgive their iniquity.}\poemline{2}{Don’t blot out their sin from your sight,}\poemline{1}{Let them be overthrown before you.}\poemline{2}{Deal with them in the time of your anger.}}}
\end{paracol}
\ChapterHeading{19}
\begin{paracol}{2}
\VersePair{19:1}{Then said the Lord to me, Go and get an earthen bottle, the work of the potter, and thou shalt bring some of the elders of the people, and of the priests;}{19:1}{Thus said Yahweh, “Go, and buy a potter’s earthen container, and take some of the elders of the people and of the elders of the priests;}
\VersePair{19:2}{and thou shalt go forth to the burial-place of the sons of their children, which is at the entrance of the gate of Charsith; and do thou read there all these words which I shall speak to thee:}{19:2}{and go out to the valley of the son of Hinnom, which is by the entry of the gate Harsith, and proclaim there the words that I will tell you.}
\VersePair{19:3}{and thou shalt say to them, Hear ye the word of the Lord, ye kings of Juda, and men of Juda, and the dwellers in Jerusalem, and they that enter in by these gates; thus saith the Lord God of Israel; Behold, I will bring evil upon this place, so that the ears of every one that hears it shall tingle.}{19:3}{Say, ‘Hear Yahweh’s word, kings of Judah and inhabitants of Jerusalem: Yahweh of Armies, the God of Israel says, “Behold, I will bring evil on this place, which whoever hears, his ears will tingle.}
\VersePair{19:4}{Because they forsook me, and profaned this place, and burnt incense in it to strange gods, which they and their fathers knew not; and the kings of Juda have filled this place with innocent blood,}{19:4}{Because they have forsaken me, and have defiled this place, and have burned incense in it to other gods that they didn’t know — they, their fathers, and the kings of Judah — and have filled this place with the blood of innocents,}
\VersePair{19:5}{and built high places for Baal, to burn their children in the fire, which things I commanded not, neither did I design them in my heart:}{19:5}{and have built the high places of Baal to burn their children in the fire for burnt offerings to Baal, which I didn’t command, nor speak, which didn’t even enter into my mind.}
\VersePair{19:6}{Therefore, behold, the days come, saith the Lord, when this place shall no more be called, The fall and burial-place of the son of Ennom, but, The burial-place of slaughter.}{19:6}{Therefore, behold, the days come,” says Yahweh, “that this place will no more be called ‘Topheth ’, nor ‘The Valley of the son of Hinnom ’, but ‘The valley of Slaughter ’.}
\VersePair{19:7}{And I will destroy the counsel of Juda and the counsel of Jerusalem in this place; and I will cast them down with the sword before their enemies, and by the hands of them that seek their lives: and I will give their dead bodies for food to the birds of the sky and to the wild beasts of the earth.}{19:7}{“‘“I will make the counsel of Judah and Jerusalem void in this place. I will cause them to fall by the sword before their enemies, and by the hand of those who seek their life. I will give their dead bodies to be food for the birds of the sky and for the animals of the earth.}
\VersePair{19:8}{And I will bring this city to desolation and make it a hissing; every one that passes by it shall scowl, and hiss because of all her plague.}{19:8}{I will make this city an astonishment and a hissing. Everyone who passes by it will be astonished and hiss because of all its plagues.}
\VersePair{19:9}{And they shall eat the flesh of their sons, and the flesh of their daughters; and they shall eat every one the flesh of his neighbour in the blockade, and in the siege wherewith their enemies shall besiege them.}{19:9}{I will cause them to eat the flesh of their sons and the flesh of their daughters. They will each eat the flesh of his friend in the siege and in the distress with which their enemies, and those who seek their life, will distress them.”’}
\VersePair{19:10}{And thou shalt break the bottle in the sight of the men that go forth with thee,}{19:10}{“Then you shall break the container in the sight of the men who go with you,}
\VersePair{19:11}{and thou shalt say, Thus saith the Lord, Thus will I break in pieces this people, and this city, even as an earthen vessel is broken in pieces which cannot be mended again.}{19:11}{and shall tell them, ‘Yahweh of Armies says: “Even so I will break this people and this city as one breaks a potter’s vessel, that can’t be made whole again. They will bury in Topheth until there is no place to bury.}
\VersePair{19:12}{Thus will I do, saith the Lord, to this place, and to the inhabitants of it, that this city may be given up, as one that is falling to ruin.}{19:12}{This is what I will do to this place,” says Yahweh, “and to its inhabitants, even making this city as Topheth.}
\VersePair{19:13}{And the houses of Jerusalem, and the houses of the kings of Juda shall be as a ruinous place, because of their uncleannesses in all the houses, wherein they burnt incense upon their roofs to all the host of heaven, and poured drink-offerings to strange gods.}{19:13}{The houses of Jerusalem and the houses of the kings of Judah, which are defiled, will be as the place of Topheth, even all the houses on whose roofs they have burned incense to all the army of the sky and have poured out drink offerings to other gods.”’”}
\VersePair{19:14}{And Jeremias came from the place of the Fall, whither the Lord had sent him to prophesy; and he stood in the court of the Lord's house; and said to all the people, Thus saith the Lord;}{19:14}{Then Jeremiah came from Topheth, where Yahweh had sent him to prophesy, and he stood in the court of Yahweh’s house, and said to all the people:}
\VersePair{19:15}{Behold, I bring upon this city, and upon all the cities belonging to it, and upon the villages of it, all the evils which I have spoken against it, because they have hardened their neck, that they might not hearken to my commands.}{19:15}{“Yahweh of Armies, the God of Israel says, ‘Behold, I will bring on this city and on all its towns all the evil that I have pronounced against it, because they have made their neck stiff, that they may not hear my words.’”}
\end{paracol}
\ChapterHeading{20}
\begin{paracol}{2}
\VersePair{20:1}{Now Paschor the son of Emmer, the priest, who also had been appointed chief of the house of the Lord, heard Jeremias prophesying these words.}{20:1}{Now Pashhur, the son of Immer the priest, who was chief officer in Yahweh’s house, heard Jeremiah prophesying these things.}
\VersePair{20:2}{And he smote him, and cast him into the dungeon which was by the gate of the upper house that was set apart, which was by the house of the Lord.}{20:2}{Then Pashhur struck Jeremiah the prophet and put him in the stocks that were in the upper gate of Benjamin, which was in Yahweh’s house.}
\VersePair{20:3}{And Paschor brought Jeremias out of the dungeon: and Jeremias said to him, The Lord has not called thy name Paschor, but Exile.}{20:3}{On the next day, Pashhur released Jeremiah out of the stocks. Then Jeremiah said to him, “Yahweh has not called your name Pashhur, but Magormissabib. \footnote{20:3: “Magormissabib” means “surrounded by terror”}}
\VersePair{20:4}{For thus saith the Lord, Behold, I will give thee up to captivity with all thy friends: and they shall fall by the sword of their enemies, and thine eyes shall see it: and I will give thee and all Juda into the hands of the king of Babylon, and they shall carry them captives, and cut them in pieces with swords.}{20:4}{For Yahweh says, ‘Behold, I will make you a terror to yourself and to all your friends. They will fall by the sword of their enemies, and your eyes will see it. I will give all Judah into the hand of the king of Babylon, and he will carry them captive to Babylon, and will kill them with the sword.}
\VersePair{20:5}{And I will give all the strength of this city, and all the labours of it, and all the treasures of the king of Juda, into the hands of his enemies, and they shall bring them to Babylon.}{20:5}{Moreover I will give all the riches of this city, and all its gains, and all its precious things, yes, I will give all the treasures of the kings of Judah into the hand of their enemies. They will make them captives, take them, and carry them to Babylon.}
\VersePair{20:6}{And thou and all the dwellers in thine house shall go into captivity: and thou shalt die in Babylon, and there thou and all thy friends shall be buried, to whom thou hast prophesied lies.}{20:6}{You, Pashhur, and all who dwell in your house will go into captivity. You will come to Babylon, and there you will die, and there you will be buried, you, and all your friends, to whom you have prophesied falsely.’”}
\VersePair{20:7}{Thou hast deceived me, O Lord, and I have been deceived: thou hast been strong, and hast prevailed: I am become a laughing-stock, I am continually mocked every day.}{20:7}{{\raggedright Yahweh, you have persuaded me, and I was persuaded. \poemline{2}{You are stronger than I, and have prevailed.}\poemline{1}{I have become a laughingstock all day.}\poemline{2}{Everyone mocks me.}}}
\VersePair{20:8}{For I will laugh with my bitter speech, I will call upon rebellion and misery: for the word of the Lord is become a reproach to me and a mockery all my days.}{20:8}{{\raggedright For as often as I speak, I cry out; \poemline{2}{I cry, “Violence and destruction!”}\poemline{1}{because Yahweh’s word has been made a reproach to me,}\poemline{2}{and a derision, all day.}}}
\VersePair{20:9}{Then I said, I will by no means name the name of the Lord, and I will no more at all speak in his name. But it was as burning fire flaming in my bones, and I am utterly weakened on all sides, and cannot bear up.}{20:9}{{\raggedright If I say that I will not make mention of him, \poemline{2}{or speak any more in his name,}\poemline{1}{then there is in my heart as it were a burning fire shut up in my bones.}\poemline{2}{I am weary with holding it in.}\poemline{2}{I can’t.}}}
\VersePair{20:10}{For I have heard the reproach of many gathering round, saying, Conspire ye, and let us conspire together against him, even all his friends: watch his intentions, if perhaps he shall be deceived, and we shall prevail against him, and we shall be avenged on him.}{20:10}{{\raggedright For I have heard the defaming of many: \poemline{2}{“Terror on every side!}\poemline{2}{Denounce, and we will denounce him!”}\poemline{1}{say all my familiar friends,}\poemline{2}{those who watch for my fall.}\poemline{1}{“Perhaps he will be persuaded,}\poemline{2}{and we will prevail against him,}\poemline{2}{and we will take our revenge on him.”}}}
\VersePair{20:11}{But the Lord was with me as a mighty man of war: therefore they persecuted me, but could not perceive anything against me; they were greatly confounded, for they perceived not their disgrace, which shall never be forgotten.}{20:11}{{\raggedright But Yahweh is with me as an awesome mighty one. \poemline{2}{Therefore my persecutors will stumble,}\poemline{2}{and they won’t prevail.}\poemline{1}{They will be utterly disappointed}\poemline{2}{because they have not dealt wisely,}\poemline{2}{even with an everlasting dishonor which will never be forgotten.}}}
\VersePair{20:12}{O Lord, that provest just deeds, understanding the reins and hearts, let me see thy vengeance upon them: for to thee I have revealed my cause.}{20:12}{{\raggedright But Yahweh of Armies, who tests the righteous, \poemline{2}{who sees the heart and the mind,}\poemline{1}{let me see your vengeance on them,}\poemline{2}{for I have revealed my cause to you.}}}
\VersePair{20:13}{Sing ye to the Lord, sing praise to him: for he has rescued the soul of the poor from the hand of evil-doers.}{20:13}{{\raggedright Sing to Yahweh! \poemline{2}{Praise Yahweh,}\poemline{2}{for he has delivered the soul of the needy from the hand of evildoers.}}}
\VersePair{20:14}{Cursed be the day wherein I was born: the day wherein my mother brought me forth, let it not be blessed.}{20:14}{{\raggedright Cursed is the day in which I was born. \poemline{2}{Don’t let the day in which my mother bore me be blessed.}}}
\VersePair{20:15}{Cursed be the man who brought the glad tidings to my father, saying, A male child is born to thee.}{20:15}{{\raggedright Cursed is the man who brought news to my father, saying, \poemline{2}{“A boy is born to you,” making him very glad.}}}
\VersePair{20:16}{Let that man rejoice as the cities which the Lord overthrew in wrath, and repented not: let him hear crying in the morning, and loud lamentation at noon;}{20:16}{{\raggedright Let that man be as the cities which Yahweh overthrew, \poemline{2}{and didn’t repent.}\poemline{1}{Let him hear a cry in the morning,}\poemline{2}{and shouting at noontime,}}}
\VersePair{20:17}{because he slew me not in the womb, and my mother became not my tomb, and her womb always great with me.}{20:17}{{\raggedright because he didn’t kill me from the womb. \poemline{2}{So my mother would have been my grave,}\poemline{2}{and her womb always great.}}}
\VersePair{20:18}{Why is it that I came forth of the womb to see troubles and distresses, and my days are spent in shame?}{20:18}{{\raggedright Why did I come out of the womb to see labor and sorrow, \poemline{2}{that my days should be consumed with shame?}}}
\end{paracol}
\ChapterHeading{21}
\begin{paracol}{2}
\VersePair{21:1}{The Word that came from the Lord to Jeremias, when King Sedekias sent to him Paschor the Son of Melchias, and Sophonias Son of Basæas, the Priest, saying,}{21:1}{The word which came to Jeremiah from Yahweh, when King Zedekiah sent to him Pashhur the son of Malchijah, and Zephaniah the son of Maaseiah, the priest, saying,}
\VersePair{21:2}{Enquire of the Lord for us; for the king of Babylon has risen up against us; if the Lord will do according to all his wonderful works, and the king shall depart from us.}{21:2}{“Please inquire of Yahweh for us; for Nebuchadnezzar king of Babylon makes war against us. Perhaps Yahweh will deal with us according to all his wondrous works, that he may withdraw from us.”}
\VersePair{21:3}{And Jeremias said to them, Thus shall ye say to Sedekias king of Juda,}{21:3}{Then Jeremiah said to them, “Tell Zedekiah:}
\VersePair{21:4}{Thus saith the Lord; Behold, I will turn back the weapons of war wherewith ye fight against the Chaldeans that have besieged you from outside the wall, and I will gather them into the midst of this city.}{21:4}{‘Yahweh, the God of Israel says, “Behold, I will turn back the weapons of war that are in your hands, with which you fight against the king of Babylon, and against the Chaldeans who besiege you outside the walls; and I will gather them into the middle of this city.}
\VersePair{21:5}{And I will fight against you with an outstretched hand and with a strong arm, with wrath and great anger.}{21:5}{I myself will fight against you with an outstretched hand and with a strong arm, even in anger, in wrath, and in great indignation.}
\VersePair{21:6}{And I will smite all the dwellers in this city, both men and cattle, with grievous pestilence: and they shall die.}{21:6}{I will strike the inhabitants of this city, both man and animal. They will die of a great pestilence.}
\VersePair{21:7}{And after this, thus saith the Lord; I will give Sedekias king of Juda, and his servants, and the people that is left in this city from the pestilence, and from the famine, and from the sword, into the hands of their enemies, that seek their lives: and they shall cut them in pieces with the edge of the sword: I will not spare them, and I will not have compassion upon them.}{21:7}{Afterward,” says Yahweh, “I will deliver Zedekiah king of Judah, his servants, and the people, even those who are left in this city from the pestilence, from the sword, and from the famine, into the hand of Nebuchadnezzar king of Babylon, and into the hand of their enemies, and into the hand of those who seek their life. He will strike them with the edge of the sword. He will not spare them, have pity, or have mercy.”’}
\VersePair{21:8}{And thou shalt say to this people, Thus saith the Lord; Behold, I have set before you the way of life, and the way of death.}{21:8}{“You shall say to this people, ‘Yahweh says: “Behold, I set before you the way of life and the way of death.}
\VersePair{21:9}{He that remains in this city shall die by the sword, and by famine: but he that goes forth to advance to the Chaldeans that have besieged you, shall live, and his life shall be to him for a spoil, and he shall live.}{21:9}{He who remains in this city will die by the sword, by the famine, and by the pestilence, but he who goes out and passes over to the Chaldeans who besiege you, he will live, and he will escape with his life.}
\VersePair{21:10}{For I have set my face against this city for evil, and not for good: it shall be delivered into the hands of the king of Babylon, and he shall consume it with fire.}{21:10}{For I have set my face on this city for evil, and not for good,” says Yahweh. “It will be given into the hand of the king of Babylon, and he will burn it with fire.”’}
\VersePair{21:11}{O house of the king of Juda, hear ye the word of the Lord.}{21:11}{“Concerning the house of the king of Judah, hear Yahweh’s word:}
\VersePair{21:12}{O house of David, thus saith the Lord; Judge judgment in the morning, and act rightly, and rescue the spoiled one from the hand of him that wrongs him, lest mine anger be kindled like fire, and it burn, and there be none to quench it.}{21:12}{{\raggedright House of David, Yahweh says, \poemline{1}{‘Execute justice in the morning,}\poemline{2}{and deliver him who is robbed out of the hand of the oppressor,}\poemline{1}{lest my wrath go out like fire,}\poemline{2}{and burn so that no one can quench it,}\poemline{2}{because of the evil of your doings.}}}
\VersePair{21:13}{Behold, I am against thee that dwellest in the valley of Sor; in the plain country, even against them that say, Who shall alarm us? or who shall enter into our habitation?}{21:13}{{\raggedright Behold, I am against you, O inhabitant of the valley, \poemline{2}{and of the rock of the plain,’ says Yahweh.}\poemline{1}{‘You that say, “Who would come down against us?”}\poemline{2}{or, “Who would enter into our homes?”}}}
\VersePair{21:14}{And I will kindle a fire in the forest thereof, and it shall devour all things round about it.}{21:14}{{\raggedright I will punish you according to the fruit of your doings,’ says Yahweh; \poemline{2}{‘and I will kindle a fire in her forest,}\poemline{2}{and it will devour all that is around her.’”}}}
\end{paracol}
\ChapterHeading{22}
\begin{paracol}{2}
\VersePair{22:1}{Thus saith the Lord; Go thou, and go down to the house of the king of Juda, and thou shalt speak there this word,}{22:1}{Yahweh said, “Go down to the house of the king of Judah, and speak this word there:}
\VersePair{22:2}{and thou shalt say, Hear the word of the Lord, O king of Juda, that sittest on the throne of David, thou, and thy house, and thy people, and they that go in at these gates:}{22:2}{‘Hear Yahweh’s word, king of Judah, who sits on David’s throne — you, your servants, and your people who enter in by these gates.}
\VersePair{22:3}{thus saith the Lord; Execute ye judgment and justice, and rescue the spoiled out of the hand of him that wrongs him: and oppress not the stranger, and orphan, and widow, and sin not, and shed no innocent blood in this place.}{22:3}{Yahweh says: “Execute justice and righteousness, and deliver him who is robbed out of the hand of the oppressor. Do no wrong. Do no violence to the foreigner, the fatherless, or the widow. Don’t shed innocent blood in this place.}
\VersePair{22:4}{For if ye will indeed perform this word, then shall there enter in by the gates of this house kings sitting upon the throne of David, and riding on chariots and horses, they, and their servants, and their people.}{22:4}{For if you do this thing indeed, then kings sitting on David’s throne will enter in by the gates of this house, riding in chariots and on horses — they, their servants, and their people.}
\VersePair{22:5}{But if ye will not perform these words, by myself have I sworn, saith the Lord, that this house shall be brought to desolation.}{22:5}{But if you will not hear these words, I swear by myself,” says Yahweh, “that this house will become a desolation.”’”}
\VersePair{22:6}{For thus saith the Lord concerning the house of the king of Juda; Thou art Galaad to me, and the head of Libanus: yet surely I will make thee a desert, even cities that shall not be inhabited:}{22:6}{{\raggedright For Yahweh says concerning the house of the king of Judah: \poemline{1}{“You are Gilead to me,}\poemline{2}{the head of Lebanon.}\poemline{1}{Yet surely I will make you a wilderness,}\poemline{2}{cities which are not inhabited.}}}
\VersePair{22:7}{and I will bring upon thee a destroying man, and his axe: and they shall cut down thy choice cedars, and cast them into the fire.}{22:7}{{\raggedright I will prepare destroyers against you, \poemline{2}{everyone with his weapons,}\poemline{1}{and they will cut down your choice cedars,}\poemline{2}{and cast them into the fire.}}}
\VersePair{22:8}{And nations shall pass through this city, and each shall say to his neighbour, Why has the Lord done thus to this great city?}{22:8}{“Many nations will pass by this city, and they will each ask his neighbor, ‘Why has Yahweh done this to this great city?’}
\VersePair{22:9}{And they shall say, Because they forsook the covenant of the Lord their God, and worshipped strange gods, and served them.}{22:9}{Then they will answer, ‘Because they abandoned the covenant of Yahweh their God, worshiped other gods, and served them.’”}
\VersePair{22:10}{Weep not for the dead, nor lament for him: weep bitterly for him that goes away: for he shall return no more, nor see his native land.}{22:10}{{\raggedright Don’t weep for the dead. \poemline{2}{Don’t bemoan him;}\poemline{1}{but weep bitterly for him who goes away,}\poemline{2}{for he will return no more,}\poemline{2}{and not see his native country.}}}
\VersePair{22:11}{For thus saith the Lord concerning Sellem the son of Josias, who reigns in the place of Josias his father, who has gone forth out of this place; He shall not return thither any more:}{22:11}{For Yahweh says touching Shallum the son of Josiah, king of Judah, who reigned instead of Josiah his father, and who went out of this place: “He won’t return there any more.}
\VersePair{22:12}{but in that place whither I have carried him captive, there shall he die, and shall see this land no more.}{22:12}{But he will die in the place where they have led him captive. He will see this land no more.”}
\VersePair{22:13}{He that builds his house not with justice, and his upper chambers not with judgment, who works by means of his neighbour for nothing, and will by no means give him his reward.}{22:13}{{\raggedright “Woe to him who builds his house by unrighteousness, \poemline{2}{and his rooms by injustice;}\poemline{1}{who uses his neighbor’s service without wages,}\poemline{2}{and doesn’t give him his hire;}}}
\VersePair{22:14}{Thou hast built for thyself a well-proportioned house, airy chambers, fitted with windows, and wainscoted with cedar, and painted with vermilion.}{22:14}{{\raggedright who says, ‘I will build myself a wide house and spacious rooms,’ \poemline{2}{and cuts out windows for himself,}\poemline{1}{with a cedar ceiling,}\poemline{2}{and painted with red.}}}
\VersePair{22:15}{Shalt thou reign, because thou art provoked with thy father Achaz? they shall not eat, and they shall not drink: it is better for thee to execute judgment and justice.}{22:15}{{\raggedright “Should you reign because you strive to excel in cedar? \poemline{2}{Didn’t your father eat and drink,}\poemline{2}{and do justice and righteousness?}\poemline{2}{Then it was well with him.}}}
\VersePair{22:16}{They understood not, they judged not the cause of the afflicted, nor the cause of the poor: is not this thy not knowing me? saith the Lord.}{22:16}{{\raggedright He judged the cause of the poor and needy; \poemline{2}{so then it was well.}\poemline{1}{Wasn’t this to know me?”}\poemline{2}{says Yahweh.}}}
\VersePair{22:17}{Behold, thine eyes are not good, nor thine heart, but they go after thy covetousness, and after the innocent blood to shed it, and after acts of injustice and slaughter, to commit them.}{22:17}{{\raggedright But your eyes and your heart are only for your covetousness, \poemline{2}{for shedding innocent blood,}\poemline{2}{for oppression, and for doing violence.”}}}
\VersePair{22:18}{Therefore thus saith the Lord concerning Joakim son of Josias, king of Juda, even concerning this man; they shall not bewail him, saying, Ah brother! neither shall they at all weep for him, saying, Alas Lord.}{22:18}{{\raggedright Therefore Yahweh says concerning Jehoiakim the son of Josiah, king of Judah: \poemline{1}{“They won’t lament for him,}\poemline{2}{saying, ‘Ah my brother!’ or, ‘Ah sister!’}\poemline{1}{They won’t lament for him,}\poemline{2}{saying ‘Ah lord!’ or, ‘Ah his glory!’}}}
\VersePair{22:19}{He shall be buried with the burial of an ass; he shall be dragged roughly along and cast outside the gate of Jerusalem.}{22:19}{{\raggedright He will be buried with the burial of a donkey, \poemline{2}{drawn and cast out beyond the gates of Jerusalem.”}}}
\VersePair{22:20}{Go up to Libanus, and cry; and utter thy voice to Basan, and cry aloud to the extremity of the sea: for all thy lovers are destroyed.}{22:20}{{\raggedright “Go up to Lebanon, and cry out. \poemline{2}{Lift up your voice in Bashan,}\poemline{1}{and cry from Abarim;}\poemline{2}{for all your lovers have been destroyed.}}}
\VersePair{22:21}{I spoke to thee on occasion of thy trespass, but thou saidst, I will not hearken. This has been thy way from thy youth, thou hast not hearkened to my voice.}{22:21}{{\raggedright I spoke to you in your prosperity, \poemline{2}{but you said, ‘I will not listen.’}\poemline{1}{This has been your way from your youth,}\poemline{2}{that you didn’t obey my voice.}}}
\VersePair{22:22}{The wind shall tend all thy shepherds, and thy lovers shall go into captivity; for then shalt thou be ashamed and disgraced because of all thy lovers.}{22:22}{{\raggedright The wind will feed all your shepherds, \poemline{2}{and your lovers will go into captivity.}\poemline{1}{Surely then you will be ashamed}\poemline{2}{and confounded for all your wickedness.}}}
\VersePair{22:23}{O thou that dwellest in Libanus, making thy nest in the cedars, thou shalt groan heavily, when pangs as of a travailing woman are come upon thee.}{22:23}{{\raggedright Inhabitant of Lebanon, \poemline{2}{who makes your nest in the cedars,}\poemline{1}{how greatly to be pitied you will be when pangs come on you,}\poemline{2}{the pain as of a woman in travail!}}}
\VersePair{22:24}{As I live, saith the Lord, though Jechonias son of Joakim king of Juda were indeed the seal upon my right hand, thence would I pluck thee;}{22:24}{“As I live,” says Yahweh, “though Coniah the son of Jehoiakim king of Judah were the signet on my right hand, I would still pluck you from there.}
\VersePair{22:25}{and I will deliver thee into the hands of them that seek thy life, before whom thou art afraid, into the hands of the Chaldeans.}{22:25}{I would give you into the hand of those who seek your life, and into the hand of them of whom you are afraid, even into the hand of Nebuchadnezzar king of Babylon, and into the hand of the Chaldeans.}
\VersePair{22:26}{And I will cast forth thee, and thy mother that bore thee, into a land where thou wast not born; and there ye shall die.}{22:26}{I will cast you out with your mother who bore you into another country, where you were not born; and there you will die.}
\VersePair{22:27}{But they shall by no means return to the land which they long for in their souls.}{22:27}{But to the land to which their soul longs to return, there they will not return.”}
\VersePair{22:28}{Jechonias is dishonoured as a good-for-nothing vessel; for he is thrown out and cast forth into a land which he knew not.}{22:28}{{\raggedright Is this man Coniah a despised broken vessel? \poemline{2}{Is he a vessel in which no one delights?}\poemline{1}{Why are they cast out, he and his offspring,}\poemline{2}{and cast into a land which they don’t know?}}}
\VersePair{22:29}{Land, land, hear the word of the Lord.}{22:29}{{\raggedright O earth, earth, earth, \poemline{2}{hear Yahweh’s word!}}}
\VersePair{22:30}{Write ye this man an outcast: for there shall none of his seed at all grow up to sit on the throne of David, or as a prince yet in Juda.}{22:30}{{\raggedright Yahweh says, \poemline{2}{“Record this man as childless,}\poemline{2}{a man who will not prosper in his days;}\poemline{1}{for no more will a man of his offspring prosper,}\poemline{2}{sitting on David’s throne}\poemline{2}{and ruling in Judah.”}}}
\end{paracol}
\ChapterHeading{23}
\begin{paracol}{2}
\VersePair{23:1}{Woe to the shepherds that destroy and scatter the sheep of their pasture!}{23:1}{“Woe to the shepherds who destroy and scatter the sheep of my pasture!” says Yahweh.}
\VersePair{23:2}{Therefore thus saith the Lord against them that tend my people; Ye have scattered my sheep, and driven them out, and ye have not visited them: behold, I will take vengeance upon you according to your evil practices.}{23:2}{Therefore Yahweh, the God of Israel, says against the shepherds who feed my people: “You have scattered my flock, driven them away, and have not visited them. Behold, I will visit on you the evil of your doings,” says Yahweh.}
\VersePair{23:3}{And I will gather in the remnant of my people in every land, whither I have driven them out, and will set them in their pasture; and they shall increase and be multiplied.}{23:3}{“I will gather the remnant of my flock out of all the countries where I have driven them, and will bring them again to their folds; and they will be fruitful and multiply.}
\VersePair{23:4}{And I will raise up shepherds to them, who shall feed them: and they shall fear no more, nor be alarmed, saith the Lord.}{23:4}{I will set up shepherds over them who will feed them. They will no longer be afraid or dismayed, neither will any be lacking,” says Yahweh.}
\VersePair{23:5}{Behold, the days come, saith the Lord, when I will raise up to David a righteous branch, and a king shall reign and understand, and shall execute judgment and righteousness on the earth.}{23:5}{{\raggedright “Behold, the days come,” says Yahweh, \poemline{2}{“that I will raise to David a righteous Branch;}\poemline{1}{and he will reign as king and deal wisely,}\poemline{2}{and will execute justice and righteousness in the land.}}}
\VersePair{23:6}{In his days both Juda shall be saved, and Israel shall dwell securely: and this is his name, which the Lord shall call him, Josedec among the prophets.}{23:6}{{\raggedright In his days Judah will be saved, \poemline{2}{and Israel will dwell safely.}\poemline{1}{This is his name by which he will be called:}\poemline{2}{Yahweh our righteousness.}}}
\VersePair{23:9}{My heart is broken within me; all my bones are shaken: I am become as a broken-down man, and as a man overcome with wine, because of the Lord, and because of the excellence of his glory.}{23:9}{{\raggedright Concerning the prophets: \poemline{1}{My heart within me is broken.}\poemline{2}{All my bones shake.}\poemline{1}{I am like a drunken man,}\poemline{2}{and like a man whom wine has overcome,}\poemline{1}{because of Yahweh,}\poemline{2}{and because of his holy words.}}}
\VersePair{23:10}{For because of these things the land mourns; the pastures of the wilderness are dried up; and their course is become evil, and so also their strength.}{23:10}{{\raggedright “For the land is full of adulterers; \poemline{2}{for because of the curse the land mourns.}\poemline{1}{The pastures of the wilderness have dried up.}\poemline{2}{Their course is evil,}\poemline{2}{and their might is not right;}}}
\VersePair{23:11}{For priest and prophet are defiled; and I have seen their iniquities in my house.}{23:11}{{\raggedright for both prophet and priest are profane. \poemline{2}{Yes, in my house I have found their wickedness,” says Yahweh.}}}
\VersePair{23:12}{Therefore let their way be to them slippery and dark: and they shall be tripped up and fall in it: for I will bring evils upon them, in the year of their visitation.}{23:12}{{\raggedright Therefore their way will be to them as slippery places in the darkness. \poemline{2}{They will be driven on,}\poemline{2}{and fall therein;}\poemline{1}{for I will bring evil on them,}\poemline{2}{even the year of their visitation,” says Yahweh.}}}
\VersePair{23:13}{And in the prophets of Samaria I have seen lawless deeds; they prophesied by Baal, and led my people Israel astray.}{23:13}{{\raggedright “I have seen folly in the prophets of Samaria. \poemline{2}{They prophesied by Baal,}\poemline{2}{and caused my people Israel to err.}}}
\VersePair{23:14}{Also in the prophets of Jerusalem I have seen horrible things: as they committed adultery, and walked in lies, and strengthened the hands of many, that they should not return each from his evil way: they are all become to me as Sodoma, and the inhabitants thereof as Gomorrha.}{23:14}{{\raggedright In the prophets of Jerusalem I have also seen a horrible thing: \poemline{2}{they commit adultery and walk in lies.}\poemline{1}{They strengthen the hands of evildoers,}\poemline{2}{so that no one returns from his wickedness.}\poemline{1}{They have all become to me as Sodom,}\poemline{2}{and its inhabitants as Gomorrah.”}}}
\VersePair{23:15}{Therefore thus saith the Lord; Behold, I will feed them with pain, and give them bitter water to drink: for from the prophets of Jerusalem has defilement gone forth into all the land.}{23:15}{{\raggedright Therefore Yahweh of Armies says concerning the prophets: \poemline{1}{“Behold, I will feed them with wormwood,}\poemline{2}{and make them drink poisoned water;}\poemline{2}{for from the prophets of Jerusalem ungodliness has gone out into all the land.”}}}
\VersePair{23:16}{Thus saith the Lord Almighty, Hearken not to the words of the prophets: for they frame a vain vision for themselves; they speak from their own heart, and not from the mouth of the Lord.}{23:16}{{\raggedright Yahweh of Armies says, \poemline{1}{“Don’t listen to the words of the prophets who prophesy to you.}\poemline{2}{They teach you vanity.}\poemline{2}{They speak a vision of their own heart,}\poemline{2}{and not out of the mouth of Yahweh.}}}
\VersePair{23:17}{They say to them that reject the word of the Lord, There shall be peace to you; and to all that walk after their own lusts, and to every one that walks in the error of his heart, they have said, No evil shall come upon thee.}{23:17}{{\raggedright They say continually to those who despise me, \poemline{2}{‘Yahweh has said, “You will have peace;”’}\poemline{1}{and to everyone who walks in the stubbornness of his own heart they say,}\poemline{2}{‘No evil will come on you.’}}}
\VersePair{23:18}{For who has stood in the counsel of the Lord, and seen his word? who has hearkened, and heard?}{23:18}{{\raggedright For who has stood in the council of Yahweh, \poemline{2}{that he should perceive and hear his word?}\poemline{2}{Who has listened to my word, and heard it?}}}
\VersePair{23:19}{Behold, there is an earthquake from the Lord, and anger proceeds to a convulsion, it shall come violently upon the ungodly.}{23:19}{{\raggedright Behold, Yahweh’s storm, his wrath, has gone out. \poemline{2}{Yes, a whirling storm!}\poemline{2}{It will burst on the head of the wicked.}}}
\VersePair{23:20}{And the Lord's wrath shall return no more, until he have accomplished it, and until he have established it, according to the purpose of his heart: at the end of the days they shall understand it.}{23:20}{{\raggedright Yahweh’s anger will not return until he has executed \poemline{2}{and performed the intents of his heart.}\poemline{2}{In the latter days, you will understand it perfectly.}}}
\VersePair{23:21}{I sent not the prophets, yet they ran: neither spoke I to them, yet they prophesied.}{23:21}{{\raggedright I didn’t send these prophets, yet they ran. \poemline{2}{I didn’t speak to them, yet they prophesied.}}}
\VersePair{23:22}{But if they had stood in my counsel, and if they had hearkened to my words, then would they have turned my people from their evil practices.}{23:22}{{\raggedright But if they had stood in my council, \poemline{2}{then they would have caused my people to hear my words,}\poemline{1}{and would have turned them from their evil way,}\poemline{2}{and from the evil of their doings.}}}
\VersePair{23:23}{I am a God nigh at hand, saith the Lord, and not a God afar off.}{23:23}{{\raggedright “Am I a God at hand,” says Yahweh, \poemline{2}{“and not a God afar off?}}}
\VersePair{23:24}{Shall any one hide himself in secret places, and I not see him? Do I not fill heaven and earth? saith the Lord.}{23:24}{{\raggedright Can anyone hide himself in secret places \poemline{2}{so that I can’t see him?” says Yahweh.}\poemline{2}{“Don’t I fill heaven and earth?” says Yahweh.}}}
\VersePair{23:25}{I have heard what the prophets say, what they prophesy in my name, saying falsely, I have seen a night vision.}{23:25}{“I have heard what the prophets have said, who prophesy lies in my name, saying, ‘I had a dream! I had a dream!’}
\VersePair{23:26}{How long shall these things be in the heart of the prophets that prophesy lies, when they prophesy the purposes of their own heart?}{23:26}{How long will this be in the heart of the prophets who prophesy lies, even the prophets of the deceit of their own heart?}
\VersePair{23:27}{who devise that men may forget my law by their dreams, which they have told every one to his neighbour, as their fathers forgot my name in the worship of Baal.}{23:27}{They intend to cause my people to forget my name by their dreams which they each tell his neighbor, as their fathers forgot my name because of Baal.}
\VersePair{23:28}{The prophet who has a dream, let him tell his dream; and he in whom is my word spoken to him, let him tell my word truly: what is the chaff to the corn? so are my words, saith the Lord.}{23:28}{The prophet who has a dream, let him tell a dream; and he who has my word, let him speak my word faithfully. What is the straw to the wheat?” says Yahweh.}
\VersePair{23:29}{Behold, are not my words as fire? saith the Lord; and as an axe cutting the rock?}{23:29}{“Isn’t my word like fire?” says Yahweh; “and like a hammer that breaks the rock in pieces?}
\VersePair{23:30}{Behold, I am therefore against the prophets, saith the Lord God, that steal my words every one from his neighbour.}{23:30}{“Therefore behold, I am against the prophets,” says Yahweh, “who each steal my words from his neighbor.}
\VersePair{23:31}{Behold, I am against the prophets that put forth prophecies of mere words, and slumber their sleep.}{23:31}{Behold, I am against the prophets,” says Yahweh, “who use their tongues, and say, ‘He says.’}
\VersePair{23:32}{Therefore, behold, I am against the prophets that prophesy false dreams, and have not told them truly, and have caused my people to err by their lies, and by their errors; yet I sent them not, and commanded them not; therefore, they shall not profit this people at all.}{23:32}{Behold, I am against those who prophesy lying dreams,” says Yahweh, “who tell them, and cause my people to err by their lies, and by their vain boasting; yet I didn’t send them or command them. They don’t profit this people at all,” says Yahweh.}
\VersePair{23:33}{And if this people, or the priest, or the prophet, should ask, What is the burden of the Lord? then thou shalt say to them, Ye are the burden, and I will dash you down, saith the Lord.}{23:33}{“When this people, or the prophet, or a priest, asks you, saying, ‘What is the message from Yahweh?’ Then you shall tell them, ‘“What message? I will cast you off,” says Yahweh.’}
\VersePair{23:34}{As for the prophet, and the priests, and the people, who shall say, The burden of the Lord, I will even take vengeance on that man, and on his house.}{23:34}{As for the prophet, the priest, and the people, who say, ‘The message from Yahweh,’ I will even punish that man and his household.}
\VersePair{23:35}{Thus shall ye say every one to his neighbour, and every one to his brother, What has the Lord answered? and, what has the Lord said?}{23:35}{You will say everyone to his neighbor, and everyone to his brother, ‘What has Yahweh answered?’ and, ‘What has Yahweh said?’}
\VersePair{23:36}{And do ye name no more the burden of the Lord; for his own word shall be a man's burden.}{23:36}{You will mention the message from Yahweh no more, for every man’s own word has become his message; for you have perverted the words of the living God, of Yahweh of Armies, our God.}
\VersePair{23:37}{But wherefore, say ye, has the Lord our God spoken?}{23:37}{You will say to the prophet, ‘What has Yahweh answered you?’ and, ‘What has Yahweh spoken?’}
\VersePair{23:38}{Therefore thus saith the Lord our God; Because ye have spoken this word, The burden of the Lord, and I sent to you, saying, Ye shall not say, The burden of the Lord;}{23:38}{Although you say, ‘The message from Yahweh,’ therefore Yahweh says: ‘Because you say this word, “The message from Yahweh,” and I have sent to you, telling you not to say, “The message from Yahweh,”}
\VersePair{23:39}{therefore, behold, I will seize, and dash down you and the city which I gave to you and your fathers.}{23:39}{therefore behold, I will utterly forget you, and I will cast you off with the city that I gave to you and to your fathers, away from my presence.}
\VersePair{23:40}{And I will bring upon you an everlasting reproach, and everlasting disgrace, which shall not be forgotten.}{23:40}{I will bring an everlasting reproach on you, and a perpetual shame, which will not be forgotten.’”}
\VersePair{23:40a}{Therefore, behold, the days come, saith the Lord, when they shall no more say, The Lord lives, who brought up the house of Israel out of the land of Egypt;}{23:40a}{}
\VersePair{23:40b}{but The Lord lives, who has gathered the whole seed of Israel from the north land, and from all the countries whither he had driven them out, and has restored them into their own land.}{23:40b}{}
\end{paracol}
\ChapterHeading{24}
\begin{paracol}{2}
\VersePair{24:1}{The Lord shewed me two baskets of figs, lying in front of the temple of the Lord, after Nabuchodonosor king of Babylon had carried captive Jechonias son of Joakim king of Juda, and the princes, and the artificers, and the prisoners, and the rich men out of Jerusalem, and had brought them to Babylon.}{24:1}{Yahweh showed me, and behold, two baskets of figs were set before Yahweh’s temple, after Nebuchadnezzar king of Babylon had carried away captive Jeconiah the son of Jehoiakim, king of Judah, and the princes of Judah, with the craftsmen and smiths, from Jerusalem, and had brought them to Babylon.}
\VersePair{24:2}{The one basket was full of very good figs, as the early figs; and the other basket was full of very bad figs, which could not be eaten, for their badness.}{24:2}{One basket had very good figs, like the figs that are first-ripe; and the other basket had very bad figs, which could not be eaten, they were so bad.}
\VersePair{24:3}{And the Lord said to me, What seest thou, Jeremias? and I said, Figs; the good figs, very good; and the bad, very bad, which cannot be eaten, for their badness.}{24:3}{Then Yahweh asked me, “What do you see, Jeremiah?” I said, “Figs. The good figs are very good, and the bad are very bad, so bad that they can’t be eaten.”}
\VersePair{24:4}{And the word of the Lord came to me, saying,}{24:4}{Yahweh’s word came to me, saying,}
\VersePair{24:5}{Thus saith the Lord, the God of Israel; As these good figs, so will I acknowledge the Jews that have been carried away captive, whom I have sent forth out of this place into the land of the Chaldeans for good.}{24:5}{“Yahweh, the God of Israel says: ‘Like these good figs, so I will regard the captives of Judah, whom I have sent out of this place into the land of the Chaldeans, as good.}
\VersePair{24:6}{And I will fix mine eyes upon them for good, and I will restore them into this land for good: and I will build them up, and not pull them down; and I will plant them, and not pluck them up.}{24:6}{For I will set my eyes on them for good, and I will bring them again to this land. I will build them, and not pull them down. I will plant them, and not pluck them up.}
\VersePair{24:7}{And I will give them a heart to know me, that I am the Lord: and they shall be to me a people, and I will be to them a God: for they shall turn to me with all their heart.}{24:7}{I will give them a heart to know me, that I am Yahweh. They will be my people, and I will be their God; for they will return to me with their whole heart.}
\VersePair{24:8}{And as the bad figs, which cannot be eaten, for their badness; thus saith the Lord, So will I deliver Sedekias king of Juda, and his nobles, and the remnant of Jerusalem, them that are left in this land, and the dwellers in Egypt.}{24:8}{“‘As the bad figs, which can’t be eaten, they are so bad,’ surely Yahweh says, ‘So I will give up Zedekiah the king of Judah, and his princes, and the remnant of Jerusalem who remain in this land, and those who dwell in the land of Egypt.}
\VersePair{24:9}{And I will cause them to be dispersed into all the kingdoms of the earth, and they shall be for a reproach, and a proverb, and an object of hatred, and a curse, in every place whither I have driven them out.}{24:9}{I will even give them up to be tossed back and forth among all the kingdoms of the earth for evil, to be a reproach and a proverb, a taunt and a curse, in all places where I will drive them.}
\VersePair{24:10}{And I will send against them famine, and pestilence, and the sword, until they are consumed from off the land which I gave them.}{24:10}{I will send the sword, the famine, and the pestilence among them, until they are consumed from off the land that I gave to them and to their fathers.’”}
\end{paracol}
\ChapterHeading{25}
\begin{paracol}{2}
\VersePair{25:1}{The Word that came to Jeremias concerning all the people of Juda in the fourth year of Joakim, son of Josias, king of Juda;}{25:1}{The word that came to Jeremiah concerning all the people of Judah, in the fourth year of Jehoiakim the son of Josiah, king of Judah ( this was the first year of Nebuchadnezzar king of Babylon),}
\VersePair{25:2}{which he spoke to all the people of Juda, and to the inhabitants of Jerusalem, saying,}{25:2}{which Jeremiah the prophet spoke to all the people of Judah, and to all the inhabitants of Jerusalem:}
\VersePair{25:3}{In the thirteenth year of Josias, son of Amos, king of Juda, even until this day for three and twenty years, I have both spoken to you, rising early and speaking,}{25:3}{From the thirteenth year of Josiah the son of Amon, king of Judah, even to this day, these twenty-three years, Yahweh’s word has come to me, and I have spoken to you, rising up early and speaking; but you have not listened.}
\VersePair{25:4}{and I sent to you my servants the prophets, sending them early; (but ye hearkened not, and listened not with your ears;) saying,}{25:4}{Yahweh has sent to you all his servants the prophets, rising up early and sending them ( but you have not listened or inclined your ear to hear ),}
\VersePair{25:5}{Turn ye every one from his evil way, and from your evil practices, and ye shall dwell in the land which I gave to you and your fathers, of old and for ever.}{25:5}{saying, “Return now everyone from his evil way, and from the evil of your doings, and dwell in the land that Yahweh has given to you and to your fathers, from of old and even forever more.}
\VersePair{25:6}{Go ye not after strange gods, to serve them, and to worship them, that ye provoke me not by the works of your hands, to do you hurt.}{25:6}{Don’t go after other gods to serve them or worship them, and don’t provoke me to anger with the work of your hands; then I will do you no harm.”}
\VersePair{25:7}{But ye hearkened not to me.}{25:7}{“Yet you have not listened to me,” says Yahweh, “that you may provoke me to anger with the work of your hands to your own hurt.”}
\VersePair{25:8}{Therefore thus saith the Lord; Since ye believed not my words,}{25:8}{Therefore Yahweh of Armies says: “Because you have not heard my words,}
\VersePair{25:9}{behold, I will send and take a family from the north, and will bring them against this land, and against the inhabitants of it, and against all the nations round about it, and I will make them utterly waste, and make them a desolation, and a hissing, and an everlasting reproach.}{25:9}{behold, I will send and take all the families of the north,” says Yahweh, “and I will send to Nebuchadnezzar the king of Babylon, my servant, and will bring them against this land, and against its inhabitants, and against all these nations around. I will utterly destroy them, and make them an astonishment, and a hissing, and perpetual desolations.}
\VersePair{25:10}{And I will destroy from among them the voice of joy, and the voice of gladness, the voice of the bridegroom, and the voice of the bride, the scent of ointment, and the light of a candle.}{25:10}{Moreover I will take from them the voice of mirth and the voice of gladness, the voice of the bridegroom and the voice of the bride, the sound of the millstones, and the light of the lamp.}
\VersePair{25:11}{And all the land shall be a desolation; and they shall serve among the Gentiles seventy years.}{25:11}{This whole land will be a desolation, and an astonishment; and these nations will serve the king of Babylon seventy years.}
\VersePair{25:12}{And when the seventy years are fulfilled, I will take vengeance on that nation, and will make them a perpetual desolation.}{25:12}{“It will happen, when seventy years are accomplished, that I will punish the king of Babylon and that nation,” says Yahweh, “for their iniquity. I will make the land of the Chaldeans desolate forever.}
\VersePair{25:13}{And I will bring upon that land all my words which I have spoken against it, even all things that are written in this book.}{25:13}{I will bring on that land all my words which I have pronounced against it, even all that is written in this book, which Jeremiah has prophesied against all the nations.}
\VersePair{25:14}{The Prophecies of Jeremias against the Nations of Ælam.}{49:34}{Yahweh’s word that came to Jeremiah the prophet concerning Elam, in the beginning of the reign of Zedekiah king of Judah, saying,}
\VersePair{25:15}{Thus saith the Lord, The bow of Ælam is broken, even the chief of their power.}{49:35}{{\raggedright “Yahweh of Armies says: \poemline{1}{‘Behold, I will break the bow of Elam,}\poemline{2}{the chief of their might.}}}
\VersePair{25:16}{And I will bring upon Ælam the four winds from the four corners of heaven, and I will disperse them toward all these winds; and there shall be no nation to which they shall not come— even the outcasts of Ælam.}{49:36}{{\raggedright I will bring on Elam the four winds from the four quarters of the sky, \poemline{2}{and will scatter them toward all those winds.}\poemline{2}{There will be no nation where the outcasts of Elam will not come.}}}
\VersePair{25:17}{And I will put them in fear before their enemies that seek their life; and I will bring evils upon them according to my great anger; and I will send forth my sword after them, until I have utterly destroyed them.}{49:37}{{\raggedright I will cause Elam to be dismayed before their enemies, \poemline{2}{and before those who seek their life.}\poemline{1}{I will bring evil on them, even my fierce anger,’ says Yahweh;}\poemline{2}{‘and I will send the sword after them,}\poemline{2}{until I have consumed them.}}}
\VersePair{25:18}{And I will set my throne in Ælam, and will send forth thence king and rulers.}{49:38}{{\raggedright I will set my throne in Elam, \poemline{2}{and will destroy from there king and princes,’ says Yahweh.}}}
\VersePair{25:19}{But it shall come to pass at the end of days, that I will turn the captivity of Ælam, saith the Lord.}{49:39}{{\raggedright ‘But it will happen in the latter days \poemline{2}{that I will reverse the captivity of Elam,’ says Yahweh.”}}}
\VersePair{25:20}{In the beginning of the reign of king Sedekias, there came this word concerning Ælam.}{49:34}{Yahweh’s word that came to Jeremiah the prophet concerning Elam, in the beginning of the reign of Zedekiah king of Judah, saying,}
\end{paracol}
\ChapterHeading{26}
\begin{paracol}{2}
\VersePair{26:2}{For Egypt, against the Power of Pharao Nechao King of Egypt, who was by the river Euphrates in Charmis, whom Nabuchodonosor king of Babylon smote in the fourth year of Joakim king of Juda.}{46:1-46:2}{Yahweh’s word which came to Jeremiah the prophet concerning the nations. Of Egypt: concerning the army of Pharaoh Necoh king of Egypt, which was by the river Euphrates in Carchemish, which Nebuchadnezzar king of Babylon struck in the fourth year of Jehoiakim the son of Josiah, king of Judah.}
\VersePair{26:3}{Take up arms and spears, and draw nigh to battle;}{46:3}{{\raggedright “Prepare the buckler and shield, \poemline{2}{and draw near to battle!}}}
\VersePair{26:4}{and harness the horses: mount, ye horsemen, and stand ready in your helmets; advance the spears, and put on your breast-plates.}{46:4}{{\raggedright Harness the horses, and get up, you horsemen, \poemline{2}{and stand up with your helmets.}\poemline{1}{Polish the spears,}\poemline{2}{put on the coats of mail.}}}
\VersePair{26:5}{Why do they fear, and turn back? even because their mighty men shall be slain: they have utterly fled, and being hemmed in they have not rallied, saith the Lord.}{46:5}{{\raggedright Why have I seen it? \poemline{2}{They are dismayed and are turned backward.}\poemline{1}{Their mighty ones are beaten down,}\poemline{2}{have fled in haste,}\poemline{2}{and don’t look back.}\poemline{1}{Terror is on every side,”}\poemline{2}{says Yahweh.}}}
\VersePair{26:6}{Let not the swift flee, and let not the mighty man escape to the north: the forces at Euphrates are become feeble, and they have fallen.}{46:6}{{\raggedright “Don’t let the swift flee away, \poemline{2}{nor the mighty man escape.}\poemline{1}{In the north by the river Euphrates}\poemline{2}{they have stumbled and fallen.}}}
\VersePair{26:7}{Who is this that shall come up as a river, and as rivers roll their waves?}{46:7}{{\raggedright “Who is this who rises up like the Nile, \poemline{2}{like rivers whose waters surge?}}}
\VersePair{26:8}{The waters of Egypt shall come up like a river: and he said, I will go up, and will cover the earth, and will destroy the dwellers in it.}{46:8}{{\raggedright Egypt rises up like the Nile, \poemline{2}{like rivers whose waters surge.}\poemline{1}{He says, ‘I will rise up. I will cover the earth.}\poemline{2}{I will destroy cities and its inhabitants.’}}}
\VersePair{26:9}{Mount ye the horses, prepare the chariots; go forth, ye warriors of the Ethiopians, and Libyans armed with shields; and mount, ye Lydians, bend the bow.}{46:9}{{\raggedright Go up, you horses! \poemline{2}{Rage, you chariots!}\poemline{1}{Let the mighty men go out:}\poemline{2}{Cush and Put, who handle the shield;}\poemline{2}{and the Ludim, who handle and bend the bow.}}}
\VersePair{26:10}{And that day shall be to the Lord our God a day of vengeance, to take vengeance on his enemies: and the sword of the Lord shall devour, and be glutted, and be drunken with their blood: for the Lord has a sacrifice from the land of the north at the river Euphrates.}{46:10}{{\raggedright For that day is of the Lord, Yahweh of Armies, \poemline{2}{a day of vengeance,}\poemline{2}{that he may avenge himself of his adversaries.}\poemline{1}{The sword will devour and be satiated,}\poemline{2}{and will drink its fill of their blood;}\poemline{2}{for the Lord, Yahweh of Armies, has a sacrifice in the north country by the river Euphrates.}}}
\VersePair{26:11}{Go up to Galaad, and take balm for the virgin daughter of Egypt: in vain hast thou multiplied thy medicines; there is no help in thee.}{46:11}{{\raggedright Go up into Gilead, and take balm, virgin daughter of Egypt. \poemline{2}{You use many medicines in vain.}\poemline{2}{There is no healing for you.}}}
\VersePair{26:12}{The nations have heard thy voice, and the land has been filled with thy cry: for the warriors have fainted fighting one against another, and both are fallen together.}{46:12}{{\raggedright The nations have heard of your shame, \poemline{2}{and the earth is full of your cry;}\poemline{2}{for the mighty man has stumbled against the mighty,}\poemline{2}{they both fall together.”}}}
\VersePair{26:13}{The Words which the Lord spoke by Jeremias, concerning the coming of the king of Babylon to smite the land of Egypt.}{46:13}{The word that Yahweh spoke to Jeremiah the prophet, how that Nebuchadnezzar king of Babylon should come and strike the land of Egypt:}
\VersePair{26:14}{Proclaim it at Magdol, and declare it at Memphis: say ye, Stand up, and prepare; for the sword has devoured thy yew-tree.}{46:14}{{\raggedright “Declare in Egypt, \poemline{2}{publish in Migdol,}\poemline{2}{and publish in Memphis and in Tahpanhes;}\poemline{1}{say, ‘Stand up, and prepare,}\poemline{2}{for the sword has devoured around you.’}}}
\VersePair{26:15}{Wherefore has Apis fled from thee? thy choice calf has not remained; for the Lord has utterly weakened him.}{46:15}{{\raggedright Why are your strong ones swept away? \poemline{2}{They didn’t stand, because Yahweh pushed them.}}}
\VersePair{26:16}{And thy multitude has fainted and fallen; and each one said to his neighbour, Let us arise, and return into our country to our people, from the Grecian sword.}{46:16}{{\raggedright He made many to stumble. \poemline{2}{Yes, they fell on one another.}\poemline{1}{They said, ‘Arise! Let’s go again to our own people,}\poemline{2}{and to the land of our birth,}\poemline{2}{from the oppressing sword.’}}}
\VersePair{26:17}{Call ye the name of Pharao Nechao king of Egypt, Saon esbeie moed.}{46:17}{{\raggedright They cried there, ‘Pharaoh king of Egypt is but a noise; \poemline{2}{he has let the appointed time pass by.’}}}
\VersePair{26:18}{As I live, saith the Lord God, he shall come as Itabyrion among the mountains, and as Carmel that is on the sea.}{46:18}{{\raggedright “As I live,” says the King, \poemline{2}{whose name is Yahweh of Armies,}\poemline{1}{“surely like Tabor among the mountains,}\poemline{2}{and like Carmel by the sea,}\poemline{2}{so he will come.}}}
\VersePair{26:19}{O daughter of Egypt dwelling at home, prepare thee stuff for removing: for Memphis shall be utterly desolate, and shall be called Woe, because there are no inhabitants in it.}{46:19}{{\raggedright You daughter who dwells in Egypt, \poemline{2}{furnish yourself to go into captivity;}\poemline{1}{for Memphis will become a desolation,}\poemline{2}{and will be burned up,}\poemline{2}{without inhabitant.}}}
\VersePair{26:20}{Egypt is a fair heifer, but destruction from the north is come upon her.}{46:20}{{\raggedright “Egypt is a very beautiful heifer; \poemline{2}{but destruction out of the north has come.}\poemline{2}{It has come.}}}
\VersePair{26:21}{Also her hired soldiers in the midst of her are as fatted calves fed in her; for they also have turned, and fled with one accord: they stood not, for the day of destruction was come upon them, and the time of their retribution.}{46:21}{{\raggedright Also her hired men in the middle of her are like calves of the stall, \poemline{2}{for they also are turned back.}\poemline{2}{They have fled away together.}\poemline{1}{They didn’t stand,}\poemline{2}{for the day of their calamity has come on them,}\poemline{2}{the time of their visitation.}}}
\VersePair{26:22}{Their voice is as that of a hissing serpent, for they go upon the sand; they shall come upon Egypt with axes, as men that cut wood.}{46:22}{{\raggedright Its sound will go like the serpent, \poemline{2}{for they will march with an army,}\poemline{2}{and come against her with axes, as wood cutters.}}}
\VersePair{26:23}{They shall cut down her forest, saith the Lord, for their number cannot at all be conjectured, for it exceeds the locust in multitude, and they are innumerable.}{46:23}{{\raggedright They will cut down her forest,” says Yahweh, \poemline{2}{“though it can’t be searched;}\poemline{1}{because they are more than the locusts,}\poemline{2}{and are innumerable.}}}
\VersePair{26:24}{The daughter of Egypt is confounded; she is delivered into the hands of a people from the north.}{46:24}{{\raggedright The daughter of Egypt will be disappointed; \poemline{2}{she will be delivered into the hand of the people of the north.”}}}
\VersePair{26:25}{Behold, I will avenge Ammon her son upon Pharao, and upon them that trust in him.}{46:25}{Yahweh of Armies, the God of Israel, says: “Behold, I will punish Amon of No, and Pharaoh, and Egypt, with her gods and her kings, even Pharaoh, and those who trust in him.}
\VersePair{26:27}{But fear not thou, my servant Jacob, neither be thou alarmed, Israel: for, behold, I will save thee from afar, and thy seed from their captivity; and Jacob shall return, and be at ease, and sleep, and there shall be no one to trouble him.}{46:27}{{\raggedright “But don’t you be afraid, Jacob my servant. \poemline{2}{Don’t be dismayed, Israel;}\poemline{1}{for, behold, I will save you from afar,}\poemline{2}{and your offspring from the land of their captivity.}\poemline{1}{Jacob will return,}\poemline{2}{and will be quiet and at ease.}\poemline{2}{No one will make him afraid.}}}
\VersePair{26:28}{Fear not thou, my servant Jacob, saith the Lord; for I am with thee: she that was without fear and in luxury, has been delivered up: for I will make a full end of every nation among whom I have thrust thee forth; but I will not cause thee to fail: yet will I chastise thee in the way of judgment, and will not hold thee entirely guiltless.}{46:28}{{\raggedright Don’t be afraid, O Jacob my servant,” says Yahweh, \poemline{2}{“for I am with you;}\poemline{2}{for I will make a full end of all the nations where I have driven you,}\poemline{1}{but I will not make a full end of you,}\poemline{2}{but I will correct you in measure,}\poemline{2}{and will in no way leave you unpunished.”}}}
\end{paracol}
\ChapterHeading{27}
\begin{paracol}{2}
\VersePair{27:1}{The Word of the Lord which He spoke against Babylon.}{50:1}{The word that Yahweh spoke concerning Babylon, concerning the land of the Chaldeans, by Jeremiah the prophet.}
\VersePair{27:2}{Proclaim ye among the Gentiles, and cause the tidings to be heard, and suppress them not: say ye, Babylon is taken, Belus is confounded; the fearless, the luxurious Mærodach is delivered up.}{50:2}{{\raggedright “Declare among the nations and publish, \poemline{2}{and set up a standard;}\poemline{2}{publish, and don’t conceal;}\poemline{1}{say, ‘Babylon has been taken,}\poemline{2}{Bel is disappointed,}\poemline{2}{Merodach is dismayed!}\poemline{1}{Her images are disappointed.}\poemline{2}{Her idols are dismayed.’}}}
\VersePair{27:3}{For a nation has come up against her from the north, he shall utterly ravage her land, and there shall be none to dwell in it, neither man nor beast.}{50:3}{{\raggedright For a nation comes up out of the north against her, \poemline{2}{which will make her land desolate,}\poemline{2}{and no one will dwell in it.}\poemline{1}{They have fled.}\poemline{2}{They are gone,}\poemline{2}{both man and animal.}}}
\VersePair{27:4}{In those days, and at that time, the children of Israel shall come, they and the children of Juda together; they shall proceed, weeping as they go, seeking the Lord their God.}{50:4}{{\raggedright “In those days, and in that time,” says Yahweh, \poemline{2}{“the children of Israel will come,}\poemline{2}{they and the children of Judah together;}\poemline{1}{they will go on their way weeping,}\poemline{2}{and will seek Yahweh their God.}}}
\VersePair{27:5}{They shall ask the way till they come to Sion, for that way shall they set their face; and they shall come and flee for refuge to the Lord their God; for the everlasting covenant shall not be forgotten.}{50:5}{{\raggedright They will inquire concerning Zion with their faces turned toward it, \poemline{2}{saying, ‘Come, and join yourselves to Yahweh in an everlasting covenant}\poemline{2}{that will not be forgotten.’}}}
\VersePair{27:6}{My people have been lost sheep: their shepherds thrust them out, they caused them to wander on the mountains: they went from mountain to hill, they forgot their resting-place.}{50:6}{{\raggedright My people have been lost sheep. \poemline{2}{Their shepherds have caused them to go astray.}\poemline{1}{They have turned them away on the mountains.}\poemline{2}{They have gone from mountain to hill.}\poemline{2}{They have forgotten their resting place.}}}
\VersePair{27:7}{All that found them consumed them: their enemies said, Let us not leave them alone, because they have sinned against the Lord: he that gathered their fathers had a pasture of righteousness.}{50:7}{{\raggedright All who found them have devoured them. \poemline{2}{Their adversaries said, ‘We are not guilty,}\poemline{1}{because they have sinned against Yahweh,}\poemline{2}{the habitation of righteousness,}\poemline{2}{even Yahweh, the hope of their fathers.’}}}
\VersePair{27:8}{Flee ye out of the midst of Babylon, and from the land of the Chaldeans, and go forth, and be as serpents before sleep.}{50:8}{{\raggedright “Flee out of the middle of Babylon! \poemline{2}{Go out of the land of the Chaldeans,}\poemline{2}{and be as the male goats before the flocks.}}}
\VersePair{27:9}{For, behold, I stir up against Babylon the gatherings of nations out of the land of the north; and they shall set themselves in array against her: thence shall she be taken, as the dart of an expert warrior shall not return empty.}{50:9}{{\raggedright For, behold, I will stir up \poemline{2}{and cause to come up against Babylon a company of great nations from the north country;}\poemline{2}{and they will set themselves in array against her.}\poemline{1}{She will be taken from there.}\poemline{2}{Their arrows will be as of an expert mighty man.}\poemline{2}{None of them will return in vain.}}}
\VersePair{27:10}{And Chaldea shall be a spoil: all that spoil her shall be satisfied.}{50:10}{{\raggedright Chaldea will be a prey. \poemline{2}{All who prey on her will be satisfied,” says Yahweh.}}}
\VersePair{27:11}{Because ye rejoiced, and boasted, while plundering mine heritage; because ye exulted as calves in the grass, and pushed with the horn as bulls.}{50:11}{{\raggedright “Because you are glad, \poemline{2}{because you rejoice,}\poemline{1}{O you who plunder my heritage,}\poemline{2}{because you are wanton as a heifer that treads out the grain,}\poemline{2}{and neigh as strong horses,}}}
\VersePair{27:12}{Your mother is greatly ashamed; your mother that bore you for prosperity is confounded: she is the last of the nations, desolate,}{50:12}{{\raggedright your mother will be utterly disappointed. \poemline{2}{She who bore you will be confounded.}\poemline{1}{Behold, she will be the least of the nations,}\poemline{2}{a wilderness, a dry land, and a desert.}}}
\VersePair{27:13}{by reason of the Lord's anger: it shall not be inhabited, but it shall be all a desolation; and every one that passes through Babylon shall scowl, and they shall hiss at all her plague.}{50:13}{{\raggedright Because of Yahweh’s wrath she won’t be inhabited, \poemline{2}{but she will be wholly desolate.}\poemline{1}{Everyone who goes by Babylon will be astonished,}\poemline{2}{and hiss at all her plagues.}}}
\VersePair{27:14}{Set yourselves in array against Babylon round about, all ye that bend the bow; shoot at her, spare not your arrows,}{50:14}{{\raggedright Set yourselves in array against Babylon all around, \poemline{2}{all you who bend the bow;}\poemline{2}{shoot at her.}\poemline{1}{Spare no arrows,}\poemline{2}{for she has sinned against Yahweh.}}}
\VersePair{27:15}{and prevail against her: her hands are weakened, her bulwarks are fallen, and her wall is broken down: for it is vengeance from God: take vengeance upon her; as she has done, do to her.}{50:15}{{\raggedright Shout against her all around. \poemline{2}{She has submitted herself.}\poemline{2}{Her bulwarks have fallen.}\poemline{1}{Her walls have been thrown down,}\poemline{2}{for it is the vengeance of Yahweh.}\poemline{1}{Take vengeance on her.}\poemline{2}{As she has done, do to her.}}}
\VersePair{27:16}{Utterly destroy seed out of Babylon, and him that holds a sickle in time of harvest: for fear of the Grecian sword, they shall return every one to his people, and every one shall flee to his own land.}{50:16}{{\raggedright Cut off the sower from Babylon, \poemline{2}{and him who handles the sickle in the time of harvest.}\poemline{1}{For fear of the oppressing sword,}\poemline{2}{they will each return to their own people,}\poemline{2}{and they will each flee to their own land.}}}
\VersePair{27:17}{Israel is a wandering sheep; the lions have driven him out: the king of Assyria first devoured him, and afterward this king of Babylon has gnawed his bones.}{50:17}{{\raggedright “Israel is a hunted sheep. \poemline{2}{The lions have driven him away.}\poemline{1}{First, the king of Assyria devoured him,}\poemline{2}{and now at last Nebuchadnezzar king of Babylon has broken his bones.”}}}
\VersePair{27:18}{Therefore thus saith the Lord; Behold, I will take vengeance on the king of Babylon, and upon his land, as I took vengeance on the king of Assyria.}{50:18}{{\raggedright Therefore Yahweh of Armies, the God of Israel, says: \poemline{1}{“Behold, I will punish the king of Babylon and his land,}\poemline{2}{as I have punished the king of Assyria.}}}
\VersePair{27:19}{And I will restore Israel to his pasture, and he shall feed on Carmel and on mount Ephraim and in Galaad, and his soul shall be satisfied.}{50:19}{{\raggedright I will bring Israel again to his pasture, \poemline{2}{and he will feed on Carmel and Bashan.}\poemline{2}{His soul will be satisfied on the hills of Ephraim and in Gilead.}}}
\VersePair{27:20}{In those days, and at that time, they shall seek for the iniquity of Israel, and there shall be none; and for the sins of Juda, and they shall not be found: for I will be merciful to them that are left}{50:20}{{\raggedright In those days, and in that time,” says Yahweh, \poemline{2}{“the iniquity of Israel will be sought for,}\poemline{2}{and there will be none,}\poemline{1}{also the sins of Judah,}\poemline{2}{and they won’t be found;}\poemline{2}{for I will pardon them whom I leave as a remnant.}}}
\VersePair{27:21}{on the land, saith the Lord. Go up against it roughly, and against them that dwell on it: avenge, O sword, and destroy utterly, saith the Lord, and do according to all that I command thee.}{50:21}{{\raggedright “Go up against the land of Merathaim, \poemline{2}{even against it, and against the inhabitants of Pekod.}\poemline{1}{Kill and utterly destroy after them,” says Yahweh,}\poemline{2}{“and do according to all that I have commanded you.}}}
\VersePair{27:22}{A sound of war, and great destruction in the land of the Chaldeans!}{50:22}{{\raggedright A sound of battle is in the land, \poemline{2}{and of great destruction.}}}
\VersePair{27:23}{How is the hammer of the whole earth broken and crushed! How is Babylon become a desolation among the nations!}{50:23}{{\raggedright How the hammer of the whole earth is cut apart and broken! \poemline{2}{How Babylon has become a desolation among the nations!}}}
\VersePair{27:24}{They shall come upon thee, and thou shalt not know it, Babylon, that thou wilt even be taken captive: thou art found and taken, because thou didst resist the Lord.}{50:24}{{\raggedright I have laid a snare for you, \poemline{2}{and you are also taken, Babylon,}\poemline{2}{and you weren’t aware.}\poemline{1}{You are found,}\poemline{2}{and also caught,}\poemline{2}{because you have fought against Yahweh.}}}
\VersePair{27:25}{The Lord has opened his treasury, and brought forth the weapons of his anger: for the Lord God has a work in the land of the Chaldeans.}{50:25}{{\raggedright Yahweh has opened his armory, \poemline{2}{and has brought out the weapons of his indignation;}\poemline{2}{for the Lord, Yahweh of Armies, has a work to do in the land of the Chaldeans.}}}
\VersePair{27:26}{For her times are come: open ye her storehouses: search her as a cave, and utterly destroy her: let there be no remnant of her.}{50:26}{{\raggedright Come against her from the farthest border. \poemline{2}{Open her storehouses.}\poemline{2}{Cast her up as heaps.}\poemline{1}{Destroy her utterly.}\poemline{2}{Let nothing of her be left.}}}
\VersePair{27:27}{Dry ye up all her fruits, and let them go down to the slaughter: woe to them! for their day is come, and the time of their retribution.}{50:27}{{\raggedright Kill all her bulls. \poemline{2}{Let them go down to the slaughter.}\poemline{1}{Woe to them! For their day has come,}\poemline{2}{the time of their visitation.}}}
\VersePair{27:28}{A voice of men fleeing and escaping from the land of Babylon, to declare to Sion the vengeance that comes from the Lord our God.}{50:28}{{\raggedright Listen to those who flee and escape out of the land of Babylon, \poemline{2}{to declare in Zion the vengeance of Yahweh our God,}\poemline{2}{the vengeance of his temple.}}}
\VersePair{27:29}{Summon many against Babylon, even every one that bends the bow: camp against her round about; let no one of her people escape: render to her according to her works; according to all that she has done, do to her: for she has resisted the Lord, the Holy God of Israel.}{50:29}{{\raggedright “Call together the archers against Babylon, \poemline{2}{all those who bend the bow.}\poemline{1}{Encamp against her all around.}\poemline{2}{Let none of it escape.}\poemline{1}{Pay her back according to her work.}\poemline{2}{According to all that she has done, do to her;}\poemline{1}{for she has been proud against Yahweh,}\poemline{2}{against the Holy One of Israel.}}}
\VersePair{27:30}{Therefore shall her young men fall in the streets, and all her warriors shall be cast down, saith the Lord.}{50:30}{{\raggedright Therefore her young men will fall in her streets. \poemline{2}{All her men of war will be brought to silence in that day,” says Yahweh.}}}
\VersePair{27:31}{Behold, I am against thee the haughty one, saith the Lord: for thy day is come, and the time of thy retribution.}{50:31}{{\raggedright “Behold, I am against you, you proud one,” says the Lord, Yahweh of Armies; \poemline{2}{“for your day has come,}\poemline{2}{the time that I will visit you.}}}
\VersePair{27:32}{And thy pride shall fail, and fall, and there shall be no one to set it up again: and I will kindle a fire in her forest, and it shall devour all things round about her.}{50:32}{{\raggedright The proud one will stumble and fall, \poemline{2}{and no one will raise him up.}\poemline{1}{I will kindle a fire in his cities,}\poemline{2}{and it will devour all who are around him.”}}}
\VersePair{27:33}{Thus saith the Lord; The children of Israel and the children of Juda have been oppressed: all they that have taken them captive have oppressed them together; for they would not let them go.}{50:33}{{\raggedright Yahweh of Armies says: “The children of Israel and the children of Judah are oppressed together. \poemline{2}{All who took them captive hold them fast.}\poemline{2}{They refuse to let them go.}}}
\VersePair{27:34}{But their Redeemer is strong; the Lord Almighty is his name: he will enter into judgment with his adversaries, that he may destroy the earth;}{50:34}{{\raggedright Their Redeemer is strong. \poemline{2}{Yahweh of Armies is his name.}\poemline{1}{He will thoroughly plead their cause,}\poemline{2}{that he may give rest to the earth,}\poemline{2}{and disquiet the inhabitants of Babylon.}}}
\VersePair{27:35}{and he will sharpen a sword against the Chaldeans, and against the inhabitants of Babylon, and upon her nobles and upon her wise men;}{50:35}{{\raggedright “A sword is on the Chaldeans,” says Yahweh, \poemline{2}{“and on the inhabitants of Babylon,}\poemline{2}{on her princes,}\poemline{2}{and on her wise men.}}}
\VersePair{27:36}{a sword upon her warriors, and they shall be weakened: a sword upon their horses, and upon their chariots:}{50:36}{{\raggedright A sword is on the boasters, \poemline{2}{and they will become fools.}\poemline{1}{A sword is on her mighty men,}\poemline{2}{and they will be dismayed.}}}
\VersePair{27:37}{a sword upon their warriors, and upon the mixed people in the midst of her; and they shall be as women: a sword upon the treasures, and they shall be scattered upon her water,}{50:37}{{\raggedright A sword is on their horses, \poemline{2}{on their chariots,}\poemline{2}{and on all the mixed people who are in the middle of her;}\poemline{2}{and they will become as women.}\poemline{1}{A sword is on her treasures,}\poemline{2}{and they will be robbed.}}}
\VersePair{27:38}{and they shall be ashamed: for it is a land of graven images; and in the islands, where they boasted.}{50:38}{{\raggedright A drought is on her waters, \poemline{2}{and they will be dried up;}\poemline{1}{for it is a land of engraved images,}\poemline{2}{and they are mad over idols.}}}
\VersePair{27:39}{Therefore shall idols dwell in the islands, and the young of monsters shall dwell in it: it shall not be inhabited any more for ever.}{50:39}{{\raggedright Therefore the wild animals of the desert \poemline{2}{with the wolves will dwell there.}\poemline{1}{The ostriches will dwell therein.}\poemline{2}{It will be inhabited no more forever,}\poemline{2}{neither will it be lived in from generation to generation.}}}
\VersePair{27:40}{As God overthrew Sodom and Gomorrha, and the cities bordering upon them, saith the Lord: no man shall dwell there, and no son of man shall sojourn there.}{50:40}{{\raggedright As when God overthrew Sodom and Gomorrah and its neighbor cities,” says Yahweh, \poemline{2}{“so no man will dwell there,}\poemline{2}{neither will any son of man live therein.}}}
\VersePair{27:41}{Behold, a people comes from the north, and a great nation, and many kings shall be stirred up from the end of the earth; holding bow and dagger:}{50:41}{{\raggedright “Behold, a people comes from the north. \poemline{2}{A great nation and many kings will be stirred up from the uttermost parts of the earth.}}}
\VersePair{27:42}{the people is fierce, and will have no mercy: their voices shall sound as the sea, they shall ride upon horses, prepared for war, like fire, against thee, O daughter of Babylon.}{50:42}{{\raggedright They take up bow and spear. \poemline{2}{They are cruel, and have no mercy.}\poemline{2}{Their voice roars like the sea.}\poemline{1}{They ride on horses,}\poemline{2}{everyone set in array,}\poemline{2}{as a man to the battle,}\poemline{2}{against you, daughter of Babylon.}}}
\VersePair{27:43}{The king of Babylon heard the sound of them, and his hands were enfeebled: anguish overcame him, pangs as of a woman in travail.}{50:43}{{\raggedright The king of Babylon has heard the news of them, \poemline{2}{and his hands become feeble.}\poemline{1}{Anguish has taken hold of him,}\poemline{2}{pains as of a woman in labor.}}}
\VersePair{27:44}{Behold, he shall come up as a lion from Jordan to Gæthan; for I will speedily drive them from her, and I will set all the youths against her: for who is like me? and who will resist me? and who is this shepherd who will stand before me?}{50:44}{{\raggedright Behold, the enemy will come up like a lion \poemline{2}{from the thickets of the Jordan against the strong habitation;}\poemline{2}{for I will suddenly make them run away from it.}\poemline{1}{Whoever is chosen,}\poemline{2}{I will appoint him over it,}\poemline{2}{for who is like me?}\poemline{1}{Who will appoint me a time?}\poemline{2}{Who is the shepherd who can stand before me?”}}}
\VersePair{27:45}{Therefore hear ye the counsel of the Lord, which he has taken against Babylon; and his devices, which he has devised upon the Chaldeans inhabiting it: surely lambs of their flock shall be destroyed: surely pasture shall be cut off from them.}{50:45}{{\raggedright Therefore hear the counsel of Yahweh \poemline{2}{that he has taken against Babylon;}\poemline{1}{and his purposes}\poemline{2}{that he has purposed against the land of the Chaldeans:}\poemline{1}{Surely they will drag them away,}\poemline{2}{even the little ones of the flock.}\poemline{2}{Surely he will make their habitation desolate over them.}}}
\VersePair{27:46}{For at the sound of the taking of Babylon the earth shall quake, and a cry shall be heard among the nations.}{50:46}{{\raggedright The earth trembles at the noise of the taking of Babylon. \poemline{2}{The cry is heard among the nations.}}}
\end{paracol}
\ChapterHeading{28}
\begin{paracol}{2}
\VersePair{28:1}{Thus saith the Lord; Behold, I stir up against Babylon, and against the Chaldeans dwelling therein, a deadly burning wind.}{51:1}{{\raggedright Yahweh says: \poemline{1}{“Behold, I will raise up against Babylon,}\poemline{2}{and against those who dwell in Lebkamai, a destroying wind.}}}
\VersePair{28:2}{And I will send forth against Babylon spoilers, and they shall spoil her, and shall ravage her land. Woe to Babylon round about her in the day of her affliction.}{51:2}{{\raggedright I will send to Babylon strangers, who will winnow her. \poemline{2}{They will empty her land;}\poemline{2}{for in the day of trouble they will be against her all around.}}}
\VersePair{28:3}{Let the archer bend his bow, and him that has armour put it on: and spare ye not her young men, but destroy ye all her host.}{51:3}{{\raggedright Against him who bends, let the archer bend his bow, \poemline{2}{also against him who lifts himself up in his coat of mail.}\poemline{1}{Don’t spare her young men!}\poemline{2}{Utterly destroy all her army!}}}
\VersePair{28:4}{And slain men shall fall in the land of the Chaldeans, and men pierced through shall fall without it.}{51:4}{{\raggedright They will fall down slain in the land of the Chaldeans, \poemline{2}{and thrust through in her streets.}}}
\VersePair{28:5}{For Israel and Juda have not been forsaken of their God, of the Lord Almighty; whereas their land was filled with iniquity against the holy things of Israel.}{51:5}{{\raggedright For Israel is not forsaken, nor Judah, by his God, \poemline{2}{by Yahweh of Armies;}\poemline{2}{though their land is full of guilt against the Holy One of Israel.}}}
\VersePair{28:6}{Flee ye out of the midst of Babylon, and deliver every one his soul: and be not overthrown in her iniquity; for it is the time of her retribution from the Lord; he is rendering to her a recompence.}{51:6}{{\raggedright “Flee out of the middle of Babylon! \poemline{2}{Everyone save his own life!}\poemline{1}{Don’t be cut off in her iniquity,}\poemline{2}{for it is the time of Yahweh’s vengeance.}\poemline{2}{He will give to her a recompense.}}}
\VersePair{28:7}{Babylon has been a golden cup in the Lord's hand, causing all the earth to be drunken: the nations have drunk of her wine; therefore they were shaken.}{51:7}{{\raggedright Babylon has been a golden cup in Yahweh’s hand, \poemline{2}{who made all the earth drunk.}\poemline{1}{The nations have drunk of her wine;}\poemline{2}{therefore the nations have gone mad.}}}
\VersePair{28:8}{And Babylon is fallen suddenly, and is broken to pieces: lament for her; take balm for her deadly wound, if by any means she may be healed.}{51:8}{{\raggedright Babylon has suddenly fallen and been destroyed! \poemline{2}{Wail for her!}\poemline{1}{Take balm for her pain.}\poemline{2}{Perhaps she may be healed.}}}
\VersePair{28:9}{We tried to heal Babylon, but she was not healed: let us forsake her, and depart every one to his own country: for her judgment has reached to the heaven, it has mounted up to the stars.}{51:9}{{\raggedright “We would have healed Babylon, \poemline{2}{but she is not healed.}\poemline{1}{Forsake her,}\poemline{2}{and let’s each go into his own country;}\poemline{1}{for her judgment reaches to heaven,}\poemline{2}{and is lifted up even to the skies.}}}
\VersePair{28:10}{The Lord has brought forth his judgment: come, and let us declare in Sion the works of the Lord our God.}{51:10}{{\raggedright ‘Yahweh has produced our righteousness. \poemline{2}{Come, and let’s declare in Zion the work of Yahweh our God.’}}}
\VersePair{28:11}{Prepare the arrows; fill the quivers: the Lord has stirred up the spirit of the king of the Medes: for his wrath is against Babylon, to destroy it utterly; for it is the Lord's vengeance, it is the vengeance of his people.}{51:11}{{\raggedright “Make the arrows sharp! \poemline{2}{Hold the shields firmly!}\poemline{1}{Yahweh has stirred up the spirit of the kings of the Medes,}\poemline{2}{because his purpose is against Babylon, to destroy it;}\poemline{1}{for it is the vengeance of Yahweh,}\poemline{2}{the vengeance of his temple.}}}
\VersePair{28:12}{Lift up a standard on the walls of Babylon, prepare the quivers, rouse the guards, prepare the weapons: for the Lord has taken the work in hand, and will execute what he has spoken against the inhabitants of Babylon,}{51:12}{{\raggedright Set up a standard against the walls of Babylon! \poemline{2}{Make the watch strong!}\poemline{1}{Set the watchmen,}\poemline{2}{and prepare the ambushes;}\poemline{1}{for Yahweh has both purposed and done}\poemline{2}{that which he spoke concerning the inhabitants of Babylon.}}}
\VersePair{28:13}{dwelling on many waters, and amidst the abundance of her treasures; thine end is come verily into thy bowels.}{51:13}{{\raggedright You who dwell on many waters, abundant in treasures, \poemline{2}{your end has come, the measure of your covetousness.}}}
\VersePair{28:14}{For the Lord has sworn by his arm, saying, I will fill thee with men as with locusts; and they that come down shall cry against thee.}{51:14}{{\raggedright Yahweh of Armies has sworn by himself, saying, \poemline{2}{‘Surely I will fill you with men,}\poemline{2}{as with locusts,}\poemline{2}{and they will lift up a shout against you.’}}}
\VersePair{28:15}{The Lord made the earth by his power, preparing the world by his wisdom, by his understanding he stretched out the heaven.}{51:15}{{\raggedright “He has made the earth by his power. \poemline{2}{He has established the world by his wisdom.}\poemline{2}{By his understanding he has stretched out the heavens.}}}
\VersePair{28:16}{At his voice he makes a sound of water in the heaven, and brings up clouds from the extremity of the earth; he makes lightnings for rain, and brings light out of his treasures.}{51:16}{{\raggedright When he utters his voice, \poemline{2}{there is a roar of waters in the heavens,}\poemline{2}{and he causes the vapors to ascend from the ends of the earth.}\poemline{1}{He makes lightning for the rain,}\poemline{2}{and brings the wind out of his treasuries.}}}
\VersePair{28:17}{Every man has completely lost understanding; every goldsmith is confounded because of his graven images: for they have cast false gods, there is no breath in them.}{51:17}{{\raggedright “Every man has become stupid and without knowledge. \poemline{2}{Every goldsmith is disappointed by his image,}\poemline{1}{for his molten images are falsehood,}\poemline{2}{and there is no breath in them.}}}
\VersePair{28:18}{They are vain works, objects of scorn; in the time of their visitation they shall perish.}{51:18}{{\raggedright They are vanity, \poemline{2}{a work of delusion.}\poemline{2}{In the time of their visitation, they will perish.}}}
\VersePair{28:19}{Not such is Jacob's portion; for he that formed all things, he is his inheritance; the Lord is his name.}{51:19}{{\raggedright The portion of Jacob is not like these, \poemline{2}{for he formed all things,}\poemline{2}{including the tribe of his inheritance.}\poemline{2}{Yahweh of Armies is his name.}}}
\VersePair{28:20}{Thou scatterest for me the weapons of war: and I will scatter nations by thee, and will destroy kings by means of thee.}{51:20}{{\raggedright “You are my battle ax and weapons of war. \poemline{2}{With you I will break the nations into pieces.}\poemline{2}{With you I will destroy kingdoms.}}}
\VersePair{28:21}{And by thee I will scatter the horse and his rider; and by thee I will scatter chariots and them that ride in them.}{51:21}{{\raggedright With you I will break in pieces \poemline{2}{the horse and his rider.}}}
\VersePair{28:22}{And by thee I will scatter youth and maid; and by thee I will scatter man and woman.}{51:22}{{\raggedright With you I will break in pieces \poemline{2}{the chariot and him who rides therein.}\poemline{1}{With you I will break in pieces}\poemline{2}{man and woman.}\poemline{1}{With you I will break in pieces}\poemline{2}{the old man and the youth.}\poemline{1}{With you I will break in pieces}\poemline{2}{the young man and the virgin.}}}
\VersePair{28:23}{And by thee I will scatter the shepherd and his flock; and by thee I will scatter the husbandman and his husbandry; and by thee I will scatter leaders and the captains.}{51:23}{{\raggedright With you I will break in pieces \poemline{2}{the shepherd and his flock.}\poemline{1}{With you I will break in pieces}\poemline{2}{the farmer and his yoke.}\poemline{1}{With you I will break in pieces}\poemline{2}{governors and deputies.}}}
\VersePair{28:24}{And I will recompense to Babylon and to all the Chaldeans that dwell there all their mischiefs that they have done to Sion before your eyes, saith the Lord.}{51:24}{“I will give to Babylon and to all the inhabitants of Chaldea all their evil that they have done in Zion in your sight,” says Yahweh.}
\VersePair{28:25}{Behold, I am against thee, the ruined mountain, that destroys the whole earth; and I will stretch out mine hand upon thee, and will roll thee down upon the rocks, and will make thee as a burnt mountain.}{51:25}{{\raggedright “Behold, I am against you, destroying mountain,” says Yahweh, \poemline{2}{“which destroys all the earth.}\poemline{1}{I will stretch out my hand on you,}\poemline{2}{roll you down from the rocks,}\poemline{2}{and will make you a burned mountain.}}}
\VersePair{28:26}{And they shall not take from thee a stone for a corner, nor a stone for a foundation: for thou shalt be a desolation for ever, saith the Lord.}{51:26}{{\raggedright They won’t take a cornerstone from you, \poemline{2}{nor a stone for foundations;}\poemline{2}{but you will be desolate forever,” says Yahweh.}}}
\VersePair{28:27}{Lift up a standard in the land, sound the trumpet among the nations, consecrate the nations against her, raise up kings against her by me, and that for the people of Achanaz; set against her engines of war; bring up against her horses as a multitude of locusts.}{51:27}{{\raggedright “Set up a standard in the land! \poemline{2}{Blow the trumpet among the nations!}\poemline{1}{Prepare the nations against her!}\poemline{2}{Call together against her the kingdoms of Ararat, Minni, and Ashkenaz!}\poemline{1}{Appoint a marshal against her!}\poemline{2}{Cause the horses to come up as the swarming locusts!}}}
\VersePair{28:28}{Bring up nations against her, even the king of the Medes and of the whole earth, his rulers, and all his captains.}{51:28}{{\raggedright Prepare against her the nations, \poemline{2}{the kings of the Medes, its governors, and all its deputies, and all the land of their dominion!}}}
\VersePair{28:29}{The earth has quaked and been troubled, because the purpose of the Lord has risen up against Babylon, to make the land of Babylon a desolation, and uninhabitable.}{51:29}{{\raggedright The land trembles and is in pain; \poemline{2}{for the purposes of Yahweh against Babylon stand,}\poemline{2}{to make the land of Babylon a desolation, without inhabitant.}}}
\VersePair{28:30}{The warrior of Babylon has failed to fight; they shall sit there in the siege; their power is broken; they are become like women; her tabernacles have been set on fire; her bars are broken.}{51:30}{{\raggedright The mighty men of Babylon have stopped fighting, \poemline{2}{they remain in their strongholds.}\poemline{1}{Their might has failed.}\poemline{2}{They have become as women.}\poemline{1}{Her dwelling places are set on fire.}\poemline{2}{Her bars are broken.}}}
\VersePair{28:31}{One shall rush, running to meet another runner, and one shall go with tidings to meet another with tidings, to bring tidings to the king of Babylon, that his city is taken.}{51:31}{{\raggedright One runner will run to meet another, \poemline{2}{and one messenger to meet another,}\poemline{2}{to show the king of Babylon that his city is taken on every quarter.}}}
\VersePair{28:32}{At the end of his passages they were taken, and his cisterns they have burnt with fire, and his warriors are going forth.}{51:32}{{\raggedright So the passages are seized. \poemline{2}{They have burned the reeds with fire.}\poemline{2}{The men of war are frightened.”}}}
\VersePair{28:33}{For thus saith the Lord, The houses of the king of Babylon shall be threshed as a floor in the season; yet a little while, and her harvest shall come.}{51:33}{{\raggedright For Yahweh of Armies, the God of Israel says: \poemline{1}{“The daughter of Babylon is like a threshing floor at the time when it is trodden.}\poemline{2}{Yet a little while, and the time of harvest comes for her.”}}}
\VersePair{28:34}{He has devoured me, he has torn me asunder, airy darkness has come upon me; Nabuchodonosor king of Babylon has swallowed me up, as a dragon has he filled his belly with my delicacies.}{51:34}{{\raggedright “Nebuchadnezzar the king of Babylon has devoured me. \poemline{2}{He has crushed me.}\poemline{2}{He has made me an empty vessel.}\poemline{1}{He has, like a monster, swallowed me up.}\poemline{2}{He has filled his mouth with my delicacies.}\poemline{2}{He has cast me out.}}}
\VersePair{28:35}{My troubles and my distresses have driven me out into Babylon, shall she that dwells in Sion say; and my blood shall be upon the Chaldeans dwelling there, shall Jerusalem say.}{51:35}{{\raggedright May the violence done to me and to my flesh be on Babylon!” \poemline{2}{the inhabitant of Zion will say; and,}\poemline{1}{“May my blood be on the inhabitants of Chaldea!”}\poemline{2}{will Jerusalem say.}}}
\VersePair{28:36}{Therefore thus saith the Lord, Behold, I will judge thine adversary, and I will execute vengeance for thee; and I will waste her sea, and dry up her fountain.}{51:36}{{\raggedright Therefore Yahweh says: \poemline{1}{“Behold, I will plead your cause,}\poemline{2}{and take vengeance for you.}\poemline{1}{I will dry up her sea,}\poemline{2}{and make her fountain dry.}}}
\VersePair{28:37}{And Babylon shall be a desolation, and shall not be inhabited.}{51:37}{{\raggedright Babylon will become heaps, \poemline{2}{a dwelling place for jackals,}\poemline{2}{an astonishment, and a hissing,}\poemline{2}{without inhabitant.}}}
\VersePair{28:38}{For they rose up together as lions, and as lions'whelps.}{51:38}{{\raggedright They will roar together like young lions. \poemline{2}{They will growl as lions’cubs.}}}
\VersePair{28:39}{In their heat I will give them a draught, and make them drunk, that they may be stupified, and sleep an everlasting sleep, and not awake, saith the Lord.}{51:39}{{\raggedright When they are inflamed, I will make their feast, \poemline{2}{and I will make them drunk,}\poemline{1}{that they may rejoice,}\poemline{2}{and sleep a perpetual sleep,}\poemline{2}{and not wake up,” says Yahweh.}}}
\VersePair{28:40}{And bring thou them down as lambs to the slaughter, and rams with kids.}{51:40}{{\raggedright “I will bring them down like lambs to the slaughter, \poemline{2}{like rams with male goats.}}}
\VersePair{28:41}{How has the boast of all the earth been taken and caught in a snare! how has Babylon become a desolation among the nations!}{51:41}{{\raggedright “How Sheshach is taken! \poemline{2}{How the praise of the whole earth is seized!}\poemline{2}{How Babylon has become a desolation among the nations!}}}
\VersePair{28:42}{The sea has come up upon Babylon with the sound of its waves, and she is covered.}{51:42}{{\raggedright The sea has come up on Babylon. \poemline{2}{She is covered with the multitude of its waves.}}}
\VersePair{28:43}{Her cities are become like a dry and trackless land; not so much as one man shall dwell in it, neither shall a son of man lodge in it.}{51:43}{{\raggedright Her cities have become a desolation, \poemline{2}{a dry land, and a desert,}\poemline{2}{a land in which no man dwells.}\poemline{2}{No son of man passes by it.}}}
\VersePair{28:44}{And I will take vengeance on Babylon, and bring forth out of her mouth what she has swallowed down, and the nations shall no more be gathered to her:}{51:44}{{\raggedright I will execute judgment on Bel in Babylon, \poemline{2}{and I will bring out of his mouth that which he has swallowed up.}\poemline{1}{The nations will not flow any more to him.}\poemline{2}{Yes, the wall of Babylon will fall.}}}
\VersePair{28:45}{}{51:45}{{\raggedright “My people, go away from the middle of her, \poemline{2}{and each of you save yourselves from Yahweh’s fierce anger.}}}
\VersePair{28:46}{}{51:46}{{\raggedright Don’t let your heart faint. \poemline{2}{Don’t fear for the news that will be heard in the land.}\poemline{1}{For news will come one year,}\poemline{2}{and after that in another year news will come,}\poemline{2}{and violence in the land,}\poemline{2}{ruler against ruler.}}}
\VersePair{28:47}{}{51:47}{{\raggedright Therefore behold, the days come that I will execute judgment on the engraved images of Babylon; \poemline{2}{and her whole land will be confounded.}\poemline{2}{All her slain will fall in the middle of her.}}}
\VersePair{28:48}{}{51:48}{{\raggedright Then the heavens and the earth, \poemline{2}{and all that is therein,}\poemline{1}{will sing for joy over Babylon;}\poemline{2}{for the destroyers will come to her from the north,” says Yahweh.}}}
\VersePair{28:49}{and in Babylon the slain men of all the earth shall fall.}{51:49}{{\raggedright “As Babylon has caused the slain of Israel to fall, \poemline{2}{so the slain of all the land will fall at Babylon.}}}
\VersePair{28:50}{Go forth of the land, ye that escape, and stay not; ye that are afar off, remember the Lord, and let Jerusalem come into your mind.}{51:50}{{\raggedright You who have escaped the sword, go! \poemline{2}{Don’t stand still!}\poemline{1}{Remember Yahweh from afar,}\poemline{2}{and let Jerusalem come into your mind.”}}}
\VersePair{28:51}{We are ashamed, because we have heard our reproach; disgrace has covered our face; aliens are come into our sanctuary, even into the house of the Lord.}{51:51}{{\raggedright “We are confounded \poemline{2}{because we have heard reproach.}\poemline{1}{Confusion has covered our faces,}\poemline{2}{for strangers have come into the sanctuaries of Yahweh’s house.”}}}
\VersePair{28:52}{Therefore, behold, the days come, saith the Lord, when I will take vengeance upon her graven images: and slain men shall fall in all her land.}{51:52}{{\raggedright “Therefore behold, the days come,” says Yahweh, \poemline{2}{“that I will execute judgment on her engraved images;}\poemline{2}{and through all her land the wounded will groan.}}}
\VersePair{28:53}{For though Babylon should go up as the heaven, and though she should strengthen her walls with her power, from me shall come they that shall destroy her, saith the Lord.}{51:53}{{\raggedright Though Babylon should mount up to the sky, \poemline{2}{and though she should fortify the height of her strength,}\poemline{2}{yet destroyers will come to her from me,” says Yahweh.}}}
\VersePair{28:54}{A sound of a cry in Babylon, and great destruction in the land of the Chaldeans:}{51:54}{{\raggedright “The sound of a cry comes from Babylon, \poemline{2}{and of great destruction from the land of the Chaldeans!}}}
\VersePair{28:55}{for the Lord has utterly destroyed Babylon, and cut off from her the great voice sounding as many waters: he has consigned her voice to destruction.}{51:55}{{\raggedright For Yahweh lays Babylon waste, \poemline{2}{and destroys out of her the great voice!}\poemline{1}{Their waves roar like many waters.}\poemline{2}{The noise of their voice is uttered.}}}
\VersePair{28:56}{For distress has come upon Babylon, her warriors are taken, their bows are useless: for God recompenses them.}{51:56}{{\raggedright For the destroyer has come on her, \poemline{2}{even on Babylon.}\poemline{1}{Her mighty men are taken.}\poemline{2}{Their bows are broken in pieces,}\poemline{1}{for Yahweh is a God of retribution.}\poemline{2}{He will surely repay.}}}
\VersePair{28:57}{The Lord recompenses, and will make her leaders and her wise men and her captains completely drunk, saith the King, the Lord Almighty is his name.}{51:57}{{\raggedright I will make her princes, her wise men, \poemline{2}{her governors, her deputies, and her mighty men drunk.}\poemline{1}{They will sleep a perpetual sleep,}\poemline{2}{and not wake up,”}\poemline{2}{says the King, whose name is Yahweh of Armies.}}}
\VersePair{28:58}{Thus saith the Lord, The wall of Babylon was made broad, but it shall be completely broken down, and her high gates shall be burnt with fire; and the peoples shall not labour in vain, nor the nations fail in their rule.}{51:58}{{\raggedright Yahweh of Armies says: \poemline{1}{“The wide walls of Babylon will be utterly overthrown.}\poemline{2}{Her high gates will be burned with fire.}\poemline{1}{The peoples will labor for vanity,}\poemline{2}{and the nations for the fire;}\poemline{2}{and they will be weary.”}}}
\VersePair{28:59}{The Word which the Lord commanded the Prophet Jeremias to say to Saræas son of Nerias, son of Maasæas, when he went from Sedekias king of Juda to Babylon, in the fourth year of his reign. And Saræas was over the bounties.}{51:59}{The word which Jeremiah the prophet commanded Seraiah the son of Neriah, the son of Mahseiah, when he went with Zedekiah the king of Judah to Babylon in the fourth year of his reign. Now Seraiah was chief quartermaster.}
\VersePair{28:60}{And Jeremias wrote in a book all the evils which should come upon Babylon, even all these words that are written against Babylon.}{51:60}{Jeremiah wrote in a book all the evil that should come on Babylon, even all these words that are written concerning Babylon.}
\VersePair{28:61}{And Jeremias said to Saræas, When thou art come to Babylon, and shalt see and read all these words;}{51:61}{Jeremiah said to Seraiah, “When you come to Babylon, then see that you read all these words,}
\VersePair{28:62}{then thou shalt say, O Lord God, thou hast spoken against this place, to destroy it, and that there should be none to dwell in it, neither man nor beast; for it shall be a desolation for ever.}{51:62}{and say, ‘Yahweh, you have spoken concerning this place, to cut it off, that no one will dwell in it, neither man nor animal, but that it will be desolate forever.’}
\VersePair{28:63}{And it shall come to pass, when thou shalt cease from reading this book, that thou shalt bind a stone upon it, and cast it into the midst of Euphrates;}{51:63}{It will be, when you have finished reading this book, that you shall bind a stone to it, and cast it into the middle of the Euphrates.}
\VersePair{28:64}{and shalt say, Thus shall Babylon sink, and not rise, because of the evils which I bring upon it.}{51:64}{Then you shall say, ‘Thus will Babylon sink, and will not rise again because of the evil that I will bring on her; and they will be weary.’” Thus far are the words of Jeremiah.}
\end{paracol}
\ChapterHeading{29}
\begin{paracol}{2}
\VersePair{29:1}{Thus saith the Lord against the Philistines;}{47:1}{Yahweh’s word that came to Jeremiah the prophet concerning the Philistines, before Pharaoh struck Gaza.}
\VersePair{29:2}{Behold, waters come up from the north, and shall become a sweeping torrent, and it shall sweep away the land, and its fulness; the city, and them that dwell in it: and men shall cry and all that dwell in the land shall howl,}{47:2}{{\raggedright Yahweh says: \poemline{1}{“Behold, waters rise up out of the north,}\poemline{2}{and will become an overflowing stream,}\poemline{1}{and will overflow the land and all that is therein,}\poemline{2}{the city and those who dwell therein.}\poemline{1}{The men will cry,}\poemline{2}{and all the inhabitants of the land will wail.}}}
\VersePair{29:3}{at the sound of his rushing, at the sound of his hoofs, and at the rattling of his chariots, at the noise of his wheels: the fathers turned not to their children because of the weakness of their hands,}{47:3}{{\raggedright At the noise of the stamping of the hoofs of his strong ones, \poemline{2}{at the rushing of his chariots,}\poemline{2}{at the rumbling of his wheels,}\poemline{1}{the fathers don’t look back for their children}\poemline{2}{because their hands are so feeble,}}}
\VersePair{29:4}{in the day that is coming to destroy all the Philistines: and I will utterly destroy Tyre and Sidon, and all the rest of their allies: for the Lord will destroy the remaining inhabitants of the islands.}{47:4}{{\raggedright because of the day that comes to destroy all the Philistines, \poemline{2}{to cut off from Tyre and Sidon every helper who remains;}\poemline{1}{for Yahweh will destroy the Philistines,}\poemline{2}{the remnant of the isle of Caphtor.}}}
\VersePair{29:5}{Baldness is come upon Gaza; Ascalon is cast away, and the remnant of the Enakim.}{47:5}{{\raggedright Baldness has come on Gaza; \poemline{2}{Ashkelon is brought to nothing.}\poemline{1}{You remnant of their valley,}\poemline{2}{how long will you cut yourself?}}}
\VersePair{29:6}{How long wilt thou smite, O sword of the Lord? how long will it be ere thou art quiet? return into thy sheath, rest, and be removed.}{47:6}{{\raggedright “‘You sword of Yahweh, how long will it be before you are quiet? \poemline{2}{Put yourself back into your scabbard;}\poemline{2}{rest, and be still.’}}}
\VersePair{29:7}{How shall it be quiet, whereas the Lord has given it a commission against Ascalon, and against the regions on the sea-coast, to awake against the remaining countries!}{47:7}{{\raggedright “How can you be quiet, \poemline{2}{since Yahweh has given you a command?}\poemline{1}{Against Ashkelon, and against the seashore,}\poemline{2}{there he has appointed it.”}}}
\end{paracol}
\ChapterHeading{30}
\begin{paracol}{2}
\VersePair{30:1}{Concerning Idumea, thus saith the Lord; There is no longer wisdom in Thæman, counsel has perished from the wise ones, their wisdom is gone,}{49:7}{{\raggedright Of Edom, Yahweh of Armies says: \poemline{1}{“Is wisdom no more in Teman?}\poemline{2}{Has counsel perished from the prudent?}\poemline{2}{Has their wisdom vanished?}}}
\VersePair{30:2}{their place has been deceived. Dig deep for a dwelling, ye that inhabit Dædam, for he has wrought grievously: I brought trouble upon him in the time at which I visited him.}{49:8}{{\raggedright Flee! Turn back! \poemline{2}{Dwell in the depths, inhabitants of Dedan;}\poemline{2}{for I will bring the calamity of Esau on him when I visit him.}}}
\VersePair{30:3}{For grape-gatherers are come, who shall not leave thee a remnant; as thieves by night, they shall lay their hand upon thy possessions.}{49:9}{{\raggedright If grape gatherers came to you, \poemline{2}{would they not leave some gleaning grapes?}\poemline{1}{If thieves came by night,}\poemline{2}{wouldn’t they steal until they had enough?}}}
\VersePair{30:4}{For I have stripped Esau, I have uncovered their secret places; they shall have no power to hide themselves, they have perished each by the hand of his brother, my neighbour, and it is impossible}{49:10}{{\raggedright But I have made Esau bare, \poemline{2}{I have uncovered his secret places,}\poemline{2}{and he will not be able to hide himself.}\poemline{1}{His offspring is destroyed,}\poemline{2}{with his brothers and his neighbors;}\poemline{2}{and he is no more.}}}
\VersePair{30:5}{for thy fatherless one to be left to live, but I shall live, and the widows trust in me.}{49:11}{{\raggedright Leave your fatherless children. \poemline{2}{I will preserve them alive.}\poemline{2}{Let your widows trust in me.”}}}
\VersePair{30:6}{For thus saith the Lord; They who were not appointed to drink the cup have drunk it; and thou shalt by no means be cleared:}{49:12}{For Yahweh says: “Behold, they to whom it didn’t pertain to drink of the cup will certainly drink; and are you he who will altogether go unpunished? You won’t go unpunished, but you will surely drink.}
\VersePair{30:7}{for by myself I have sworn, saith the Lord, that thou shalt be in the midst of her an impassable land, and a reproach, and a curse; and all her cities shall be desert for ever.}{49:13}{For I have sworn by myself,” says Yahweh, “that Bozrah will become an astonishment, a reproach, a waste, and a curse. All its cities will be perpetual wastes.”}
\VersePair{30:8}{I have heard a report from the Lord, and he has sent messengers to the nations, saying, Assemble yourselves, and come against her; rise ye up to war.}{49:14}{{\raggedright I have heard news from Yahweh, \poemline{2}{and an ambassador is sent among the nations,}\poemline{1}{saying, “Gather yourselves together!}\poemline{2}{Come against her!}\poemline{2}{Rise up to the battle!”}}}
\VersePair{30:9}{I have made thee small among the nations, utterly contemptible among men.}{49:15}{{\raggedright “For, behold, I have made you small among the nations, \poemline{2}{and despised among men.}}}
\VersePair{30:10}{Thine insolence has risen up against thee, the fierceness of thine heart has burst the holes of the rocks, it has seized upon the strength of a lofty hill; for as an eagle he set his nest on high: thence will I bring thee down.}{49:16}{{\raggedright As for your terror, \poemline{2}{the pride of your heart has deceived you,}\poemline{1}{O you who dwell in the clefts of the rock,}\poemline{2}{who hold the height of the hill,}\poemline{1}{though you should make your nest as high as the eagle,}\poemline{2}{I will bring you down from there,” says Yahweh.}}}
\VersePair{30:11}{And Idumea shall be a desert: every one that passes by shall hiss at it.}{49:17}{{\raggedright “Edom will become an astonishment. \poemline{2}{Everyone who passes by it will be astonished,}\poemline{2}{and will hiss at all its plagues.}}}
\VersePair{30:12}{As Sodom was overthrown and Gomorrha and they that sojourned in her, saith the Lord Almighty, no man shall dwell there, nor shall any son of man inhabit there.}{49:18}{{\raggedright As in the overthrow of Sodom and Gomorrah and its neighbor cities,” says Yahweh, \poemline{2}{“no man will dwell there,}\poemline{2}{neither will any son of man live therein.}}}
\VersePair{30:13}{Behold, he shall come up as a lion out of the midst of Jordan to the place of Ætham: for I will speedily drive them from it, and do ye set the young men against her: for who is like me? and who will withstand me? and who is this shepherd, who shall confront me?}{49:19}{{\raggedright “Behold, he will come up like a lion from the pride of the Jordan against the strong habitation; \poemline{2}{for I will suddenly make them run away from it,}\poemline{1}{and whoever is chosen,}\poemline{2}{I will appoint him over it.}\poemline{1}{For who is like me?}\poemline{2}{Who will appoint me a time?}\poemline{2}{Who is the shepherd who will stand before me?”}}}
\VersePair{30:14}{Therefore hear ye the counsel of the Lord, which he has framed against Idumea; and his device, which he has devised against the inhabitants of Thæman: surely the least of the sheep shall be swept off; surely their dwelling shall be made desolate for them.}{49:20}{{\raggedright Therefore hear the counsel of Yahweh, that he has taken against Edom, \poemline{2}{and his purposes that he has purposed against the inhabitants of Teman:}\poemline{1}{Surely they will drag them away,}\poemline{2}{the little ones of the flock.}\poemline{2}{Surely he will make their habitation desolate over them.}}}
\VersePair{30:15}{For at the sound of their fall the earth was scared, and the cry of the sea was not heard.}{49:21}{{\raggedright The earth trembles at the noise of their fall; \poemline{2}{there is a cry, the noise which is heard in the Red Sea.}}}
\VersePair{30:16}{Behold, he shall look upon her as an eagle, and spread forth his wings over her strongholds; and the heart of the mighty men of Idumea shall be in that day as the heart of a woman in her pangs.}{49:22}{{\raggedright Behold, he will come up and fly as the eagle, \poemline{2}{and spread out his wings against Bozrah.}\poemline{2}{The heart of the mighty men of Edom at that day will be as the heart of a woman in her pangs.}}}
\VersePair{30:17}{Concerning the sons of Ammon thus saith the Lord, Are there no sons in Israel? or have they no one to succeed them? wherefore has Melchol inherited Galaad, and why shall their people dwell in their cities?}{49:1}{{\raggedright Of the children of Ammon. Yahweh says: \poemline{1}{“Has Israel no sons?}\poemline{2}{Has he no heir?}\poemline{1}{Why then does Malcam possess Gad,}\poemline{2}{and his people dwell in its cities?}}}
\VersePair{30:18}{Therefore, behold, the days come, saith the Lord, when I will cause to be heard in Rabbath a tumult of wars; and they shall become a waste and ruined place, and her altars shall be burned with fire; then shall Israel succeed to his dominion.}{49:2}{{\raggedright Therefore behold, the days come,” \poemline{2}{says Yahweh,}\poemline{1}{“that I will cause an alarm of war to be heard against Rabbah of the children of Ammon,}\poemline{2}{and it will become a desolate heap,}\poemline{2}{and her daughters will be burned with fire;}\poemline{1}{then Israel will possess those who possessed him,”}\poemline{2}{says Yahweh.}}}
\VersePair{30:19}{Howl, O Esebon, for Gai has perished; cry, ye daughters of Rabbath, gird yourselves with sackclothes, and lament; for Melchol shall go into banishment, his priests and his princes together.}{49:3}{{\raggedright “Wail, Heshbon, for Ai is laid waste! \poemline{2}{Cry, you daughters of Rabbah!}\poemline{1}{Clothe yourself in sackcloth.}\poemline{2}{Lament, and run back and forth among the fences;}\poemline{1}{for Malcam will go into captivity,}\poemline{2}{his priests and his princes together.}}}
\VersePair{30:20}{Why do ye exult in the plains of the Enakim, thou haughty daughter, that trustest in thy treasures, that sayest, Who shall come in to me?}{49:4}{{\raggedright Why do you boast in the valleys, \poemline{2}{your flowing valley, backsliding daughter?}\poemline{1}{You trusted in her treasures,}\poemline{2}{saying, ‘Who will come to me?’}}}
\VersePair{30:21}{Behold, I will bring terror upon thee, saith the Lord, from all the country round about thee; and ye shall be scattered every one right before him, and there is none to gather you.}{49:5}{{\raggedright Behold, I will bring a terror on you,” \poemline{2}{says the Lord, Yahweh of Armies,}\poemline{1}{“from all who are around you.}\poemline{2}{All of you will be driven completely out,}\poemline{2}{and there will be no one to gather together the fugitives.}}}
\VersePair{30:22}{}{49:6}{{\raggedright “But afterward I will reverse the captivity of the children of Ammon,” \poemline{2}{says Yahweh.}}}
\VersePair{30:23}{Concerning Kedar the Queen of the Palace, whom Nabuchodonosor King of Babylon smote, thus saith the Lord; Arise ye, and go up to Kedar, and fill the sons of Kedem.}{49:28}{{\raggedright Of Kedar, and of the kingdoms of Hazor, which Nebuchadnezzar king of Babylon struck, Yahweh says: \poemline{1}{“Arise, go up to Kedar,}\poemline{2}{and destroy the children of the east.}}}
\VersePair{30:24}{They shall take their tents and their sheep, they shall take for themselves their garments, and all their baggage and their camels; and summon ye destruction against them from every side.}{49:29}{{\raggedright They will take their tents and their flocks. \poemline{2}{they will carry away for themselves their curtains,}\poemline{2}{all their vessels, and their camels;}\poemline{2}{and they will cry to them, ‘Terror on every side!’}}}
\VersePair{30:25}{Flee ye, dig very deep for a dwelling-place, ye that dwell in the palace; for the king of Babylon has framed a counsel, and devised a device against you.}{49:30}{{\raggedright Flee! \poemline{2}{Wander far off!}\poemline{1}{Dwell in the depths, you inhabitants of Hazor,” says Yahweh;}\poemline{2}{“for Nebuchadnezzar king of Babylon has taken counsel against you,}\poemline{2}{and has conceived a purpose against you.}}}
\VersePair{30:26}{Rise up, and go up against a nation settled and dwelling at ease, who have no doors, nor bolts, nor bars, who dwell alone.}{49:31}{{\raggedright Arise! Go up to a nation that is at ease, \poemline{2}{that dwells without care,” says Yahweh;}\poemline{2}{“that has neither gates nor bars,}\poemline{2}{that dwells alone.}}}
\VersePair{30:27}{And their camels shall be a spoil, and the multitude of their cattle shall be destroyed: and I will scatter them as chaff with every wind, having their hair cut about their foreheads, I will bring on their overthrow from all sides, saith the Lord.}{49:32}{{\raggedright Their camels will be a booty, \poemline{2}{and the multitude of their livestock a plunder.}\poemline{1}{I will scatter to all winds those who have the corners of their beards cut off;}\poemline{2}{and I will bring their calamity from every side of them,”}\poemline{2}{says Yahweh.}}}
\VersePair{30:28}{And the palace shall be a resting-place for ostriches, and desolate for ever: no man shall abide there, and no son of man shall dwell there.}{49:33}{{\raggedright Hazor will be a dwelling place of jackals, \poemline{2}{a desolation forever.}\poemline{1}{No man will dwell there,}\poemline{2}{neither will any son of man live therein.”}}}
\VersePair{30:29}{Concerning Damascus. Emath is brought to shame, and Arphath: for they have heard an evil report: they are amazed, they are angry, they shall be utterly unable to rest.}{49:23}{{\raggedright Of Damascus: \poemline{1}{“Hamath and Arpad are confounded,}\poemline{2}{for they have heard evil news.}\poemline{2}{They have melted away.}\poemline{1}{There is sorrow on the sea.}\poemline{2}{It can’t be quiet.}}}
\VersePair{30:30}{Damascus is utterly weakened, she is put to flight; trembling has seized upon her.}{49:24}{{\raggedright Damascus has grown feeble, \poemline{2}{she turns herself to flee,}\poemline{2}{and trembling has seized her.}\poemline{1}{Anguish and sorrows have taken hold of her,}\poemline{2}{as of a woman in travail.}}}
\VersePair{30:31}{How has she not left my city, they have loved the village?}{49:25}{{\raggedright How is the city of praise not forsaken, \poemline{2}{the city of my joy?}}}
\VersePair{30:32}{Therefore shall the young men fall in thy streets, and all thy warriors shall fall, saith the Lord.}{49:26}{{\raggedright Therefore her young men will fall in her streets, \poemline{2}{and all the men of war will be brought to silence in that day,”}\poemline{2}{says Yahweh of Armies.}}}
\VersePair{30:33}{And I will kindle a fire in the wall of Damascus, and it shall devour the streets of the son of Ader.}{49:27}{{\raggedright “I will kindle a fire in the wall of Damascus, \poemline{2}{and it will devour the palaces of Ben Hadad.”}}}
\end{paracol}
\ChapterHeading{31}
\begin{paracol}{2}
\VersePair{31:1}{Thus has the Lord said concerning Moab, Woe to Nabau! for it has perished: Cariathaim is taken: Amath and Agath are put to shame.}{48:1}{{\raggedright Of Moab. Yahweh of Armies, the God of Israel, says: \poemline{1}{“Woe to Nebo!}\poemline{2}{For it is laid waste.}\poemline{1}{Kiriathaim is disappointed.}\poemline{2}{It is taken.}\poemline{1}{Misgab is put to shame}\poemline{2}{and broken down. \footnote{48:1: or, The stronghold}}}}
\VersePair{31:2}{There is no longer any healing for Moab, nor glorying in Esebon: he has devised evils against her: we have cut her off from being a nation, and she shall be completely still: after thee shall go a sword;}{48:2}{{\raggedright The praise of Moab is no more. \poemline{2}{In Heshbon they have devised evil against her:}\poemline{2}{‘Come! Let’s cut her off from being a nation.’}\poemline{1}{You also, Madmen, will be brought to silence.}\poemline{2}{The sword will pursue you.}}}
\VersePair{31:3}{for there is a voice of men crying out of Oronaim, destruction and great ruin.}{48:3}{{\raggedright The sound of a cry from Horonaim, \poemline{2}{desolation and great destruction!}}}
\VersePair{31:4}{Moab is ruined, proclaim it to Zogora:}{48:4}{{\raggedright Moab is destroyed. \poemline{2}{Her little ones have caused a cry to be heard.}}}
\VersePair{31:5}{for Aloth is filled with weeping: one shall go up weeping by the way of Oronaim; ye have heard a cry of destruction.}{48:5}{{\raggedright For they will go up by the ascent of Luhith with continual weeping. \poemline{2}{For at the descent of Horonaim they have heard the distress of the cry of destruction.}}}
\VersePair{31:6}{Flee ye, and save your lives, and ye shall be as a wild ass in the desert.}{48:6}{{\raggedright Flee! Save your lives! \poemline{2}{Be like the juniper bush in the wilderness.}}}
\VersePair{31:7}{Since thou hast trusted in thy strong-hold, therefore thou shalt be taken: and Chamos shall go forth into captivity, and his priests, and his princes together.}{48:7}{{\raggedright For, because you have trusted in your works and in your treasures, \poemline{2}{you also will be taken.}\poemline{1}{Chemosh will go out into captivity,}\poemline{2}{his priests and his princes together.}}}
\VersePair{31:8}{And destruction shall come upon every city, it shall by no means escape; the valley also shall perish, and the plain country shall be completely destroyed, as the Lord has said.}{48:8}{{\raggedright The destroyer will come on every city, \poemline{2}{and no city will escape;}\poemline{1}{the valley also will perish,}\poemline{2}{and the plain will be destroyed, as Yahweh has spoken.}}}
\VersePair{31:9}{Set marks upon Moab, for she shall be touched with a plague-spot, and all her cities shall become desolate; whence shall there be an inhabitant for her?}{48:9}{{\raggedright Give wings to Moab, \poemline{2}{that she may fly and get herself away:}\poemline{1}{and her cities will become a desolation,}\poemline{2}{without anyone to dwell in them.}}}
\VersePair{31:10}{Cursed is the man that does the works of the Lord carelessly, keeping back his sword from blood.}{48:10}{{\raggedright “Cursed is he who does the work of Yahweh negligently; \poemline{2}{and cursed is he who keeps back his sword from blood.}}}
\VersePair{31:11}{Moab has been at ease from a child, and trusted in his glory; he has not poured out his liquor from vessel to vessel, and has not gone into banishment, therefore his taste remained in him, and his smell departed not.}{48:11}{{\raggedright “Moab has been at ease from his youth, \poemline{2}{and he has settled on his dregs,}\poemline{1}{and has not been emptied from vessel to vessel,}\poemline{2}{neither has he gone into captivity;}\poemline{1}{therefore his taste remains in him,}\poemline{2}{and his scent is not changed.}}}
\VersePair{31:12}{Therefore, behold, his days come, saith the Lord, when I shall send upon him bad leaders, and they shall lead him astray, and they shall utterly break in pieces his possessions, and shall cut his horns asunder.}{48:12}{{\raggedright Therefore behold, the days come,” says Yahweh, \poemline{2}{“that I will send to him those who pour off,}\poemline{1}{and they will pour him off;}\poemline{2}{and they will empty his vessels,}\poemline{2}{and break their containers in pieces.}}}
\VersePair{31:13}{And Moab shall be ashamed of Chamos, as the house of Israel was ashamed of Bæthel their hope, having trusted in them.}{48:13}{{\raggedright Moab will be ashamed of Chemosh, \poemline{2}{as the house of Israel was ashamed of Bethel, their confidence.}}}
\VersePair{31:14}{How will ye say, We are strong, and men strong for war?}{48:14}{{\raggedright “How do you say, ‘We are mighty men, \poemline{2}{and valiant men for the war ’?}}}
\VersePair{31:15}{Moab is ruined, even his city, and his choice young men have gone down to slaughter.}{48:15}{{\raggedright Moab is laid waste, \poemline{2}{and they have gone up into his cities,}\poemline{2}{and his chosen young men have gone down to the slaughter,”}\poemline{2}{says the King, whose name is Yahweh of Armies.}}}
\VersePair{31:16}{The day of Moab is near at hand, and his iniquity moves swiftly to vengeance.}{48:16}{{\raggedright “The calamity of Moab is near to come, \poemline{2}{and his affliction hurries fast.}}}
\VersePair{31:17}{Shake the head at him, all ye that are round about him; all of you utter his name; say ye, How is the glorious staff broken to pieces, the rod of magnificence!}{48:17}{{\raggedright All you who are around him, bemoan him; \poemline{2}{and all you who know his name, say,}\poemline{1}{‘How the strong staff is broken,}\poemline{2}{the beautiful rod!’}}}
\VersePair{31:18}{Come down from thy glory, and sit down in a damp place: Dæbon shall be broken, because Moab is destroyed: there has gone up against thee one to ravage thy strong-hold.}{48:18}{{\raggedright “You daughter who dwells in Dibon, \poemline{2}{come down from your glory,}\poemline{2}{and sit in thirst;}\poemline{1}{for the destroyer of Moab has come up against you.}\poemline{2}{He has destroyed your strongholds.}}}
\VersePair{31:19}{Stand by the way, and look, thou that dwellest in Arer; and ask him that is fleeing, and him that escapes, and say, What has happened?}{48:19}{{\raggedright Inhabitant of Aroer, stand by the way and watch. \poemline{2}{Ask him who flees, and her who escapes;}\poemline{2}{say, ‘What has been done?’}}}
\VersePair{31:20}{Moab is put to shame, because he is broken: howl and cry; proclaim in Arnon, that Moab has perished.}{48:20}{{\raggedright Moab is disappointed; \poemline{2}{for it is broken down.}\poemline{1}{Wail and cry!}\poemline{2}{Tell it by the Arnon, that Moab is laid waste.}}}
\VersePair{31:21}{And judgment is coming against the land of Misor, upon Chelon, and Rephas, and Mophas,}{48:21}{{\raggedright Judgment has come on the plain country— \poemline{2}{on Holon, on Jahzah, on Mephaath,}}}
\VersePair{31:22}{and upon Dæbon, and upon Nabau, and upon the house of Dæthlathaim,}{48:22}{on Dibon, on Nebo, on Beth Diblathaim,}
\VersePair{31:23}{and upon Cariathaim, and upon the house of Gæmol, and upon the house of Maon,}{48:23}{on Kiriathaim, on Beth Gamul, on Beth Meon,}
\VersePair{31:24}{and upon Carioth, and upon Bosor, and upon all the cities of Moab, far and near.}{48:24}{{\raggedright on Kerioth, on Bozrah, \poemline{2}{and on all the cities of the land of Moab, far or near.}}}
\VersePair{31:25}{The horn of Moab is broken, and his arm is crushed.}{48:25}{{\raggedright The horn of Moab is cut off, \poemline{2}{and his arm is broken,” says Yahweh.}}}
\VersePair{31:26}{Make ye him drunk; for he has magnified himself against the Lord: and Moab shall clap with his hand, and shall be also himself a laughing-stock.}{48:26}{{\raggedright “Make him drunk, \poemline{2}{for he magnified himself against Yahweh.}\poemline{1}{Moab will wallow in his vomit,}\poemline{2}{and he also will be in derision.}}}
\VersePair{31:27}{For surely Israel was to thee a laughing-stock, and was found among thy thefts, because thou didst fight against him.}{48:27}{{\raggedright For wasn’t Israel a derision to you? \poemline{2}{Was he found among thieves?}\poemline{1}{For as often as you speak of him,}\poemline{2}{you shake your head.}}}
\VersePair{31:28}{The inhabitants of Moab have left the cities, and dwelt in rocks; they have become as doves nestling in rocks, at the mouth of a cave.}{48:28}{{\raggedright You inhabitants of Moab, leave the cities, and dwell in the rock. \poemline{2}{Be like the dove that makes her nest over the mouth of the abyss.}}}
\VersePair{31:29}{And I have heard of the pride of Moab, he has greatly heightened his pride and his haughtiness, and his heart has been lifted up.}{48:29}{{\raggedright “We have heard of the pride of Moab. \poemline{2}{He is very proud in his loftiness, his pride,}\poemline{2}{his arrogance, and the arrogance of his heart.}}}
\VersePair{31:30}{But I know his works: is it not enough for him? has he not done thus?}{48:30}{{\raggedright I know his wrath,” says Yahweh, “that it is nothing; \poemline{2}{his boastings have done nothing.}}}
\VersePair{31:31}{Therefore howl ye for Moab on all sides; cry out against the shorn men in a gloomy place. I will weep for thee,}{48:31}{{\raggedright Therefore I will wail for Moab. \poemline{2}{Yes, I will cry out for all Moab.}\poemline{2}{They will mourn for the men of Kir Heres.}}}
\VersePair{31:32}{O vine of Aserema, as with the weeping of Jazer: thy branches are gone over the sea, they reached the cities of Jazer: destruction has come upon thy fruits, and upon thy grape-gatherers.}{48:32}{{\raggedright With more than the weeping of Jazer \poemline{2}{I will weep for you, vine of Sibmah.}\poemline{1}{Your branches passed over the sea.}\poemline{2}{They reached even to the sea of Jazer.}\poemline{1}{The destroyer has fallen on your summer fruits}\poemline{2}{and on your vintage.}}}
\VersePair{31:33}{Joy and gladness have been utterly swept off the land of Moab: and though there was wine in thy presses, in the morning they trod it not, neither in the evening did they raise the cry of joy.}{48:33}{{\raggedright Gladness and joy is taken away from the fruitful field \poemline{2}{and from the land of Moab.}\poemline{1}{I have caused wine to cease from the wine presses.}\poemline{2}{No one will tread with shouting.}\poemline{2}{The shouting will be no shouting.}}}
\VersePair{31:34}{From the cry of Esebon even to Ætam their cities uttered their voice, from Zogor to Oronaim, and their tidings as a heifer of three years old, for the water also of Nebrin shall be dried up.}{48:34}{{\raggedright From the cry of Heshbon even to Elealeh, \poemline{2}{even to Jahaz they have uttered their voice,}\poemline{2}{from Zoar even to Horonaim, to Eglath Shelishiyah;}\poemline{2}{for the waters of Nimrim will also become desolate.}}}
\VersePair{31:35}{And I will destroy Moab, saith the Lord, as he comes up to the altar, and burns incense to his gods.}{48:35}{{\raggedright Moreover I will cause to cease in Moab,” says Yahweh, \poemline{2}{“him who offers in the high place,}\poemline{2}{and him who burns incense to his gods.}}}
\VersePair{31:36}{Therefore the heart of Moab shall sound as pipes, my heart shall sound as a pipe for the shorn men; forasmuch as what every man has gained has perished from him.}{48:36}{{\raggedright Therefore my heart sounds for Moab like flutes, \poemline{2}{and my heart sounds like flutes for the men of Kir Heres.}\poemline{2}{Therefore the abundance that he has gotten has perished.}}}
\VersePair{31:37}{They shall all have their heads shaved in every place, and every beard shall be shaved; and all hands shall beat the breasts, and on all loins shall be sackcloth.}{48:37}{{\raggedright For every head is bald, \poemline{2}{and every beard clipped.}\poemline{1}{There are cuttings on all the hands,}\poemline{2}{and sackcloth on the waist.}}}
\VersePair{31:38}{And on all the housetops of Moab, and in his streets shall be mourning: for I have broken him, saith the Lord, as a vessel, which is useless.}{48:38}{{\raggedright On all the housetops of Moab, \poemline{2}{and in its streets, there is lamentation everywhere;}\poemline{2}{for I have broken Moab like a vessel in which no one delights,” says Yahweh.}}}
\VersePair{31:39}{How has he changed! how has Moab turned his back! Moab is put to shame, and become a laughing-stock, and an object of anger to all that are round about him.}{48:39}{{\raggedright “How it is broken down! \poemline{2}{How they wail!}\poemline{1}{How Moab has turned the back with shame!}\poemline{2}{So will Moab become a derision}\poemline{2}{and a terror to all who are around him.”}}}
\VersePair{31:40}{For thus said the Lord;}{48:40}{{\raggedright For Yahweh says: “Behold, he will fly as an eagle, \poemline{2}{and will spread out his wings against Moab.}}}
\VersePair{31:41}{Carioth is taken, and the strong-holds have been taken together.}{48:41}{{\raggedright Kerioth is taken, \poemline{2}{and the strongholds are seized.}\poemline{1}{The heart of the mighty men of Moab at that day}\poemline{2}{will be as the heart of a woman in her pangs.}}}
\VersePair{31:42}{And Moab shall perish from being a multitude, because he has magnified himself against the Lord.}{48:42}{{\raggedright Moab will be destroyed from being a people, \poemline{2}{because he has magnified himself against Yahweh.}}}
\VersePair{31:43}{A snare, and fear, and the pit, are upon thee, O inhabitant of Moab.}{48:43}{{\raggedright Terror, the pit, and the snare are on you, \poemline{2}{inhabitant of Moab,” says Yahweh.}}}
\VersePair{31:44}{He that flees from the terror shall fall into the pit, and he that comes up out of the pit shall even be taken in the snare: for I will bring these things upon Moab in the year of their visitation.}{48:44}{{\raggedright “He who flees from the terror will fall into the pit; \poemline{2}{and he who gets up out of the pit will be taken in the snare,}\poemline{1}{for I will bring on him, even on Moab,}\poemline{2}{the year of their visitation,” says Yahweh.}}}
\VersePair{31:45}{}{48:45}{{\raggedright “Those who fled stand without strength under the shadow of Heshbon; \poemline{2}{for a fire has gone out of Heshbon,}\poemline{2}{and a flame from the middle of Sihon,}\poemline{1}{and has devoured the corner of Moab,}\poemline{2}{and the crown of the head of the tumultuous ones.}}}
\VersePair{31:46}{}{48:46}{{\raggedright Woe to you, O Moab! \poemline{2}{The people of Chemosh are undone;}\poemline{1}{for your sons are taken away captive,}\poemline{2}{and your daughters into captivity.}}}
\end{paracol}
\ChapterHeading{32}
\begin{paracol}{2}
\VersePair{32:15}{Thus said the Lord God of Israel; Take the cup of this unmixed wine from mine hand, and thou shalt cause all the nations to drink, to whom I send thee.}{25:15}{For Yahweh, the God of Israel, says to me: “Take this cup of the wine of wrath from my hand, and cause all the nations to whom I send you to drink it.}
\VersePair{32:16}{And they shall drink, and vomit, and be mad, because of the sword which I send among them.}{25:16}{They will drink, and reel back and forth, and be insane, because of the sword that I will send among them.”}
\VersePair{32:17}{So I took the cup out of the Lord's hand, and caused the nations to whom the Lord sent me to drink:}{25:17}{Then I took the cup at Yahweh’s hand, and made all the nations to drink, to whom Yahweh had sent me:}
\VersePair{32:18}{Jerusalem, and the cities of Juda, and the kings of Juda, and his princes, to make them a desert place, a desolation, and a hissing;}{25:18}{Jerusalem, and the cities of Judah, with its kings and its princes, to make them a desolation, an astonishment, a hissing, and a curse, as it is today;}
\VersePair{32:19}{and Pharao king of Egypt, and his servants, and his nobles, and all his people;}{25:19}{Pharaoh king of Egypt, with his servants, his princes, and all his people;}
\VersePair{32:20}{and all the mingled people, and all the kings of the Philistines, and Ascalon, and Gaza, and Accaron, and the remnant of Azotus,}{25:20}{and all the mixed people, and all the kings of the land of Uz, all the kings of the Philistines, Ashkelon, Gaza, Ekron, and the remnant of Ashdod;}
\VersePair{32:21}{and Idumea, and the land of Moab, and the children of Ammon,}{25:21}{Edom, Moab, and the children of Ammon;}
\VersePair{32:22}{and the kings of Tyre, and the kings of Sidon, and the kings in the country beyond the sea,}{25:22}{and all the kings of Tyre, all the kings of Sidon, and the kings of the isle which is beyond the sea;}
\VersePair{32:23}{and Dædan, and Thæman, and Ros, and every one that is shaved round about the face,}{25:23}{Dedan, Tema, Buz, and all who have the corners of their beard cut off;}
\VersePair{32:24}{and all the mingled people lodging in the wilderness,}{25:24}{and all the kings of Arabia, all the kings of the mixed people who dwell in the wilderness;}
\VersePair{32:25}{and all the kings of Ælam, and all the kings of the Persians,}{25:25}{and all the kings of Zimri, all the kings of Elam, and all the kings of the Medes;}
\VersePair{32:26}{and all the kings from the north, the far and the near, each one with his brother, and all the kingdoms which are on the face of the earth.}{25:26}{and all the kings of the north, far and near, one with another; and all the kingdoms of the world, which are on the surface of the earth. The king of Sheshach will drink after them.}
\VersePair{32:27}{And thou shalt say to them, Thus said the Lord Almighty; Drink ye, be ye drunken; and ye shall vomit, and shall fall, and shall in nowise rise, because of the sword which I send among you.}{25:27}{“You shall tell them, ‘Yahweh of Armies, the God of Israel says: “Drink, and be drunk, vomit, fall, and rise no more, because of the sword which I will send among you.”’}
\VersePair{32:28}{And it shall come to pass, when they refuse to take the cup out of thine hand, to drink it, that thou shalt say, Thus said the Lord; Ye shall surely drink.}{25:28}{It shall be, if they refuse to take the cup at your hand to drink, then you shall tell them, ‘Yahweh of Armies says: “You shall surely drink.}
\VersePair{32:29}{For I am beginning to afflict the city whereon my name is called, and ye shall by no means be held guiltless: for I am calling a sword upon all that dwell upon the earth.}{25:29}{For, behold, I begin to work evil at the city which is called by my name; and should you be utterly unpunished? You will not be unpunished; for I will call for a sword on all the inhabitants of the earth, says Yahweh of Armies.”’}
\VersePair{32:30}{And thou shalt prophesy against them these words, and shalt say, The Lord shall speak from on high, from his sanctuary he will utter his voice; he will pronounce a declaration on his place; and these shall answer like men gathering grapes: and destruction is coming on them that dwell on the earth,}{25:30}{{\raggedright “Therefore prophesy against them all these words, and tell them, \poemline{1}{“‘Yahweh will roar from on high,}\poemline{2}{and utter his voice from his holy habitation.}\poemline{2}{He will mightily roar against his fold.}\poemline{1}{He will give a shout, as those who tread grapes,}\poemline{2}{against all the inhabitants of the earth.}}}
\VersePair{32:31}{even upon the extreme part of the earth; for the Lord has a controversy with the nations, he is pleading with all flesh, and the ungodly are given to the sword, saith the Lord.}{25:31}{{\raggedright A noise will come even to the end of the earth; \poemline{2}{for Yahweh has a controversy with the nations.}\poemline{1}{He will enter into judgment with all flesh.}\poemline{2}{As for the wicked, he will give them to the sword,”’ says Yahweh.”}}}
\VersePair{32:32}{Thus said the Lord; Behold, evils are proceeding from nation to nation, and a great whirlwind goes forth from the end of the earth.}{25:32}{{\raggedright Yahweh of Armies says, \poemline{1}{“Behold, evil will go out from nation to nation,}\poemline{2}{and a great storm will be raised up from the uttermost parts of the earth.”}}}
\VersePair{32:33}{And the slain of the Lord shall be in the day of the Lord from one end of the earth even to the other end of the earth: they shall not be buried; they shall be as dung on the face of the earth.}{25:33}{The slain of Yahweh will be at that day from one end of the earth even to the other end of the earth. They won’t be lamented. They won’t be gathered or buried. They will be dung on the surface of the ground.}
\VersePair{32:34}{Howl, ye shepherds, and cry; and lament, ye rams of the flock: for your days have been completed for slaughter, and ye shall fall as the choice rams.}{25:34}{{\raggedright Wail, you shepherds, and cry. \poemline{2}{Wallow in dust, you leader of the flock;}\poemline{1}{for the days of your slaughter and of your dispersions have fully come,}\poemline{2}{and you will fall like fine pottery.}}}
\VersePair{32:35}{And flight shall perish from the shepherds, and safety from the rams of the flock.}{25:35}{{\raggedright The shepherds will have no way to flee. \poemline{2}{The leader of the flock will have no escape.}}}
\VersePair{32:36}{A voice of the crying of the shepherds, and a moaning of the sheep and the rams: for the Lord has destroyed their pastures.}{25:36}{{\raggedright A voice of the cry of the shepherds, \poemline{2}{and the wailing of the leader of the flock,}\poemline{2}{for Yahweh destroys their pasture.}}}
\VersePair{32:37}{And the peaceable abodes that remain shall be destroyed before the fierceness of my anger.}{25:37}{{\raggedright The peaceful folds are brought to silence \poemline{2}{because of the fierce anger of Yahweh.}}}
\VersePair{32:38}{He has forsaken his lair, as a lion: for their land is become desolate before the great sword.}{25:38}{{\raggedright He has left his covert, as the lion; \poemline{2}{for their land has become an astonishment because of the fierceness of the oppression,}\poemline{2}{and because of his fierce anger.}}}
\end{paracol}
\ChapterHeading{33}
\begin{paracol}{2}
\VersePair{33:1}{In the beginning of the Reign of King Joakim Son of Josias there came this Word from the Lord.}{26:1}{In the beginning of the reign of Jehoiakim the son of Josiah, king of Judah, this word came from Yahweh:}
\VersePair{33:2}{Thus said the Lord; Stand in the court of the Lord's house, and thou shalt declare to all the Jews, and to all that come to worship in the house of the Lord, all the words which I commanded thee to speak to them; abate not one word.}{26:2}{“Yahweh says: ‘Stand in the court of Yahweh’s house, and speak to all the cities of Judah which come to worship in Yahweh’s house, all the words that I command you to speak to them. Don’t omit a word.}
\VersePair{33:3}{Peradventure they will hear, and turn every one from his evil way: then I will cease from the evils which I purpose to do to them, because of their evil practices.}{26:3}{It may be they will listen, and every man turn from his evil way, that I may relent from the evil which I intend to do to them because of the evil of their doings.’”}
\VersePair{33:4}{And thou shalt say, Thus said the Lord; If ye will not hearken to me, to walk in my statutes which I set before you,}{26:4}{You shall tell them, “Yahweh says: ‘If you will not listen to me, to walk in my law which I have set before you,}
\VersePair{33:5}{to hearken to the words of my servants the prophets, whom I send to you early in the morning; yea, I sent them, but ye hearkened not to me;}{26:5}{to listen to the words of my servants the prophets whom I send to you, even rising up early and sending them — but you have not listened —}
\VersePair{33:6}{then will I make this house as Selo, and I will make this city a curse to all the nations of all the earth.}{26:6}{then I will make this house like Shiloh, and will make this city a curse to all the nations of the earth.’”}
\VersePair{33:7}{And the priests, and the false prophets, and all the people heard Jeremias speaking these words in the house of the Lord.}{26:7}{The priests and the prophets and all the people heard Jeremiah speaking these words in Yahweh’s house.}
\VersePair{33:8}{And it came to pass, when Jeremias had ceased speaking all that the Lord had ordered him to speak to all the people, that the priests and the false prophets and all the people took him, saying,}{26:8}{When Jeremiah had finished speaking all that Yahweh had commanded him to speak to all the people, the priests and the prophets and all the people seized him, saying, “You shall surely die!}
\VersePair{33:9}{Thou shalt surely die, because thou hast prophesied in the name of the Lord, saying, This house shall be as Selo, and this city shall be made quite destitute of inhabitants. And all the people assembled against Jeremias in the house of the Lord.}{26:9}{Why have you prophesied in Yahweh’s name, saying, ‘This house will be like Shiloh, and this city will be desolate, without inhabitant ’?” All the people were crowded around Jeremiah in Yahweh’s house.}
\VersePair{33:10}{And the princes of Juda heard this word, and they went up out of the house of the king to the house of the Lord, and sat in the entrance of the new gate.}{26:10}{When the princes of Judah heard these things, they came up from the king’s house to Yahweh’s house; and they sat in the entry of the new gate of Yahweh’s house.}
\VersePair{33:11}{Then the priests and the false prophets said to the princes and to all the people, The judgment of death is due to this man; because he has prophesied against this city, as ye have heard with your ears.}{26:11}{Then the priests and the prophets spoke to the princes and to all the people, saying, “This man is worthy of death, for he has prophesied against this city, as you have heard with your ears.”}
\VersePair{33:12}{Then Jeremias spoke to the princes, and to all the people, saying, The Lord sent me to prophesy against this house and against this city, all the words which ye have heard.}{26:12}{Then Jeremiah spoke to all the princes and to all the people, saying, “Yahweh sent me to prophesy against this house and against this city all the words that you have heard.}
\VersePair{33:13}{And now amend your ways and your works, and hearken to the voice of the Lord; and the Lord shall cease from the evils which he has pronounced against you.}{26:13}{Now therefore amend your ways and your doings, and obey Yahweh your God’s voice; then Yahweh will relent from the evil that he has pronounced against you.}
\VersePair{33:14}{And behold, I am in your hands; do to me as is expedient, and as it is best for you.}{26:14}{But as for me, behold, I am in your hand. Do with me what is good and right in your eyes.}
\VersePair{33:15}{But know for a certainty, that if ye slay me, ye bring innocent blood upon yourselves, and upon this city, and upon them that dwell in it; for in truth the Lord has sent me to you to speak in your ears all these words.}{26:15}{Only know for certain that if you put me to death, you will bring innocent blood on yourselves, on this city, and on its inhabitants; for in truth Yahweh has sent me to you to speak all these words in your ears.”}
\VersePair{33:16}{Then the princes and all the people said to the priests and to the false prophets; Judgment of death is not due to this man; for he has spoken to us in the name of the Lord our God.}{26:16}{Then the princes and all the people said to the priests and to the prophets: “This man is not worthy of death; for he has spoken to us in the name of Yahweh our God.”}
\VersePair{33:17}{And there rose up men of the elders of the land, and said to all the assembly of the people,}{26:17}{Then certain of the elders of the land rose up, and spoke to all the assembly of the people, saying,}
\VersePair{33:18}{Michæas the Morathite lived in the days of Ezekias king of Juda, and said to all the people of Juda, Thus saith the Lord; Sion shall be ploughed as a field, and Jerusalem shall become a desolation, and the mountain of the house shall be a thicket of trees.}{26:18}{{\raggedright “Micah the Morashtite prophesied in the days of Hezekiah king of Judah; and he spoke to all the people of Judah, saying, ‘Yahweh of Armies says: \poemline{1}{“‘Zion will be plowed as a field,}\poemline{2}{and Jerusalem will become heaps,}\poemline{2}{and the mountain of the house as the high places of a forest.’}}}
\VersePair{33:19}{Did Ezekias and all Juda in any way slay him? Was it not that they feared the Lord, and they made supplication before the Lord, and the Lord ceased from the evils which he had pronounced against them? whereas we have wrought great evil against our own souls.}{26:19}{Did Hezekiah king of Judah and all Judah put him to death? Didn’t he fear Yahweh, and entreat the favor of Yahweh, and Yahweh relented of the disaster which he had pronounced against them? We would commit great evil against our own souls that way!”}
\VersePair{33:20}{And there was another man prophesying in the name of the Lord, Urias the son of Samæas of Cariathiarim; and he prophesied concerning this land according to all the words of Jeremias.}{26:20}{There was also a man who prophesied in Yahweh’s name, Uriah the son of Shemaiah of Kiriath Jearim; and he prophesied against this city and against this land according to all the words of Jeremiah.}
\VersePair{33:21}{And king Joakim and all the princes heard all his words, and sought to slay him; and Urias heard it and went into Egypt.}{26:21}{When Jehoiakim the king, with all his mighty men and all the princes heard his words, the king sought to put him to death; but when Uriah heard it, he was afraid, and fled, and went into Egypt.}
\VersePair{33:22}{And the king sent men into Egypt;}{26:22}{Then Jehoiakim the king sent Elnathan the son of Achbor and certain men with him into Egypt.}
\VersePair{33:23}{and they brought him thence, and brought him in to the king; and he smote him with the sword, and cast him into the sepulchre of the children of his people.}{26:23}{They fetched Uriah out of Egypt and brought him to Jehoiakim the king, who killed him with the sword and cast his dead body into the graves of the common people.}
\VersePair{33:24}{Nevertheless the hand of Achicam son of Saphan was with Jeremias, to prevent his being delivered into the hands of the people, or being killed.}{26:24}{But the hand of Ahikam the son of Shaphan was with Jeremiah, so that they didn’t give him into the hand of the people to put him to death.}
\end{paracol}
\ChapterHeading{34}
\begin{paracol}{2}
\VersePair{34:1}{}{27:1}{In the beginning of the reign of Jehoiakim the son of Josiah, king of Judah, this word came to Jeremiah from Yahweh, saying,}
\VersePair{34:2}{Thus said the Lord; Make to thyself bonds and yokes, and put them about thy neck,}{27:2}{Yahweh says to me: “Make bonds and bars, and put them on your neck.}
\VersePair{34:3}{and thou shalt send them to the king of Idumea, and to the king of Moab, and to the king of the children of Ammon, and to the king of Tyre, and to the king of Sidon, by the hands of their messengers that come to meet them at Jerusalem to Sedekias king of Juda.}{27:3}{Then send them to the king of Edom, to the king of Moab, to the king of the children of Ammon, to the king of Tyre, and to the king of Sidon, by the hand of the messengers who come to Jerusalem to Zedekiah king of Judah.}
\VersePair{34:4}{And thou shalt commission them to say to their lords, Thus said the Lord God of Israel; Thus shall ye say to your lords;}{27:4}{Give them a command to their masters, saying, ‘Yahweh of Armies, the God of Israel says, “You shall tell your masters:}
\VersePair{34:5}{I have made the earth by my great power, and with my high arm, and I will give it to whomsoever it shall seem good in mine eyes.}{27:5}{‘I have made the earth, the men, and the animals that are on the surface of the earth by my great power and by my outstretched arm. I give it to whom it seems right to me.}
\VersePair{34:6}{I gave the earth to Nabuchodonosor king of Babylon to serve him, and the wild beasts of the field to labour for him.}{27:6}{Now I have given all these lands into the hand of Nebuchadnezzar the king of Babylon, my servant. I have also given the animals of the field to him to serve him.}
\VersePair{34:8}{And the nation and kingdom, all that shall not put their neck under the yoke of the king of Babylon, with sword and famine will I visit them, saith the Lord, until they are consumed by his hand.}{27:8}{“‘“‘It will happen that I will punish the nation and the kingdom which will not serve the same Nebuchadnezzar king of Babylon, and that will not put their neck under the yoke of the king of Babylon,’ says Yahweh, ‘with the sword, with famine, and with pestilence, until I have consumed them by his hand.}
\VersePair{34:9}{And hearken ye not to your false prophets, nor to them that divine to you, nor to them that foretel events by dreams to you, nor to your auguries, nor your sorcerers, that say, Ye shall by no means work for the king of Babylon:}{27:9}{But as for you, don’t listen to your prophets, to your diviners, to your dreams, to your soothsayers, or to your sorcerers, who speak to you, saying, “You shall not serve the king of Babylon;”}
\VersePair{34:10}{for they prophesy lies to you, to remove you far from your land.}{27:10}{for they prophesy a lie to you, to remove you far from your land, so that I would drive you out, and you would perish.}
\VersePair{34:11}{But the nation which shall put its neck under the yoke of the king of Babylon, and serve him, I will even leave it upon its land, and it shall serve him, and dwell in it.}{27:11}{But the nation that brings their neck under the yoke of the king of Babylon and serves him, that nation I will let remain in their own land,’ says Yahweh; ‘and they will till it and dwell in it.’”’”}
\VersePair{34:12}{I spoke also to Sedekias king of Juda according to all these words, saying, Put your neck into the yoke,}{27:12}{I spoke to Zedekiah king of Judah according to all these words, saying, “Bring your necks under the yoke of the king of Babylon, and serve him and his people, and live.}
\VersePair{34:14}{and serve the king of Babylon. For they prophesy unrighteous words to you,}{27:14}{Don’t listen to the words of the prophets who speak to you, saying, ‘You shall not serve the king of Babylon;’ for they prophesy a lie to you.}
\VersePair{34:15}{for I sent them not, saith the Lord; and they prophesy in my name unjustly, that I might destroy you, and ye should perish, and your prophets, who unrighteously prophesy lies to you.}{27:15}{For I have not sent them,” says Yahweh, “but they prophesy falsely in my name; that I may drive you out, and that you may perish, you, and the prophets who prophesy to you.”}
\VersePair{34:16}{I spoke to you, and to all this people, and to the priests, saying, Thus said the Lord; Hearken not to the words of the prophets that prophesy to you, saying, Behold, the vessels of the Lord's house shall return from Babylon: for they prophesy to you unrighteous words. I sent them not.}{27:16}{Also I spoke to the priests and to all this people, saying, Yahweh says, “Don’t listen to the words of your prophets who prophesy to you, saying, ‘Behold, the vessels of Yahweh’s house will now shortly be brought again from Babylon;’ for they prophesy a lie to you.}
\VersePair{34:18}{If they are prophets, and if the word of the Lord is in them, let them meet me, for thus has the Lord said.}{27:18}{But if they are prophets, and if Yahweh’s word is with them, let them now make intercession to Yahweh of Armies, that the vessels which are left in Yahweh’s house, in the house of the king of Judah, and at Jerusalem, don’t go to Babylon.}
\VersePair{34:19}{And as for the remaining vessels,}{27:19}{For Yahweh of Armies says concerning the pillars, concerning the sea, concerning the bases, and concerning the rest of the vessels that are left in this city,}
\VersePair{34:20}{which the king of Babylon took not, when he carried Jechonias prisoner out of Jerusalem,}{27:20}{which Nebuchadnezzar king of Babylon didn’t take when he carried away captive Jeconiah the son of Jehoiakim, king of Judah, from Jerusalem to Babylon, and all the nobles of Judah and Jerusalem —}
\VersePair{34:22}{they shall go into Babylon, saith the Lord.}{27:22}{‘They will be carried to Babylon, and there they will be, until the day that I visit them,’ says Yahweh; ‘then I will bring them up, and restore them to this place.’”}
\end{paracol}
\ChapterHeading{35}
\begin{paracol}{2}
\VersePair{35:1}{And it came to pass in the fourth year of Sedekias king of Juda, in the fifth month, that Ananias the false prophet, the son of Azor, from Gabaon, spoke to me in the house of the Lord, in the sight of the priests and all the people, saying,}{28:1}{That same year, in the beginning of the reign of Zedekiah king of Judah, in the fourth year, in the fifth month, Hananiah the son of Azzur, the prophet, who was of Gibeon, spoke to me in Yahweh’s house, in the presence of the priests and of all the people, saying,}
\VersePair{35:2}{Thus saith the Lord; I have broken the yoke of the king of Babylon.}{28:2}{“Yahweh of Armies, the God of Israel, says, ‘I have broken the yoke of the king of Babylon.}
\VersePair{35:3}{Yet two full years, and I will return into this place the vessels of the house of the Lord,}{28:3}{Within two full years I will bring again into this place all the vessels of Yahweh’s house that Nebuchadnezzar king of Babylon took away from this place and carried to Babylon.}
\VersePair{35:4}{and Jechonias, and the captivity of Juda: for I will break the yoke of the king of Babylon.}{28:4}{I will bring again to this place Jeconiah the son of Jehoiakim, king of Judah, with all the captives of Judah, who went to Babylon,’ says Yahweh; ‘for I will break the yoke of the king of Babylon.’”}
\VersePair{35:5}{Then Jeremias spoke to Ananias in the sight of all the people, and in the sight of the priests that stood in the house of the Lord,}{28:5}{Then the prophet Jeremiah said to the prophet Hananiah in the presence of the priests, and in the presence of all the people who stood in Yahweh’s house,}
\VersePair{35:6}{and Jeremias said, May the Lord indeed do thus; may he confirm thy word which thou dost prophesy, to return the vessels of the house of the Lord, and all the captivity, out of Babylon to this place.}{28:6}{even the prophet Jeremiah said, “Amen! May Yahweh do so. May Yahweh perform your words which you have prophesied, to bring again the vessels of Yahweh’s house, and all those who are captives, from Babylon to this place.}
\VersePair{35:7}{Nevertheless hear ye the word of the Lord which I speak in your ears, and in the ears of all the people.}{28:7}{Nevertheless listen now to this word that I speak in your ears, and in the ears of all the people:}
\VersePair{35:8}{The prophets that were before me and before you of old, also prophesied over much country, and against great kingdoms, concerning war.}{28:8}{The prophets who have been before me and before you of old prophesied against many countries, and against great kingdoms, of war, of evil, and of pestilence.}
\VersePair{35:9}{As for the prophet that has prophesied for peace, when the word has come to pass, they shall know the prophet whom the Lord has sent them in truth.}{28:9}{As for the prophet who prophesies of peace, when the word of the prophet happens, then the prophet will be known, that Yahweh has truly sent him.”}
\VersePair{35:10}{Then Ananias took the yokes from the neck of Jeremias in the sight of all the people, and broke them to pieces.}{28:10}{Then Hananiah the prophet took the bar from off the prophet Jeremiah’s neck, and broke it.}
\VersePair{35:11}{And Ananias spoke in the presence of all the people, saying, Thus said the Lord; Thus will I break the yoke of the king of Babylon from the necks of all the nations. And Jeremias went his way.}{28:11}{Hananiah spoke in the presence of all the people, saying, “Yahweh says: ‘Even so I will break the yoke of Nebuchadnezzar king of Babylon from off the neck of all the nations within two full years.’” Then the prophet Jeremiah went his way.}
\VersePair{35:12}{And the word of the Lord came to Jeremias, after that Ananias had broken the yokes off his neck, saying,}{28:12}{Then Yahweh’s word came to Jeremiah, after Hananiah the prophet had broken the bar from off the neck of the prophet Jeremiah, saying,}
\VersePair{35:13}{Go and speak to Ananias, saying, Thus saith the Lord; Thou hast broken the yokes of wood; but I will make instead of them yokes of iron.}{28:13}{“Go, and tell Hananiah, saying, ‘Yahweh says, “You have broken the bars of wood, but you have made in their place bars of iron.”}
\VersePair{35:14}{For thus said the Lord, I have put a yoke of iron on the neck of all the nations, that they may serve the king of Babylon.}{28:14}{For Yahweh of Armies, the God of Israel says, “I have put a yoke of iron on the neck of all these nations, that they may serve Nebuchadnezzar king of Babylon; and they will serve him. I have also given him the animals of the field.”’”}
\VersePair{35:15}{And Jeremias said to Ananias, The Lord has not sent thee; and thou hast caused this people to trust in unrighteousness.}{28:15}{Then the prophet Jeremiah said to Hananiah the prophet, “Listen, Hananiah! Yahweh has not sent you, but you make this people trust in a lie.}
\VersePair{35:16}{Therefore thus said the Lord: Behold, I will cast thee off from the face of the earth: this year thou shalt die.}{28:16}{Therefore Yahweh says, ‘Behold, I will send you away from off the surface of the earth. This year you will die, because you have spoken rebellion against Yahweh.’”}
\VersePair{35:17}{So he died in the seventh month.}{28:17}{So Hananiah the prophet died the same year in the seventh month.}
\end{paracol}
\ChapterHeading{36}
\begin{paracol}{2}
\VersePair{36:1}{And these are the words of the book which Jeremias sent from Jerusalem to the elders of the captivity, and to the priests, and to the false prophets, even an epistle to Babylon for the captivity, and to all the people;}{29:1}{Now these are the words of the letter that Jeremiah the prophet sent from Jerusalem to the residue of the elders of the captivity, and to the priests, to the prophets, and to all the people whom Nebuchadnezzar had carried away captive from Jerusalem to Babylon,}
\VersePair{36:2}{(after the departure of Jechonias the king and the queen, and the eunuchs, and every freeman, and bondman, and artificer, out of Jerusalem;)}{29:2}{( after Jeconiah the king, the queen mother, the eunuchs, the princes of Judah and Jerusalem, the craftsmen, and the smiths had departed from Jerusalem ),}
\VersePair{36:3}{by the hand of Eleasan son of Saphan, and Gamarias son of Chelcias, (whom Sedekias king of Juda sent to the king of Babylon to Babylon) saying,}{29:3}{by the hand of Elasah the son of Shaphan and Gemariah the son of Hilkiah, (whom Zedekiah king of Judah sent to Babylon to Nebuchadnezzar king of Babylon). It said:}
\VersePair{36:4}{Thus said the Lord God of Israel concerning the captivity which I caused to be carried away from Jerusalem;}{29:4}{Yahweh of Armies, the God of Israel, says to all the captives whom I have caused to be carried away captive from Jerusalem to Babylon:}
\VersePair{36:5}{Build ye houses, and inhabit them; and plant gardens, and eat the fruits thereof;}{29:5}{“Build houses and dwell in them. Plant gardens and eat their fruit.}
\VersePair{36:6}{and take ye wives, and beget sons and daughters; and take wives for your sons, and give your daughters to husbands, and be multiplied, and be not diminished.}{29:6}{Take wives and father sons and daughters. Take wives for your sons, and give your daughters to husbands, that they may bear sons and daughters. Multiply there, and don’t be diminished.}
\VersePair{36:7}{And seek the peace of the land into which I have carried you captive, and ye shall pray to the Lord for the people: for in its peace ye shall have peace.}{29:7}{Seek the peace of the city where I have caused you to be carried away captive, and pray to Yahweh for it; for in its peace you will have peace.”}
\VersePair{36:8}{For thus saith the Lord; Let not the false prophets that are among you persuade you, and let not your diviners persuade you, and hearken not to your dreams which ye dream.}{29:8}{For Yahweh of Armies, the God of Israel says: “Don’t let your prophets who are among you and your diviners deceive you. Don’t listen to your dreams which you cause to be dreamed.}
\VersePair{36:9}{For they prophesy to you unrighteous words in my name; and I sent them not.}{29:9}{For they prophesy falsely to you in my name. I have not sent them,” says Yahweh.}
\VersePair{36:10}{For thus said the Lord; When seventy years shall be on the point of being accomplished at Babylon, I will visit you, and will confirm my words to you, to bring back your people to this place.}{29:10}{For Yahweh says, “After seventy years are accomplished for Babylon, I will visit you and perform my good word toward you, in causing you to return to this place.}
\VersePair{36:11}{And I will devise for you a device of peace, and not evil, to bestow upon you these good things.}{29:11}{For I know the thoughts that I think toward you,” says Yahweh, “thoughts of peace, and not of evil, to give you hope and a future.}
\VersePair{36:12}{And do ye pray to me, and I will hearken to you: and do ye earnestly seek me, and ye shall find me;}{29:12}{You shall call on me, and you shall go and pray to me, and I will listen to you.}
\VersePair{36:13}{for ye shall seek me with your whole heart.}{29:13}{You shall seek me and find me, when you search for me with all your heart.}
\VersePair{36:14}{And I will appear to you:}{29:14}{I will be found by you,” says Yahweh, “and I will turn again your captivity, and I will gather you from all the nations, and from all the places where I have driven you,” says Yahweh. “I will bring you again to the place from where I caused you to be carried away captive.”}
\VersePair{36:15}{whereas ye said, The Lord has appointed for us prophets in Babylon:}{29:15}{Because you have said, “Yahweh has raised us up prophets in Babylon,”}
\VersePair{36:16}{}{29:16}{Yahweh says concerning the king who sits on David’s throne, and concerning all the people who dwell in this city, your brothers who haven’t gone with you into captivity,}
\VersePair{36:17}{}{29:17}{Yahweh of Armies says: “Behold, I will send on them the sword, the famine, and the pestilence, and will make them like rotten figs that can’t be eaten, they are so bad.}
\VersePair{36:18}{}{29:18}{I will pursue after them with the sword, with the famine, and with the pestilence, and will deliver them to be tossed back and forth among all the kingdoms of the earth, to be an object of horror, an astonishment, a hissing, and a reproach among all the nations where I have driven them,}
\VersePair{36:19}{}{29:19}{because they have not listened to my words,” says Yahweh, “with which I sent to them my servants the prophets, rising up early and sending them; but you would not hear,” says Yahweh.}
\VersePair{36:20}{}{29:20}{Hear therefore Yahweh’s word, all you captives whom I have sent away from Jerusalem to Babylon.}
\VersePair{36:21}{Thus saith the Lord concerning Achiab, and concerning Sedekias; Behold, I will deliver them into the hands of the king of Babylon; and he shall smite them in your sight.}{29:21}{Yahweh of Armies, the God of Israel, says concerning Ahab the son of Kolaiah, and concerning Zedekiah the son of Maaseiah, who prophesy a lie to you in my name: “Behold, I will deliver them into the hand of Nebuchadnezzar king of Babylon; and he will kill them before your eyes.}
\VersePair{36:22}{And they shall make of them a curse in all the captivity of Juda in Babylon, saying, The Lord do to thee as he did to Sedekias, and as he did to Achiab, whom the king of Babylon fried in the fire;}{29:22}{A curse will be taken up about them by all the captives of Judah who are in Babylon, saying, ‘Yahweh make you like Zedekiah and like Ahab, whom the king of Babylon roasted in the fire;’}
\VersePair{36:23}{because of the iniquity which they wrought in Israel, and because they committed adultery with the wives of their fellow-citizens, and spoke a word in my name, which I did not command them to speak, and I am witness, saith the Lord.}{29:23}{because they have done foolish things in Israel, and have committed adultery with their neighbors’wives, and have spoken words in my name falsely, which I didn’t command them. I am he who knows, and am witness,” says Yahweh.}
\VersePair{36:24}{And to Samæas the Ælamite thou shalt say,}{29:24}{Concerning Shemaiah the Nehelamite you shall speak, saying,}
\VersePair{36:25}{I sent thee not in my name: and to Sophonias the priest the son of Maasæas say thou,}{29:25}{“Yahweh of Armies, the God of Israel, says, ‘Because you have sent letters in your own name to all the people who are at Jerusalem, and to Zephaniah the son of Maaseiah, the priest, and to all the priests, saying,}
\VersePair{36:26}{The Lord has made thee priest in the place of Jodae the priest, to be ruler in the house of the Lord over every prophet, and to every madman, and thou shalt put them in prison, and into the dungeon.}{29:26}{“Yahweh has made you priest in the place of Jehoiada the priest, that there may be officers in Yahweh’s house, for every man who is crazy and makes himself a prophet, that you should put him in the stocks and in shackles.}
\VersePair{36:27}{And now wherefore have ye reviled together Jeremias of Anathoth, who prophesied to you?}{29:27}{Now therefore, why have you not rebuked Jeremiah of Anathoth, who makes himself a prophet to you,}
\VersePair{36:28}{Did he not send for this purpose? for in the course of this month he sent to you to Babylon, saying, It is far off: build ye houses, and inhabit them; and plant gardens, and eat the fruit of them.}{29:28}{because he has sent to us in Babylon, saying, The captivity is long. Build houses, and dwell in them. Plant gardens, and eat their fruit?”’”}
\VersePair{36:29}{And Sophonias read the book in the ears of Jeremias.}{29:29}{Zephaniah the priest read this letter in the hearing of Jeremiah the prophet.}
\VersePair{36:30}{Then the word of the Lord came to Jeremias, saying,}{29:30}{Then Yahweh’s word came to Jeremiah, saying,}
\VersePair{36:31}{Send to the captivity, saying, Thus saith the Lord concerning Samæas the Ælamite, Since Samæas has prophesied to you, and I sent him not, and he has made you to trust in iniquity,}{29:31}{“Send to all of the captives, saying, ‘Yahweh says concerning Shemaiah the Nehelamite: “Because Shemaiah has prophesied to you, and I didn’t send him, and he has caused you to trust in a lie,”}
\VersePair{36:32}{therefore thus saith the Lord; Behold, I will visit Samæas, and his family: and there shall not be a man of them in the midst of you to see the good which I will do to you: they shall not see it.}{29:32}{therefore Yahweh says, “Behold, I will punish Shemaiah the Nehelamite and his offspring. He will not have a man to dwell among this people. He won’t see the good that I will do to my people,” says Yahweh, “because he has spoken rebellion against Yahweh.”’”}
\end{paracol}
\ChapterHeading{37}
\begin{paracol}{2}
\VersePair{37:1}{The Word that came to Jeremias from the Lord, saying,}{30:1}{The word that came to Jeremiah from Yahweh, saying,}
\VersePair{37:2}{Thus speaks the Lord God of Israel, saying, Write all the words which I have spoken to thee in a book.}{30:2}{“Yahweh, the God of Israel, says, ‘Write all the words that I have spoken to you in a book.}
\VersePair{37:3}{For, behold, the days come, saith the Lord, when I will bring back the captivity of my people Israel and Juda, said the Lord: and I will bring them back to the land which I gave to their fathers, and they shall be lords of it.}{30:3}{For, behold, the days come,’ says Yahweh, ‘that I will reverse the captivity of my people Israel and Judah,’ says Yahweh. ‘I will cause them to return to the land that I gave to their fathers, and they will possess it.’”}
\VersePair{37:4}{And these are the Words which the Lord spoke concerning Israel and Juda:}{30:4}{These are the words that Yahweh spoke concerning Israel and concerning Judah.}
\VersePair{37:5}{Thus said the Lord: Ye shall hear a sound of fear, there is fear, and there is not peace.}{30:5}{{\raggedright For Yahweh says: \poemline{1}{“We have heard a voice of trembling;}\poemline{2}{a voice of fear, and not of peace.}}}
\VersePair{37:6}{Enquire, and see if a male has born a child? and ask concerning the fear, wherein they shall hold their loins, and look for safety: for I have seen every man, and his hands are on his loins; their faces are turned to paleness.}{30:6}{{\raggedright Ask now, and see whether a man travails with child. \poemline{2}{Why do I see every man with his hands on his waist, as a woman in travail,}\poemline{2}{and all faces are turned pale?}}}
\VersePair{37:7}{For that day is great, and there is not such another; and it is a time of straitness to Jacob; but he shall be saved out of it.}{30:7}{{\raggedright Alas, for that day is great, so that none is like it! \poemline{2}{It is even the time of Jacob’s trouble;}\poemline{2}{but he will be saved out of it.}}}
\VersePair{37:8}{In that day, said the Lord, I will break the yoke off their neck, and will burst their bonds, and they shall no longer serve strangers:}{30:8}{{\raggedright It will come to pass in that day, says Yahweh of Armies, that I will break his yoke from off your neck, \poemline{2}{and will burst your bonds.}\poemline{2}{Strangers will no more make them their bondservants;}}}
\VersePair{37:9}{but they shall serve the Lord their God; and I will raise up to them David their king.}{30:9}{{\raggedright but they will serve Yahweh their God, \poemline{2}{and David their king,}\poemline{2}{whom I will raise up to them.}}}
\VersePair{37:10}{}{30:10}{{\raggedright Therefore don’t be afraid, O Jacob my servant, says Yahweh. \poemline{2}{Don’t be dismayed, Israel.}\poemline{1}{For, behold, I will save you from afar,}\poemline{2}{and save your offspring from the land of their captivity.}\poemline{1}{Jacob will return,}\poemline{2}{and will be quiet and at ease.}\poemline{2}{No one will make him afraid.}}}
\VersePair{37:11}{}{30:11}{{\raggedright For I am with you, says Yahweh, to save you; \poemline{2}{for I will make a full end of all the nations where I have scattered you,}\poemline{2}{but I will not make a full end of you;}\poemline{1}{but I will correct you in measure,}\poemline{2}{and will in no way leave you unpunished.”}}}
\VersePair{37:12}{Thus saith the Lord; I have brought on thee destruction; thy stroke is painful.}{30:12}{{\raggedright For Yahweh says, \poemline{1}{“Your hurt is incurable.}\poemline{2}{Your wound is grievous.}}}
\VersePair{37:13}{There is none to judge thy cause: thou hast been painfully treated for healing, there is no help for thee.}{30:13}{{\raggedright There is no one to plead your cause, \poemline{2}{that you may be bound up.}\poemline{2}{You have no healing medicines.}}}
\VersePair{37:14}{All thy friends have forgotten thee; they shall not ask about thee at all, for I have smitten thee with the stroke of an enemy, even severe correction: thy sins have abounded above all thine iniquity.}{30:14}{{\raggedright All your lovers have forgotten you. \poemline{2}{They don’t seek you.}\poemline{1}{For I have wounded you with the wound of an enemy,}\poemline{2}{with the chastisement of a cruel one,}\poemline{1}{for the greatness of your iniquity,}\poemline{2}{because your sins were increased.}}}
\VersePair{37:16}{Therefore all that devour thee shall be eaten, and all thine enemies shall eat all their own flesh. Thy sins have abounded beyond the multitude of thine iniquities, therefore they have done these things to thee. And they that spoil thee shall become a spoil, and I will give up to be plundered all that have plundered thee.}{30:16}{{\raggedright Therefore all those who devour you will be devoured. \poemline{2}{All your adversaries, everyone of them, will go into captivity.}\poemline{1}{Those who plunder you will be plunder.}\poemline{2}{I will make all who prey on you become prey.}}}
\VersePair{37:17}{For I will bring about thy healing, I will heal thee of thy grievous wound, saith the Lord; for thou art called Dispersed: she is your prey, for no one seeks after her.}{30:17}{{\raggedright For I will restore health to you, \poemline{2}{and I will heal you of your wounds,” says Yahweh,}\poemline{1}{“because they have called you an outcast,}\poemline{2}{saying, ‘It is Zion, whom no man seeks after.’”}}}
\VersePair{37:18}{Thus said the Lord; Behold, I will turn the captivity of Jacob, and will have pity upon his prisoners; and the city shall be built upon her hill, and the people shall settle after their manner.}{30:18}{{\raggedright Yahweh says: \poemline{1}{“Behold, I will reverse the captivity of Jacob’s tents,}\poemline{2}{and have compassion on his dwelling places.}\poemline{1}{The city will be built on its own hill,}\poemline{2}{and the palace will be inhabited in its own place.}}}
\VersePair{37:19}{And there shall go forth from them singers, even the sound of men making merry: and I will multiply them, and they shall not at all be diminished.}{30:19}{{\raggedright Thanksgiving will proceed out of them \poemline{2}{with the voice of those who make merry.}\poemline{1}{I will multiply them,}\poemline{2}{and they will not be few;}\poemline{1}{I will also glorify them,}\poemline{2}{and they will not be small.}}}
\VersePair{37:20}{And their sons shall go in as before, and their testimonies shall be established before me, and I will visit them that afflict them.}{30:20}{{\raggedright Their children also will be as before, \poemline{2}{and their congregation will be established before me.}\poemline{2}{I will punish all who oppress them.}}}
\VersePair{37:21}{And their mighty ones shall be over them, and their prince shall proceed of themselves; and I will gather them, and they shall return to me: for who is this that has set his heart to return to me? saith the Lord.}{30:21}{{\raggedright Their prince will be one of them, \poemline{2}{and their ruler will proceed from among them.}\poemline{1}{I will cause him to draw near,}\poemline{2}{and he will approach me;}\poemline{2}{for who is he who has had boldness to approach me?” says Yahweh.}}}
\VersePair{37:23}{For the wrathful anger of the Lord has gone forth, even a whirlwind of anger has gone forth: it shall come upon the ungodly.}{30:23}{{\raggedright Behold, Yahweh’s storm, his wrath, has gone out, \poemline{2}{a sweeping storm;}\poemline{2}{it will burst on the head of the wicked.}}}
\VersePair{37:24}{The fierce anger of the Lord shall not return, until he shall execute it, and until he shall establish the purpose of his heart: in the latter days ye shall know these things.}{30:24}{{\raggedright The fierce anger of Yahweh will not return until he has accomplished, \poemline{2}{and until he has performed the intentions of his heart.}\poemline{2}{In the latter days you will understand it.”}}}
\end{paracol}
\ChapterHeading{38}
\begin{paracol}{2}
\VersePair{38:1}{At that time, saith the Lord, I will be a God to the family of Israel, and they shall be to me a people.}{31:1}{“At that time,” says Yahweh, “I will be the God of all the families of Israel, and they will be my people.”}
\VersePair{38:2}{Thus saith the Lord, I found him warm in the wilderness with them that were slain with the sword: go ye and destroy not Israel.}{31:2}{Yahweh says, “The people who survive the sword found favor in the wilderness; even Israel, when I went to cause him to rest.”}
\VersePair{38:3}{The Lord appeared to him from afar, saying, I have loved thee with an everlasting love: therefore have I drawn thee in compassion.}{31:3}{{\raggedright Yahweh appeared of old to me, saying, \poemline{1}{“Yes, I have loved you with an everlasting love.}\poemline{2}{Therefore I have drawn you with loving kindness.}}}
\VersePair{38:4}{For I will build thee, and thou shalt be built, O virgin of Israel: thou shalt yet take thy timbrel, and go forth with the party of them that make merry.}{31:4}{{\raggedright I will build you again, \poemline{2}{and you will be built, O virgin of Israel.}\poemline{1}{You will again be adorned with your tambourines,}\poemline{2}{and will go out in the dances of those who make merry.}}}
\VersePair{38:5}{For ye have planted vineyards on the mountains of Samaria: plant ye, and praise.}{31:5}{{\raggedright Again you will plant vineyards on the mountains of Samaria. \poemline{2}{The planters will plant,}\poemline{2}{and will enjoy its fruit.}}}
\VersePair{38:6}{For it is a day when those that plead on the mountains of Ephraim shall call, saying, Arise ye, and go up to Sion to the Lord your God.}{31:6}{{\raggedright For there will be a day that the watchmen on the hills of Ephraim cry, \poemline{2}{‘Arise! Let’s go up to Zion to Yahweh our God.’”}}}
\VersePair{38:7}{For thus saith the Lord to Jacob; Rejoice ye, and exult over the head of the nations: make proclamation, and praise ye: say, The Lord has delivered his people, the remnant of Israel.}{31:7}{{\raggedright For Yahweh says, \poemline{1}{“Sing with gladness for Jacob,}\poemline{2}{and shout for the chief of the nations.}\poemline{1}{Publish, praise, and say,}\poemline{2}{‘Yahweh, save your people,}\poemline{2}{the remnant of Israel!’}}}
\VersePair{38:8}{Behold, I bring them from the north, and will gather them from the end of the earth to the feast of the passover: and the people shall beget a great multitude, and they shall return hither.}{31:8}{{\raggedright Behold, I will bring them from the north country, \poemline{2}{and gather them from the uttermost parts of the earth,}\poemline{1}{along with the blind and the lame,}\poemline{2}{the woman with child and her who travails with child together.}\poemline{2}{They will return as a great company.}}}
\VersePair{38:9}{They went forth with weeping, and I will bring them back with consolation, causing them to lodge by the channels of waters in a straight way, and they shall not err in it: for I am become a father to Israel, and Ephraim is my first-born.}{31:9}{{\raggedright They will come with weeping. \poemline{2}{I will lead them with petitions.}\poemline{1}{I will cause them to walk by rivers of waters,}\poemline{2}{in a straight way in which they won’t stumble;}\poemline{1}{for I am a father to Israel.}\poemline{2}{Ephraim is my firstborn.}}}
\VersePair{38:10}{Hear the words of the Lord, ye nations, and proclaim them to the islands afar off; say, He that scattered Israel will also gather him, and keep him as one that feeds his flock.}{31:10}{{\raggedright “Hear Yahweh’s word, you nations, \poemline{2}{and declare it in the distant islands. Say,}\poemline{1}{‘He who scattered Israel will gather him,}\poemline{2}{and keep him, as a shepherd does his flock.’}}}
\VersePair{38:11}{For the Lord has ransomed Jacob, he has rescued him out of the hand of them that were stronger than he.}{31:11}{{\raggedright For Yahweh has ransomed Jacob, \poemline{2}{and redeemed him from the hand of him who was stronger than he.}}}
\VersePair{38:12}{And they shall come, and shall rejoice in the mount of Sion, and shall come to the good things of the Lord, even to a land of corn, and wine, and fruits, and cattle, and sheep: and their soul shall be as a fruitful tree; and they shall hunger no more.}{31:12}{{\raggedright They will come and sing in the height of Zion, \poemline{2}{and will flow to the goodness of Yahweh,}\poemline{1}{to the grain, to the new wine, to the oil,}\poemline{2}{and to the young of the flock and of the herd.}\poemline{1}{Their soul will be as a watered garden.}\poemline{2}{They will not sorrow any more at all.}}}
\VersePair{38:13}{Then shall the virgins rejoice in the assembly of youth, and the old men shall rejoice; and I will turn their mourning into joy, and will make them merry.}{31:13}{{\raggedright Then the virgin will rejoice in the dance, \poemline{2}{the young men and the old together;}\poemline{1}{for I will turn their mourning into joy,}\poemline{2}{and will comfort them, and make them rejoice from their sorrow.}}}
\VersePair{38:14}{I will expand and cheer with wine the soul of the priests the sons of Levi, and my people shall be satisfied with my good things: thus saith the Lord.}{31:14}{{\raggedright I will satiate the soul of the priests with fatness, \poemline{2}{and my people will be satisfied with my goodness,” says Yahweh.}}}
\VersePair{38:15}{A voice was heard in Rama, of lamentation, and of weeping, and wailing; Rachel would not cease weeping for her children, because they are not.}{31:15}{{\raggedright Yahweh says: \poemline{1}{“A voice is heard in Ramah,}\poemline{2}{lamentation and bitter weeping,}\poemline{1}{Rachel weeping for her children.}\poemline{2}{She refuses to be comforted for her children,}\poemline{2}{because they are no more.”}}}
\VersePair{38:16}{Thus saith the Lord; Let thy voice cease from weeping, and thine eyes from thy tears: for there is a reward for thy works; and they shall return from the land of thine enemies.}{31:16}{{\raggedright Yahweh says: \poemline{1}{“Refrain your voice from weeping,}\poemline{2}{and your eyes from tears,}\poemline{2}{for your work will be rewarded,” says Yahweh.}\poemline{2}{“They will come again from the land of the enemy.}}}
\VersePair{38:17}{There shall be an abiding home for thy children.}{31:17}{{\raggedright There is hope for your latter end,” says Yahweh. \poemline{2}{“Your children will come again to their own territory.}}}
\VersePair{38:18}{I have heard the sound of Ephraim lamenting, and saying, Thou hast chastened me, and I was chastened; I as a calf was not willingly taught: turn thou me, and I shall turn; for thou art the Lord my God.}{31:18}{{\raggedright “I have surely heard Ephraim grieving thus, \poemline{2}{‘You have chastised me,}\poemline{2}{and I was chastised, as an untrained calf.}\poemline{1}{Turn me, and I will be turned,}\poemline{2}{for you are Yahweh my God.}}}
\VersePair{38:19}{For after my captivity I repented; and after I knew, I groaned for the day of shame, and shewed thee that I bore reproach from my youth.}{31:19}{{\raggedright Surely after that I was turned. \poemline{2}{I repented.}\poemline{1}{After that I was instructed.}\poemline{2}{I struck my thigh.}\poemline{1}{I was ashamed, yes, even confounded,}\poemline{2}{because I bore the reproach of my youth.’}}}
\VersePair{38:20}{Ephraim is a beloved son, a pleasing child to me: for because my words are in him, I will surely remember him: therefore I made haste to help him; I will surely have mercy upon him, saith the Lord.}{31:20}{{\raggedright Is Ephraim my dear son? \poemline{2}{Is he a darling child?}\poemline{1}{For as often as I speak against him,}\poemline{2}{I still earnestly remember him.}\poemline{1}{Therefore my heart yearns for him.}\poemline{2}{I will surely have mercy on him,” says Yahweh.}}}
\VersePair{38:21}{Prepare thyself, O Sion; execute vengeance; look to thy ways: return, O virgin of Israel, by the way by which thou wentest, return mourning to thy cities.}{31:21}{{\raggedright “Set up road signs. \poemline{2}{Make guideposts.}\poemline{1}{Set your heart toward the highway,}\poemline{2}{even the way by which you went.}\poemline{1}{Turn again, virgin of Israel.}\poemline{2}{Turn again to these your cities.}}}
\VersePair{38:22}{How long, O disgraced daughter, wilt thou turn away? for the Lord has created safety for a new plantation: men shall go about in safety.}{31:22}{{\raggedright How long will you go here and there, \poemline{2}{you backsliding daughter?}\poemline{1}{For Yahweh has created a new thing in the earth:}\poemline{2}{a woman will encompass a man.”}}}
\VersePair{38:23}{For thus saith the Lord; They shall yet speak this word in the land of Juda, and in the cities thereof, when I shall turn his captivity; blessed be the Lord on his righteous holy mountain!}{31:23}{Yahweh of Armies, the God of Israel, says: “Yet again they will use this speech in the land of Judah and in its cities, when I reverse their captivity: ‘Yahweh bless you, habitation of righteousness, mountain of holiness.’}
\VersePair{38:24}{And there shall be dwellers in the cities of Juda, and in all his land, together with the husbandman, and the shepherd shall go forth with the flock.}{31:24}{Judah and all its cities will dwell therein together, the farmers, and those who go about with flocks.}
\VersePair{38:25}{For I have saturated every thirsting soul, and filled every hungry soul.}{31:25}{For I have satiated the weary soul, and I have replenished every sorrowful soul.”}
\VersePair{38:26}{Therefore I awaked, and beheld; and my sleep was sweet to me.}{31:26}{On this I awakened, and saw; and my sleep was sweet to me.}
\VersePair{38:27}{Therefore, behold, the days come, saith the Lord, when I will sow the house of Israel and the house of Juda with the seed of man, and the seed of beast.}{31:27}{“Behold, the days come,” says Yahweh, “that I will sow the house of Israel and the house of Judah with the seed of man and with the seed of animal.}
\VersePair{38:28}{And it shall come to pass, that as I watched over them, to pull down, and to afflict, so will I watch over them, to build, and to plant, saith the Lord.}{31:28}{It will happen that, like as I have watched over them to pluck up and to break down and to overthrow and to destroy and to afflict, so I will watch over them to build and to plant,” says Yahweh.}
\VersePair{38:29}{In those days they shall certainly not say, The fathers ate a sour grape, and the children's teeth were set on edge.}{31:29}{{\raggedright “In those days they will say no more, \poemline{1}{“‘The fathers have eaten sour grapes,}\poemline{2}{and the children’s teeth are set on edge.’}}}
\VersePair{38:30}{But every one shall die in his own sin; and the teeth of him that eats the sour grape shall be set on edge.}{31:30}{But everyone will die for his own iniquity. Every man who eats the sour grapes, his teeth will be set on edge.}
\VersePair{38:31}{Behold, the days come, saith the Lord, when I will make a new covenant with the house of Israel, and with the house of Juda:}{31:31}{“Behold, the days come,” says Yahweh, “that I will make a new covenant with the house of Israel, and with the house of Judah,}
\VersePair{38:32}{not according to the covenant which I made with their fathers in the day when I took hold of their hand to bring them out of the land of Egypt; for they abode not in my covenant, and I disregarded them, saith the Lord.}{31:32}{not according to the covenant that I made with their fathers in the day that I took them by the hand to bring them out of the land of Egypt, which covenant of mine they broke, although I was a husband to them,” says Yahweh.}
\VersePair{38:33}{For this is my covenant which I will make with the house of Israel; after those days, saith the Lord, I will surely put my laws into their mind, and write them on their hearts; and I will be to them a God, and they shall be to me a people.}{31:33}{{\raggedright “But this is the covenant that I will make with the house of Israel after those days,” says Yahweh: \poemline{1}{“I will put my law in their inward parts,}\poemline{2}{and I will write it in their heart.}\poemline{1}{I will be their God,}\poemline{2}{and they shall be my people.}}}
\VersePair{38:34}{And they shall not at all teach every one his fellow citizen, and every one his brother, saying, Know the Lord: for all shall know me, from the least of them to the greatest of them: for I will be merciful to their iniquities, and their sins I will remember no more.}{31:34}{{\raggedright They will no longer each teach his neighbor, \poemline{2}{and every man teach his brother, saying, ‘Know Yahweh;’}\poemline{1}{for they will all know me,}\poemline{2}{from their least to their greatest,” says Yahweh,}\poemline{1}{“for I will forgive their iniquity,}\poemline{2}{and I will remember their sin no more.”}}}
\VersePair{38:35}{Though the sky should be raised to a greater height, saith the Lord, and though the ground of the earth should be sunk lower beneath, yet I will not cast off the family of Israel, saith the Lord, for all that they have done.}{31:37}{{\raggedright Yahweh says: “If heaven above can be measured, \poemline{2}{and the foundations of the earth searched out beneath,}\poemline{2}{then I will also cast off all the offspring of Israel for all that they have done,” says Yahweh.}}}
\VersePair{38:36}{Thus saith the Lord, who gives the sun for a light by day, the moon and the stars for a light by night, and makes a roaring in the sea, so that the waves thereof roar; the Lord Almighty is his name:}{31:35}{{\raggedright Yahweh, who gives the sun for a light by day, \poemline{2}{and the ordinances of the moon and of the stars for a light by night,}\poemline{1}{who stirs up the sea, so that its waves roar —}\poemline{2}{Yahweh of Armies is his name, says:}}}
\VersePair{38:37}{if these ordinances cease from before me, saith the Lord, then shall the family of Israel cease to be a nation before me for ever.}{31:36}{{\raggedright “If these ordinances depart from before me,” says Yahweh, \poemline{2}{“then the offspring of Israel also will cease from being a nation before me forever.”}}}
\VersePair{38:38}{Behold, the days come, saith the Lord, when the city shall be built to the Lord from the tower of Anameel to the gate of the corner.}{31:38}{“Behold, the days come,” says Yahweh, “that the city will be built to Yahweh from the tower of Hananel to the gate of the corner.}
\VersePair{38:39}{And the measurement of it shall proceed in front of them as far as the hills of Gareb, and it shall be compassed with a circular wall of choice stones.}{31:39}{The measuring line will go out further straight onward to the hill Gareb, and will turn toward Goah.}
\VersePair{38:40}{And all the Asaremoth even to Nachal Kedron, as far as the corner of the horse-gate eastward, shall be holiness to the Lord; and it shall not fail any more, and shall not be destroyed for ever.}{31:40}{The whole valley of the dead bodies and of the ashes, and all the fields to the brook Kidron, to the corner of the horse gate toward the east, will be holy to Yahweh. It will not be plucked up or thrown down any more forever.”}
\end{paracol}
\ChapterHeading{39}
\begin{paracol}{2}
\VersePair{39:1}{The word that came from the Lord to Jeremias in the tenth year of king Sedekias, this is the eighteenth year of king Nabuchodonosor king of Babylon.}{32:1}{This is the word that came to Jeremiah from Yahweh in the tenth year of Zedekiah king of Judah, which was the eighteenth year of Nebuchadnezzar.}
\VersePair{39:2}{And the host of the king of Babylon had made a rampart against Jerusalem: and Jeremias was kept in the court of the prison, which is in the king's house;}{32:2}{Now at that time the king of Babylon’s army was besieging Jerusalem. Jeremiah the prophet was shut up in the court of the guard, which was in the king of Judah’s house.}
\VersePair{39:3}{in which king Sedekias had shut him up, saying, Wherefore dost thou prophesy, saying, Thus saith the Lord, Behold, I will give this city into the hands of the king of Babylon, and he shall take it;}{32:3}{For Zedekiah king of Judah had shut him up, saying, “Why do you prophesy, and say, ‘Yahweh says, “Behold, I will give this city into the hand of the king of Babylon, and he will take it;}
\VersePair{39:4}{and Sedekias shall by no means be delivered out of the hand of the Chaldeans, for he shall certainly be given up into the hands of the king of Babylon, and his mouth shall speak to his mouth, and his eyes shall look upon his eyes;}{32:4}{and Zedekiah king of Judah won’t escape out of the hand of the Chaldeans, but will surely be delivered into the hand of the king of Babylon, and will speak with him mouth to mouth, and his eyes will see his eyes;}
\VersePair{39:5}{and Sedekias shall go into Babylon, and dwell there?}{32:5}{and he will bring Zedekiah to Babylon, and he will be there until I visit him,” says Yahweh, “though you fight with the Chaldeans, you will not prosper ”’?”}
\VersePair{39:6}{And the Word of the Lord came to Jeremias, saying,}{32:6}{Jeremiah said, “Yahweh’s word came to me, saying,}
\VersePair{39:7}{Behold, Anameel the son of Salom thy father's brother is coming to thee, saying, Buy thee my field that is in Anathoth: for thou hast the right to take it as a purchase.}{32:7}{‘Behold, Hanamel the son of Shallum your uncle will come to you, saying, “Buy my field that is in Anathoth; for the right of redemption is yours to buy it.”’”}
\VersePair{39:8}{So Anameel the son of Salom my father's brother came to me into the court of the prison, and said, Buy thee my field that is in the land of Benjamin, in Anathoth: for thou hast a right to buy it, and thou art the elder. So I knew that it was the word of the Lord.}{32:8}{“So Hanamel my uncle’s son came to me in the court of the guard according to Yahweh’s word, and said to me, ‘Please buy my field that is in Anathoth, which is in the land of Benjamin; for the right of inheritance is yours, and the redemption is yours. Buy it for yourself.’ “Then I knew that this was Yahweh’s word.}
\VersePair{39:9}{And I bought the field of Anameel the son of my father's brother, and I weighed him seventeen shekels of silver.}{32:9}{I bought the field that was in Anathoth of Hanamel my uncle’s son, and weighed him the money, even seventeen shekels of silver. \footnote{32:9: A shekel is about 10 grams or about 0.35 ounces.}}
\VersePair{39:10}{And I wrote it in a book, and sealed it, and took the testimony of witnesses, and weighed the money in the balance.}{32:10}{I signed the deed, sealed it, called witnesses, and weighed the money in the balances to him.}
\VersePair{39:11}{And I took the book of the purchase that was sealed;}{32:11}{So I took the deed of the purchase, both that which was sealed, containing the terms and conditions, and that which was open;}
\VersePair{39:12}{and I gave it to Baruch son of Nerias, son of Maasæas, in the sight of Anameel my father's brother's son, and in the sight of the men that stood by and wrote in the book of the purchase, and in the sight of the Jews that were in the court of the prison.}{32:12}{and I delivered the deed of the purchase to Baruch the son of Neriah, the son of Mahseiah, in the presence of Hanamel my uncle’s son, and in the presence of the witnesses who signed the deed of the purchase, before all the Jews who sat in the court of the guard.}
\VersePair{39:13}{And I charged Baruch in their presence, saying, Thus saith the Lord Almighty;}{32:13}{“I commanded Baruch before them, saying,}
\VersePair{39:14}{Take this book of the purchase, and the book that has been read; and thou shalt put it into an earthen vessel, that it may remain many days.}{32:14}{Yahweh of Armies, the God of Israel, says: ‘Take these deeds, this deed of the purchase which is sealed, and this deed which is open, and put them in an earthen vessel, that they may last many days.’}
\VersePair{39:15}{For thus saith the Lord; There shall yet be bought fields and houses and vineyards in this land.}{32:15}{For Yahweh of Armies, the God of Israel says: ‘Houses and fields and vineyards will yet again be bought in this land.’}
\VersePair{39:16}{And I prayed to the Lord after I had given the book of the purchase to Baruch the son of Nerias, saying,}{32:16}{Now after I had delivered the deed of the purchase to Baruch the son of Neriah, I prayed to Yahweh, saying,}
\VersePair{39:17}{O ever living Lord! thou hast made the heaven and the earth by thy great power, and with thy high and lofty arm: nothing can be hidden from thee.}{32:17}{“Ah Lord Yahweh! Behold, you have made the heavens and the earth by your great power and by your outstretched arm. There is nothing too hard for you.}
\VersePair{39:18}{Granting mercy to thousands, and recompensing the sins of the fathers into the bosoms of their children after them: the great, the strong God;}{32:18}{You show loving kindness to thousands, and repay the iniquity of the fathers into the bosom of their children after them. The great, the mighty God, Yahweh of Armies is your name:}
\VersePair{39:19}{the Lord of great counsel, and mighty in deeds, the great Almighty God, and Lord of great name: thine eyes are upon the ways of the children of men, to give to every one according to his way:}{32:19}{great in counsel, and mighty in work; whose eyes are open to all the ways of the children of men, to give everyone according to his ways, and according to the fruit of his doings;}
\VersePair{39:20}{who hast wrought signs and wonders in the land of Egypt even to this day, and in Israel, and among the inhabitants of the earth; and thou didst make for thyself a name, as at this day;}{32:20}{who performed signs and wonders in the land of Egypt, even to this day, both in Israel and among other men; and made yourself a name, as it is today;}
\VersePair{39:21}{and thou didst bring out thy people Israel out of the land of Egypt with signs, and with wonders, with a mighty hand, and with a high arm, and with great sights;}{32:21}{and brought your people Israel out of the land of Egypt with signs, with wonders, with a strong hand, with an outstretched arm, and with great terror;}
\VersePair{39:22}{and thou gavest them this land, which thou didst swear to give to their fathers, a land flowing with milk and honey;}{32:22}{and gave them this land, which you swore to their fathers to give them, a land flowing with milk and honey.}
\VersePair{39:23}{and they went in, and took it; but they hearkened not to thy voice, and walked not in thine ordinances; they did none of the things which thou didst command them, and they caused all these calamities to happen to them.}{32:23}{They came in and possessed it, but they didn’t obey your voice and didn’t walk in your law. They have done nothing of all that you commanded them to do. Therefore you have caused all this evil to come upon them.}
\VersePair{39:24}{Behold, a multitude is come against the city to take it; and the city is given into the hands of the Chaldeans that fight against it, by the power of the sword, and the famine: as thou hast spoken, so has it happened.}{32:24}{“Behold, siege ramps have been built against the city to take it. The city is given into the hand of the Chaldeans who fight against it, because of the sword, of the famine, and of the pestilence. What you have spoken has happened. Behold, you see it.}
\VersePair{39:25}{And thou sayest to me, Buy thee the field for money; and I wrote a book, and sealed it, and took the testimony of witnesses: and the city is given into the hands of the Chaldeans.}{32:25}{You have said to me, Lord Yahweh, ‘Buy the field for money, and call witnesses;’ whereas the city is given into the hand of the Chaldeans.”}
\VersePair{39:26}{And the word of the Lord came to me, saying,}{32:26}{Then Yahweh’s word came to Jeremiah, saying,}
\VersePair{39:27}{I am the Lord, the God of all flesh: shall anything be hidden from me!}{32:27}{“Behold, I am Yahweh, the God of all flesh. Is there anything too hard for me?}
\VersePair{39:28}{Therefore thus saith the Lord God of Israel; This city shall certainly be delivered into the hands of the king of Babylon, and he shall take it:}{32:28}{Therefore Yahweh says: Behold, I will give this city into the hand of the Chaldeans, and into the hand of Nebuchadnezzar king of Babylon, and he will take it.}
\VersePair{39:29}{and the Chaldeans shall come to war against this city, and they shall burn this city with fire, and shall burn down the houses wherein they burnt incense on the roofs thereof to Baal, and poured drink-offerings to other gods, to provoke me.}{32:29}{The Chaldeans, who fight against this city, will come and set this city on fire, and burn it with the houses on whose roofs they have offered incense to Baal, and poured out drink offerings to other gods, to provoke me to anger.}
\VersePair{39:30}{For the children of Israel and the children of Juda alone did evil in my sight from their youth.}{32:30}{“For the children of Israel and the children of Judah have done only that which was evil in my sight from their youth; for the children of Israel have only provoked me to anger with the work of their hands, says Yahweh.}
\VersePair{39:31}{For this city was obnoxious to my anger and my wrath, from the day that they built it even to this day; that I should remove it from my presence,}{32:31}{For this city has been to me a provocation of my anger and of my wrath from the day that they built it even to this day, so that I should remove it from before my face,}
\VersePair{39:32}{because of all the wickedness of the children of Israel and Juda, which they wrought to provoke me, they and their kings, and their princes, and their priests, and their prophets, the men of Juda, and the dwellers in Jerusalem.}{32:32}{because of all the evil of the children of Israel and of the children of Judah, which they have done to provoke me to anger — they, their kings, their princes, their priests, their prophets, the men of Judah, and the inhabitants of Jerusalem.}
\VersePair{39:33}{And they turned the back to me, and not the face: whereas I taught them early in the morning, but they hearkened no more to receive instruction.}{32:33}{They have turned their backs to me, and not their faces. Although I taught them, rising up early and teaching them, yet they have not listened to receive instruction.}
\VersePair{39:34}{And they set their pollutions in the house, on which my name was called, by their uncleannesses.}{32:34}{But they set their abominations in the house which is called by my name, to defile it.}
\VersePair{39:35}{And they built to Baal the altars that are in the valley of the son of Ennom, to offer their sons and their daughters to king Moloch; which things I commanded them not, neither came it into my mind that they should do this abomination, to cause Juda to sin.}{32:35}{They built the high places of Baal, which are in the valley of the son of Hinnom, to cause their sons and their daughters to pass through fire to Molech, which I didn’t command them. It didn’t even come into my mind, that they should do this abomination, to cause Judah to sin.”}
\VersePair{39:36}{And now thus has the Lord God of Israel said concerning this city, of which thou sayest, It shall be delivered into the hands of the king of Babylon by the sword, and by famine, and banishment.}{32:36}{Now therefore Yahweh, the God of Israel, says concerning this city, about which you say, “It is given into the hand of the king of Babylon by the sword, by the famine, and by the pestilence:”}
\VersePair{39:37}{Behold, I will gather them out of every land, where I have scattered them in my anger, and my wrath, and great fury; and I will bring them back into this place, and will cause them to dwell safely:}{32:37}{“Behold, I will gather them out of all the countries where I have driven them in my anger, and in my wrath, and in great indignation; and I will bring them again to this place. I will cause them to dwell safely.}
\VersePair{39:38}{and they shall be to me a people, and I will be to them a God.}{32:38}{Then they will be my people, and I will be their God.}
\VersePair{39:39}{And I will give them another way, and another heart, to fear me continually, and that for good to them and their children after them.}{32:39}{I will give them one heart and one way, that they may fear me forever, for their good and the good of their children after them.}
\VersePair{39:40}{And I will make with them an everlasting covenant, which I will by no means turn away from them, and I will put my fear into their heart, that they may not depart from me.}{32:40}{I will make an everlasting covenant with them, that I will not turn away from following them, to do them good. I will put my fear in their hearts, that they may not depart from me.}
\VersePair{39:41}{And I will visit them to do them good, and I will plant them in this land in faithfulness, and with all my heart, and with all my soul.}{32:41}{Yes, I will rejoice over them to do them good, and I will plant them in this land assuredly with my whole heart and with my whole soul.”}
\VersePair{39:42}{For thus saith the Lord; As I have brought upon this people all these great evils, so will I bring upon them all the good things which I pronounced upon them.}{32:42}{For Yahweh says: “Just as I have brought all this great evil on this people, so I will bring on them all the good that I have promised them.}
\VersePair{39:43}{And there shall yet be fields bought in the land, of which thou sayest, It shall be destitute of man and beast; and they are delivered into the hands of the Chaldeans.}{32:43}{Fields will be bought in this land, about which you say, ‘It is desolate, without man or animal. It is given into the hand of the Chaldeans.’}
\VersePair{39:44}{And they shall buy fields for money, and thou shalt write a book, and seal it, and shalt take the testimony of witnesses in the land of Benjamin, and round about Jerusalem, and in the cities of Juda, and in the cities of the mountain, and in the cities of the plain, and in the cities of the south: for I will turn their captivity.}{32:44}{Men will buy fields for money, sign the deeds, seal them, and call witnesses, in the land of Benjamin, and in the places around Jerusalem, in the cities of Judah, in the cities of the hill country, in the cities of the lowland, and in the cities of the South; for I will cause their captivity to be reversed,” says Yahweh.}
\end{paracol}
\ChapterHeading{40}
\begin{paracol}{2}
\VersePair{40:1}{And the word of the Lord came to Jeremias the second time, when he was yet bound in the court of the prison, saying,}{33:1}{Moreover Yahweh’s word came to Jeremiah the second time, while he was still locked up in the court of the guard, saying,}
\VersePair{40:2}{Thus saith the Lord, who made the earth and formed it, to establish it; the Lord is his name;}{33:2}{“Yahweh who does it, Yahweh who forms it to establish it — Yahweh is his name, says:}
\VersePair{40:3}{Cry to me, and I will answer thee, and I will declare to thee great and mighty things, which thou knowest not.}{33:3}{‘Call to me, and I will answer you, and will show you great and difficult things, which you don’t know.’}
\VersePair{40:4}{For thus saith the Lord concerning the houses of this city, and concerning the houses of the king of Juda, which have been pulled down for mounds and fortifications,}{33:4}{For Yahweh, the God of Israel, says concerning the houses of this city and concerning the houses of the kings of Judah, which are broken down to make a defense against the mounds and against the sword:}
\VersePair{40:5}{to fight against the Chaldeans, and to fill it with the corpses of men, whom I smote in mine anger and my wrath, and turned away my face from them, for all their wickedness:}{33:5}{‘While men come to fight with the Chaldeans, and to fill them with the dead bodies of men, whom I have killed in my anger and in my wrath, and for all whose wickedness I have hidden my face from this city,}
\VersePair{40:6}{Behold, I bring upon her healing and cure, and I will show myself to them, and will heal her, and make both peace and security.}{33:6}{behold, I will bring it health and healing, and I will cure them; and I will reveal to them abundance of peace and truth.}
\VersePair{40:7}{And I will turn the captivity of Juda, and the captivity of Israel, and will build them, even as before.}{33:7}{I will restore the fortunes of Judah and Israel, and will build them as at the first.}
\VersePair{40:8}{And I will cleanse them from all their iniquities, whereby they have sinned against me, and will not remember their sins, whereby they have sinned against me, and revolted from me.}{33:8}{I will cleanse them from all their iniquity by which they have sinned against me. I will pardon all their iniquities by which they have sinned against me and by which they have transgressed against me.}
\VersePair{40:9}{And it shall be for joy and praise, and for glory to all the people of the earth, who shall hear all the good that I will do: and they shall fear and be provoked for all the good things and for all the peace which I will bring upon them.}{33:9}{This city will be to me for a name of joy, for praise, and for glory, before all the nations of the earth, which will hear all the good that I do to them, and will fear and tremble for all the good and for all the peace that I provide to it.’”}
\VersePair{40:10}{Thus saith the Lord; There shall yet be heard in this place, of which ye say, It is destitute of men and cattle, in the cities of Juda, and in the streets of Jerusalem, the places that have been made desolate for want of men and cattle,}{33:10}{Yahweh says: “Yet again there will be heard in this place, about which you say, ‘It is waste, without man and without animal, even in the cities of Judah, and in the streets of Jerusalem, that are desolate, without man and without inhabitant and without animal,’}
\VersePair{40:11}{the voice of gladness, and the voice of joy, the voice of the bridegroom, and the voice of the bride, the voice of men saying, Give thanks to the Lord Almighty: for the Lord is good; for his mercy endures for ever: and they shall bring gifts into the house of the Lord; for I will turn all the captivity of that land as before, said the Lord.}{33:11}{the voice of joy and the voice of gladness, the voice of the bridegroom and the voice of the bride, the voice of those who say, ‘Give thanks to Yahweh of Armies, for Yahweh is good, for his loving kindness endures forever;’ who bring thanksgiving into Yahweh’s house. For I will cause the captivity of the land to be reversed as at the first,” says Yahweh.}
\VersePair{40:12}{Thus saith the Lord of hosts; There shall yet be in this place, that is desert for want of man and beast, in all the cities thereof, resting-places for shepherds causing their flocks to lie down.}{33:12}{Yahweh of Armies says: “Yet again there will be in this place, which is waste, without man and without animal, and in all its cities, a habitation of shepherds causing their flocks to lie down.}
\VersePair{40:13}{In the cities of the hill country, and in the cities of the valley, and in the cities of the south, and in the land of Benjamin, and in the cities round about Jerusalem, and in the cities of Juda, flocks shall yet pass under the hand of him that numbers them, saith the Lord.}{33:13}{In the cities of the hill country, in the cities of the lowland, in the cities of the South, in the land of Benjamin, in the places around Jerusalem, and in the cities of Judah, the flocks will again pass under the hands of him who counts them,” says Yahweh.}
\VersePair{40:14}{}{33:14}{“Behold, the days come,” says Yahweh, “that I will perform that good word which I have spoken concerning the house of Israel and concerning the house of Judah.}
\VersePair{40:15}{}{33:15}{{\raggedright “In those days and at that time, \poemline{2}{I will cause a Branch of righteousness to grow up to David.}\poemline{2}{He will execute justice and righteousness in the land.}}}
\VersePair{40:16}{}{33:16}{{\raggedright In those days Judah will be saved, \poemline{2}{and Jerusalem will dwell safely.}\poemline{1}{This is the name by which she will be called:}\poemline{2}{Yahweh our righteousness.”}}}
\VersePair{40:17}{}{33:17}{For Yahweh says: “David will never lack a man to sit on the throne of the house of Israel.}
\VersePair{40:18}{}{33:18}{The Levitical priests won’t lack a man before me to offer burnt offerings, to burn meal offerings, and to do sacrifice continually.”}
\VersePair{40:19}{}{33:19}{Yahweh’s word came to Jeremiah, saying,}
\VersePair{40:20}{}{33:20}{“Yahweh says: ‘If you can break my covenant of the day and my covenant of the night, so that there will not be day and night in their time,}
\VersePair{40:21}{}{33:21}{then my covenant could also be broken with David my servant, that he won’t have a son to reign on his throne; and with the Levitical priests, my ministers.}
\VersePair{40:22}{}{33:22}{As the army of the sky can’t be counted, and the sand of the sea can’t be measured, so I will multiply the offspring of David my servant and the Levites who minister to me.’”}
\VersePair{40:23}{}{33:23}{Yahweh’s word came to Jeremiah, saying,}
\VersePair{40:24}{}{33:24}{“Don’t consider what this people has spoken, saying, ‘Has Yahweh cast off the two families which he chose?’ Thus they despise my people, that they should be no more a nation before them.”}
\VersePair{40:25}{}{33:25}{Yahweh says: “If my covenant of day and night fails, if I have not appointed the ordinances of heaven and earth,}
\VersePair{40:26}{}{33:26}{then I will also cast away the offspring of Jacob, and of David my servant, so that I will not take of his offspring to be rulers over the offspring of Abraham, Isaac, and Jacob; for I will cause their captivity to be reversed and will have mercy on them.”}
\end{paracol}
\ChapterHeading{41}
\begin{paracol}{2}
\VersePair{41:1}{The word that came to Jeremias from the Lord (now Nabuchodonosor king of Babylon, and all his army, and all the country of his dominion, were warring against Jerusalem, and against all the cities of Juda,) saying,}{34:1}{The word which came to Jeremiah from Yahweh, when Nebuchadnezzar king of Babylon, with all his army, all the kingdoms of the earth that were under his dominion, and all the peoples, were fighting against Jerusalem and against all its cities, saying:}
\VersePair{41:2}{Thus has the Lord said; Go to Sedekias king of Juda, and thou shalt say to him, Thus has the Lord said, This city shall certainly be delivered into the hands of the king of Babylon, and he shall take it, and shall burn it with fire:}{34:2}{“Yahweh, the God of Israel, says, ‘Go, and speak to Zedekiah king of Judah, and tell him, Yahweh says, “Behold, I will give this city into the hand of the king of Babylon and he will burn it with fire.}
\VersePair{41:3}{and thou shalt not escape out of his hand, but shalt certainly be taken, and shalt be given into his hands; and thine eyes shall see his eyes, and thou shalt enter into Babylon.}{34:3}{You won’t escape out of his hand, but will surely be taken and delivered into his hand. Your eyes will see the eyes of the king of Babylon, and he will speak with you mouth to mouth. You will go to Babylon.”’}
\VersePair{41:4}{But hear the word of the Lord, O Sedekias king of Juda; Thus saith the Lord,}{34:4}{“Yet hear Yahweh’s word, O Zedekiah king of Judah. Yahweh says concerning you, ‘You won’t die by the sword.}
\VersePair{41:5}{Thou shalt die in peace: and as they wept for thy fathers that reigned before thee, they shall weep also for thee, saying, Ah lord! and they shall lament for thee down to the grave: for I have spoken the word, said the Lord.}{34:5}{You will die in peace; and with the burnings of your fathers, the former kings who were before you, so they will make a burning for you. They will lament you, saying, “Ah Lord!” for I have spoken the word,’ says Yahweh.”}
\VersePair{41:6}{And Jeremias spoke to king Sedekias all these words in Jerusalem.}{34:6}{Then Jeremiah the prophet spoke all these words to Zedekiah king of Judah in Jerusalem,}
\VersePair{41:7}{And the host of the king of Babylon warred against Jerusalem, and against the cities of Juda, and against Lachis, and against Azeca: for these strong cities were left among the cities of Juda.}{34:7}{when the king of Babylon’s army was fighting against Jerusalem and against all the cities of Judah that were left, against Lachish and against Azekah; for these alone remained of the cities of Judah as fortified cities.}
\VersePair{41:8}{The word that came from the Lord to Jeremias, after king Sedekias had concluded a covenant with the people, to proclaim a release;}{34:8}{The word came to Jeremiah from Yahweh, after King Zedekiah had made a covenant with all the people who were at Jerusalem, to proclaim liberty to them,}
\VersePair{41:9}{that every one should set at liberty his servant, and every one his handmaid, the Hebrew man and Hebrew woman, that no man of Juda should be a bondman.}{34:9}{that every man should let his male servant, and every man his female servant, who is a Hebrew or a Hebrewess, go free, that no one should make bondservants of them, of a Jew his brother.}
\VersePair{41:10}{Then all the nobles, and all the people who had entered into the covenant, engaging to set free every one his man-servant, and every one his maid, turned,}{34:10}{All the princes and all the people obeyed who had entered into the covenant, that everyone should let his male servant and everyone his female servant go free, that no one should make bondservants of them any more. They obeyed and let them go,}
\VersePair{41:11}{and gave them over to be men-servants and maid-servants.}{34:11}{but afterwards they turned, and caused the servants and the handmaids whom they had let go free to return, and brought them into subjection for servants and for handmaids.}
\VersePair{41:12}{And the word of the Lord came to Jeremias, saying,}{34:12}{Therefore Yahweh’s word came to Jeremiah from Yahweh, saying,}
\VersePair{41:13}{Thus saith the Lord; I made a covenant with your fathers in the day wherein I took them out of the land of Egypt, out of the house of bondage, saying,}{34:13}{“Yahweh, the God of Israel, says: ‘I made a covenant with your fathers in the day that I brought them out of the land of Egypt, out of the house of bondage, saying:}
\VersePair{41:14}{When six years are accomplished, thou shalt set free thy brother the Hebrew, who shall be sold to thee: for he shall serve thee six years, and then thou shalt let him go free: but they hearkened not to me, and inclined not their ear.}{34:14}{At the end of seven years, every man of you shall release his brother who is a Hebrew, who has been sold to you, and has served you six years. You shall let him go free from you. But your fathers didn’t listen to me, and didn’t incline their ear.}
\VersePair{41:15}{And this day they turned to do that which was right in my sight, to proclaim every one the release of his neighbour; and they had concluded a covenant before me, in the house whereon my name is called.}{34:15}{You had now turned, and had done that which is right in my eyes, in every man proclaiming liberty to his neighbor. You had made a covenant before me in the house which is called by my name;}
\VersePair{41:16}{But ye turned and profaned my name, to bring back every one his servant, and every one his handmaid, whom ye had sent forth free and at their own disposal, to be to you men-servants and maid-servants.}{34:16}{but you turned and profaned my name, and every man caused his servant and every man his handmaid, whom you had let go free at their pleasure, to return. You brought them into subjection, to be to you for servants and for handmaids.’”}
\VersePair{41:17}{Therefore thus said the Lord; Ye have not hearkened to me, to proclaim a release every one to his neighbour: behold, I proclaim a release to you, to the sword, and to the pestilence, and to the famine; and I will give you up to dispersion among all the kingdoms of the earth.}{34:17}{Therefore Yahweh says: “You have not listened to me, to proclaim liberty, every man to his brother, and every man to his neighbor. Behold, I proclaim to you a liberty,” says Yahweh, “to the sword, to the pestilence, and to the famine. I will make you be tossed back and forth among all the kingdoms of the earth.}
\VersePair{41:18}{And I will give the men that have transgressed my covenant, who have not kept my covenant, which they made before me, the calf which they prepared to sacrifice with it,}{34:18}{I will give the men who have transgressed my covenant, who have not performed the words of the covenant which they made before me when they cut the calf in two and passed between its parts:}
\VersePair{41:19}{the princes of Juda, and the men in power, and the priests, and the people;}{34:19}{the princes of Judah, the princes of Jerusalem, the eunuchs, the priests, and all the people of the land, who passed between the parts of the calf.}
\VersePair{41:20}{I will even give them to their enemies, and their carcases shall be food for the birds of the sky and for the wild beasts of the earth.}{34:20}{I will even give them into the hand of their enemies and into the hand of those who seek their life. Their dead bodies will be food for the birds of the sky and for the animals of the earth.}
\VersePair{41:21}{And I will give Sedekias king of Judea, and their princes, into the hands of their enemies, and the host of the king of Babylon shall come upon them that run away from them.}{34:21}{“I will give Zedekiah king of Judah and his princes into the hands of their enemies, into the hands of those who seek their life and into the hands of the king of Babylon’s army, who has gone away from you.}
\VersePair{41:22}{Behold, I will give command, saith the Lord, and will bring them back to this land; and they shall fight against it, and take it, and burn it with fire, and the cities of Juda; and I will make them desolate without inhabitants.}{34:22}{Behold, I will command,” says Yahweh, “and cause them to return to this city. They will fight against it, take it, and burn it with fire. I will make the cities of Judah a desolation, without inhabitant.”}
\end{paracol}
\ChapterHeading{42}
\begin{paracol}{2}
\VersePair{42:1}{The Word that came to Jeremias from the Lord in the days of Joakim, king of Juda, saying,}{35:1}{The word which came to Jeremiah from Yahweh in the days of Jehoiakim the son of Josiah, king of Judah, saying,}
\VersePair{42:2}{Go to the house of the Archabin, and thou shalt bring them to the house of the Lord, into one of the courts, and give them wine to drink.}{35:2}{“Go to the house of the Rechabites, and speak to them, and bring them into Yahweh’s house, into one of the rooms, and give them wine to drink.”}
\VersePair{42:3}{So I brought forth Jechonias the son of Jeremin the son of Chabasin, and his brethren, and his sons, and all the family of the Archabin;}{35:3}{Then I took Jaazaniah the son of Jeremiah, the son of Habazziniah, with his brothers, all his sons, and the whole house of the Rechabites;}
\VersePair{42:4}{and I brought them into the house of the Lord, into the chamber of the sons of Joanan, the son of Ananias, the son of Godolias, a man of God, who dwells near the house of the princes that are over the house of Maasæas the son of Selom, who kept the court.}{35:4}{and I brought them into Yahweh’s house, into the room of the sons of Hanan the son of Igdaliah, the man of God, which was by the room of the princes, which was above the room of Maaseiah the son of Shallum, the keeper of the threshold.}
\VersePair{42:5}{And I set before them a jar of wine, and cups, and I said, Drink ye wine.}{35:5}{I set before the sons of the house of the Rechabites bowls full of wine, and cups; and I said to them, “Drink wine!”}
\VersePair{42:6}{But they said, We will on no account drink wine, for our father Jonadab the son of Rechab commanded us, saying, Ye shall on no account drink wine, neither ye, nor your sons for ever:}{35:6}{But they said, “We will drink no wine; for Jonadab the son of Rechab, our father, commanded us, saying, ‘You shall drink no wine, neither you nor your children, forever.}
\VersePair{42:7}{nor shall ye at all build houses, nor sow any seed, nor shall ye have a vineyard: for ye shall dwell in tents all your days; that ye may live many days upon the land, in which ye sojourn.}{35:7}{You shall not build a house, sow seed, plant a vineyard, or have any; but all your days you shall dwell in tents, that you may live many days in the land in which you live as nomads.’}
\VersePair{42:8}{And we hearkened to the voice of Jonadab our father, so as to drink no wine all our days, we, and our wives, and our sons, and our daughters;}{35:8}{We have obeyed the voice of Jonadab the son of Rechab, our father, in all that he commanded us, to drink no wine all our days, we, our wives, our sons, or our daughters;}
\VersePair{42:9}{and so as to build no houses to dwell in: and we have had no vineyard, nor field, nor seed:}{35:9}{and not to build houses for ourselves to dwell in. We have no vineyard, field, or seed;}
\VersePair{42:10}{but we have dwelt in tents, and have hearkened, and done according to all that Jonadab our father commanded us.}{35:10}{but we have lived in tents, and have obeyed, and done according to all that Jonadab our father commanded us.}
\VersePair{42:11}{And it came to pass, when Nabuchodonosor came up against the land, that we said we would come in; and we entered into Jerusalem, for fear of the host of the Chaldeans, and for fear of the host of the Assyrians: and we dwelt there.}{35:11}{But when Nebuchadnezzar king of Babylon came up into the land, we said, ‘Come! Let’s go to Jerusalem for fear of the army of the Chaldeans, and for fear of the army of the Syrians; so we will dwell at Jerusalem.’”}
\VersePair{42:12}{And the word of the Lord came to me, saying,}{35:12}{Then Yahweh’s word came to Jeremiah, saying,}
\VersePair{42:13}{Thus saith the Lord, Go, and say to the men of Juda, and to them that dwell in Jerusalem, Will ye not receive correction to hearken to my words?}{35:13}{“Yahweh of Armies, the God of Israel, says: ‘Go and tell the men of Judah and the inhabitants of Jerusalem, “Will you not receive instruction to listen to my words?” says Yahweh.}
\VersePair{42:14}{The sons of Jonadab the son of Rechab have kept the word which he commanded his children, that they should drink no wine; and they have not drunk it: but I spoke to you early, and ye hearkened not.}{35:14}{“The words of Jonadab the son of Rechab that he commanded his sons, not to drink wine, are performed; and to this day they drink none, for they obey their father’s commandment; but I have spoken to you, rising up early and speaking, and you have not listened to me.}
\VersePair{42:15}{And I sent to you my servants the prophets, saying, Turn ye every one from his evil way, and amend your practices, and go not after other gods to serve them, and ye shall dwell upon the land which I gave to you and to your fathers: but ye inclined not your ears, and hearkened not.}{35:15}{I have sent also to you all my servants the prophets, rising up early and sending them, saying, ‘Every one of you must return now from his evil way, amend your doings, and don’t go after other gods to serve them. Then you will dwell in the land which I have given to you and to your fathers;’ but you have not inclined your ear, nor listened to me.}
\VersePair{42:16}{But the sons of Jonadab the son of Rechab have kept the command of their father; but this people has not hearkened to me.}{35:16}{The sons of Jonadab the son of Rechab have performed the commandment of their father which he commanded them, but this people has not listened to me.”’}
\VersePair{42:17}{Therefore thus saith the Lord; Behold, I will bring upon Juda and upon the inhabitants of Jerusalem all the evils which I pronounced against them.}{35:17}{“Therefore Yahweh, the God of Armies, the God of Israel, says: ‘Behold, I will bring on Judah and on all the inhabitants of Jerusalem all the evil that I have pronounced against them, because I have spoken to them, but they have not heard; and I have called to them, but they have not answered.’”}
\VersePair{42:18}{Therefore thus saith the Lord; Since the sons of Jonadab the son of Rechab have hearkened to the command of their father, to do as their father commanded them:}{35:18}{Jeremiah said to the house of the Rechabites, “Yahweh of Armies, the God of Israel, says: ‘Because you have obeyed the commandment of Jonadab your father, and kept all his precepts, and done according to all that he commanded you,’}
\VersePair{42:19}{there shall never be wanting a man of the sons of Jonadab the son of Rechab to stand before my face while the earth remains.}{35:19}{therefore Yahweh of Armies, the God of Israel, says: ‘Jonadab the son of Rechab will not lack a man to stand before me forever.’”}
\end{paracol}
\ChapterHeading{43}
\begin{paracol}{2}
\VersePair{43:1}{In the fourth Year of Joakim son of Josias king of Juda, the word of the Lord came to me, saying,}{36:1}{In the fourth year of Jehoiakim the son of Josiah, king of Judah, this word came to Jeremiah from Yahweh, saying,}
\VersePair{43:2}{Take thee a roll of a book, and write upon it all the words which I spoke to thee against Jerusalem, and against Juda, and against all the nations, from the day when I spoke to thee, from the days of Josias king of Juda, even to this day.}{36:2}{“Take a scroll of a book, and write in it all the words that I have spoken to you against Israel, against Judah, and against all the nations, from the day I spoke to you, from the days of Josiah even to this day.}
\VersePair{43:3}{Perhaps the house of Juda will hear all the evils which I purpose to do to them; that they may turn from their evil way; and so I will be merciful to their iniquities and their sins.}{36:3}{It may be that the house of Judah will hear all the evil which I intend to do to them, that they may each return from his evil way; that I may forgive their iniquity and their sin.”}
\VersePair{43:4}{So Jeremias called Baruch the son of Nerias: and he wrote from the mouth of Jeremias all the words of the Lord, which he had spoken to him, on a roll of a book.}{36:4}{Then Jeremiah called Baruch the son of Neriah; and Baruch wrote from the mouth of Jeremiah all Yahweh’s words, which he had spoken to him, on a scroll of a book.}
\VersePair{43:5}{And Jeremias commanded Baruch, saying, I am in prison; I cannot enter into the house of the Lord:}{36:5}{Jeremiah commanded Baruch, saying, “I am restricted. I can’t go into Yahweh’s house.}
\VersePair{43:6}{so thou shalt read in this roll in the ears of the people in the house of the Lord, on the fast day; and in the ears of all Juda that come out of their cities, thou shalt read to them.}{36:6}{Therefore you go, and read from the scroll which you have written from my mouth, Yahweh’s words, in the ears of the people in Yahweh’s house on the fast day. Also you shall read them in the ears of all Judah who come out of their cities.}
\VersePair{43:7}{Peradventure their supplication will come before the Lord, and they will turn from their evil way: for great is the wrath and the anger of the Lord, which he has pronounced against this people.}{36:7}{It may be they will present their supplication before Yahweh, and will each return from his evil way; for Yahweh has pronounced great anger and wrath against this people.”}
\VersePair{43:8}{And Baruch did according to all that Jeremias commanded him—reading in the book the words of the Lord in the Lord's house.}{36:8}{Baruch the son of Neriah did according to all that Jeremiah the prophet commanded him, reading in the book Yahweh’s words in Yahweh’s house.}
\VersePair{43:9}{And it came to pass in the eighth year of king Joakim, in the ninth month, all the people in Jerusalem, and the house of Juda, proclaimed a fast before the Lord.}{36:9}{Now in the fifth year of Jehoiakim the son of Josiah, king of Judah, in the ninth month, all the people in Jerusalem and all the people who came from the cities of Judah to Jerusalem, proclaimed a fast before Yahweh.}
\VersePair{43:10}{And Baruch read in the book the words of Jeremias in the house of the Lord, in the house of Gamarias son of Saphan the scribe, in the upper court, in the entrance of the new gate of the house of the Lord, and in the ears of all the people.}{36:10}{Then Baruch read the words of Jeremiah from the book in Yahweh’s house, in the room of Gemariah the son of Shaphan the scribe, in the upper court, at the entry of the new gate of Yahweh’s house, in the ears of all the people.}
\VersePair{43:11}{And Michæas the son of Gamarias the son of Saphan heard all the words of the Lord, out of the book.}{36:11}{When Micaiah the son of Gemariah, the son of Shaphan, had heard out of the book all Yahweh’s words,}
\VersePair{43:12}{And he went down to the king's house, into the house of the scribe: and, behold, there were sitting there all the princes, Elisama the scribe, and Dalæas the son of Selemias, and Jonathan the son of Acchobor, and Gamarias the son of Saphan, and Sedekias the son of Ananias, and all the princes.}{36:12}{he went down into the king’s house, into the scribe’s room; and behold, all the princes were sitting there, Elishama the scribe, Delaiah the son of Shemaiah, Elnathan the son of Achbor, Gemariah the son of Shaphan, Zedekiah the son of Hananiah, and all the princes.}
\VersePair{43:13}{And Michæas reported to them all the words which he had heard Baruch reading in the ears of the people.}{36:13}{Then Micaiah declared to them all the words that he had heard, when Baruch read the book in the ears of the people.}
\VersePair{43:14}{And all the princes sent to Baruch son of Nerias Judin the son of Nathanias, the son of Selemias, the son of Chusi, saying, Take in thine hand the roll in which thou readest in the ears of the people, and come. So Baruch took the roll, and went down to them.}{36:14}{Therefore all the princes sent Jehudi the son of Nethaniah, the son of Shelemiah, the son of Cushi, to Baruch, saying, “Take in your hand the scroll in which you have read in the ears of the people, and come.” So Baruch the son of Neriah took the scroll in his hand, and came to them.}
\VersePair{43:15}{And they said to him, Read it again in our ears. And Baruch read it.}{36:15}{They said to him, “Sit down now, and read it in our hearing.” So Baruch read it in their hearing.}
\VersePair{43:16}{And it came to pass, when they had heard all the words, that they took counsel each with his neighbour, and said, Let us by all means tell the king all these words.}{36:16}{Now when they had heard all the words, they turned in fear one toward another, and said to Baruch, “We will surely tell the king of all these words.”}
\VersePair{43:17}{And they asked Baruch, saying, Where didst thou write all these words?}{36:17}{They asked Baruch, saying, “Tell us now, how did you write all these words at his mouth?”}
\VersePair{43:18}{And Baruch said, Jeremias told me from his own mouth all these words, and I wrote them in a book.}{36:18}{Then Baruch answered them, “He dictated all these words to me with his mouth, and I wrote them with ink in the book.”}
\VersePair{43:19}{And they said to Baruch, Go, and hide, thou and Jeremias; let no man know where ye are.}{36:19}{Then the princes said to Baruch, “You and Jeremiah go hide. Don’t let anyone know where you are.”}
\VersePair{43:20}{And they went in to the king into the court, and gave the roll to one to keep in the house of Elisama; and they told the king all these words.}{36:20}{They went in to the king into the court, but they had laid up the scroll in the room of Elishama the scribe. Then they told all the words in the hearing of the king.}
\VersePair{43:21}{And the king sent Judin to fetch the roll: and he took it out of the house of Elisama: and Judin read in the ears of the king, and in the ears of all the princes who stood round the king.}{36:21}{So the king sent Jehudi to get the scroll, and he took it out of the room of Elishama the scribe. Jehudi read it in the hearing of the king, and in the hearing of all the princes who stood beside the king.}
\VersePair{43:22}{Now the king was sitting in the winter house: and there was a fire on the hearth before him.}{36:22}{Now the king was sitting in the winter house in the ninth month, and there was a fire in the brazier burning before him.}
\VersePair{43:23}{And it came to pass when Judin had read three or four leaves, he cut them off with a penknife, and cast them into the fire that was on the hearth, until the whole roll was consumed in the fire that was on the hearth.}{36:23}{When Jehudi had read three or four columns, the king cut it with the penknife, and cast it into the fire that was in the brazier, until all the scroll was consumed in the fire that was in the brazier.}
\VersePair{43:24}{And the king and his servants that heard all these words sought not the Lord, and rent not their garments.}{36:24}{The king and his servants who heard all these words were not afraid, and didn’t tear their garments.}
\VersePair{43:25}{But Elnathan and Godolias suggested to the king that he should burn the roll.}{36:25}{Moreover Elnathan and Delaiah and Gemariah had made intercession to the king that he would not burn the scroll; but he would not listen to them.}
\VersePair{43:26}{And the king commanded Jeremeel the king's son, and Saræas the son of Esriel, to take Baruch and Jeremias: but they were hidden.}{36:26}{The king commanded Jerahmeel the king’s son, and Seraiah the son of Azriel, and Shelemiah the son of Abdeel, to arrest Baruch the scribe and Jeremiah the prophet; but Yahweh hid them.}
\VersePair{43:27}{Then the word of the Lord came to Jeremias, after the king had burnt the roll, even all the words which Baruch wrote from the mouth of Jeremias, saying,}{36:27}{Then Yahweh’s word came to Jeremiah, after the king had burned the scroll, and the words which Baruch wrote at the mouth of Jeremiah, saying,}
\VersePair{43:28}{Again take thou another roll, and write all the words that were on the roll, which king Joakim has burnt.}{36:28}{“Take again another scroll, and write in it all the former words that were in the first scroll, which Jehoiakim the king of Judah has burned.}
\VersePair{43:29}{And thou shalt say, Thus saith the Lord; Thou hast burnt this roll, saying, Why hast thou written therein, saying, The king of Babylon shall certainly come in, and destroy this land, and man and cattle shall fail from off it?}{36:29}{Concerning Jehoiakim king of Judah you shall say, ‘Yahweh says: “You have burned this scroll, saying, ‘Why have you written therein, saying, “The king of Babylon will certainly come and destroy this land, and will cause to cease from there man and animal ”?’”}
\VersePair{43:30}{Therefore thus saith the Lord concerning Joakim king of Juda; He shall not have a man to sit on the throne of David: and his carcase shall be cast forth in the heat by day, and in the frost by night.}{36:30}{Therefore Yahweh says concerning Jehoiakim king of Judah: “He will have no one to sit on David’s throne. His dead body will be cast out in the day to the heat, and in the night to the frost.}
\VersePair{43:31}{And I will visit him, and his family, and his servants: and I will bring upon him, and upon the inhabitants of Jerusalem, and upon the land of Juda, all the evils which I spoke of to them; and they hearkened not.}{36:31}{I will punish him, his offspring, and his servants for their iniquity. I will bring on them, on the inhabitants of Jerusalem, and on the men of Judah, all the evil that I have pronounced against them, but they didn’t listen.”’”}
\VersePair{43:32}{And Baruch took another roll, and wrote upon it from the mouth of Jeremias all the words of the book which Joakim had burnt: and there were yet more words added to it like the former.}{36:32}{Then Jeremiah took another scroll, and gave it to Baruch the scribe, the son of Neriah, who wrote therein from the mouth of Jeremiah all the words of the book which Jehoiakim king of Judah had burned in the fire; and many similar words were added to them.}
\end{paracol}
\ChapterHeading{44}
\begin{paracol}{2}
\VersePair{44:1}{And Sedekias the son of Josias reigned instead of Joakim, whom Nabuchodonosor appointed to reign over Juda.}{37:1}{Zedekiah the son of Josiah reigned as king instead of Coniah the son of Jehoiakim, whom Nebuchadnezzar king of Babylon made king in the land of Judah.}
\VersePair{44:2}{And he and his servants and the people of the land hearkened not to the words of the Lord, which he spoke by Jeremias.}{37:2}{But neither he, nor his servants, nor the people of the land, listened to Yahweh’s words, which he spoke by the prophet Jeremiah.}
\VersePair{44:3}{And king Sedekias sent Joachal son of Selemias and Sophonias the priest son of Maasæas to Jeremias, saying, Pray now for us to the Lord.}{37:3}{Zedekiah the king sent Jehucal the son of Shelemiah and Zephaniah the son of Maaseiah, the priest, to the prophet Jeremiah, saying, “Pray now to Yahweh our God for us.”}
\VersePair{44:4}{Now Jeremias came and went through the midst of the city: for they had not put him into the house of the prison.}{37:4}{Now Jeremiah came in and went out among the people, for they had not put him into prison.}
\VersePair{44:5}{And the host of Pharao was come forth out of Egypt; and the Chaldeans heard the report of them, and they went up from Jerusalem.}{37:5}{Pharaoh’s army had come out of Egypt; and when the Chaldeans who were besieging Jerusalem heard news of them, they withdrew from Jerusalem.}
\VersePair{44:6}{And the word of the Lord came to Jeremias, saying,}{37:6}{Then Yahweh’s word came to the prophet Jeremiah, saying,}
\VersePair{44:7}{Thus said the Lord; Thus shalt thou say to the king of Juda who sent to thee, to seek me; Behold, the army of Pharao which is come forth to help you: they shall return to the land of Egypt:}{37:7}{“Yahweh, the God of Israel, says, ‘You shall tell the king of Judah, who sent you to me to inquire of me: “Behold, Pharaoh’s army, which has come out to help you, will return to Egypt into their own land.}
\VersePair{44:8}{and the Chaldeans themselves shall turn again, and fight against this city, and take it, and burn it with fire.}{37:8}{The Chaldeans will come again, and fight against this city. They will take it and burn it with fire.”’}
\VersePair{44:9}{For thus saith the Lord; Suppose not in your hearts, saying, The Chaldeans will certainly depart from us: for they shall not depart.}{37:9}{“Yahweh says, ‘Don’t deceive yourselves, saying, “The Chaldeans will surely depart from us;” for they will not depart.}
\VersePair{44:10}{And though ye should smite the whole host of the Chaldeans that fight against you, and there should be left a few wounded men, these should rise up each in his place, and burn this city with fire.}{37:10}{For though you had struck the whole army of the Chaldeans who fight against you, and only wounded men remained among them, they would each rise up in his tent and burn this city with fire.’”}
\VersePair{44:11}{And it came to pass, when the host of the Chaldeans had gone up from Jerusalem for fear of the host of Pharao,}{37:11}{When the army of the Chaldeans had withdrawn from Jerusalem for fear of Pharaoh’s army,}
\VersePair{44:12}{that Jeremias went forth from Jerusalem to go into the land of Benjamin, to buy thence a property in the midst of the people.}{37:12}{then Jeremiah went out of Jerusalem to go into the land of Benjamin, to receive his portion there, in the middle of the people.}
\VersePair{44:13}{And he was in the gate of Benjamin, and there was there a man with whom he lodged, Saruia the son of Selemias, the son of Ananias; and he caught Jeremias, saying, Thou art fleeing to the Chaldeans.}{37:13}{When he was in Benjamin’s gate, a captain of the guard was there, whose name was Irijah, the son of Shelemiah, the son of Hananiah; and he seized Jeremiah the prophet, saying, “You are defecting to the Chaldeans!”}
\VersePair{44:14}{And he said, It is false; I do not flee to the Chaldeans. But he hearkened not to him; and Saruia caught Jeremias, and brought him to the princes.}{37:14}{Then Jeremiah said, “That is false! I am not defecting to the Chaldeans.” But he didn’t listen to him; so Irijah seized Jeremiah, and brought him to the princes.}
\VersePair{44:15}{And the princes were very angry with Jeremias, and smote him, and sent him into the house of Jonathan the scribe: for they had made this a prison.}{37:15}{The princes were angry with Jeremiah, and struck him, and put him in prison in the house of Jonathan the scribe; for they had made that the prison.}
\VersePair{44:16}{So Jeremias came into the dungeon, and into the cells, and he remained there many days.}{37:16}{When Jeremiah had come into the dungeon house and into the cells, and Jeremiah had remained there many days,}
\VersePair{44:17}{Then Sedekias sent, and called him; and the king asked him secretly, saying, Is there a word from the Lord? and he said, There is: thou shalt be delivered into the hands of the king of Babylon.}{37:17}{then Zedekiah the king sent and had him brought out. The king asked him secretly in his house, “Is there any word from Yahweh?” Jeremiah said, “There is.” He also said, “You will be delivered into the hand of the king of Babylon.”}
\VersePair{44:18}{And Jeremias said to the king, Wherein have I wronged thee, or thy servants, or this people, that thou puttest me in prison?}{37:18}{Moreover Jeremiah said to King Zedekiah, “How have I sinned against you, against your servants, or against this people, that you have put me in prison?}
\VersePair{44:19}{And where are your prophets who prophesied to you saying, The king of Babylon shall not come against this land?}{37:19}{Now where are your prophets who prophesied to you, saying, ‘The king of Babylon will not come against you, nor against this land’?}
\VersePair{44:20}{Now therefore, my lord the king, let my supplication come before thy face: and why dost thou send me back to the house of Jonathan the scribe? and let me not on any account die there.}{37:20}{Now please hear, my lord the king: please let my supplication be presented before you, that you not cause me to return to the house of Jonathan the scribe, lest I die there.”}
\VersePair{44:21}{Then the king commanded, and they cast him into the prison, and gave him a loaf a day out of the place where they bake, until the bread failed out of the city. So Jeremias continued in the court of the prison.}{37:21}{Then Zedekiah the king commanded, and they committed Jeremiah into the court of the guard. They gave him daily a loaf of bread out of the bakers’street, until all the bread in the city was gone. Thus Jeremiah remained in the court of the guard.}
\end{paracol}
\ChapterHeading{45}
\begin{paracol}{2}
\VersePair{45:1}{And Saphanias the son of Nathan, and Godolias the son of Paschor, and Joachal the son of Semelias, heard the words which Jeremias spoke to the people, saying,}{38:1}{Shephatiah the son of Mattan, Gedaliah the son of Pashhur, Jucal the son of Shelemiah, and Pashhur the son of Malchijah heard the words that Jeremiah spoke to all the people, saying,}
\VersePair{45:2}{Thus saith the Lord; He that remains in this city shall die by the sword, and by the famine: but he that goes out to the Chaldeans shall live; and his soul shall be given him for a found treasure, and he shall live.}{38:2}{“Yahweh says, ‘He who remains in this city will die by the sword, by the famine, and by the pestilence, but he who goes out to the Chaldeans will live. He will escape with his life and he will live.’}
\VersePair{45:3}{For thus saith the Lord; This city shall certainly be delivered into the hands of the host of the king of Babylon, and they shall take it.}{38:3}{Yahweh says, ‘This city will surely be given into the hand of the army of the king of Babylon, and he will take it.’”}
\VersePair{45:4}{And they said to the king, Let that man, we pray thee, be slain, for he weakens the hands of the fighting men that are left in the city, and the hands of all the people, speaking to them according to these words: for this man does not prophesy peace to this people, but evil.}{38:4}{Then the princes said to the king, “Please let this man be put to death, because he weakens the hands of the men of war who remain in this city, and the hands of all the people, in speaking such words to them; for this man doesn’t seek the welfare of this people, but harm.”}
\VersePair{45:5}{Then the king said, Behold, he is in your hands. For the king could not resist them.}{38:5}{Zedekiah the king said, “Behold, he is in your hand; for the king can’t do anything to oppose you.”}
\VersePair{45:6}{And they cast him into the dungeon of Melchias the king's son, which was in the court of the prison; and they let him down into the pit: and there was no water in the pit, but mire: and he was in the mire.}{38:6}{Then they took Jeremiah and threw him into the dungeon of Malchijah the king’s son, that was in the court of the guard. They let down Jeremiah with cords. In the dungeon there was no water, but mire; and Jeremiah sank in the mire.}
\VersePair{45:7}{And Abdemelech the Ethiopian heard, (now he was in the king's household,) that they had put Jeremias into the dungeon; and the king was in the gate of Benjamin:}{38:7}{Now when Ebedmelech the Ethiopian, a eunuch, who was in the king’s house, heard that they had put Jeremiah in the dungeon ( the king was then sitting in Benjamin’s gate ),}
\VersePair{45:8}{and he went forth to him, and spoke to the king and said,}{38:8}{Ebedmelech went out of the king’s house, and spoke to the king, saying,}
\VersePair{45:9}{Thou hast done evil in what thou hast done to slay this man with hunger: for there is no more bread in the city.}{38:9}{“My lord the king, these men have done evil in all that they have done to Jeremiah the prophet, whom they have cast into the dungeon. He is likely to die in the place where he is, because of the famine; for there is no more bread in the city.”}
\VersePair{45:10}{And the king commanded Abdemelech, saying, Take with thee hence thirty men, and bring him up out of the dungeon, that he die not.}{38:10}{Then the king commanded Ebedmelech the Ethiopian, saying, “Take from here thirty men with you, and take up Jeremiah the prophet out of the dungeon, before he dies.”}
\VersePair{45:11}{So Abdemelech took the men, and went into the underground part of the king's house, and took thence old rags and old ropes, and threw them to Jeremias into the dungeon.}{38:11}{So Ebedmelech took the men with him, and went into the house of the king under the treasury, and took from there rags and worn-out garments, and let them down by cords into the dungeon to Jeremiah.}
\VersePair{45:12}{And he said, Put these under the ropes. And Jeremias did so.}{38:12}{Ebedmelech the Ethiopian said to Jeremiah, “Now put these rags and worn-out garments under your armpits under the cords.” Jeremiah did so.}
\VersePair{45:13}{And they drew him with the ropes, and lifted him out of the dungeon: and Jeremias remained in the court of the prison.}{38:13}{So they lifted Jeremiah up with the cords, and took him up out of the dungeon; and Jeremiah remained in the court of the guard.}
\VersePair{45:14}{Then the king sent, and called him to himself into the house of Aselisel, which was in the house of the Lord: and the king said to him, I will ask thee a question, and I pray thee hide nothing from me.}{38:14}{Then Zedekiah the king sent and took Jeremiah the prophet to himself into the third entry that is in Yahweh’s house. Then the king said to Jeremiah, “I will ask you something. Hide nothing from me.”}
\VersePair{45:15}{And Jeremias said to the king, If I tell thee, wilt thou not certainly put me to death? and if I give thee counsel, thou wilt not at all hearken to me.}{38:15}{Then Jeremiah said to Zedekiah, “If I declare it to you, will you not surely put me to death? If I give you counsel, you will not listen to me.”}
\VersePair{45:16}{And the king swore to him, saying, As the Lord lives who gave us this soul, I will not slay thee, neither will I give thee into the hands of these men.}{38:16}{So Zedekiah the king swore secretly to Jeremiah, saying, “As Yahweh lives, who made our souls, I will not put you to death, neither will I give you into the hand of these men who seek your life.”}
\VersePair{45:17}{And Jeremias said to him, Thus saith the Lord; If thou wilt indeed go forth to the captains of the king of Babylon, thy soul shall live, and this city shall certainly not be burnt with fire; and thou shalt live, and thy house.}{38:17}{Then Jeremiah said to Zedekiah, “Yahweh, the God of Armies, the God of Israel, says: ‘If you will go out to the king of Babylon’s princes, then your soul will live, and this city will not be burned with fire. You will live, along with your house.}
\VersePair{45:18}{But if thou wilt not go forth, this city shall be delivered into the hands of the Chaldeans, and they shall burn it with fire, and thou shalt by no means escape.}{38:18}{But if you will not go out to the king of Babylon’s princes, then this city will be given into the hand of the Chaldeans, and they will burn it with fire, and you won’t escape out of their hand.’”}
\VersePair{45:19}{And the king said to Jeremias, I consider the Jews that have gone over to the Chaldeans, lest they deliver me into their hands, and they mock me.}{38:19}{Zedekiah the king said to Jeremiah, “I am afraid of the Jews who have defected to the Chaldeans, lest they deliver me into their hand, and they mock me.”}
\VersePair{45:20}{And Jeremias said, They shall in no wise deliver thee up. Hear the word of the Lord which I speak to thee; and it shall be better for thee, and thy soul shall live.}{38:20}{But Jeremiah said, “They won’t deliver you. Obey, I beg you, Yahweh’s voice, in that which I speak to you; so it will be well with you, and your soul will live.}
\VersePair{45:21}{But if thou wilt not go forth, this is the word which the Lord has shewn me.}{38:21}{But if you refuse to go out, this is the word that Yahweh has shown me:}
\VersePair{45:22}{And, behold, all the women that are left in the house of the king of Juda were brought forth to the princes of the king of Babylon; and they said, The men who were at peace with thee have deceived thee, and will prevail against thee; and they shall cause thy foot to slide and fail, they have turned back from thee.}{38:22}{{\raggedright ‘Behold, all the women who are left in the king of Judah’s house will be brought out to the king of Babylon’s princes, and those women will say, \poemline{1}{“Your familiar friends have turned on you,}\poemline{2}{and have prevailed over you.}\poemline{1}{Your feet are sunk in the mire,}\poemline{2}{they have turned away from you.”}}}
\VersePair{45:23}{And they shall bring forth thy wives and thy children to the Chaldeans: and thou shalt by no means escape, for thou shalt be taken by the hand of the king of Babylon, and this city shall be burnt.}{38:23}{They will bring out all your wives and your children to the Chaldeans. You won’t escape out of their hand, but will be taken by the hand of the king of Babylon. You will cause this city to be burned with fire.’”}
\VersePair{45:24}{Then the king said to him, Let no man know any of these words, and certainly thou shalt not die.}{38:24}{Then Zedekiah said to Jeremiah, “Let no man know of these words, and you won’t die.}
\VersePair{45:25}{And if the princes shall hear that I have spoken to thee, and they come to thee, and say to thee, Tell us, what said the king to thee? hide it not from us, and we will in no wise slay thee, and what said the king to thee?}{38:25}{But if the princes hear that I have talked with you, and they come to you, and tell you, ‘Declare to us now what you have said to the king; don’t hide it from us, and we will not put you to death; also tell us what the king said to you;’}
\VersePair{45:26}{Then thou shalt say to them, I brought my supplication before the presence of the king, that he would not send me back into the house of Jonathan, that I should die there.}{38:26}{then you shall tell them, ‘I presented my supplication before the king, that he would not cause me to return to Jonathan’s house, to die there.’”}
\VersePair{45:27}{And all the princes came to Jeremias, and asked him: and he told them according to all these words, which the king had commanded him. And they were silent, because the word of the Lord was not heard.}{38:27}{Then all the princes came to Jeremiah, and asked him; and he told them according to all these words that the king had commanded. So they stopped speaking with him, for the matter was not perceived.}
\VersePair{45:28}{And Jeremias remained in the court of the prison, until the time when Jerusalem was taken.}{38:28}{So Jeremiah stayed in the court of the guard until the day that Jerusalem was taken.}
\end{paracol}
\ChapterHeading{46}
\begin{paracol}{2}
\VersePair{46:1}{And it came to pass in the ninth month of Sedekias king of Juda, that Nabuchodonosor king of Babylon came, and all his host, against Jerusalem, and they besieged it.}{39:1}{In the ninth year of Zedekiah king of Judah, in the tenth month, Nebuchadnezzar king of Babylon and all his army came against Jerusalem, and besieged it.}
\VersePair{46:2}{And in the eleventh year of Sedekias, in the fourth month, on the ninth day of the month, the city was broken up.}{39:2}{In the eleventh year of Zedekiah, in the fourth month, the ninth day of the month, a breach was made in the city.}
\VersePair{46:3}{And all the leaders of the king of Babylon went in, and sat in the middle gate, Marganasar, and Samagoth, and Nabusachar, and Nabusaris, Nagargas, Naserrabamath, and the rest of the leaders of the king of Babylon,}{39:3}{All the princes of the king of Babylon came in, and sat in the middle gate: Nergal Sharezer, Samgarnebo, Sarsechim the Rabsaris, Nergal Sharezer the Rabmag, with all the rest of the princes of the king of Babylon.}
\VersePair{46:4}{}{39:4}{When Zedekiah the king of Judah and all the men of war saw them, then they fled and went out of the city by night, by the way of the king’s garden, through the gate between the two walls; and he went out toward the Arabah.}
\VersePair{46:5}{}{39:5}{But the army of the Chaldeans pursued them, and overtook Zedekiah in the plains of Jericho. When they had taken him, they brought him up to Nebuchadnezzar king of Babylon to Riblah in the land of Hamath; and he pronounced judgment on him.}
\VersePair{46:6}{}{39:6}{Then the king of Babylon killed Zedekiah’s sons in Riblah before his eyes. The king of Babylon also killed all the nobles of Judah.}
\VersePair{46:7}{}{39:7}{Moreover he put out Zedekiah’s eyes and bound him in fetters, to carry him to Babylon.}
\VersePair{46:8}{}{39:8}{The Chaldeans burned the king’s house and the people’s houses with fire and broke down the walls of Jerusalem.}
\VersePair{46:9}{}{39:9}{Then Nebuzaradan the captain of the guard carried away captive into Babylon the rest of the people who remained in the city, the deserters also who fell away to him, and the rest of the people who remained.}
\VersePair{46:10}{}{39:10}{But Nebuzaradan the captain of the guard left of the poor of the people, who had nothing, in the land of Judah, and gave them vineyards and fields at the same time.}
\VersePair{46:11}{}{39:11}{Now Nebuchadnezzar king of Babylon commanded Nebuzaradan the captain of the guard concerning Jeremiah, saying,}
\VersePair{46:12}{}{39:12}{“Take him and take care of him. Do him no harm; but do to him even as he tells you.”}
\VersePair{46:13}{}{39:13}{So Nebuzaradan the captain of the guard, Nebushazban, Rabsaris, and Nergal Sharezer, Rabmag, and all the chief officers of the king of Babylon}
\VersePair{46:14}{and they sent, and took Jeremias out of the court of the prison, and gave him in charge to Godolias the son of Achicam, the son of Saphan: and they brought him out, and he sat in the midst of the people.}{39:14}{sent and took Jeremiah out of the court of the guard, and committed him to Gedaliah the son of Ahikam, the son of Shaphan, that he should bring him home. So he lived among the people.}
\VersePair{46:15}{And the word of the Lord came to Jeremias in the court of the prison, saying,}{39:15}{Now Yahweh’s word came to Jeremiah while he was shut up in the court of the guard, saying,}
\VersePair{46:16}{Go and say to Abdemelech the Ethiopian, Thus said the Lord God of Israel; Behold, I will bring my words upon this city for evil, and not for good.}{39:16}{“Go, and speak to Ebedmelech the Ethiopian, saying, ‘Yahweh of Armies, the God of Israel, says: “Behold, I will bring my words on this city for evil, and not for good; and they will be accomplished before you in that day.}
\VersePair{46:17}{But I will save thee in that day, and I will by no means deliver thee into the hands of the men before whom thou art afraid.}{39:17}{But I will deliver you in that day,” says Yahweh; “and you will not be given into the hand of the men of whom you are afraid.}
\VersePair{46:18}{For I will surely save thee, and thou shalt by no means fall by the sword; and thou shalt find thy life, because thou didst trust in me, saith the Lord.}{39:18}{For I will surely save you. You won’t fall by the sword, but you will escape with your life, because you have put your trust in me,” says Yahweh.’”}
\end{paracol}
\ChapterHeading{47}
\begin{paracol}{2}
\VersePair{47:1}{The word that came from the Lord to Jeremias, after that Nabuzardan the captain of the guard had let him go out of Rama, when he had taken him in manacles in the midst of the captivity of Juda, even those who were carried to Babylon.}{40:1}{The word which came to Jeremiah from Yahweh, after Nebuzaradan the captain of the guard had let him go from Ramah, when he had taken him being bound in chains among all the captives of Jerusalem and Judah who were carried away captive to Babylon.}
\VersePair{47:2}{And the chief captain of the guard took him, and said to him, The Lord thy God has pronounced all these evils upon this place:}{40:2}{The captain of the guard took Jeremiah and said to him, “Yahweh your God pronounced this evil on this place;}
\VersePair{47:3}{and the Lord has done it; because ye sinned against him, and hearkened not to his voice.}{40:3}{and Yahweh has brought it, and done according as he spoke. Because you have sinned against Yahweh, and have not obeyed his voice, therefore this thing has come on you.}
\VersePair{47:4}{Behold, I have loosed thee from the manacles that were upon thine hands. If it seem good to thee to go with me to Babylon, then will I set mine eyes upon thee.}{40:4}{Now, behold, I release you today from the chains which are on your hand. If it seems good to you to come with me into Babylon, come, and I will take care of you; but if it seems bad to you to come with me into Babylon, don’t. Behold, all the land is before you. Where it seems good and right to you to go, go there.”}
\VersePair{47:5}{But if not, depart; return to Godolias the son of Achicam, the son of Saphan, whom the king of Babylon has appointed governor in the land of Juda, and dwell with him in the midst of the people in the land of Juda: to whatsoever places it seems good in thine eyes to go, do thou even go. And the captain of the guard made him presents, and let him go.}{40:5}{Now while he had not yet gone back, “Go back then,” he said, “to Gedaliah the son of Ahikam, the son of Shaphan, whom the king of Babylon has made governor over the cities of Judah, and dwell with him among the people; or go wherever it seems right to you to go.” So the captain of the guard gave him food and a present, and let him go.}
\VersePair{47:6}{And he came to Godolias to Massepha, and dwelt in the midst of his people that was left in the land.}{40:6}{Then Jeremiah went to Gedaliah the son of Ahikam to Mizpah, and lived with him among the people who were left in the land.}
\VersePair{47:7}{And all the leaders of the host that was in the country, they and their men, heard that the king of Babylon had appointed Godolias governor in the land, and they committed to him the men and their wives, whom Nabuchodonosor had not removed to Babylon.}{40:7}{Now when all the captains of the forces who were in the fields, even they and their men, heard that the king of Babylon had made Gedaliah the son of Ahikam governor in the land, and had committed to him men, women, children, and of the poorest of the land, of those who were not carried away captive to Babylon,}
\VersePair{47:8}{And there came to Godolias to Massepha Ismael the son of Nathanias, and Joanan son of Caree, and Saræas the son of Thanaemeth, and the sons of Jophe the Netophathite, and Ezonias son of the Mochathite, they and their men.}{40:8}{then Ishmael the son of Nethaniah, and Johanan and Jonathan the sons of Kareah, and Seraiah the son of Tanhumeth, and the sons of Ephai the Netophathite, and Jezaniah the son of the Maacathite, they and their men came to Gedaliah to Mizpah.}
\VersePair{47:9}{And Godolias swore to them and to their men, saying, Be not afraid before the children of the Chaldeans: dwell in the land, and serve the king of Babylon, and it shall be better for you.}{40:9}{Gedaliah the son of Ahikam the son of Shaphan swore to them and to their men, saying, “Don’t be afraid to serve the Chaldeans. Dwell in the land, and serve the king of Babylon, and it will be well with you.}
\VersePair{47:10}{And, behold, I dwell in your presence at Massepha, to stand before the Chaldeans who shall come against you: and do ye gather grapes, and fruits, and oil, and put them into your vessels, and dwell in the cities which ye have obtained possession of.}{40:10}{As for me, behold, I will dwell at Mizpah, to stand before the Chaldeans who will come to us; but you, gather wine and summer fruits and oil, and put them in your vessels, and dwell in your cities that you have taken.”}
\VersePair{47:11}{And all the Jews that were in Moab, and among the children of Ammon, and those that were in Idumea, and those that were in all the rest of the country, heard that the king of Babylon had granted a remnant to Juda, and that he had appointed over them Godolias the son of Achicam.}{40:11}{Likewise when all the Jews who were in Moab, and among the children of Ammon, and in Edom, and who were in all the countries, heard that the king of Babylon had left a remnant of Judah, and that he had set over them Gedaliah the son of Ahikam, the son of Shaphan,}
\VersePair{47:12}{And they came to Godolias into the land of Juda, to Massepha, and gathered grapes, and very much summer fruit, and oil.}{40:12}{then all the Jews returned out of all places where they were driven, and came to the land of Judah, to Gedaliah, to Mizpah, and gathered very much wine and summer fruits.}
\VersePair{47:13}{And Joanan the son of Caree, and all the leaders of the host, who were in the fields, came to Godolias to Massepha,}{40:13}{Moreover Johanan the son of Kareah, and all the captains of the forces who were in the fields, came to Gedaliah to Mizpah,}
\VersePair{47:14}{and said to him, Dost thou indeed know that king Beleissa son of Ammon has sent Ismael to thee to slay thee? But Godolias believed them not.}{40:14}{and said to him, “Do you know that Baalis the king of the children of Ammon has sent Ishmael the son of Nethaniah to take your life?” But Gedaliah the son of Ahikam didn’t believe them.}
\VersePair{47:15}{And Joanan said to Godolias secretly in Massepha, I will go now and smite Ismael, and let no man know it; lest he slay thee, and all the Jews that are gathered to thee be dispersed, and the remnant of Juda perish.}{40:15}{Then Johanan the son of Kareah spoke to Gedaliah in Mizpah secretly, saying, “Please let me go, and I will kill Ishmael the son of Nethaniah, and no man will know it. Why should he take your life, that all the Jews who are gathered to you should be scattered, and the remnant of Judah perish?”}
\VersePair{47:16}{But Godolias said to Joanan, Do not the thing, for thou speakest lies concerning Ismael.}{40:16}{But Gedaliah the son of Ahikam said to Johanan the son of Kareah, “You shall not do this thing, for you speak falsely of Ishmael.”}
\end{paracol}
\ChapterHeading{48}
\begin{paracol}{2}
\VersePair{48:1}{Now it came to pass in the seventh month that Ismael the son of Nathanias the son of Eleasa of the seed royal, came, and ten men with him, to Godolias to Massepha: and they ate bread there together.}{41:1}{Now in the seventh month, Ishmael the son of Nethaniah, the son of Elishama, of the royal offspring and one of the chief officers of the king, and ten men with him, came to Gedaliah the son of Ahikam to Mizpah; and there they ate bread together in Mizpah.}
\VersePair{48:2}{And Ismael rose up, and the ten men that were with him, and smote Godolias, whom the king of Babylon had appointed governor over the land,}{41:2}{Then Ishmael the son of Nethaniah arose, and the ten men who were with him, and struck Gedaliah the son of Ahikam the son of Shaphan with the sword and killed him, whom the king of Babylon had made governor over the land.}
\VersePair{48:3}{and all the Jews that were with him in Massepha, and all the Chaldeans that were found there.}{41:3}{Ishmael also killed all the Jews who were with Gedaliah at Mizpah, and the Chaldean men of war who were found there.}
\VersePair{48:4}{And it came to pass on the second day after he had smitten Godolias, and no man knew of it,}{41:4}{The second day after he had killed Gedaliah, and no man knew it,}
\VersePair{48:5}{that there came men from Sychem, and from Salem, and from Samaria, even eighty men, having their beards shaven, and their clothes rent, and beating their breasts, and they had manna and frankincense in their hands, to bring them into the house of the Lord.}{41:5}{men came from Shechem, from Shiloh, and from Samaria, even eighty men, having their beards shaved and their clothes torn, and having cut themselves, with meal offerings and frankincense in their hand, to bring them to Yahweh’s house.}
\VersePair{48:6}{And Ismael went out to meet them; and they went on and wept: and he said to them, Come in to Godolias.}{41:6}{Ishmael the son of Nethaniah went out from Mizpah to meet them, weeping all along as he went, and as he met them, he said to them, “Come to Gedaliah the son of Ahikam.”}
\VersePair{48:7}{And it came to pass, when they had entered into the midst of the city, that he slew them and cast them into a pit.}{41:7}{It was so, when they came into the middle of the city, that Ishmael the son of Nethaniah killed them, and cast them into the middle of the pit, he, and the men who were with him.}
\VersePair{48:8}{But ten men were found there, and they said to Ismael, Slay us not: for we have treasures in the field, wheat and barley, honey and oil. So he passed by, and slew them not in the midst of their brethren.}{41:8}{But ten men were found among those who said to Ishmael, “Don’t kill us; for we have stores hidden in the field, of wheat, and of barley, and of oil, and of honey.” So he stopped, and didn’t kill them among their brothers.}
\VersePair{48:9}{Now the pit into which Ismael cast all whom he smote, is the great pit, which king Asa had made for fear of Baasa king of Israel: even this Ismael filled with slain men.}{41:9}{Now the pit in which Ishmael cast all the dead bodies of the men whom he had killed, by the side of Gedaliah ( this was that which Asa the king had made for fear of Baasha king of Israel ), Ishmael the son of Nethaniah filled it with those who were killed.}
\VersePair{48:10}{And Ismael brought back all the people that were left in Massepha, and the king's daughter, whom the captain of the guard had committed in charge to Godolias the son of Achicam: and he went away beyond the children of Ammon.}{41:10}{Then Ishmael carried away captive all of the people who were left in Mizpah, even the king’s daughters, and all the people who remained in Mizpah, whom Nebuzaradan the captain of the guard had committed to Gedaliah the son of Ahikam. Ishmael the son of Nethaniah carried them away captive, and departed to go over to the children of Ammon.}
\VersePair{48:11}{And Joanan the son of Careë, and all the leaders of the host that were with him, heard of all the evil deeds which Ismael had done.}{41:11}{But when Johanan the son of Kareah, and all the captains of the forces who were with him, heard of all the evil that Ishmael the son of Nethaniah had done,}
\VersePair{48:12}{And they brought all their army, and went to fight against him, and found him near much water in Gabaon.}{41:12}{then they took all the men, and went to fight with Ishmael the son of Nethaniah, and found him by the great waters that are in Gibeon.}
\VersePair{48:13}{And it came to pass, when all the people that was with Ismael saw Joanan, and the leaders of the host that was with him,}{41:13}{Now when all the people who were with Ishmael saw Johanan the son of Kareah, and all the captains of the forces who were with him, then they were glad.}
\VersePair{48:14}{that they returned to Joanan.}{41:14}{So all the people who Ishmael had carried away captive from Mizpah turned about and came back, and went to Johanan the son of Kareah.}
\VersePair{48:15}{But Ismael escaped with eight men and went to the children of Ammon.}{41:15}{But Ishmael the son of Nethaniah escaped from Johanan with eight men, and went to the children of Ammon.}
\VersePair{48:16}{And Joanan, and all the leaders of the host that were with him, took all the remnant of the people, whom he had brought back from Ismael, mighty men in war, and the women, and the other property, and the eunuchs, whom they had brought back from Gabaon:}{41:16}{Then Johanan the son of Kareah and all the captains of the forces who were with him took all the remnant of the people whom he had recovered from Ishmael the son of Nethaniah, from Mizpah, after he had killed Gedaliah the son of Ahikam— the men of war, with the women, the children, and the eunuchs, whom he had brought back from Gibeon.}
\VersePair{48:17}{and they departed, and dwelt in Gaberoch-amaa, that is by Bethleem, to go into Egypt, for fear of the Chaldeans:}{41:17}{They departed and lived in Geruth Chimham, which is by Bethlehem, to go to enter into Egypt}
\VersePair{48:18}{for they were afraid of them, because Ismael had smitten Godolias, whom the king of Babylon made governor in the land.}{41:18}{because of the Chaldeans; for they were afraid of them, because Ishmael the son of Nethaniah had killed Gedaliah the son of Ahikam, whom the king of Babylon made governor over the land.}
\end{paracol}
\ChapterHeading{49}
\begin{paracol}{2}
\VersePair{49:1}{Then came all the leaders of the host, and Joanan, and Azarias the son of Maasæas, and all the people great and small,}{42:1}{Then all the captains of the forces, and Johanan the son of Kareah, and Jezaniah the son of Hoshaiah, and all the people from the least even to the greatest, came near,}
\VersePair{49:2}{to Jeremias the prophet, and said to him, Let now our supplication come before thy face, and pray thou to the Lord thy God for this remnant; for we are left few out of many, as thine eyes see.}{42:2}{and said to Jeremiah the prophet, “Please let our supplication be presented before you, and pray for us to Yahweh your God, even for all this remnant, for we are left but a few of many, as your eyes see us,}
\VersePair{49:3}{And let the Lord thy God declare to us the way wherein we should walk, and the thing which we should do.}{42:3}{that Yahweh your God may show us the way in which we should walk, and the things that we should do.”}
\VersePair{49:4}{And Jeremias said to them, I have heard you; behold, I will pray for you to the Lord our God, according to your words; and it shall come to pass, that whatsoever word the Lord God shall answer, I will declare it to you; I will not hide anything from you.}{42:4}{Then Jeremiah the prophet said to them, “I have heard you. Behold, I will pray to Yahweh your God according to your words; and it will happen that whatever thing Yahweh answers you, I will declare it to you. I will keep nothing back from you.”}
\VersePair{49:5}{And they said to Jeremias, Let the Lord be between us for a just and faithful witness, if we do not according to every word which the Lord shall send to us.}{42:5}{Then they said to Jeremiah, “May Yahweh be a true and faithful witness among us, if we don’t do according to all the word with which Yahweh your God sends you to tell us.}
\VersePair{49:6}{And whether it be good, or whether it be evil, we will hearken to the voice of the Lord our God, to whom we send thee; that it may be well with us, because we shall hearken to the voice of the Lord our God.}{42:6}{Whether it is good, or whether it is bad, we will obey the voice of Yahweh our God, to whom we send you; that it may be well with us, when we obey the voice of Yahweh our God.”}
\VersePair{49:7}{And it came to pass after ten days, that the word of the Lord came to Jeremias.}{42:7}{After ten days, Yahweh’s word came to Jeremiah.}
\VersePair{49:8}{And he called Joanan, and the leaders of the host, and all the people from the least even to the greatest,}{42:8}{Then he called Johanan the son of Kareah, and all the captains of the forces who were with him, and all the people from the least even to the greatest,}
\VersePair{49:9}{and he said to them, Thus saith the Lord;}{42:9}{and said to them, “Yahweh, the God of Israel, to whom you sent me to present your supplication before him, says:}
\VersePair{49:10}{If ye will indeed dwell in this land, I will build you, and will not pull you down, but will plant you, and in no wise pluck you up: for I have ceased from the calamities which I brought upon you.}{42:10}{‘If you will still live in this land, then I will build you, and not pull you down, and I will plant you, and not pluck you up; for I grieve over the distress that I have brought on you.}
\VersePair{49:11}{Be not afraid of the king of Babylon, of whom ye are afraid; be not afraid of him, saith the Lord: for I am with you, to deliver you, and save you out of their hand.}{42:11}{Don’t be afraid of the king of Babylon, of whom you are afraid. Don’t be afraid of him,’ says Yahweh, ‘for I am with you to save you, and to deliver you from his hand.}
\VersePair{49:12}{And I will grant you mercy, and pity you, and will restore you to your land.}{42:12}{I will grant you mercy, that he may have mercy on you, and cause you to return to your own land.}
\VersePair{49:13}{But if ye say, We will not dwell in this land, that we may not hearken to the voice of the Lord;}{42:13}{“‘But if you say, “We will not dwell in this land,” so that you don’t obey Yahweh your God’s voice,}
\VersePair{49:14}{for we will go into the land of Egypt, and we shall see no war, and shall not hear the sound of a trumpet, and we shall not hunger for bread; and there we will dwell:}{42:14}{saying, “No, but we will go into the land of Egypt, where we will see no war, nor hear the sound of the trumpet, nor have hunger of bread; and there we will dwell;”’}
\VersePair{49:15}{then hear the word of the Lord; thus saith the Lord;}{42:15}{now therefore hear Yahweh’s word, O remnant of Judah! Yahweh of Armies, the God of Israel, says, ‘If you indeed set your faces to enter into Egypt, and go to live there,}
\VersePair{49:16}{If ye set your face toward Egypt, and go in there to dwell; then it shall be, that the sword which ye fear shall find you in the land of Egypt, and the famine to which ye have regard, shall overtake you, coming after you in Egypt; and there ye shall die.}{42:16}{then it will happen that the sword, which you fear, will overtake you there in the land of Egypt; and the famine, about which you are afraid, will follow close behind you there in Egypt; and you will die there.}
\VersePair{49:17}{And all the men, and all the strangers who have set their face toward the land of Egypt to dwell there, shall be consumed by the sword, and by the famine: and there shall not one of them escape from the evils which I bring upon them.}{42:17}{So will it be with all the men who set their faces to go into Egypt to live there. They will die by the sword, by the famine, and by the pestilence. None of them will remain or escape from the evil that I will bring on them.’}
\VersePair{49:18}{For thus saith the Lord; As my wrath has dropped upon the inhabitants of Jerusalem, so shall my wrath drop upon you, when ye have entered into Egypt: and ye shall be a desolation, and under the power of others, and a curse and a reproach: and ye shall no more see this place.}{42:18}{For Yahweh of Armies, the God of Israel, says: ‘As my anger and my wrath has been poured out on the inhabitants of Jerusalem, so my wrath will be poured out on you, when you enter into Egypt; and you will be an object of horror, an astonishment, a curse, and a reproach; and you will see this place no more.’}
\VersePair{49:19}{These are the words which the Lord has spoken concerning you the remnant of Juda; Enter ye not into Egypt: and now know ye for a certainty,}{42:19}{“Yahweh has spoken concerning you, remnant of Judah, ‘Don’t go into Egypt!’ Know certainly that I have testified to you today.}
\VersePair{49:20}{that ye have wrought wickedness in your hearts, when ye sent me, saying, Pray thou for us to the Lord; and according to all that the Lord shall speak to thee we will do.}{42:20}{For you have dealt deceitfully against your own souls; for you sent me to Yahweh your God, saying, ‘Pray for us to Yahweh our God; and according to all that Yahweh our God says, so declare to us, and we will do it.’}
\VersePair{49:21}{And ye have not hearkened to the voice of the Lord, with which he sent me to you.}{42:21}{I have declared it to you today; but you have not obeyed Yahweh your God’s voice in anything for which he has sent me to you.}
\VersePair{49:22}{Now therefore ye shall perish by sword and by famine, in the place which ye desire to go into to dwell there.}{42:22}{Now therefore know certainly that you will die by the sword, by the famine, and by the pestilence in the place where you desire to go to live.”}
\end{paracol}
\ChapterHeading{50}
\begin{paracol}{2}
\VersePair{50:1}{And it came to pass, when Jeremias ceased speaking to the people all the words of the Lord, for which the Lord had sent him to them, even all these words,}{43:1}{When Jeremiah had finished speaking to all the people all the words of Yahweh their God, with which Yahweh their God had sent him to them, even all these words,}
\VersePair{50:2}{that Azarias son of Maasæas spoke, and Joanan, the son of Careë, and all the men who had spoken to Jeremias, saying, It is false: the Lord has not sent thee to us, saying, Enter not into Egypt to dwell there:}{43:2}{then Azariah the son of Hoshaiah, Johanan the son of Kareah, and all the proud men spoke, saying to Jeremiah, “You speak falsely. Yahweh our God has not sent you to say, ‘You shall not go into Egypt to live there;’}
\VersePair{50:3}{but Baruch the son of Nerias sets thee against us, that thou mayest deliver us into the hands of the Chaldeans, to kill us, and that we should be carried away captives to Babylon.}{43:3}{but Baruch the son of Neriah has turned you against us, to deliver us into the hand of the Chaldeans, that they may put us to death or carry us away captive to Babylon.”}
\VersePair{50:4}{So Joanan, and all the leaders of the host, and all the people, refused to hearken to the voice of the Lord, to dwell in the land of Juda.}{43:4}{So Johanan the son of Kareah, and all the captains of the forces, and all the people, didn’t obey Yahweh’s voice, to dwell in the land of Judah.}
\VersePair{50:5}{And Joanan, and all the leaders of the host, took all the remnant of Juda, who had returned to dwell in the land;}{43:5}{But Johanan the son of Kareah and all the captains of the forces took all the remnant of Judah, who had returned from all the nations where they had been driven, to live in the land of Judah —}
\VersePair{50:6}{the mighty men, and the women, and the children that were left, and the daughters of the king, and the souls which Nabuzardan had left with Godolias the son of Achicam, and Jeremias the prophet, and Baruch the son of Nerias.}{43:6}{the men, the women, the children, the king’s daughters, and every person who Nebuzaradan the captain of the guard had left with Gedaliah the son of Ahikam, the son of Shaphan; and Jeremiah the prophet, and Baruch the son of Neriah.}
\VersePair{50:7}{And they came into Egypt: for they hearkened not to the voice of the Lord: and they entered into Taphnas.}{43:7}{They came into the land of Egypt, for they didn’t obey Yahweh’s voice; and they came to Tahpanhes.}
\VersePair{50:8}{And the word of the Lord came to Jeremias in Taphnas, saying,}{43:8}{Then Yahweh’s word came to Jeremiah in Tahpanhes, saying,}
\VersePair{50:9}{Take thee great stones, and hide them in the entrance, at the gate of the house of Pharao in Taphnas, in the sight of the men of Juda:}{43:9}{“Take great stones in your hand and hide them in mortar in the brick work which is at the entry of Pharaoh’s house in Tahpanhes, in the sight of the men of Judah.}
\VersePair{50:10}{and thou shalt say, Thus has the Lord said; Behold, I will send, and will bring Nabuchodonosor king of Babylon, and he shall place his throne upon these stones which thou hast hidden, and he shall lift up weapons against them.}{43:10}{Tell them, Yahweh of Armies, the God of Israel, says: ‘Behold, I will send and take Nebuchadnezzar the king of Babylon, my servant, and will set his throne on these stones that I have hidden; and he will spread his royal pavilion over them.}
\VersePair{50:11}{And he shall enter in, and smite the land of Egypt, delivering some for death to death; and some for captivity to captivity; and some for the sword to the sword.}{43:11}{He will come, and will strike the land of Egypt; such as are for death will be put to death, and such as are for captivity to captivity, and such as are for the sword to the sword.}
\VersePair{50:12}{And he shall kindle a fire in the houses of their gods, and shall burn them, and shall carry them away captives: and shall search the land of Egypt, as a shepherd searches his garment; and he shall go forth in peace.}{43:12}{I will kindle a fire in the houses of the gods of Egypt. He will burn them, and carry them away captive. He will array himself with the land of Egypt, as a shepherd puts on his garment; and he will go out from there in peace.}
\VersePair{50:13}{And he shall break to pieces the pillars of Heliopolis that are in On, and shall burn their houses with fire.}{43:13}{He will also break the pillars of Beth Shemesh that is in the land of Egypt; and he will burn the houses of the gods of Egypt with fire.’”}
\end{paracol}
\ChapterHeading{51}
\begin{paracol}{2}
\VersePair{51:1}{The Word that came to Jeremias for all the Jews dwelling in the land of Egypt, and for those settled in Magdolo and in Taphnas, and in the land of Pathura, saying,}{44:1}{The word that came to Jeremiah concerning all the Jews who lived in the land of Egypt, who lived at Migdol, and at Tahpanhes, and at Memphis, and in the country of Pathros, saying,}
\VersePair{51:2}{Thus has the Lord God of Israel said; Ye have seen all the evils which I have brought upon Jerusalem, and upon the cities of Juda; and, behold, they are desolate without inhabitants,}{44:2}{“Yahweh of Armies, the God of Israel, says: ‘You have seen all the evil that I have brought on Jerusalem, and on all the cities of Judah. Behold, today they are a desolation, and no man dwells in them,}
\VersePair{51:3}{because of their wickedness, which they have wrought to provoke me, by going to burn incense to other gods, whom ye knew not.}{44:3}{because of their wickedness which they have committed to provoke me to anger, in that they went to burn incense, to serve other gods that they didn’t know, neither they, nor you, nor your fathers.}
\VersePair{51:4}{Yet I sent to you my servants the prophets early in the morning, and I sent, saying, Do not ye this abominable thing which I hate.}{44:4}{However I sent to you all my servants the prophets, rising up early and sending them, saying, “Oh, don’t do this abominable thing that I hate.”}
\VersePair{51:5}{But they hearkened not to me, and inclined not their ear to turn from their wickedness, so as not to burn incense to strange gods.}{44:5}{But they didn’t listen and didn’t incline their ear. They didn’t turn from their wickedness, to stop burning incense to other gods.}
\VersePair{51:6}{So mine anger and my wrath dropped upon them, and was kindled in the gates of Juda, and in the streets of Jerusalem; and they became a desolation and a waste, as at this day.}{44:6}{Therefore my wrath and my anger was poured out, and was kindled in the cities of Judah and in the streets of Jerusalem; and they are wasted and desolate, as it is today.’}
\VersePair{51:7}{And now thus has the Lord Almighty said, Wherefore do ye commit these great evils against your souls? to cut off man and woman of you, infant and suckling from the midst of Juda, to the end that not one of you should be left;}{44:7}{“Therefore now Yahweh, the God of Armies, the God of Israel, says: ‘Why do you commit great evil against your own souls, to cut off from yourselves man and woman, infant and nursing child out of the middle of Judah, to leave yourselves no one remaining,}
\VersePair{51:8}{by provoking me with the works of your hands, to burn incense to other gods in the land of Egypt, into which ye entered to dwell there, that ye might be cut off, and that ye might become a curse and a reproach among all the nations of the earth?}{44:8}{in that you provoke me to anger with the works of your hands, burning incense to other gods in the land of Egypt where you have gone to live, that you may be cut off, and that you may be a curse and a reproach among all the nations of the earth?}
\VersePair{51:9}{Have ye forgotten the sins of your fathers, and the sins of the kings of Juda, and the sins of your princes, and the sins of your wives, which they wrought in the land of Juda, and in the streets of Jerusalem?}{44:9}{Have you forgotten the wickedness of your fathers, the wickedness of the kings of Judah, the wickedness of their wives, your own wickedness, and the wickedness of your wives which they committed in the land of Judah and in the streets of Jerusalem?}
\VersePair{51:10}{And have not ceased even to this day, and they have not kept to my ordinances, which I set before their fathers.}{44:10}{They are not humbled even to this day, neither have they feared, nor walked in my law, nor in my statutes, that I set before you and before your fathers.’}
\VersePair{51:11}{Therefore thus saith the Lord; Behold, I do set my face against you}{44:11}{“Therefore Yahweh of Armies, the God of Israel, says: ‘Behold, I will set my face against you for evil, even to cut off all Judah.}
\VersePair{51:12}{to destroy all the remnant that are in Egypt; and they shall fall by the sword, and by famine, and shall be consumed small and great: and they shall be for reproach, and for destruction, and for a curse.}{44:12}{I will take the remnant of Judah that have set their faces to go into the land of Egypt to live there, and they will all be consumed. They will fall in the land of Egypt. They will be consumed by the sword and by the famine. They will die, from the least even to the greatest, by the sword and by the famine. They will be an object of horror, an astonishment, a curse, and a reproach.}
\VersePair{51:13}{And I will visit them that dwell in the land of Egypt, as I have visited Jerusalem, with sword and with famine:}{44:13}{For I will punish those who dwell in the land of Egypt, as I have punished Jerusalem, by the sword, by the famine, and by the pestilence;}
\VersePair{51:14}{and there shall not one be preserved of the remnant of Juda that sojourn in the land of Egypt, to return to the land of Juda, to which they hope in their hearts to return: they shall not return, but only they that escape.}{44:14}{so that none of the remnant of Judah, who have gone into the land of Egypt to live there, will escape or be left to return into the land of Judah, to which they have a desire to return to dwell there; for no one will return except those who will escape.’”}
\VersePair{51:15}{Then all the men that knew that their wives burned incense, and all the women, a great multitude, and all the people that dwelt in the land of Egypt, in Pathura, answered Jeremias, saying,}{44:15}{Then all the men who knew that their wives burned incense to other gods, and all the women who stood by, a great assembly, even all the people who lived in the land of Egypt, in Pathros, answered Jeremiah, saying,}
\VersePair{51:16}{As for the word which thou hast spoken to us in the name of the Lord, we will not hearken to thee.}{44:16}{“As for the word that you have spoken to us in Yahweh’s name, we will not listen to you.}
\VersePair{51:17}{For we will surely perform every word that shall proceed out of our mouth, to burn incense to the queen of heaven, and to pour drink-offerings to her, as we and our fathers have done, and our kings and princes, in the cities of Juda, and in the streets of Jerusalem: and so we were filled with bread, and were well, and saw no evils.}{44:17}{But we will certainly perform every word that has gone out of our mouth, to burn incense to the queen of the sky and to pour out drink offerings to her, as we have done, we and our fathers, our kings and our princes, in the cities of Judah and in the streets of Jerusalem; for then we had plenty of food, and were well, and saw no evil.}
\VersePair{51:18}{But since we left off to burn incense to the queen of heaven, we have all been brought low, and have been consumed by sword and by famine.}{44:18}{But since we stopped burning incense to the queen of the sky, and pouring out drink offerings to her, we have lacked all things, and have been consumed by the sword and by the famine.”}
\VersePair{51:19}{And whereas we burned incense to the queen of heaven, and poured drink-offerings to her, did we make cakes to her, and pour drink-offerings to her, without our husbands?}{44:19}{The women said, “When we burned incense to the queen of the sky and poured out drink offerings to her, did we make her cakes to worship her, and pour out drink offerings to her, without our husbands?”}
\VersePair{51:20}{Then Jeremias answered all the people, the mighty men, and the women, and all the people that returned him these words for answer, saying,}{44:20}{Then Jeremiah said to all the people — to the men and to the women, even to all the people who had given him an answer, saying,}
\VersePair{51:21}{Did not the Lord remember the incense which ye burned in the cities of Juda, and in the streets of Jerusalem, ye, and your fathers, and your kings, and your princes, and the people of the land? and came it not into his heart?}{44:21}{“The incense that you burned in the cities of Judah, and in the streets of Jerusalem, you and your fathers, your kings and your princes, and the people of the land, didn’t Yahweh remember them, and didn’t it come into his mind?}
\VersePair{51:22}{And the Lord could no longer bear you, because of the wickedness of your doings, and because of your abominations which ye wrought; and so your land became a desolation and a waste, and a curse, as at this day;}{44:22}{Thus Yahweh could no longer bear it, because of the evil of your doings and because of the abominations which you have committed. Therefore your land has become a desolation, an astonishment, and a curse, without inhabitant, as it is today.}
\VersePair{51:23}{because of your burning incense, and because of the things wherein ye sinned against the Lord: and ye have not hearkened to the voice of the Lord, and have not walked in his ordinances, and in his law, and in his testimonies; and so these evils have come upon you.}{44:23}{Because you have burned incense and because you have sinned against Yahweh, and have not obeyed Yahweh’s voice, nor walked in his law, nor in his statutes, nor in his testimonies; therefore this evil has happened to you, as it is today.”}
\VersePair{51:24}{And Jeremias said to the people, and to the women, Hear ye the word of the Lord.}{44:24}{Moreover Jeremiah said to all the people, including all the women, “Hear Yahweh’s word, all Judah who are in the land of Egypt!}
\VersePair{51:25}{Thus has the Lord God of Israel said; Ye women have spoken with your mouth, and ye fulfilled it with your hands, saying, We will surely perform our vows that we have vowed, to burn incense to the queen of heaven, and to pour drink-offerings to her: full well did ye keep to your vows, and ye have indeed performed them.}{44:25}{Yahweh of Armies, the God of Israel, says, ‘You and your wives have both spoken with your mouths, and with your hands have fulfilled it, saying, “We will surely perform our vows that we have vowed, to burn incense to the queen of the sky, and to pour out drink offerings to her.” “‘Establish then your vows, and perform your vows.’}
\VersePair{51:26}{Therefore hear ye the word of the Lord, all Jews dwelling in the land of Egypt; Behold, I have sworn by my great name, saith the Lord, my name shall no longer be in the mouth of every Jew to say, The Lord lives, in all the land of Egypt.}{44:26}{“Therefore hear Yahweh’s word, all Judah who dwell in the land of Egypt: ‘Behold, I have sworn by my great name,’ says Yahweh, ‘that my name will no more be named in the mouth of any man of Judah in all the land of Egypt, saying, “As the Lord Yahweh lives.”}
\VersePair{51:27}{For I have watched over them, to hurt them, and not to do them good: and all the Jews dwelling in the land of Egypt shall perish by sword and by famine, until they are utterly consumed.}{44:27}{Behold, I watch over them for evil, and not for good; and all the men of Judah who are in the land of Egypt will be consumed by the sword and by the famine, until they are all gone.}
\VersePair{51:28}{And they that escape the sword shall return to the land of Juda few in number, and the remnant of Juda, who have continued in the land of Egypt to dwell there, shall know whose word shall stand.}{44:28}{Those who escape the sword will return out of the land of Egypt into the land of Judah few in number. All the remnant of Judah, who have gone into the land of Egypt to live there, will know whose word will stand, mine or theirs.}
\VersePair{51:29}{And this shall be a sign to you, that I will visit you for evil.}{44:29}{“‘This will be the sign to you,’ says Yahweh, ‘that I will punish you in this place, that you may know that my words will surely stand against you for evil.’}
\VersePair{51:30}{Thus said the Lord; Behold, I will give Uaphres king of Egypt into the hands of his enemy, and into the hands of one that seeks his life; as I gave Sedekias king of Juda into the hands of Nabuchodonosor king of Babylon, his enemy, and who sought his life.}{44:30}{Yahweh says, ‘Behold, I will give Pharaoh Hophra king of Egypt into the hand of his enemies and into the hand of those who seek his life, just as I gave Zedekiah king of Judah into the hand of Nebuchadnezzar king of Babylon, who was his enemy and sought his life.’”}
\VersePair{51:31}{The Word which Jeremias the Prophet spoke to Baruch son of Nerias, when he wrote these words in the book from the mouth of Jeremias, in the fourth year of Joakim the son of Josias king of Juda.}{45:1}{The message that Jeremiah the prophet spoke to Baruch the son of Neriah, when he wrote these words in a book at the mouth of Jeremiah, in the fourth year of Jehoiakim the son of Josiah, king of Judah, saying,}
\VersePair{51:32}{Thus has the Lord said to thee, O Baruch.}{45:2}{“Yahweh, the God of Israel, says to you, Baruch:}
\VersePair{51:33}{Whereas thou hast said, Alas! alas! for the Lord has laid a grievous trouble upon me; I lay down in groaning, I found no rest;}{45:3}{‘You said, “Woe is me now! For Yahweh has added sorrow to my pain! I am weary with my groaning, and I find no rest.”’}
\VersePair{51:34}{say thou to him, Thus saith the Lord; Behold, I pull down those whom I have built up, and I pluck up those whom I have planted.}{45:4}{“You shall tell him, Yahweh says: ‘Behold, that which I have built, I will break down, and that which I have planted I will pluck up; and this in the whole land.}
\VersePair{51:35}{And wilt thou seek great things for thyself? seek them not: for, behold, I bring evil upon all flesh, saith the Lord: but I will give to thee thy life for a spoil in every place whither thou shalt go.}{45:5}{Do you seek great things for yourself? Don’t seek them; for, behold, I will bring evil on all flesh,’ says Yahweh, ‘but I will let you escape with your life wherever you go.’”}
\end{paracol}
\ChapterHeading{52}
\begin{paracol}{2}
\VersePair{52:1}{It was the twenty-first year of Sedekias, when he began to reign, and he reigned eleven years in Jerusalem. And his mother's name was Amitaal, the daughter of Jeremias, of Lobena.}{52:1}{Zedekiah was twenty-one years old when he began to reign. He reigned eleven years in Jerusalem. His mother’s name was Hamutal the daughter of Jeremiah of Libnah.}
\VersePair{52:2}{}{52:2}{He did that which was evil in Yahweh’s sight, according to all that Jehoiakim had done.}
\VersePair{52:3}{}{52:3}{For through Yahweh’s anger this happened in Jerusalem and Judah, until he had cast them out from his presence. Zedekiah rebelled against the king of Babylon.}
\VersePair{52:4}{And it came to pass in the ninth year of his reign, in the ninth month, on the tenth day of the month, that Nabuchodonosor king of Babylon came, and all his host, against Jerusalem, and they made a rampart round it, and built a wall round about it with large stones.}{52:4}{In the ninth year of his reign, in the tenth month, in the tenth day of the month, Nebuchadnezzar king of Babylon came, he and all his army, against Jerusalem, and encamped against it; and they built forts against it round about.}
\VersePair{52:5}{So the city was besieged, until the eleventh year of king Sedekias,}{52:5}{So the city was besieged to the eleventh year of King Zedekiah.}
\VersePair{52:6}{on the ninth day of the month, and then the famine was severe in the city, and there was no bread for the people of the land.}{52:6}{In the fourth month, in the ninth day of the month, the famine was severe in the city, so that there was no bread for the people of the land.}
\VersePair{52:7}{And the city was broken up, and all the men of war went out by night by the way of the gate, between the wall and the outworks, which were by the king's garden; and the Chaldeans were by the city round about; and they went by the way leading to the wilderness.}{52:7}{Then a breach was made in the city, and all the men of war fled, and went out of the city by night by the way of the gate between the two walls, which was by the king’s garden. Now the Chaldeans were against the city all around. The men of war went toward the Arabah,}
\VersePair{52:8}{But the host of the Chaldeans pursued after the king, and overtook him in the country beyond Jericho; and all his servants were dispersed from about him.}{52:8}{but the army of the Chaldeans pursued the king, and overtook Zedekiah in the plains of Jericho; and all his army was scattered from him.}
\VersePair{52:9}{And they took the king, and brought him to the king of Babylon to Deblatha, and he judged him.}{52:9}{Then they took the king, and carried him up to the king of Babylon to Riblah in the land of Hamath; and he pronounced judgment on him.}
\VersePair{52:10}{And the king of Babylon slew the sons of Sedekias before his eyes; and he slew all the princes of Juda in Deblatha.}{52:10}{The king of Babylon killed the sons of Zedekiah before his eyes. He also killed all the princes of Judah in Riblah.}
\VersePair{52:11}{And he put out the eyes of Sedekias, and bound him in fetters; and the king of Babylon brought him to Babylon, and put him into the grinding-house, until the day when he died.}{52:11}{He put out the eyes of Zedekiah; and the king of Babylon bound him in fetters, and carried him to Babylon, and put him in prison until the day of his death.}
\VersePair{52:12}{And in the fifth month, on the tenth day of the month, Nabuzardan the captain of the guard, who waited on the king of Babylon, came to Jerusalem;}{52:12}{Now in the fifth month, in the tenth day of the month, which was the nineteenth year of King Nebuchadnezzar, king of Babylon, Nebuzaradan the captain of the guard, who stood before the king of Babylon, came into Jerusalem.}
\VersePair{52:13}{and he burnt the house of the Lord, and the king's house; and all the houses of the city, and every great house he burnt with fire.}{52:13}{He burned Yahweh’s house, and the king’s house; and all the houses of Jerusalem, even every great house, he burned with fire.}
\VersePair{52:14}{And the host of the Chaldeans that was with the captain of the guard pulled down all the wall of Jerusalem round about.}{52:14}{All the army of the Chaldeans, who were with the captain of the guard, broke down all the walls of Jerusalem all around.}
\VersePair{52:16}{But the captain of the guard left the remnant of the people to be vinedressers and husbandmen.}{52:16}{But Nebuzaradan the captain of the guard left of the poorest of the land to be vineyard keepers and farmers.}
\VersePair{52:17}{And the Chaldeans broke in pieces the brazen pillars that were in the house of the Lord, and the bases, and the brazen sea that was in the house of the Lord, and they took the brass thereof, and carried it away to Babylon.}{52:17}{The Chaldeans broke the pillars of bronze that were in Yahweh’s house and the bases and the bronze sea that were in Yahweh’s house in pieces, and carried all of their bronze to Babylon.}
\VersePair{52:18}{Also the rim, and the bowls, and the flesh-hooks, and all the brazen vessels, wherewith they ministered;}{52:18}{They also took away the pots, the shovels, the snuffers, the basins, the spoons, and all the vessels of bronze with which they ministered.}
\VersePair{52:19}{and the basons, and the snuffers, and the oil-funnels, and the candlesticks, and the censers, and the cups, the golden, of gold, and the silver, of silver, the captain of the guard took away.}{52:19}{The captain of the guard took away the cups, the fire pans, the basins, the pots, the lamp stands, the spoons, and the bowls; that which was of gold, as gold, and that which was of silver, as silver.}
\VersePair{52:20}{And the two pillars, and the one sea, and the twelve brazen oxen under the sea, which things king Solomon made for the house of the Lord; the brass of which articles was without weight.}{52:20}{They took the two pillars, the one sea, and the twelve bronze bulls that were under the bases, which King Solomon had made for Yahweh’s house. The bronze of all these vessels was without weight.}
\VersePair{52:21}{And as for the pillars, the height of one pillar was thirty-five cubits; and a line of twelve cubits compassed it round; and the thickness of it all round was four fingers.}{52:21}{As for the pillars, the height of the one pillar was eighteen cubits; and a line of twelve cubits encircled it; and its thickness was four fingers. It was hollow. \footnote{52:21: A cubit is the length from the tip of the middle finger to the elbow on a man’s arm, or about 18 inches or 46 centimeters.}}
\VersePair{52:22}{And there was a brazen chapiter upon them, and the length was five cubits, even the height of one chapiter; and there were on the chapiter round about network and pomegranates, all of brass: and correspondingly the second pillar had eight pomegranates to a cubit for the twelve cubits.}{52:22}{A capital of bronze was on it; and the height of the one capital was five cubits, with network and pomegranates on the capital all around, all of bronze. The second pillar also had the same, with pomegranates. \footnote{52:22: A cubit is the length from the tip of the middle finger to the elbow on a man’s arm, or about 18 inches or 46 centimeters.}}
\VersePair{52:23}{And the pomegranates were ninety-six on a side; and all the pomegranates on the network round about were a hundred.}{52:23}{There were ninety-six pomegranates on the sides; all the pomegranates were one hundred on the network all around.}
\VersePair{52:24}{And the captain of the guard took the chief priest, and the second priest, and those that kept the way;}{52:24}{The captain of the guard took Seraiah the chief priest, and Zephaniah the second priest, and the three keepers of the threshold,}
\VersePair{52:25}{and one eunuch, who was over the men of war, and seven men of renown, who were in the king's presence, that were found in the city; and the scribe of the forces, who did the part of a scribe to the people of the land; and sixty men of the people of the land, who were found in the midst of the city.}{52:25}{and out of the city he took an officer who was set over the men of war; and seven men of those who saw the king’s face, who were found in the city; and the scribe of the captain of the army, who mustered the people of the land; and sixty men of the people of the land, who were found in the middle of the city.}
\VersePair{52:26}{And Nabuzardan the captain of the king's guard took them, and brought them to the king of Babylon to Deblatha.}{52:26}{Nebuzaradan the captain of the guard took them, and brought them to the king of Babylon to Riblah.}
\VersePair{52:27}{And the king of Babylon smote them in Deblatha, in the land of Æmath.}{52:27}{The king of Babylon struck them, and put them to death at Riblah in the land of Hamath. So Judah was carried away captive out of his land.}
\VersePair{52:28}{}{52:28}{{\raggedright This is the number of the people whom Nebuchadnezzar carried away captive: \poemline{1}{in the seventh year, three thousand twenty-three Jews;}}}
\VersePair{52:29}{}{52:29}{in the eighteenth year of Nebuchadnezzar, he carried away captive from Jerusalem eight hundred thirty-two persons;}
\VersePair{52:30}{}{52:30}{{\raggedright in the twenty-third year of Nebuchadnezzar, Nebuzaradan the captain of the guard carried away captive of the Jews seven hundred forty-five people. \poemline{1}{All the people numbered four thousand six hundred.}}}
\VersePair{52:31}{And it came to pass in the thirty-seventh year after that Joakim king of Juda had been carried away captive, in the twelfth month, on the four and twentieth day of the month, that Ulæmadachar king of Babylon, in the year in which he began to reign, raised the head of Joakim king of Juda, and shaved him, and brought him out of the house where he was kept,}{52:31}{In the thirty-seventh year of the captivity of Jehoiachin king of Judah, in the twelfth month, in the twenty-fifth day of the month, Evilmerodach king of Babylon, in the first year of his reign, lifted up the head of Jehoiachin king of Judah, and released him from prison.}
\VersePair{52:32}{and spoke kindly to him, and set his throne above the kings that were with him in Babylon,}{52:32}{He spoke kindly to him, and set his throne above the throne of the kings who were with him in Babylon,}
\VersePair{52:33}{and changed his prison garments: and he ate bread continually before him all the days that he lived.}{52:33}{and changed his prison garments. Jehoiachin ate bread before him continually all the days of his life.}
\VersePair{52:34}{And his appointed portion was given him continually by the king of Babylon from day to day, until the day when he died.}{52:34}{For his allowance, there was a continual allowance given him by the king of Babylon, every day a portion until the day of his death, all the days of his life.}
\end{paracol}
\end{document}
